### 2018-05-24 abstract.tex
initial port from grove-paper repo

% !TEX root = palettes-paper.tex

\begin{abstract}

Palettes... \cite{popl-paper}

\end{abstract}

### 2018-10-10 abstract.tex

@@
 % !TEX root = palettes-paper.tex
+
+%% no citations in Abstract
+%%
+%% \cite{popl-paper}
 
@@
 
-Palettes... \cite{popl-paper}
+We present \emph{nested palettes}, an approach to programming custom graphical
+user interfaces (GUIs) that are used to build program expressions---to ``fill
+holes'' within an incomplete program.
+%
+Nested palettes serve as an alternative and complementary code editing mechanism
+to traditional text-editing.
+%
+Compared to other ``hybrid'' text-and-GUI code editors, nested palettes provide
+several unique capabilities.
+%
+First, a palette persists even after user interactions produce a complete
+expression that fills the hole, so that the programmer can return to the GUI to
+make subsequent changes.
+%
+Second, palettes for smaller expression types can appear nested within those for
+larger expressions, allowing palettes to be composed in complex ways.
+%
+Third, palettes have access to the execution environment---even when the overall
+program is not yet complete---and can, thus, provide live feedback to the
+programmer while working to fill an incomplete expression.
+
+This paper investigates the design and implementation of nested palettes in
+several steps.
+%
+First, we propose a formal framework for nested palettes within Hazelnut, a
+foundational semantics for live programming that provides continuous code editor
+services through a combination of structured edits and dynamic evaluation of
+incomplete programs.
+%
+Second, to demonstrate the expressiveness of our framework, we describe several
+example palettes, many of which are informed by a prior study that identified
+desirable use cases for hybrid editors.
+%
+Lastly, we provide a prototype implementation of nested palettes within Hazel, a
+programming environment based on the Hazelnut theory extended with palettes.
+%
+Based on our experience, we believe nested palettes are an extensible and
+expressive approach for mixing text- and GUI-based code editing.
 

### 2019-02-26 abstract.tex
-nested

--- 2018-05-24 abstract.tex
@@
 
+%% no citations in Abstract
+%%
+%% \cite{popl-paper}
+
 \begin{abstract}
+%
+\vspace{2.00in} %% TODO leave room for authors/copyright
+%
+We present \emph{palette expressions}, an approach to programming custom
+graphical user interfaces (GUIs) that are used to build expressions---to ``fill
+holes''---within an incomplete program.
+%
+Palette expressions serve as an alternative and complementary code editing
+mechanism to traditional text-editing.
+%
+Compared to other ``hybrid'' text-and-GUI code editors, palette expressions
+provide several unique capabilities.
+%
+First, a palette persists even after user interactions produce a complete
+expression that fills the hole, so that the programmer can return to the GUI to
+make subsequent changes.
+%
+Second, palettes for smaller expression types can appear nested within those for
+larger expressions, allowing palettes to be composed in complex ways.
+%
+Third, palettes have access to the execution environment---even when the overall
+program is not yet complete---and can, thus, provide live feedback to the
+programmer while working to fill an incomplete expression.
 
-Palettes... \cite{popl-paper}
+This paper investigates the design and implementation of palettes in several
+steps.
+%
+First, we propose a formal framework for palettes within Hazelnut, a
+foundational semantics for live programming that provides continuous code editor
+services through a combination of structured edits and dynamic evaluation of
+incomplete programs.
+%
+Second, to demonstrate the expressiveness of our framework, we describe several
+example palettes, many of which are informed by a prior study that identified
+desirable use cases for hybrid editors.
+%
+Lastly, we provide a prototype implementation of palettes within Hazel, a
+programming environment based on the Hazelnut theory extended with palettes.
+%
+Based on our experience, we believe palettes are an extensible and expressive
+approach for mixing text- and GUI-based code editing.
 

### 2019-02-26 intro.tex
wip Intro from October

\section{Introduction}
\label{sec:intro}

Programs are most often represented as text.
%
This representation format provides expert programmers access to a variety of
text editors and text edit actions---to affect fine-grained control over
programming decisions---and a large ecosystem of other text-based
utilities---for example, for line-based file differencing of source code
versions.
%
On the other hand, the flexibility and low-level nature of text comes at some
costs.
%
For experts, some tedious and manual edits could lead to
inefficiency---especially because many routine code editing operations do not
require the full flexibility of text~\citep{XXX}.
%
For novices, the large class of syntax errors that stem from text-edits presents
a steep learning curve.

Motivated to overcome the disadvantages of represent programs as text, a variety
of alternative code editing interfaces have been investigated.
%
At the opposite end of text are \emph{visual programming languages}, which often
completely represent a program with graphical elements~\citep{XXX}.
%
To forgo text completely, such approaches often target domain-specific
languages, such as dataflow programing~\citep{XXX} and pedagogical languages
that are, by design, restricted to relatively few building blocks.
%
Other \emph{structure}, or \emph{projectional} editors, still use a significant
amount of text to render programs, but forgo text-edits in favor of
\emph{structure edit actions} which transform the program (represented as a tree
or some other structure), sidestepping the danger of invalid intermediate states
of concrete syntax.

On the spectrum somewhere between fully text- and fully structure-based are
``hybrid'' editors, which augment text with additional ways to visualize and
manipulate the structure of the program.
%
Victor Scrubbing, APX~\citep{APX}, Sketch-n-Sketch~\citep{sns-pldi}, and Carbide
IDE~\citep{XXX}, among others, allow numeric values to be ``scrubbed'' by
directly manipulating sliders rather than text-editing numeric literals.
%
Barista~\cite{Barista} is a hybrid Java editor where custom \emph{structure
view} GUIs provide alternate representations of expressions instead of text.
%
For example, an arithmetic expression may be rendered with mathematical symbols,
a method may be accompanied by interactive documentation with input-output
examples, and structures may be selectively collapsed, expanded, or zoomed.
%
Graphite~\citep{Graphite} allows custom GUIs---called \emph{palettes}---to help
the programmer fill missing expressions---``holes''---in the program.
%
For example, a color palette can provide visual previews of different candidate
color values, and a regular expression palette can show input-output examples
for different candidate regular expressions.

The GUI representations and interactions enabled by the above hybrid editors are
useful, but several limitations likely preclude wider utility:
%
the types of expressions that benefit from alternative GUI editors are limited
to
%
privileged types that have baked-in interfaces~\citep{XXX}
%
or just of base type~\citep{XXX};
%
the GUI view is ephemeral, in that it disappears once it has been used to
generate an expression~\citep{XXX}; or
%
expressions generated by the GUI are not deeply integrated with the static type
system and interpreter of the language~\citep{XXX,XXX,XXX}.

%% refactoring
%% 
%% DNDRefactoring~\citep{DNDRefactoring}
%% Deuce~\citep{sns-deuce}

\parahead{Persistent, Composable, and Live GUIs for Filling Holes}

compared to the ``simple'' palettes

extend palettes with: persistence, composition, and live feedback

macro systems allow alternative syntactic (i.e. string) representations, to
expand into underlying expressions

palatable

by analogy to these ``literal'' macro systems, palettes are ``graphical
macros'': through interaction with the user, the GUI generates the underlying
expression.

TLMs~\cite{TLMs}

\parahead{Contributions and Paper Outline}

design for palettes. specifically, within the Hazelnut Live framework,
which is a system to address the gap problem that arises in traditional editor
frontends.

demonstration of the expressiveness of the approach through a series of
examples, many of which are drawn from the user study mentioned before.

a prototype implementation of palettes within Hazel, which currently
supports several examples using a core calculus with minimal features. the
implementation provides a clear path for scaling up to larger, full-featured
syntactic programming conveniences, as well as further UI design in future work.

hazelnut~\citep{Hazelnut,HazelnutLive}

### 2019-02-26 evaluation.tex
rough outline notes from Friday

\section{Evaluation}

\subsection{Implementation}

implementation in \Hazel{}: palette macros; structure edit actions; expression formula bar; inline layout

implementation in \sns{}: palette functions; text editor; pop-up menu

\subsection{Examples}

\rkc{old list of examples below:}

\subsubsection{Boolean}
Checkbox

\subsubsection{Matrix}
Simple example

\subsubsection{Grade Cutoffs}
Uses liveness more obviously

\subsubsection{Table}
Uses type reflection

\subsubsection{Pixel Art}
Cool example

\subsubsection{Regex}
Similar to Graphite paper, can cite the empirical study we did there

\subsubsection{Forms}
Show off composition + mention full-screening stuff

\subsubsection{Equation Editor, Judgement Editor and Category Diagrams}
Things that PL people like

\subsubsection{TikZ diagrams}
...

### 2019-02-26 overview.tex
rough outline notes from Friday

\section{Overview Examples: Using Palettes}
\label{sec:overview}

\begin{figure*}

\includegraphics[scale=0.50]{images/MrSmileyFace.png}

\caption{Example Palettes in Hazel.}
\end{figure*}

\begin{itemize}

\item figure with 2/3 Hazel palettes:

  \begin{enumerate}
    \item table (synthetic)
    \item grade cutoffs (analytic)
    \item color? (analytic)
  \end{enumerate}

\item highlight persistence

\item highlight composition

\item two dimensions for palette definitions:

  \begin{enumerate}
    \item macros (expressions) vs. functions (values)
    \item analytic (annotated) vs. synthetic (unannotated)
      \begin{itemize}
        \item synthetic requires type splices
        \item an ``untyped'' version that is like synthetic but relies on casts rather than type splices?
      \end{itemize}
  \end{enumerate}

\item live environments (for view function)

  \begin{itemize}
    \item hole for palette application, run, collect hole closures, pass to view
  \end{itemize}

\item two prototype UIs:

  \begin{enumerate}
    \item \Hazel{}: (palette macros;) structure edit actions; expression formula bar; inline layout
    \item \sns{}: (palette functions;) text editor; pop-up menu
  \end{enumerate}

\item (nice-to-have: parameterized palettes; partial application)

\item (nice-to-have: palette definitions in source lang)

\end{itemize}

### 2019-02-26 overview.tex
more fiddling with outline

@@
 
-\item highlight persistence
+\item emphasize persistence
 
-\item highlight composition
+\item emphasize composition
 
-\item two dimensions for palette definitions:
+\item emphasize liveness
 
-  \begin{enumerate}
-    \item macros (expressions) vs. functions (values)
-    \item analytic (annotated) vs. synthetic (unannotated)
-      \begin{itemize}
-        \item synthetic requires type splices
-        \item an ``untyped'' version that is like synthetic but relies on casts rather than type splices?
-      \end{itemize}
-  \end{enumerate}
-
-\item live environments (for view function)
+\item mention analytic (annotated) vs. synthetic (unannotated) palettes
 
   \begin{itemize}
-    \item hole for palette application, run, collect hole closures, pass to view
+    \item stick to palette macros for now
+    \item introduce palette functions in ``Overview Examples: Defining Palettes''
+          or ``Formal System''
   \end{itemize}
 
-\item two prototype UIs:
-
-  \begin{enumerate}
-    \item \Hazel{}: (palette macros;) structure edit actions; expression formula bar; inline layout
-    \item \sns{}: (palette functions;) text editor; pop-up menu
-  \end{enumerate}
-
-\item (nice-to-have: parameterized palettes; partial application)
-
-\item (nice-to-have: palette definitions in source lang)
-
 \end{itemize}

### 2020-05-10 abstract.tex
more fiddling with outline
reorg, abstract, 2/3 of intro

--- 2018-05-24 abstract.tex
@@
-% !TEX root = palettes-paper.tex
+% !TEX root = livelits-paper.tex
+
+%% no citations in Abstract
+%%
+%% \cite{popl-paper}
 
 \begin{abstract}
-
-Palettes... \cite{popl-paper}
-
+    Although textual program manipulation is powerful, 
+    some types of expressions are more naturally represented and manipulated
+    graphically.     
+    Examples include expressions computing   
+    colors,
+    animation parameters,
+    musical sequences,
+    tabular data, 
+    plot parameters,
+    GUI widgets, 
+    and many other domain-specific data structures.
+    This paper introduces \emph{live literals}, or \emph{livelits}, which allow
+    a programmer to fill holes of types like these by directly manipulating a corresponding    
+    GUI embedded persistently into the code. 
+    Livelits are compositional: the GUI can itself contain
+    spliced expressions, which in turn can contain other livelits. Splices are typed, and the system 
+    requires that the livelit treat each splice hygienically. 
+    Livelits are also live: they can provide {immediate feedback} about the dynamic implications 
+    of the programmer's choices 
+    even when the spliced expressions mention bound variables, because the system gathers closures associated with the hole that the livelit is tasked with filling.
+    This paper details the livelits mechanism with several non-trivial case studies. 
+    It then defines a typed lambda calculus that captures the essence of the mechanism.     
+    Our implementation extends Hazel, a web-based live programming environment able to assign
+    static and dynamic meaning to programs with holes. 
 \end{abstract}

### 2020-05-10 intro.tex
more fiddling with outline
reorg, abstract, 2/3 of intro

@@
-\section{Introduction}
-\label{sec:intro}
+\section{Introduction}\label{sec:intro}
+In the popular imagination, programmers spend their days interacting with screens of flowing text. 
+Textual program editors are powerful and expressive user interfaces, 
+so it is little wonder that they remain dominant decades after the teletype. 
+However, textual user interfaces are not the best tool for every job. In particular, there are many 
+data types for which a non-textual  
+user interface is situationally more appropriate.
 
+For example, consider the following record type\footnote{We will assume that we are in an ML-family language, e.g.
+Elm (TODO cite), for the remainder of this paper, though the ideas are general.} classifying colors:
+\begin{lstlisting}[numbers=none]
+type Color = { r: Int, g: Int, b: Int, a : Int }
+\end{lstlisting}
+It is possible to select a particular color by entering  
+an expression of this type using a text editor.
+% \begin{lstlisting}[numbers=none]
+% let bgcolor : color = { r: 255, g: 178, b: 45, a }
+% \end{lstlisting}
+The problem is that this \emph{de facto} user interface for color selection offers the user no immediate feedback 
+and limited editing affordances. 
+It is difficult for the programmer, or anyone subsequently reading the code, to know which color is represented 
+and to interactively tweak that color. 
+Better supporting these activities would be particularly useful when engaging in live and exploratory programming 
+in domains like web design, audiovisual production, and interactive data analysis.
+
+Indeed, practitioners in domains like these commonly turn to   
+graphical end-user applications---%
+image and video editors, spreadsheets, music composition software, and so on---%
+that do provide live feedback and 
+direct manipulation affordances. 
+The trade-off is that these end-user applications have limited or no support for programmatic abstraction and composition. 
+It is difficult, for example, to bind a
+color to a variable for use in multiple places, or to darken a color by passing it to a function, or to organize and distribute color schemes using a package manager.
+Moreover, it is difficult to compose affordances in ways that the application designer has not 
+anticipated, e.g. to use a slider to manipulate the alpha channel of a color 
+if the native color palette provides only a text box.
+
+This paper is motivated by the conviction that it is possible to resolve this apparent tension between 
+programmatic and direct manipulation by designing a programming system that 
+is able to surface GUIs for working with data types for which 
+they are useful, while also retaining support for textual program manipulation  
+and the full array of abstraction and composition mechanisms
+available in modern general-purpose programming languages.
+
+\subsection{Background}
+We are not, of course, the first to attempt to integrate   
+programming languages and direct manipulation user interfaces. 
+Prior work on projectional editing
+and active code completion, summarized in Sec. \ref{sec:related-work}, 
+has also considered the problem of entering expressions 
+of certain types, like \li{Color}, 
+using specialized GUIs integrated into a erprogram editor. 
+The prior work most closely related to this paper is the {Graphite} system for Java \cite{DBLP:conf/icse/OmarYLM12}.
+Graphite allows library providers to associate GUIs, called \emph{palettes}, with types. 
+The programming environment offers the programmer the option, via the code completion menu, 
+to activate the associated GUI wherever an expression of the corresponding type is needed, 
+i.e. wherever there is a \emph{hole} of that type in the program text. 
+It is the GUI's job to fill the hole, i.e. to generate an 
+appropriate expression to fill the hole. 
+Figure~\ref{fig:color}, reproduced from this prior work, demonstrates an example of a simple color chooser.
+
+This prior study extensively evaluated this mechanism. Most notable, the study 
+performed a survey of 450 developers and found
+that a mechanism of this sort would be valued. The prior work also solicited use cases from these
+developers, gather N examples that the authors grouped heuristically into a number of categories. Some examples:
+
+\dots
+
+This paper takes this extensive empirical evaluation as sufficient to establish the utility and broad applicability of a mechanism
+of a mechanism of this sort. Our focus in this paper is on addressing a number of non-trivial technical 
+deficiencies that limit the ability of library providers to implement useful palettes.
+
+\begin{figure*}
+\begin{center}
+\includegraphics[width=35pc]{color_palette.png}\end{center}
+\caption{This figure, reproduced from the prior work \cite{DBLP:conf/icse/OmarYLM12}, shows (a) a simple example GUI associated with the \texttt{Color} type, and (b) the code generated by this GUI once the ephemeral interaction is complete.}
+\label{fig:color}
+\end{figure*}
+
+\begin{figure*}[t!]
+\begin{center}
+\includegraphics[width=34pc]{cutoffs-mockup.png}\end{center}
+\caption{A mockup of a livelit for adjusting grade cutoffs. Livelits are persistent and can access the live environment.}
+\label{fig:cutoffs}
+\end{figure*}
+
+There are two major limitations with this prior work. 
+First, GUIs in Graphite are \textbf{ephemeral}, i.e.
+they disappear once the initial interaction is complete, leaving behind only the textual representation.
+This means that feedback and assistance is available only to the programmer
+that first writes the code. 
+% Graphite does include an \emph{ad hoc} mechanism that allows palettes to parse user-selected code when the GUI loads, but this requires significant additional effort.
+
+Second, Graphite does not provide any compositional way to {enter subexpressions within the GUI}.
+This implies that Graphite's GUIs can only generate \textbf{closed expressions}.
+For example, in Figure~\ref{fig:color}, there is no way to specify that the \texttt{R}, \texttt{G}, or \texttt{B} values of the color being entered should be computed by a specified expression---only closed colors are supported.
+% This limitation precludes a variety of other perhaps more interesting use cases, e.g. GUIs for entering tabular data where each entry in
+% the table is an arbitrary expression of the appropriate type. It also precludes nested composition of GUIs, e.g. a table where some entries are numbers generated by manipulating a slider GUI.
+
+\subsection{Summary of Contributions}
+Four things:
+1. open extensibility
+
+2. persistent
+
+3. compositional
+
+4. live
+
+Begin in Sec.~2 with an overview using a  case study involving several non-trivial livelits, starting 
+from the client's perspective and then considering the livelit provider's perpsective in Sec.~3.
+
+In Sec.~4, we provide a formal account of the mechanism, rooted in Hazelnut Live.
+
+In Sec.~5, we describe some extensions of the core mechanism.
+
+In Sec.~6, we outline our implementation of livelits within Hazel.
+
+In Sec.~7, we describe related work. Finally, we conclude in Sec.~8 with a discussion of 
+limitations and future work.
+
+\section{Livelits}
+To address these limitations, we introduce live literals, or livelits.
+Figure~\ref{fig:cutoffs} shows a mockup of the definition and application of a livelit named \li{$grade_cutoffs} for adjusting grade cutoffs, represented as values of  record type \li{grade_cutoffs}.
+Like textual literal forms (e.g.
+list literals), livelits are alternative representations of expressions of the associated type \cite{DBLP:journals/pacmpl/OmarA18}.
+% From the perspective of the remainder of the program, \li{cutoffs} is a value of type \li{grade_cutoffs} like any other.
+We are implementing livelits in Hazel (\url{hazel.org}), a live functional programming environment with support for typed holes \cite{popl-paper}. We plan to perform a live demo.
+
+The definition of
+the livelit, outlined in Figure~\ref{fig:cutoffs}, follows the Elm architecture,
+i.e. there are types representing the abstract model and the messages that the 
+GUI generates.
+Livelits are \textbf{persistent} rather than ephemeral, i.e.
+the model is recorded in the underlying syntax tree.
+The \li{view} function generates the GUI, implemented using HTML, on demand (so the view is not persisted).
+Another function, \li{to_exp}, is responsible for generating the underlying expression, called the \emph{expansion}, from the \li{model}.
+While other projectional editors, e.g. those generated by Citrus \cite{DBLP:conf/uist/KoM05}, also support persistent GUIs in code, they are not user-extensible.
+
+The GUI can itself contain typed holes, represented by the \li{HtmlHole} constructor. In this case, there is a single hole
+for entering an expression of type \li{list(float)}, i.e. the list of weighted
+averages. In this example, the user has filled this hole with a variable,
+\li{weighted_averages} (the definition of which precedes the content of this figure and is not shown). In other words,
+livelits support \textbf{open expressions} and therefore interact cleanly with 
+standard abstraction mechanisms, i.e. they can appear under binders. We follow the reasoning principles for literals with spliced sub-expressions established  by \citet{DBLP:journals/pacmpl/OmarA18}.
+
+The main complication when dealing with open expressions relates to how live feedback
+is to be generated. Given just the symbolic expression in the hole, it would be 
+impossible to plot (as orange dots) the actual data from the list that the variable refers to. To resolve this issue, the system evaluates the program 
+as if it the livelits were empty holes, 
+relying on the support for evaluating incomplete programs described recently 
+by \citet{DBLP:journals/pacmpl/OmarVCH19}. The result of evaluation is an expression containing
+hole closures, i.e. holes equipped with environments. 
+%Hazel allows the user to interactively select which closure
+%interests them if multiple closures for a given hole appear in the result (e.g. if a livelit appears inside a function called multiple times). 
+The livelit 
+can then evaluate the expression in the hole against the closure selected by the user (not shown) using \li{Env.run}.
+
+%%%%%
+
+\todo{this is the WIP one from last year}{}
 Programs are most often represented as text.
@@
 
+prototype UIs:
+
+  \begin{enumerate}
+    \item \Hazel{}: (palette macros;) structure edit actions; expression formula bar; inline layout
+    \item \sns{}: (palette functions;) text editor; pop-up menu
+  \end{enumerate}
+
 hazelnut~\citep{Hazelnut,HazelnutLive}

### 2020-05-10 livelits-paper.tex
reorg, abstract, 2/3 of intro

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}

%% % From https://pldi19.sigplan.org/track/pldi-2019-papers#Call-for-Papers
%% \documentclass[sigplan,10pt,review,anonymous]{acmart}
%% \settopmatter{printfolios=true,printccs=false,printacmref=false}

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review,anonymous,nonacm]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
\acmYear{2018}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
% \bibliographystyle{ACM-Reference-Format}
%% Citation style
%\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

%% Cyrus packages
\usepackage{microtype}
\usepackage{mdframed}
\usepackage{colortab}
\usepackage{mathpartir}
\usepackage{enumitem}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{leftidx}
\usepackage{todonotes}
\usepackage{xspace}
\usepackage{wrapfig}

\usepackage{listings}%
\lstloadlanguages{ML}
\lstset{tabsize=2, 
basicstyle=\footnotesize\ttfamily, 
% keywordstyle=\sffamily,
commentstyle=\itshape\ttfamily\color{gray}, 
stringstyle=\ttfamily\color{purple},
mathescape=false,escapechar=\#,
numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML, showspaces=false,showstringspaces=false,xleftmargin=15pt, 
morekeywords={string, float, int},
classoffset=0,belowskip=\smallskipamount, aboveskip=\smallskipamount,
moredelim=**[is][\color{red}]{SSTR}{ESTR}
}
\newcommand{\li}[1]{\lstinline[basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]{#1}}
\newcommand{\lismall}[1]{\lstinline[basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]{#1}}

%% Joshua Dunfield macros
\def\OPTIONConf{1}%
\usepackage{joshuadunfield}

%% Can remove this eventually
\usepackage{blindtext}

\usepackage{enumitem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Matt says: Cyrus, this package `adjustbox` seems directly related
%% to the `clipbox` error; To get rid of the error, I moved it last
%% (after other usepackages) and I added the line just above it, which
%% permits it to redefine `clipbox` (apparently also defined in
%% `pstricks`, and due to latex's complete lack of namespace
%% management, these would otherwise names clash).
\let\clipbox\relax
\usepackage[export]{adjustbox}% http://ctan.org/pkg/adjustbox
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\usepackage{draftwatermark}
%\SetWatermarkText{DRAFT}
%\SetWatermarkScale{1}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% A macro for the name of the system being described by ``this paper''
\newcommand{\HazelnutLive}{\textsf{Hazelnut Live}\xspace}
\newcommand{\Hazelnut}{\textsf{Hazelnut}\xspace}
% The mockup, work-in-progress system.
\newcommand{\Hazel}{\textsf{Hazel}\xspace}

% \newtheorem{theorem}{Theorem}[chapter]
% \newtheorem{lemma}[theorem]{Lemma}
% \newtheorem{corollary}[theorem]{Corollary}
% \newtheorem{definition}[theorem]{Definition}
% \newtheorem{assumption}[theorem]{Assumption}
% \newtheorem{condition}[theorem]{Condition}

\newtheoremstyle{slplain}% name
  {.15\baselineskip\@plus.1\baselineskip\@minus.1\baselineskip}% Space above
  {.15\baselineskip\@plus.1\baselineskip\@minus.1\baselineskip}% Space below
  {\slshape}% Body font
  {\parindent}%Indent amount (empty = no indent, \parindent = para indent)
  {\bfseries}%  Thm head font
  {.}%       Punctuation after thm head
  { }%      Space after thm head: " " = normal interword space;
        %       \newline = linebreak
  {}%       Thm head spec
\theoremstyle{slplain}
\newtheorem{thm}{Theorem}  % Numbered with the equation counter
\numberwithin{thm}{section}
\newtheorem{defn}[thm]{Definition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
% \newtheorem{cor}[section]{Corollary}     
% \newtheorem{lem}[section]{Lemma}         
% \newtheorem{prop}[section]{Proposition}

% \setlength{\abovedisplayskip}{0pt}
% \setlength{\belowdisplayskip}{0pt}
% \setlength{\abovedisplayshortskip}{0pt}
% \setlength{\belowdisplayshortskip}{0pt}

\makeatletter\if@ACM@journal\makeatother
%% Journal information (used by PACMPL format)
%% Supplied to authors by publisher for camera-ready submission
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmYear{2018}
\acmMonth{3}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\else\makeatother
%% Conference information (used by SIGPLAN proceedings format)
%% Supplied to authors by publisher for camera-ready submission
% \acmConference[]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}

\acmYear{2018}
\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\fi

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission
\setcopyright{none}             %% For review submission
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

\fancyfoot{} % suppresses the footer (also need \thispagestyle{empty} after \maketitle below)

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

\input{commands}
\input{macros}

\setlength{\abovecaptionskip}{4pt plus 3pt minus 2pt} % Chosen fairly arbitrarily
\setlength{\belowcaptionskip}{-4pt plus 3pt minus 2pt} % Chosen fairly arbitrarily

\begin{document}

%% Title information
\title
  [Filling Typed Holes with Live GUIs]
  {Filling Typed Holes with Live GUIs}
  %% [Programming by Direct Manipulation of Palettes]
  %% {Functional Programming by\\Direct Manipulation of Live Palette Expressions}

                                        %% when present, will be used in
                                        %% header instead of Full Title.
% \titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
% \subtitle{Subtitle}                     %% \subtitle is optional
% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{Cyrus Omar}
% \authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  % \position{Position1}
  % \department{Department1}              %% \department is recommended
  \institution{University of Michigan}            %% \institution is required
  % \streetaddress{Street1 Address1}
  % \city{City1}
  % \state{State1}
  % \postcode{Post-Code1}
  % \country{Country1}
}
\email{comar@cs.uchicago.edu}          %% \email is recommended

\author{Nick Collins}
% \authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  % \position{Position1}
  % \department{Department1}              %% \department is recommended
  \institution{University of Chicago}            %% \institution is required
  % \streetaddress{Street1 Address1}
  % \city{City1}
  % \state{State1}
  % \postcode{Post-Code1}
  % \country{Country1}
}
\email{nickmc@uchicago.edu}          %% \email is recommended

\author{David Moon}
% \authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  % \position{Position1}
  % \department{Department1}              %% \department is recommended
  \institution{University of Colorado Boulder}            %% \institution is required
  % \streetaddress{Street1 Address1}
  % \city{City1}
  % \state{State1}
  % \postcode{Post-Code1}
  % \country{Country1}
}
\email{domo9239@colorado.edu}          %% \email is recommended

\author{Ian Voysey}
% \authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  % \position{Position1}
  % \department{Department1}              %% \department is recommended
  \institution{Carnegie Mellon University}            %% \institution is required
  % \streetaddress{Street1 Address1}
  % \city{City1}
  % \state{State1}
  % \postcode{Post-Code1}
  % \country{Country1}
}
\email{iev@cs.cmu.edu}          %% \email is recommended

\author{Ravi Chugh}
% \authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  % \position{Position1}
  % \department{Department1}              %% \department is recommended
  \institution{University of Chicago}            %% \institution is required
  % \streetaddress{Street1 Address1}
  % \city{City1}
  % \state{State1}
  % \postcode{Post-Code1}
  % \country{Country1}
}
\email{rchugh@cs.uchicago.edu}          %% \email is recommended

\author{Ravi Chugh}
% \authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  % \position{Position1}
  % \department{Department1}              %% \department is recommended
  \institution{University of Chicago}            %% \institution is required
  % \streetaddress{Street1 Address1}
  % \city{City1}
  % \state{State1}
  % \postcode{Post-Code1}
  % \country{Country1}
}
\email{rchugh@cs.uchicago.edu}          %% \email is recommended

% %% Author with two affiliations and emails.
% \author{First2 Last2}
% \authornote{with author2 note}          %% \authornote is optional;
%                                         %% can be repeated if necessary
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
% \affiliation{
%   \position{Position2a}
%   \department{Department2a}             %% \department is recommended
%   \institution{Institution2a}           %% \institution is required
%   \streetaddress{Street2a Address2a}
%   \city{City2a}
%   \state{State2a}
%   \postcode{Post-Code2a}
%   \country{Country2a}
% }
% \email{first2.last2@inst2a.com}         %% \email is recommended
% \affiliation{
%   \position{Position2b}
%   \department{Department2b}             %% \department is recommended
%   \institution{Institution2b}           %% \institution is required
%   \streetaddress{Street3b Address2b}
%   \city{City2b}
%   \state{State2b}
%   \postcode{Post-Code2b}
%   \country{Country2b}
% }
% \email{first2.last2@inst2b.org}         %% \email is recommended

%% Paper note
%% The \thanks command may be used to create a "paper note" ---
%% similar to a title note or an author note, but not explicitly
%% associated with a particular element.  It will appear immediately
%% above the permission/copyright statement.
% \thanks{with paper note}                %% \thanks is optional
                                        %% can be repeated if necesary
                                        %% contents suppressed with 'anonymous'

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\input{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
% \begin{CCSXML}
% <ccs2012>
% <concept>
% <concept_id>10011007.10011006.10011008</concept_id>
% <concept_desc>Software and its engineering~General programming languages</concept_desc>
% <concept_significance>500</concept_significance>
% </concept>
% <concept>
% <concept_id>10003456.10003457.10003521.10003525</concept_id>
% <concept_desc>Social and professional topics~History of programming languages</concept_desc>
% <concept_significance>300</concept_significance>
% </concept>
% </ccs2012>
% \end{CCSXML}

% \ccsdesc[500]{Software and its engineering~General programming languages}
% \ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
% \keywords{keyword1, keyword2, keyword3}  %% \keywords is optional
% \keywords{Live Programming, Palettes, Hazel, Code Editors}

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle
% \thispagestyle{empty} % suppresses the footer

% \section{TODO List}

% \begin{itemize}

% \item[\cmark] palette macros
% \item action semantics
% \item \Hazel{} in \Hazel{} (sums, productions, recursive types)
% \item palette functions
% \item live palettes
% \item synthetic (or fancy record-type-level-computation) (or not-fancy untyped) palettes
% \item parameterized palettes
% \item palette-specific actions
% \item option to render palette macro expansion as expression or value
% \item \Hazel{} UI: ``formula bar'' for expressions
% \item \sns{} implementation
% \item examples!

% \end{itemize}

\input{intro}

\input{overview}

%% \section{Defining Palettes}

\section{Overview Examples: Defining Palettes}

Palette definitions take the following general form:
\begin{lstlisting}
palette $name 
  (arg1 : t1) 
  ... 
  (argn : tn) 
  at t 
  implementation P in package pkg;
\end{lstlisting}
The static semantics requires the following:
\begin{enumerate}
\item \li{arg1} ... \li{argn} for \li{n} $\geq 0$ are distinct labels
\item \li{t1} ...\li{t1} are valid types
\item \li{t} is a valid type
\item and \li{P} identifies a module that can be loaded from package \li{pkg} such that \begin{lstlisting}
P : IPalette
    with type args = { 
      arg1 : HoleRef.t<t1>, 
      ..., 
      argn : HoleRef.t<tn> 
    } 
    with type output = t
\end{lstlisting}
where \li{IPalette} is defined in Fig.~\ref{fig:IPalette} and \li{HoleRef} in Fig.~\ref{fig:HoleRefs}.
\end{enumerate}

\begin{figure}
\begin{lstlisting}
module type IPalette = {
  type args
  type output
  type model
  type msg
  val init    : args -> HRG.t(model)
  val update  : (model, msg) -> HRG.t(model)
  val view    : model -> HRE.t(Html.t(msg))
  val compute : model -> HRE.t(output)
}
\end{lstlisting}
\caption{\li{IPalette} module type}
\label{fig:IPalette}
\end{figure}

\begin{figure}
\begin{lstlisting}
(* Abstract type of hole refs *)
module HoleRef : {
  type t('a)
}

(* HRG is the hole ref generation monad *)
module HRG : {
  type t('a)
  val fresh('a) : t(HoleRef.t('a))
  val bind : t('a) -> 
             ('a -> t('b)) -> 
             t('b)
  val return : 'a -> t('a)
}

(* HRE is the hole ref evaluation monad *)
module HRE : {
  type t('a) 
  type result('a) = Value('a * Exp)
                  | Indet(Exp)
  val eval('a) : HoleRef.t('a) -> 
                 t(result('a))
  val bind : t('a) -> 
             ('a -> t('b)) -> 
             t('b)
  val return : 'a -> t('a)
}
\end{lstlisting}
\caption{Modules for working with hole refs}
\label{fig:HoleRefs}
\end{figure}

\subsection{Background: Elm architecture}

\subsection{GUIs with Typed Holes}

\subsection{Palette-Specific Actions?}

\subsection{Reasoning Principles}

\subsection{Deriving Palettes from Type Definitions?}

\clearpage
\input{formal-system}

\clearpage
\input{evaluation}

\clearpage
\section{Related Work}
Graphite, Relit, projectional editors, notebooks / Mathematica,

\url{https://twitter.com/johnregehr/status/1095018518737637376} -- ascii art in source code

\url{http://idl.cs.washington.edu/papers/lyra/} - plotting UIs

\begin{figure*}
\begin{lstlisting}
                  Graphite    Projectional Editors    TLMs     Live Palette Expressions
interactive       Y           Y                       N        Y
extensible        Y           N                       Y        Y
persistent        N           Y                       Y        Y
compositional     N           N                       Y        Y
live              N           N                       N        Y
formalized        N           N                       Y        Y
\end{lstlisting}
\caption{Related Work Table}
\label{fig:related-work}
\end{figure*}

\section{Discussion}

\parahead{User Interface Design}

\begin{itemize}

\item
%
our \Hazel{} and \sns{} prototypes demonstrated two simple approaches to palette
layout: inline and floating, respectively.
%
might also want user-controlled layout and pinning.

\end{itemize}

\parahead{Deriving Palettes from Type Definitions}

\begin{itemize}

\item \cite{SSS}

\end{itemize}

\section{Conclusion}

\begin{quote}
%
\textit{
%
``The arithmetical symbols are written diagrams and the geometrical figures are graphic formulas.''
%
}

\vspace{3pt}

\hfill{}--- David Hilbert~\cite{XXX}
\end{quote}

%\clearpage
\bibliography{references,all.short,hazel_NSF}

\clearpage
\appendix
\input{fig-expand-appendix}

\end{document}

### 2020-05-14 abstract.tex
near-finished intro + more detailed outline for remainder of paper

@@
     Although textual program manipulation is powerful, 
-    some types of expressions are more naturally represented and manipulated
+    some types of expressions are more naturally manipulated
     graphically.     
@@
     This paper introduces \emph{live literals}, or \emph{livelits}, which allow
-    a programmer to fill holes of types like these by directly manipulating a corresponding    
+    a programmer to fill holes of types like these by directly manipulating a     
     GUI embedded persistently into the code. 
     Livelits are compositional: the GUI can itself contain
-    spliced expressions, which in turn can contain other livelits. Splices are typed, and the system 
+    spliced expressions, which in turn can contain other livelits. 
+    These splices are typed, and the system 
     requires that the livelit treat each splice hygienically. 
@@
     even when the spliced expressions mention bound variables, because the system gathers closures associated with the hole that the livelit is tasked with filling.
-    This paper details the livelits mechanism with several non-trivial case studies. 
-    It then defines a typed lambda calculus that captures the essence of the mechanism.     
+    This paper introduces livelits with several non-trivial case studies that exercise these capabilities. 
     Our implementation extends Hazel, a web-based live programming environment able to assign
     static and dynamic meaning to programs with holes. 
+    We then define a typed lambda calculus that captures the essence of livelits, which
+    operate as graphical macros.     
+    The formalism is mechanized in Agda.
 \end{abstract}

### 2020-05-14 intro.tex
near-finished intro + more detailed outline for remainder of paper

--- 2019-02-26 intro.tex
@@
-\section{Introduction}
-\label{sec:intro}
+\section{Introduction}\label{sec:intro}
+Textual program editors are powerful and expressive user interfaces, 
+so it is little wonder that they remain dominant decades after the teletype. 
+However, textual user interfaces are not the best tool for every job. 
+In particular, there are many 
+data types for which a non-textual  
+user interface may situationally be more appropriate \cite{Graphite}.
 
+For example, consider the following record type%
+\footnote{We will assume that we are in an ML-family language, 
+e.g. Elm \cite{Elm,ElmArchitecture,czaplicki2012elm}, 
+for the remainder of this paper, though the ideas are general.} 
+classifying colors using an RGBA encoding:
+\begin{lstlisting}[numbers=none]
+type Color = { r: Int, g: Int, b: Int, a: Int }
+\end{lstlisting}
+It is possible to select a particular color by entering  
+an expression of this type using a text editor:
+\begin{lstlisting}[numbers=none]
+let bgcolor : Color = { r: 255, g: 178, b: 45, a: 100 }
+\end{lstlisting}
+The problem is that this \emph{de facto} user interface for color selection 
+offers the user no immediate feedback 
+and limited editing affordances.
+It is difficult for the programmer, or anyone subsequently reading the code, to know which color is represented 
+and to interactively tweak that color. 
+Better support for activities like these would be particularly helpful when engaging in 
+live and exploratory programming in domains like web design, audio-visual production, 
+and interactive data analysis.
+
+Indeed, practitioners in domains like these commonly eschew general-purpose programming 
+in favor of graphical end-user applications, like %
+image and video editors, spreadsheets, and music composition software, 
+that do provide live feedback and 
+direct manipulation affordances, e.g. color palettes, visual timelines, and interactive plots. 
+The trade-off is that these end-user applications have limited support for abstraction and composition. 
+It is difficult, for example, to bind a
+color to a variable for use in multiple locations, 
+or to compute a darkened color by passing it to a function, 
+or to organize and distribute color schemes using a package manager.
+Moreover, it is difficult to compose affordances in ways that the application designer has not 
+anticipated, 
+e.g. to use a slider to manipulate the alpha channel 
+if the native color palette provides only a text box\todo{better example?}{}.
+
+This paper aims to resolve this apparent tension between 
+programmatic and direct manipulation interfaces by designing a programming system that 
+is able to surface GUIs when working with the data types for which 
+they are useful, while retaining support for textual program manipulation  
+and the full array of abstraction and composition mechanisms
+available in modern general-purpose programming languages.
+
+\subsection{Background}
+\begin{figure*}
+  \begin{center}
+  \includegraphics[width=35pc]{color-palettes.png}\end{center}
+  \caption{
+  (a) This figure, adapted from the prior work of \citet{Graphite}, 
+  shows a Graphite palette associated with the \texttt{Color} class. 
+  The client can specify the RGBA components only using literal numbers. 
+  (b) A similar Hazel livelit for colors uses live splices for the RGBA components,
+  so the client can enter any Hazel expression. 
+  In this case, the client has entered a variable into the RGB 
+  splices, which itself was defined using a slider livelit. 
+  The livelit can evaluate splices under its closure to display the 
+  computed color. 
+  The value of \li{defaultColor} is determined by the livelit via a macro expansion step.}
+  \label{fig:color}
+\end{figure*}
+
+We are not, of course, the first to integrate direct manipulation user interfaces
+into general-purpose programming systems.
+Prior work on projectional editing
+and active code completion, summarized in Sec. \ref{sec:related-work}\todo{sections}{}, 
+has also considered the problem of entering expressions 
+of certain types, like \li{Color}, 
+using specialized GUIs integrated into the editor. 
+The prior work most closely related to this paper is the {Graphite} system for Java \cite{Graphite}.
+Graphite allows library providers to associate a GUIs, called \emph{palettes}, with types (using Java's class annotations). 
+Wherever an expression of a type equipped with a palette is needed, 
+i.e. wherever there is a \emph{hole} of that type in the program, 
+the programming environment offers the programmer the option, via the code completion menu (not shown), 
+to activate the palette. 
+It is the palette's job to generate an 
+appropriate expression to fill the hole once the interaction is finished. 
+Figure~\ref{fig:color}(a)\todo{use subfigure?}{}, adapted from this prior work, demonstrates a simple color palette.
+When the user presses the \li{Enter} key, the Java code \li{new Color(0, 0, 128)}\todo{correct code}{} is inserted at the cursor (not shown) and the palette disappears.
+
+\citet{Graphite} extensively evaluated this mechanism. 
+Most notably, a survey of approximately 450 developers found that 
+most developers viewed the proposed mechanism favorably and stated that they 
+would use a suitable palette some or all of the time if one was available.%
+\footnote{When presented with a color palette, 
+many participants remarked that they rarely entered colors directly into Java code, 
+but rather into stylesheets or other files governed by a domain-specific language. 
+Other palettes, e.g. a palette
+that supported regular expression construction and testing, were viewed as more 
+suitable for Java code. 
+The mechanisms being considered are suitable both for general-purpose languages
+and typed domain-specific languages, many of which can be embedded into modern 
+general-purpose languages, e.g. Typed CSS in Reason. TODO: CITE}
+
+This survey also solicited a large number use cases from the survey participants, 
+which the authors grouped into several broad categories. Notable examples include:
+\begin{itemize}
+  \item \textbf{graphical elements}, such as brushes, fonts, polygon editors, user interface widgets, and 3D primitives;
+  \item \textbf{data analysis primitives}, such as plot setups and database queries; 
+  \item \textbf{data structures} such as matrices, maps, queues, and others for which specialized notation (e.g. JSON) may be more suitable than general-purpose notation; and
+  \item \textbf{data transformations} for which sensory feedback could aid in understanding, 
+  e.g. audio filters, image transformations, and animations with parameters such as speed and shape.
+\end{itemize}
+
+We take this extensive empirical evaluation as sufficient to establish 
+the utility and broad applicability of a mechanism of this sort.
+
+\subsection{Contributions}
+
+Our focus in this paper is on addressing a number of non-trivial technical 
+deficiencies that limit both palette providers and palette clients in Graphite. 
+The system of \emph{live literals}, or \emph{livelits}, introduced in
+this work and demonstrated in Fig.~\ref{fig:color}(b), differs from Graphite  
+in that it achieves each of the following goals.
+\begin{enumerate}
+  \setlength\itemsep{0.5em}
+  \item \textbf{Decentralized Extensibility}: 
+    In Graphite, there can be at most one palette per type, 
+    and it must be associated with that type at its definition site, i.e. by the provider of that type.
+    By contrast, livelit definitions stand apart from type definitions 
+    and clients invoke livelits explicitly by name (names are prefixed with the \li{\$} symbol, 
+    e.g. \li{\$color} and \li{\$slider} in Fig.~\ref{fig:color}(b), to distinguish
+    livelit names from standard variables), so these limitations do not apply.
+    We call this \emph{decentralized extensibility}.
+  \item \textbf{Persistence}: Graphite palettes are {ephemeral}, 
+  i.e. they disappear once the initial interaction is complete, 
+  leaving behind only the generated textual code.\footnote{Graphite does include an \emph{ad hoc} mechanism that allows palettes to parse the code that is selected in the editor 
+  when the palette loads, but this requires that each palette implement a parser for the subset of Java used in the code that it generates,
+  and therefore this mechanism is quite brittle. It is also difficult to persist GUI state that is not included in the generated code.}
+  This means that only the programmer that initially enters the expression 
+  benefits from the feedback and affordances provided by the palette. 
+  By contrast, livelits are persistent. We use a pure functional model-view-update architecture 
+  (i.e. the Elm architecture\todo{what to cite?}{}) as our GUI framework, 
+  so only the model needs to be persisted, rather than the full GUI state.  
+  We use the word ``literal'' rather than ``palette'' because, with support for persistence, livelits
+  are more analogous to literal notation, e.g. list literals, than palettes as they exist in graphical end-user 
+  applications.
+
+  \item \textbf{Compositionality}: 
+  Graphite does not provide any way to {enter subexpressions within a palette}.
+  At best, they can contain text boxes, but expressions entered into these text boxes 
+  do not have any editor support (which creates many problems, one of which is that they cannot themselves be generated by palettes). 
+  This implies that palettes can only reasonably generate {closed expressions}. 
+  For example, in Figure~\ref{fig:color}(a), there is no way to specify that a color component 
+  should be computed by a specified expression---only literal numbers are supported, so the color palette
+  is useful only for generating simple color constants.
+  By contrast, livelits allow subexpressions, which we call \emph{splices} following
+  the terminology of \citet{TLMs} who developed reasoning principles for splicing in a purely textual setting.
+  Fig.~\ref{fig:color}(b) demonstrates splicing: the client defines a variable, \li{grayLevel}, 
+  for each of the red, green, and blue components, and places a slider inline in the splice governing 
+  the alpha (A) component. In Sec.~X\todo{section}{}, we will see more sophisticated examples.
+  \item \textbf{Liveness}: Graphite's palettes can only provide static feedback, because they do not have 
+  access to run-time information. By contrast, livelits can request the evaluation result for splices,
+  using it to provide live feedback, i.e. feedback that takes into account the run-time environments 
+  around the livelit. For example, in Fig.~\ref{fig:color}, the color displayed requires evaluating the RGBA 
+  component splices to numeric values. We will see more sophisticated examples later in the paper, including examples
+  that appear inside functions.
+  \item \textbf{Parameterization}: 
+    \cyrus{... this would be really nice to include + demonstrate with slider bounds and later data flow + 
+    we need it to do anything non-trivial with hygiene 
+    ...}
+\end{enumerate}
+\clearpage
+
+The remainder of the paper is organized as follows. We begin in Sec.~2 with two case studies 
+involving several non-trivial livelits, focusing on the client programmer's perspective. 
+We then consider the livelit provider's perpsective in Sec.~3.
+In Sec.~4, we provide a formal account of the mechanism as a typed lambda calculus. 
+This calculus formally specifies the essential aspects of livelit representation and 
+the livelit expansion process, independent of the particularities of our implementation, 
+and establishes essential type safety and hygiene properties. 
+In Sec.~5, we describe some extensions of the core mechanism\todo{what are the extensions?}{}.
+In Sec.~6, we outline our implementation of livelits within Hazel.
+In Sec.~7, we situate livelits relative to other related work. 
+Finally, we conclude in Sec.~8 with a discussion of some present limitations and future work.\todo{section refs}{}
+
+\section{Case Studies}
+\subsection{Hazel Overview}
+Brief overview (1/2 page) on Hazel (what do holes look like? structure editing, but not in too much detail,
+live evaluation)
+
+\subsection{Live Grading}
+\subsubsection{Overview}
+Talk about the scenario
+
+Live matrix: what is a matrix structurally? (data frame? list of lists?) + how does liveness work
++ overview of hygiene guarantees
+
+Grade cutoffs: how does it work wrt liveness? (do we use a parameter here instead of a splice?) -- 
+focus on domain-specific benefits here.
+
+\subsection{Image Transformation Pipeline}
+Show example of an image transformation pipeline going through livelits
+
+This might also benefit from Parameterization
+
+Show multiple calls with different example images + closure selector UIs
+
+Go into more detail about how evaluation works + fill-and-resume mechanics (and efficiency nod)
+
+Talk about probes?
+
+\subsection{Additional Examples}
+It would be nice to have a gallery-style figure and a brief overview of some other case studies
+and how they exercise the novel features of the livelits mechanism. Maybe some statistics on how
+many lines of code it took.
+
+\section{Livelit Implementation}
+Which livelit implementation do we want to show? Maybe grade cutoffs? Slider as simple intro + color chooser?
+
+Talk about each component of a livelit definition:
+\begin{itemize}
+  \item livelit name (talk about binding structure)
+  \item expansion type
+  \item model type
+  \item action type (mention integration into Hazel's undo system)
+  \item initial model
+  \item update function (monadic actions; talk about color palette update; SpliceRefs)
+  \item view function (monadic actions; talk about GUI stuff? tab order? dimensions? mostly defer to Elm for details)
+  \item expand function (should we call it fill?)
+\end{itemize}
+
+Hygiene from a provider's perspective -- have to use parameters to explicitly capture definition site bindings.
+
+\section{A Typed Livelit Calculus}
+Should we talk about anything other than expansion + hygiene, i.e. parameterization, closure collection, closure resumption?
+None of that is mechanized, but that may be okay. Closure collection and resumption seem to be worth formalizing.
+
+Note: probably just going to omit the whole livelit function thing
+
+Subsection about Agda mechanization (Ian + Nick?)
+
+\section{Extensions}
+Is this section necessary?
+
+\section{Implementation Considerations}
+What should we talk about here? Layout of livelits?
+
+\section{Related Work}
+Cyrus has a list somewhere, ask if you want to work on this section
+
+\section{Discussion and Conclusion}
+Future work:
+\begin{itemize}
+  \item pattern matching
+  \item type splices
+  \item module splices
+  \item explicitly make bindings available in splices
+  \item better UI for closure provenance / connect with control flow / call stack better
+  \item deriving livelits from type definitions
+  \item bidirectional evaluation ala SnS
+  \item integration with structure editing more cleanly
+  \item \dots
+  \item collaborative editing?
+  \item side effects?
+  \item big vision: authoring environment based on livelits
+\end{itemize}
+\clearpage
+\appendix
+\section{Misc. Material from Prior Draft}
+
+To address these limitations, we introduce live literals, or livelits.
+Figure~\ref{fig:cutoffs} shows a mockup of the definition and application of a livelit named \li{$grade_cutoffs} for adjusting grade cutoffs, represented as values of  record type \li{grade_cutoffs}.
+Like textual literal forms (e.g.
+list literals), livelits are alternative representations of expressions of the associated type \cite{DBLP:journals/pacmpl/OmarA18}.
+% From the perspective of the remainder of the program, \li{cutoffs} is a value of type \li{grade_cutoffs} like any other.
+We are implementing livelits in Hazel (\url{hazel.org}), a live functional programming environment with support for typed holes \cite{popl-paper}. We plan to perform a live demo.
+
+The definition of
+the livelit, outlined in Figure~\ref{fig:cutoffs}, follows the Elm architecture,
+i.e. there are types representing the abstract model and the messages that the 
+GUI generates.
+Livelits are \textbf{persistent} rather than ephemeral, i.e.
+the model is recorded in the underlying syntax tree.
+The \li{view} function generates the GUI, implemented using HTML, on demand (so the view is not persisted).
+Another function, \li{to_exp}, is responsible for generating the underlying expression, called the \emph{expansion}, from the \li{model}.
+While other projectional editors, e.g. those generated by Citrus \cite{DBLP:conf/uist/KoM05}, also support persistent GUIs in code, they are not user-extensible.
+
+The GUI can itself contain typed holes, represented by the \li{HtmlHole} constructor. In this case, there is a single hole
+for entering an expression of type \li{list(float)}, i.e. the list of weighted
+averages. In this example, the user has filled this hole with a variable,
+\li{weighted_averages} (the definition of which precedes the content of this figure and is not shown). In other words,
+livelits support \textbf{open expressions} and therefore interact cleanly with 
+standard abstraction mechanisms, i.e. they can appear under binders. We follow the reasoning principles for literals with spliced sub-expressions established  by \citet{DBLP:journals/pacmpl/OmarA18}.
+
+The main complication when dealing with open expressions relates to how live feedback
+is to be generated. Given just the symbolic expression in the hole, it would be 
+impossible to plot (as orange dots) the actual data from the list that the variable refers to. To resolve this issue, the system evaluates the program 
+as if it the livelits were empty holes, 
+relying on the support for evaluating incomplete programs described recently 
+by \citet{DBLP:journals/pacmpl/OmarVCH19}. The result of evaluation is an expression containing
+hole closures, i.e. holes equipped with environments. 
+%Hazel allows the user to interactively select which closure
+%interests them if multiple closures for a given hole appear in the result (e.g. if a livelit appears inside a function called multiple times). 
+The livelit 
+can then evaluate the expression in the hole against the closure selected by the user (not shown) using \li{Env.run}.
+
+%%%%%
+
+\todo{this is the WIP one from last year}{}
 Programs are most often represented as text.
@@
 
+prototype UIs:
+
+  \begin{enumerate}
+    \item \Hazel{}: (palette macros;) structure edit actions; expression formula bar; inline layout
+    \item \sns{}: (palette functions;) text editor; pop-up menu
+  \end{enumerate}
+
 hazelnut~\citep{Hazelnut,HazelnutLive}
+
+\begin{figure*}[t!]
+  \begin{center}
+  \includegraphics[width=34pc]{cutoffs-mockup.png}\end{center}
+  \caption{A mockup of a livelit for adjusting grade cutoffs. Livelits are persistent and can access the live environment.}
+  \label{fig:cutoffs}
+  \end{figure*}

### 2020-05-14 livelits-paper.tex
Update author info
near-finished intro + more detailed outline for remainder of paper

@@
 \PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
-
-%% % From https://pldi19.sigplan.org/track/pldi-2019-papers#Call-for-Papers
-%% \documentclass[sigplan,10pt,review,anonymous]{acmart}
-%% \settopmatter{printfolios=true,printccs=false,printacmref=false}
-
-%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
 \documentclass[acmsmall,review,anonymous,nonacm]{acmart}
@@
 
+\newcommand{\cyrus}[1]{{\color{blue} #1}}
+
 \usepackage{listings}%
@@
 numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML, showspaces=false,showstringspaces=false,xleftmargin=15pt, 
-morekeywords={string, float, int},
+morekeywords={string, float, int, Int, Float, String},
 classoffset=0,belowskip=\smallskipamount, aboveskip=\smallskipamount,

### 2020-05-15 intro.tex
add missing images
cleaned up contribution summary

@@
 
-Our focus in this paper is on addressing a number of non-trivial technical 
+Our focus in this paper is on addressing several non-trivial technical 
 deficiencies that limit both palette providers and palette clients in Graphite. 
 The system of \emph{live literals}, or \emph{livelits}, introduced in
-this work and demonstrated in Fig.~\ref{fig:color}(b), differs from Graphite  
-in that it achieves each of the following goals.
+this paper and demonstrated in Fig.~\ref{fig:color}(b), differs from Graphite  
+in each of the following ways.
 \begin{enumerate}
@@
   \item \textbf{Decentralized Extensibility}: 
-    In Graphite, there can be at most one palette per type, 
-    and it must be associated with that type at its definition site, i.e. by the provider of that type.
-    By contrast, livelit definitions stand apart from type definitions 
-    and clients invoke livelits explicitly by name (names are prefixed with the \li{\$} symbol, 
-    e.g. \li{\$color} and \li{\$slider} in Fig.~\ref{fig:color}(b), to distinguish
-    livelit names from standard variables), so these limitations do not apply.
+    Graphite allows only the provider of a type to define a corresponding palette. 
+    By contrast, livelit definitions are orthogonal to type definitions. 
+    Clients invoke livelits explicitly by name (names are prefixed by the \li{\$} symbol, 
+    e.g. \li{\$color} and \li{\$slider} in Fig.~\ref{fig:color}(b)).
     We call this \emph{decentralized extensibility}.
-  \item \textbf{Persistence}: Graphite palettes are {ephemeral}, 
-  i.e. they disappear once the initial interaction is complete, 
-  leaving behind only the generated textual code.\footnote{Graphite does include an \emph{ad hoc} mechanism that allows palettes to parse the code that is selected in the editor 
-  when the palette loads, but this requires that each palette implement a parser for the subset of Java used in the code that it generates,
-  and therefore this mechanism is quite brittle. It is also difficult to persist GUI state that is not included in the generated code.}
-  This means that only the programmer that initially enters the expression 
-  benefits from the feedback and affordances provided by the palette. 
+  \item \textbf{Persistence}: Graphite's palettes are {ephemeral}, 
+  i.e. they disappear after the initial interaction, 
+  leaving behind only the generated textual code.%
+  \footnote{Graphite does include an \emph{ad hoc} mechanism that 
+  allows palettes to parse the code that is selected in the editor 
+  when the palette loads, but this requires that each palette implement 
+  a parser for the subset of Java used in the code that it generates,
+  and therefore this mechanism is quite brittle. It is also difficult 
+  to persist GUI state that is not included in the generated code.}
+  Consequently, only the programmer that initially enters the expression 
+  benefits from the feedback and affordances that the palette provides.
   By contrast, livelits are persistent. We use a pure functional model-view-update architecture 
@@
   so only the model needs to be persisted, rather than the full GUI state.  
-  We use the word ``literal'' rather than ``palette'' because, with support for persistence, livelits
-  are more analogous to literal notation, e.g. list literals, than palettes as they exist in graphical end-user 
-  applications.
+  We use the word ``literal'' rather than ``palette'' because, with persistence, livelits
+  are more analogous to literal notation, e.g. list literals.
 
@@
   Graphite does not provide any way to {enter subexpressions within a palette}.
-  At best, they can contain text boxes, but expressions entered into these text boxes 
+  At best, palettes can contain text boxes, but expressions entered into these text boxes 
   do not have any editor support (which creates many problems, one of which is that they cannot themselves be generated by palettes). 
   This implies that palettes can only reasonably generate {closed expressions}. 
-  For example, in Figure~\ref{fig:color}(a), there is no way to specify that a color component 
-  should be computed by a specified expression---only literal numbers are supported, so the color palette
+  For example, in Figure~\ref{fig:color}(a), the color palette
   is useful only for generating simple color constants. 
-  By contrast, livelits allow subexpressions, which we call \emph{splices} following
-  the terminology of \citet{TLMs} who developed reasoning principles for splicing in a purely textual setting.
-  Fig.~\ref{fig:color}(b) demonstrates splicing: the client defines a variable, \li{grayLevel}, 
-  for each of the red, green, and blue components, and places a slider inline in the splice governing 
-  the alpha (A) component. In Sec.~X\todo{section}{}, we will see more sophisticated examples.
-  \item \textbf{Liveness}: Graphite's palettes can only provide static feedback, because they do not have 
-  access to run-time information. By contrast, livelits can request the evaluation result for splices,
-  using it to provide live feedback, i.e. feedback that takes into account the run-time environments 
-  around the livelit. For example, in Fig.~\ref{fig:color}, the color displayed requires evaluating the RGBA 
-  component splices to numeric values. We will see more sophisticated examples later in the paper, including examples
-  that appear inside functions.
+  By contrast, livelits have first-class support for subexpressions, which we call \emph{splices} (following
+  the terminology of \citet{TLMs} who developed reasoning principles for splicing in a purely textual setting).
+  Fig.~\ref{fig:color}(b) demonstrates splicing: the client is able to define a variable, \li{grayLevel}, 
+  to capture the relationship between the red, green, and blue components.
+  The client is also able to use a slider inline in the splice governing 
+  the alpha (A) component.
+  \item \textbf{Liveness}: Graphite's palettes do not have 
+  access to the run-time environment. By contrast, livelits can evaluate splices
+  and therefore provide feedback related to the run-time behavior of the expression being generated. 
+  For example, in Fig.~\ref{fig:color}(b), the color that is displayed is determined by evaluating the RGBA 
+  component splices to numeric values.
+  Evaluation occurs in a run-time environment (i.e. closure) determined by 
+  leaving the hole being filled by the livelit temporarily unfilled and then evaluating
+  using a two-phased variation on the semantics for incomplete programs developed by \citet{HazelnutLive}. 
+  We will see more sophisticated examples later in the paper, including a livelit 
+  that appears inside a function that is called multiple times, leading to multiple closures for the client 
+  select from.
   \item \textbf{Parameterization}: 
@@
 
-\section{Livelit Implementation}
+\section{Livelit Definitions}
 Which livelit implementation do we want to show? Maybe grade cutoffs? Slider as simple intro + color chooser?
@@
   \item expansion type
-  \item model type
+  \item model type (must be serializable)
   \item action type (mention integration into Hazel's undo system)
@@
   \item side effects?
+  \item full screening livelits (or compositions thereof) as a way to create end-user workflows
   \item big vision: authoring environment based on livelits

### 2020-05-15 intro.tex
added discussion of parameterization to intro

@@
   select from.
-  \item \textbf{Parameterization}: 
-    \cyrus{... this would be really nice to include + demonstrate with slider bounds and later data flow + 
-    we need it to do anything non-trivial with hygiene 
-    ...}
+  \item \textbf{Parameterization}: Finally, it is possible to define parameterized families of livelits. 
+  For example, the \li{\$slider} livelit in Fig.~\ref{fig:color}(b) is parameterized by the lower and upper bound. 
+  Parameters behave like splices, differing in that (1) livelits support partial parameter application, and (2) 
+  livelit applications can be abbreviated. For example, it is possible to define a livelit named \li{\$percentage} 
+  that operates as a \li{\$slider} from \li{0} to \li{100} by abbreviation:
+  \begin{lstlisting}[numbers=none]
+  let livelit $percentage = $slider(0, 100) in ...
+  \end{lstlisting}
+  Parameterization also underlies the hygiene mechanism, as we will discuss in Sec. 4\todo{sec}{}.
 \end{enumerate}

### 2020-05-15 livelits-paper.tex
added discussion of parameterization to intro

@@
 numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML, showspaces=false,showstringspaces=false,xleftmargin=15pt, 
-morekeywords={string, float, int, Int, Float, String},
+morekeywords={string, float, int, Int, Float, String, livelit},
 classoffset=0,belowskip=\smallskipamount, aboveskip=\smallskipamount,
@@
   % \department{Department1}              %% \department is recommended
-  \institution{University of Colorado Boulder}            %% \institution is required
+  \institution{University of Michigan}            %% \institution is required
   % \streetaddress{Street1 Address1}
@@
 }
-\email{domo9239@colorado.edu}          %% \email is recommended
+\email{dmoo@umich.edu}          %% \email is recommended
 

### 2020-05-15 intro.tex
move parameterization next to compositionality

@@
   the alpha (A) component.
+
+  \item \textbf{Parameterization}: It is possible to define parameterized families of livelits. 
+  For example, \li{\$slider} in Fig.~\ref{fig:color}(b) is parameterized by the lower and upper bound. 
+  Parameters behave like splices, differing in that (1) livelits support partial parameter application, and (2) 
+  livelit applications can be abbreviated. For example, it is possible to define a livelit named \li{\$percentage} 
+  that operates as a \li{\$slider} from \li{0} to \li{100} by scoped abbreviation:
+  \begin{lstlisting}[numbers=none]
+  let livelit $percentage = $slider(0, 100) in ...
+  \end{lstlisting}
+  Parameterization also underlies the hygiene mechanism, as we will discuss in Sec. 4\todo{sec}{}.
+
   \item \textbf{Liveness}: Graphite's palettes do not have 
@@
   We will see more sophisticated examples later in the paper, including a livelit 
-  that appears inside a function that is called multiple times, leading to multiple closures for the client 
+  that appears inside a function that is called multiple times, leading to multiple closures for the client to 
   select from.
-  \item \textbf{Parameterization}: Finally, it is possible to define parameterized families of livelits. 
-  For example, the \li{\$slider} livelit in Fig.~\ref{fig:color}(b) is parameterized by the lower and upper bound. 
-  Parameters behave like splices, differing in that (1) livelits support partial parameter application, and (2) 
-  livelit applications can be abbreviated. For example, it is possible to define a livelit named \li{\$percentage} 
-  that operates as a \li{\$slider} from \li{0} to \li{100} by abbreviation:
-  \begin{lstlisting}[numbers=none]
-  let livelit $percentage = $slider(0, 100) in ...
-  \end{lstlisting}
-  Parameterization also underlies the hygiene mechanism, as we will discuss in Sec. 4\todo{sec}{}.
 \end{enumerate}

### 2020-05-24 intro.tex
more tweaks to intro

@@
 for the remainder of this paper, though the ideas are general.} 
-classifying colors using an RGBA encoding:
+classifying colors using the RGBA encoding:
 \begin{lstlisting}[numbers=none]
@@
 and limited editing affordances.
-It is difficult for the programmer, or anyone subsequently reading the code, to know which color is represented 
+It is difficult for the programmer, or anyone subsequently reading or modifying the code, to know which color is represented 
 and to interactively tweak that color. 
-Better support for activities like these would be particularly helpful when engaging in 
-live and exploratory programming in domains like web design, audio-visual production, 
-and interactive data analysis.
+Better support for activities like these would be particularly helpful for users engaging in 
+live and exploratory programming in domains like web design, media production, 
+and data analysis.
 
@@
 image and video editors, spreadsheets, and music composition software, 
-that do provide live feedback and 
+in part because these applications provide domain-specific forms of live feedback and 
 direct manipulation affordances, e.g. color palettes, visual timelines, and interactive plots. 
-The trade-off is that these end-user applications have limited support for abstraction and composition. 
+The trade-off is that these applications have limited support for abstraction and composition. 
 It is difficult, for example, to bind a
 color to a variable for use in multiple locations, 
-or to compute a darkened color by passing it to a function, 
+or to computationally darken a color by passing it to a function, 
 or to organize and distribute color schemes using a package manager.
-Moreover, it is difficult to compose affordances in ways that the application designer has not 
-anticipated, 
+Moreover, it is difficult to add new feedback and affordances and to compose 
+affordances in ways that the application designer has not anticipated, 
 e.g. to use a slider to manipulate the alpha channel 
@@
 they are useful, while retaining support for textual program manipulation  
-and the full array of abstraction and composition mechanisms
-available in modern general-purpose programming languages.
+and the abstraction and composition mechanisms
+available in modern general-purpose languages.
 
@@
   so the client can enter any Hazel expression. 
-  In this case, the client has entered a variable into the RGB 
-  splices, which itself was defined using a slider livelit. 
-  The livelit can evaluate splices under its closure to display the 
+  In this case, the client has entered a variable, itself defined using a slider livelit, into the RGB 
+  splices. The A splice is filled directly with a slider livelit. 
+  The color livelit evaluates splices under its closure to display the 
   computed color. 
@@
 of certain types, like \li{Color}, 
-using specialized GUIs integrated into the editor. 
+using specialized GUIs integrated into an editor. 
 The prior work most closely related to this paper is the {Graphite} system for Java \cite{Graphite}.
-Graphite allows library providers to associate a GUIs, called \emph{palettes}, with types (using Java's class annotations). 
+Graphite allows library providers to associate GUIs, called \emph{palettes}, with types (using Java's class annotations). 
 Wherever an expression of a type equipped with a palette is needed, 
 i.e. wherever there is a \emph{hole} of that type in the program, 
-the programming environment offers the programmer the option, via the code completion menu (not shown), 
+the programming environment offers the programmer the option, via the code completion menu, 
 to activate the palette. 
-It is the palette's job to generate an 
-appropriate expression to fill the hole once the interaction is finished. 
-Figure~\ref{fig:color}(a)\todo{use subfigure?}{}, adapted from this prior work, demonstrates a simple color palette.
+Once the interaction is finished, it is the palette's job to generate an 
+appropriate Java expression to fill the hole. 
+Figure~\ref{fig:color}(a)\todo{use subfigure?}{}, adapted from this prior work, demonstrates a simple color palette invoked using Graphite.
 When the user presses the \li{Enter} key, the Java code \li{new Color(0, 0, 128)}\todo{correct code}{} is inserted at the cursor (not shown) and the palette disappears.
@@
 Most notably, a survey of approximately 450 developers found that 
-most developers viewed the proposed mechanism favorably and stated that they 
+most of these developers viewed the proposed mechanism favorably and stated that they 
 would use a suitable palette some or all of the time if one was available.%
@@
 
-This survey also solicited a large number use cases from the survey participants, 
+This survey also solicited a large number of use cases from the survey participants, 
 which the authors grouped into several broad categories. Notable examples include:
@@
   \item \textbf{data analysis primitives}, such as plot setups and database queries; 
-  \item \textbf{data structures} such as matrices, maps, queues, and others for which specialized notation (e.g. JSON) may be more suitable than general-purpose notation; and
+  \item \textbf{data structures} such as matrices, maps, queues, and others for which specialized notation (e.g. JSON) and affordances may be more suitable than general-purpose mechanisms; and
   \item \textbf{data transformations} for which sensory feedback could aid in understanding, 
@@
 
-We take this extensive empirical evaluation as sufficient to establish 
+We take this extensive prior empirical evaluation as sufficient to establish 
 the utility and broad applicability of a mechanism of this sort.
@@
 deficiencies that limit both palette providers and palette clients in Graphite. 
-The system of \emph{live literals}, or \emph{livelits}, introduced in
-this paper and demonstrated in Fig.~\ref{fig:color}(b), differs from Graphite  
+The system of \emph{live literals}, or \emph{livelits} introduced in
+this paper and demonstrated in Fig.~\ref{fig:color}(b) improves upon Graphite  
 in each of the following ways.
@@
   \item \textbf{Decentralized Extensibility}: 
-    Graphite allows only the provider of a type to define a corresponding palette. 
+    In Graphite, only one palette can be associated with a type, and this must
+    be done at the definition site.
     By contrast, livelit definitions are orthogonal to type definitions. 
-    Clients invoke livelits explicitly by name (names are prefixed by the \li{\$} symbol, 
-    e.g. \li{\$color} and \li{\$slider} in Fig.~\ref{fig:color}(b)).
+    Clients invoke livelits explicitly via their names, which are prefixed by the \li{\$} character, 
+    e.g. \li{\$color} and \li{\$slider} in Fig.~\ref{fig:color}(b).
     We call this \emph{decentralized extensibility}.
@@
   i.e. they disappear after the initial interaction, 
-  leaving behind only the generated textual code.%
+  leaving behind only the generated textual code.
+  Consequently, only the programmer that initially enters the expression 
+  benefits from the feedback and affordances that the palette provides.%
   \footnote{Graphite does include an \emph{ad hoc} mechanism that 
@@
   to persist GUI state that is not included in the generated code.}
-  Consequently, only the programmer that initially enters the expression 
-  benefits from the feedback and affordances that the palette provides.
   By contrast, livelits are persistent. We use a pure functional model-view-update architecture 
@@
   We use the word ``literal'' rather than ``palette'' because, with persistence, livelits
-  are more analogous to literal notation, e.g. list literals.
+  operate more like graphical literal notation, e.g. list literals.
 
@@
   By contrast, livelits have first-class support for subexpressions, which we call \emph{splices} (following
-  the terminology of \citet{TLMs} who developed reasoning principles for splicing in a purely textual setting).
+  the terminology of \citet{TLMs} who developed reasoning principles for splicing in a strictly textual setting).
   Fig.~\ref{fig:color}(b) demonstrates splicing: the client is able to define a variable, \li{grayLevel}, 
@@
   For example, \li{\$slider} in Fig.~\ref{fig:color}(b) is parameterized by the lower and upper bound. 
-  Parameters behave like splices, differing in that (1) livelits support partial parameter application, and (2) 
-  livelit applications can be abbreviated. For example, it is possible to define a livelit named \li{\$percentage} 
+  Parameters behave like splices, differing in that livelit applications can be abbreviated, and abbreviations
+  support partial application. For example, it is possible to define a livelit named \li{\$percentage} 
   that operates as a \li{\$slider} from \li{0} to \li{100} by scoped abbreviation:
@@
   We will see more sophisticated examples later in the paper, including a livelit 
-  that appears inside a function that is called multiple times, leading to multiple closures for the client to 
-  select from.
+  that appears inside a function that is called multiple times, leading to multiple closures that the client can 
+  interactively select between.
 \end{enumerate}
-\clearpage
 
-The remainder of the paper is organized as follows. We begin in Sec.~2 with two case studies 
-involving several non-trivial livelits, focusing on the client programmer's perspective. 
-We then consider the livelit provider's perpsective in Sec.~3.
+The remainder of this paper is organized as follows. We begin in Sec.~2 with two case studies 
+involving several non-trivial livelits, focusing on the livelit client's perspective. 
+We then consider the livelit provider's perspective in Sec.~3.
 In Sec.~4, we provide a formal account of the mechanism as a typed lambda calculus. 
-This calculus formally specifies the essential aspects of livelit representation and 
-the livelit expansion process, independent of the particularities of our implementation, 
+This calculus formally specifies the essential aspects of livelit definition, representation, expansion, 
+and run-time closure collection independent of the particularities of our implementation, 
 and establishes essential type safety and hygiene properties. 
@@
 Finally, we conclude in Sec.~8 with a discussion of some present limitations and future work.\todo{section refs}{}
-
-\section{Case Studies}
-\subsection{Hazel Overview}
-Brief overview (1/2 page) on Hazel (what do holes look like? structure editing, but not in too much detail,
-live evaluation)
-
-\subsection{Live Grading}
-\subsubsection{Overview}
-Talk about the scenario
-
-Live matrix: what is a matrix structurally? (data frame? list of lists?) + how does liveness work
-+ overview of hygiene guarantees
-
-Grade cutoffs: how does it work wrt liveness? (do we use a parameter here instead of a splice?) -- 
-focus on domain-specific benefits here.
-
-\subsection{Image Transformation Pipeline}
-Show example of an image transformation pipeline going through livelits
-
-This might also benefit from Parameterization
-
-Show multiple calls with different example images + closure selector UIs
-
-Go into more detail about how evaluation works + fill-and-resume mechanics (and efficiency nod)
-
-Talk about probes?
-
-\subsection{Additional Examples}
-It would be nice to have a gallery-style figure and a brief overview of some other case studies
-and how they exercise the novel features of the livelits mechanism. Maybe some statistics on how
-many lines of code it took.
-
-\section{Livelit Definitions}
-Which livelit implementation do we want to show? Maybe grade cutoffs? Slider as simple intro + color chooser?
-
-Talk about each component of a livelit definition:
-\begin{itemize}
-  \item livelit name (talk about binding structure)
-  \item expansion type
-  \item model type (must be serializable)
-  \item action type (mention integration into Hazel's undo system)
-  \item initial model
-  \item update function (monadic actions; talk about color palette update; SpliceRefs)
-  \item view function (monadic actions; talk about GUI stuff? tab order? dimensions? mostly defer to Elm for details)
-  \item expand function (should we call it fill?)
-\end{itemize}
-
-Hygiene from a provider's perspective -- have to use parameters to explicitly capture definition site bindings.
-
-\section{A Typed Livelit Calculus}
-Should we talk about anything other than expansion + hygiene, i.e. parameterization, closure collection, closure resumption?
-None of that is mechanized, but that may be okay. Closure collection and resumption seem to be worth formalizing.
-
-Note: probably just going to omit the whole livelit function thing
-
-Subsection about Agda mechanization (Ian + Nick?)
-
-\section{Extensions}
-Is this section necessary?
-
-\section{Implementation Considerations}
-What should we talk about here? Layout of livelits?
-
-\section{Related Work}
-Cyrus has a list somewhere, ask if you want to work on this section
-
-\section{Discussion and Conclusion}
-Future work:
-\begin{itemize}
-  \item pattern matching
-  \item type splices
-  \item module splices
-  \item explicitly make bindings available in splices
-  \item better UI for closure provenance / connect with control flow / call stack better
-  \item deriving livelits from type definitions
-  \item bidirectional evaluation ala SnS
-  \item integration with structure editing more cleanly
-  \item \dots
-  \item collaborative editing?
-  \item side effects?
-  \item full screening livelits (or compositions thereof) as a way to create end-user workflows
-  \item big vision: authoring environment based on livelits
-\end{itemize}
-\clearpage
-\appendix
-\section{Misc. Material from Prior Draft}
-
-To address these limitations, we introduce live literals, or livelits.
-Figure~\ref{fig:cutoffs} shows a mockup of the definition and application of a livelit named \li{$grade_cutoffs} for adjusting grade cutoffs, represented as values of  record type \li{grade_cutoffs}.
-Like textual literal forms (e.g.
-list literals), livelits are alternative representations of expressions of the associated type \cite{DBLP:journals/pacmpl/OmarA18}.
-% From the perspective of the remainder of the program, \li{cutoffs} is a value of type \li{grade_cutoffs} like any other.
-We are implementing livelits in Hazel (\url{hazel.org}), a live functional programming environment with support for typed holes \cite{popl-paper}. We plan to perform a live demo.
-
-The definition of
-the livelit, outlined in Figure~\ref{fig:cutoffs}, follows the Elm architecture,
-i.e. there are types representing the abstract model and the messages that the 
-GUI generates.
-Livelits are \textbf{persistent} rather than ephemeral, i.e.
-the model is recorded in the underlying syntax tree.
-The \li{view} function generates the GUI, implemented using HTML, on demand (so the view is not persisted).
-Another function, \li{to_exp}, is responsible for generating the underlying expression, called the \emph{expansion}, from the \li{model}.
-While other projectional editors, e.g. those generated by Citrus \cite{DBLP:conf/uist/KoM05}, also support persistent GUIs in code, they are not user-extensible.
-
-The GUI can itself contain typed holes, represented by the \li{HtmlHole} constructor. In this case, there is a single hole
-for entering an expression of type \li{list(float)}, i.e. the list of weighted
-averages. In this example, the user has filled this hole with a variable,
-\li{weighted_averages} (the definition of which precedes the content of this figure and is not shown). In other words,
-livelits support \textbf{open expressions} and therefore interact cleanly with 
-standard abstraction mechanisms, i.e. they can appear under binders. We follow the reasoning principles for literals with spliced sub-expressions established  by \citet{DBLP:journals/pacmpl/OmarA18}.
-
-The main complication when dealing with open expressions relates to how live feedback
-is to be generated. Given just the symbolic expression in the hole, it would be 
-impossible to plot (as orange dots) the actual data from the list that the variable refers to. To resolve this issue, the system evaluates the program 
-as if it the livelits were empty holes, 
-relying on the support for evaluating incomplete programs described recently 
-by \citet{DBLP:journals/pacmpl/OmarVCH19}. The result of evaluation is an expression containing
-hole closures, i.e. holes equipped with environments. 
-%Hazel allows the user to interactively select which closure
-%interests them if multiple closures for a given hole appear in the result (e.g. if a livelit appears inside a function called multiple times). 
-The livelit 
-can then evaluate the expression in the hole against the closure selected by the user (not shown) using \li{Env.run}.
-
-%%%%%
-
-\todo{this is the WIP one from last year}{}
-Programs are most often represented as text.
-%
-This representation format provides expert programmers access to a variety of
-text editors and text edit actions---to affect fine-grained control over
-programming decisions---and a large ecosystem of other text-based
-utilities---for example, for line-based file differencing of source code
-versions.
-%
-On the other hand, the flexibility and low-level nature of text comes at some
-costs.
-%
-For experts, some tedious and manual edits could lead to
-inefficiency---especially because many routine code editing operations do not
-require the full flexibility of text~\citep{XXX}.
-%
-For novices, the large class of syntax errors that stem from text-edits presents
-a steep learning curve.
-
-Motivated to overcome the disadvantages of represent programs as text, a variety
-of alternative code editing interfaces have been investigated.
-%
-At the opposite end of text are \emph{visual programming languages}, which often
-completely represent a program with graphical elements~\citep{XXX}.
-%
-To forgo text completely, such approaches often target domain-specific
-languages, such as dataflow programing~\citep{XXX} and pedagogical languages
-that are, by design, restricted to relatively few building blocks.
-%
-Other \emph{structure}, or \emph{projectional} editors, still use a significant
-amount of text to render programs, but forgo text-edits in favor of
-\emph{structure edit actions} which transform the program (represented as a tree
-or some other structure), sidestepping the danger of invalid intermediate states
-of concrete syntax.
-
-On the spectrum somewhere between fully text- and fully structure-based are
-``hybrid'' editors, which augment text with additional ways to visualize and
-manipulate the structure of the program.
-%
-Victor Scrubbing, APX~\citep{APX}, Sketch-n-Sketch~\citep{sns-pldi}, and Carbide
-IDE~\citep{XXX}, among others, allow numeric values to be ``scrubbed'' by
-directly manipulating sliders rather than text-editing numeric literals.
-%
-Barista~\cite{Barista} is a hybrid Java editor where custom \emph{structure
-view} GUIs provide alternate representations of expressions instead of text.
-%
-For example, an arithmetic expression may be rendered with mathematical symbols,
-a method may be accompanied by interactive documentation with input-output
-examples, and structures may be selectively collapsed, expanded, or zoomed.
-%
-Graphite~\citep{Graphite} allows custom GUIs---called \emph{palettes}---to help
-the programmer fill missing expressions---``holes''---in the program.
-%
-For example, a color palette can provide visual previews of different candidate
-color values, and a regular expression palette can show input-output examples
-for different candidate regular expressions.
-
-The GUI representations and interactions enabled by the above hybrid editors are
-useful, but several limitations likely preclude wider utility:
-%
-the types of expressions that benefit from alternative GUI editors are limited
-to
-%
-privileged types that have baked-in interfaces~\citep{XXX}
-%
-or just of base type~\citep{XXX};
-%
-the GUI view is ephemeral, in that it disappears once it has been used to
-generate an expression~\citep{XXX}; or
-%
-expressions generated by the GUI are not deeply integrated with the static type
-system and interpreter of the language~\citep{XXX,XXX,XXX}.
-
-%% refactoring
-%% 
-%% DNDRefactoring~\citep{DNDRefactoring}
-%% Deuce~\citep{sns-deuce}
-
-\parahead{Persistent, Composable, and Live GUIs for Filling Holes}
-
-compared to the ``simple'' palettes
-
-extend palettes with: persistence, composition, and live feedback
-
-macro systems allow alternative syntactic (i.e. string) representations, to
-expand into underlying expressions
-
-palatable
-
-by analogy to these ``literal'' macro systems, palettes are ``graphical
-macros'': through interaction with the user, the GUI generates the underlying
-expression.
-
-TLMs~\cite{TLMs}
-
-\parahead{Contributions and Paper Outline}
-
-design for palettes. specifically, within the Hazelnut Live framework,
-which is a system to address the gap problem that arises in traditional editor
-frontends.
-
-demonstration of the expressiveness of the approach through a series of
-examples, many of which are drawn from the user study mentioned before.
-
-a prototype implementation of palettes within Hazel, which currently
-supports several examples using a core calculus with minimal features. the
-implementation provides a clear path for scaling up to larger, full-featured
-syntactic programming conveniences, as well as further UI design in future work.
-
-prototype UIs:
-
-  \begin{enumerate}
-    \item \Hazel{}: (palette macros;) structure edit actions; expression formula bar; inline layout
-    \item \sns{}: (palette functions;) text editor; pop-up menu
-  \end{enumerate}
-
-hazelnut~\citep{Hazelnut,HazelnutLive}
-
-\begin{figure*}[t!]
-  \begin{center}
-  \includegraphics[width=34pc]{cutoffs-mockup.png}\end{center}
-  \caption{A mockup of a livelit for adjusting grade cutoffs. Livelits are persistent and can access the live environment.}
-  \label{fig:cutoffs}
-  \end{figure*}

### 2020-05-24 case-studies.tex
separated everything into separate files

\section{Case Studies}\label{sec:case-studies}
\subsection{Hazel Overview}\label{sec:hazel-overview}
Brief overview (1/2 page) on Hazel (what do holes look like? structure editing, but not in too much detail,
live evaluation)

\subsection{Live Grading}\label{sec:live-grading}
\subsubsection{Overview}\label{sec:live-grading-overview}
Talk about the scenario

Live matrix: what is a matrix structurally? (data frame? list of lists?) + how does liveness work
+ overview of hygiene guarantees

Grade cutoffs: how does it work wrt liveness? (do we use a parameter here instead of a splice?) -- 
focus on domain-specific benefits here.

\subsection{Image Transformation Pipeline}\label{sec:transformation}
Show example of an image transformation pipeline going through livelits

This might also benefit from Parameterization

Show multiple calls with different example images + closure selector UIs

Go into more detail about how evaluation works + fill-and-resume mechanics (and efficiency nod)

Talk about probes?

\subsection{Additional Examples}\label{sec:additional-examples}
It would be nice to have a gallery-style figure and a brief overview of some other case studies
and how they exercise the novel features of the livelits mechanism. Maybe some statistics on how
many lines of code it took.

### 2020-05-24 discussion.tex
separated everything into separate files

\section{Discussion and Conclusion}\label{sec:discussion}\label{sec:conclusion}
Future work:
\begin{itemize}
  \item pattern matching
  \item type splices
  \item module splices
  \item explicitly make bindings available in splices
  \item better UI for closure provenance / connect with control flow / call stack better
  \item deriving livelits from type definitions
  \item bidirectional evaluation ala SnS
  \item integration with structure editing more cleanly
  \item \dots
  \item collaborative editing?
  \item side effects?
  \item full screening livelits (or compositions thereof) as a way to create end-user workflows
  \item big vision: authoring environment based on livelits
\end{itemize}

### 2020-05-24 implementation.tex
separated everything into separate files

\section{Implementation}\label{sec:implementation}
What should we talk about here? Layout of livelits?

### 2020-05-24 livelit-calculus.tex
separated everything into separate files

\section{LiveLittle: A Typed Livelit Calculus}\label{sec:livelit-calculus}
Should we talk about anything other than expansion + hygiene, i.e. parameterization, closure collection, closure resumption?
None of that is mechanized, but that may be okay. Closure collection and resumption seem to be worth formalizing.

Note: probably just going to omit the whole livelit function thing

Subsection about Agda mechanization (Ian + Nick?)

### 2020-05-24 livelit-definitions.tex
separated everything into separate files

\section{Livelit Definitions}\label{sec:livelit-definitions}
Which livelit implementation do we want to show? Maybe grade cutoffs? Slider as simple intro + color chooser?

Talk about each component of a livelit definition:
\begin{itemize}
  \item livelit name (talk about binding structure)
  \item expansion type
  \item model type (must be serializable)
  \item action type (mention integration into Hazel's undo system)
  \item initial model
  \item update function (monadic actions; talk about color palette update; SpliceRefs)
  \item view function (monadic actions; talk about GUI stuff? tab order? dimensions? mostly defer to Elm for details)
  \item expand function (should we call it fill?)
\end{itemize}

Hygiene from a provider's perspective -- have to use parameters to explicitly capture definition site bindings.

### 2020-05-24 livelits-paper.tex
separated everything into separate files

@@
 \input{intro}
+\input{case-studies}
+\input{livelit-definitions}
+\input{livelit-calculus}
+\input{extensions}
+\input{implementation}
+\input{related-work}
+\input{discussion}
 
-\input{overview}
+\clearpage
+\appendix
+\section{Misc. Material from Prior Draft}
+
+To address these limitations, we introduce live literals, or livelits.
+Figure~\ref{fig:cutoffs} shows a mockup of the definition and application of a livelit named \li{$grade_cutoffs} for adjusting grade cutoffs, represented as values of  record type \li{grade_cutoffs}.
+Like textual literal forms (e.g.
+list literals), livelits are alternative representations of expressions of the associated type \cite{DBLP:journals/pacmpl/OmarA18}.
+% From the perspective of the remainder of the program, \li{cutoffs} is a value of type \li{grade_cutoffs} like any other.
+We are implementing livelits in Hazel (\url{hazel.org}), a live functional programming environment with support for typed holes \cite{popl-paper}. We plan to perform a live demo.
+
+The definition of
+the livelit, outlined in Figure~\ref{fig:cutoffs}, follows the Elm architecture,
+i.e. there are types representing the abstract model and the messages that the 
+GUI generates.
+Livelits are \textbf{persistent} rather than ephemeral, i.e.
+the model is recorded in the underlying syntax tree.
+The \li{view} function generates the GUI, implemented using HTML, on demand (so the view is not persisted).
+Another function, \li{to_exp}, is responsible for generating the underlying expression, called the \emph{expansion}, from the \li{model}.
+While other projectional editors, e.g. those generated by Citrus \cite{DBLP:conf/uist/KoM05}, also support persistent GUIs in code, they are not user-extensible.
+
+The GUI can itself contain typed holes, represented by the \li{HtmlHole} constructor. In this case, there is a single hole
+for entering an expression of type \li{list(float)}, i.e. the list of weighted
+averages. In this example, the user has filled this hole with a variable,
+\li{weighted_averages} (the definition of which precedes the content of this figure and is not shown). In other words,
+livelits support \textbf{open expressions} and therefore interact cleanly with 
+standard abstraction mechanisms, i.e. they can appear under binders. We follow the reasoning principles for literals with spliced sub-expressions established  by \citet{DBLP:journals/pacmpl/OmarA18}.
+
+The main complication when dealing with open expressions relates to how live feedback
+is to be generated. Given just the symbolic expression in the hole, it would be 
+impossible to plot (as orange dots) the actual data from the list that the variable refers to. To resolve this issue, the system evaluates the program 
+as if it the livelits were empty holes, 
+relying on the support for evaluating incomplete programs described recently 
+by \citet{DBLP:journals/pacmpl/OmarVCH19}. The result of evaluation is an expression containing
+hole closures, i.e. holes equipped with environments. 
+%Hazel allows the user to interactively select which closure
+%interests them if multiple closures for a given hole appear in the result (e.g. if a livelit appears inside a function called multiple times). 
+The livelit 
+can then evaluate the expression in the hole against the closure selected by the user (not shown) using \li{Env.run}.
+
+%%%%%
+
+\todo{this is the WIP one from last year}{}
+Programs are most often represented as text.
+%
+This representation format provides expert programmers access to a variety of
+text editors and text edit actions---to affect fine-grained control over
+programming decisions---and a large ecosystem of other text-based
+utilities---for example, for line-based file differencing of source code
+versions.
+%
+On the other hand, the flexibility and low-level nature of text comes at some
+costs.
+%
+For experts, some tedious and manual edits could lead to
+inefficiency---especially because many routine code editing operations do not
+require the full flexibility of text~\citep{XXX}.
+%
+For novices, the large class of syntax errors that stem from text-edits presents
+a steep learning curve.
+
+Motivated to overcome the disadvantages of represent programs as text, a variety
+of alternative code editing interfaces have been investigated.
+%
+At the opposite end of text are \emph{visual programming languages}, which often
+completely represent a program with graphical elements~\citep{XXX}.
+%
+To forgo text completely, such approaches often target domain-specific
+languages, such as dataflow programing~\citep{XXX} and pedagogical languages
+that are, by design, restricted to relatively few building blocks.
+%
+Other \emph{structure}, or \emph{projectional} editors, still use a significant
+amount of text to render programs, but forgo text-edits in favor of
+\emph{structure edit actions} which transform the program (represented as a tree
+or some other structure), sidestepping the danger of invalid intermediate states
+of concrete syntax.
+
+On the spectrum somewhere between fully text- and fully structure-based are
+``hybrid'' editors, which augment text with additional ways to visualize and
+manipulate the structure of the program.
+%
+Victor Scrubbing, APX~\citep{APX}, Sketch-n-Sketch~\citep{sns-pldi}, and Carbide
+IDE~\citep{XXX}, among others, allow numeric values to be ``scrubbed'' by
+directly manipulating sliders rather than text-editing numeric literals.
+%
+Barista~\cite{Barista} is a hybrid Java editor where custom \emph{structure
+view} GUIs provide alternate representations of expressions instead of text.
+%
+For example, an arithmetic expression may be rendered with mathematical symbols,
+a method may be accompanied by interactive documentation with input-output
+examples, and structures may be selectively collapsed, expanded, or zoomed.
+%
+Graphite~\citep{Graphite} allows custom GUIs---called \emph{palettes}---to help
+the programmer fill missing expressions---``holes''---in the program.
+%
+For example, a color palette can provide visual previews of different candidate
+color values, and a regular expression palette can show input-output examples
+for different candidate regular expressions.
+
+The GUI representations and interactions enabled by the above hybrid editors are
+useful, but several limitations likely preclude wider utility:
+%
+the types of expressions that benefit from alternative GUI editors are limited
+to
+%
+privileged types that have baked-in interfaces~\citep{XXX}
+%
+or just of base type~\citep{XXX};
+%
+the GUI view is ephemeral, in that it disappears once it has been used to
+generate an expression~\citep{XXX}; or
+%
+expressions generated by the GUI are not deeply integrated with the static type
+system and interpreter of the language~\citep{XXX,XXX,XXX}.
+
+%% refactoring
+%% 
+%% DNDRefactoring~\citep{DNDRefactoring}
+%% Deuce~\citep{sns-deuce}
+
+\parahead{Persistent, Composable, and Live GUIs for Filling Holes}
+
+compared to the ``simple'' palettes
+
+extend palettes with: persistence, composition, and live feedback
+
+macro systems allow alternative syntactic (i.e. string) representations, to
+expand into underlying expressions
+
+palatable
+
+by analogy to these ``literal'' macro systems, palettes are ``graphical
+macros'': through interaction with the user, the GUI generates the underlying
+expression.
+
+TLMs~\cite{TLMs}
+
+\parahead{Contributions and Paper Outline}
+
+design for palettes. specifically, within the Hazelnut Live framework,
+which is a system to address the gap problem that arises in traditional editor
+frontends.
+
+demonstration of the expressiveness of the approach through a series of
+examples, many of which are drawn from the user study mentioned before.
+
+a prototype implementation of palettes within Hazel, which currently
+supports several examples using a core calculus with minimal features. the
+implementation provides a clear path for scaling up to larger, full-featured
+syntactic programming conveniences, as well as further UI design in future work.
+
+prototype UIs:
+
+  \begin{enumerate}
+    \item \Hazel{}: (palette macros;) structure edit actions; expression formula bar; inline layout
+    \item \sns{}: (palette functions;) text editor; pop-up menu
+  \end{enumerate}
+
+hazelnut~\citep{Hazelnut,HazelnutLive}
+
+\begin{figure*}[t!]
+  \begin{center}
+  \includegraphics[width=34pc]{cutoffs-mockup.png}\end{center}
+  \caption{A mockup of a livelit for adjusting grade cutoffs. Livelits are persistent and can access the live environment.}
+  \label{fig:cutoffs}
+  \end{figure*}
 
 %% \section{Defining Palettes}
-
+\input{overview}
 \section{Overview Examples: Defining Palettes}

### 2020-05-24 related-work.tex
separated everything into separate files

\section{Related Work}\label{sec:related-work}
Cyrus has a list somewhere, ask if you want to work on this section

### 2020-05-24 intro.tex
updated outline

@@
 Prior work on projectional editing
-and active code completion, summarized in Sec. \ref{sec:related-work}\todo{sections}{}, 
+and active code completion, summarized in Sec. \ref{sec:related-work}, 
 has also considered the problem of entering expressions 
@@
   \end{lstlisting}
-  Parameterization also underlies the hygiene mechanism, as we will discuss in Sec. 4\todo{sec}{}.
+  Parameterization also underlies the hygiene mechanism, as we will discuss in Sec.~\ref{sec:livelit-definitions}.
 
@@
 
-The remainder of this paper is organized as follows. We begin in Sec.~2 with two case studies 
+The remainder of this paper is organized as follows. We begin in Sec.~\ref{sec:case-studies} with two case studies 
 involving several non-trivial livelits, focusing on the livelit client's perspective. 
-We then consider the livelit provider's perspective in Sec.~3.
-In Sec.~4, we provide a formal account of the mechanism as a typed lambda calculus. 
+In Sec.~\ref{sec:livelit-definitions}, we consider the livelit provider's perspective by detailing the livelit
+definition mechanism by example. 
+In Sec.~\ref{sec:livelit-calculus}, we provide a formal account of the livelit mechanism as a typed lambda calculus. 
 This calculus formally specifies the essential aspects of livelit definition, representation, expansion, 
-and run-time closure collection independent of the particularities of our implementation, 
-and establishes essential type safety and hygiene properties. 
-In Sec.~5, we describe some extensions of the core mechanism\todo{what are the extensions?}{}.
-In Sec.~6, we outline our implementation of livelits within Hazel.
-In Sec.~7, we situate livelits relative to other related work. 
-Finally, we conclude in Sec.~8 with a discussion of some present limitations and future work.\todo{section refs}{}
+and run-time closure collection independent of the particularities of GUI framework and other implementation details. 
+This calculus and the key metatheoretic properties has been mechanized in Agda.
+In Sec.~\ref{sec:extensions}, we describe some extensions of the core calculus\todo{what are the extensions?}{}.
+In Sec.~\ref{sec:implementation}, we provide a more detailed account of our implementation of livelits within Hazel, 
+and discuss several implementation details that would need to be considered by other implementations.  
+In Sec.~\ref{sec:related-work}, we compare livelits to related work. 
+Finally, we conclude in Sec.~\ref{sec:discussion} with a discussion of some present limitations and future work.

### 2020-05-24 case-studies.tex
added updated textual mockup for grading case study + beginning portions of Sec 2

@@
 \section{Case Studies}\label{sec:case-studies}
-\subsection{Hazel Overview}\label{sec:hazel-overview}
-Brief overview (1/2 page) on Hazel (what do holes look like? structure editing, but not in too much detail,
-live evaluation)
+This section introduces livelits from the client programmer's perspective by way of 
+two domain-specific case studies:
+a course grading case study in Sec.~\ref{sec:live-grading} 
+and an image transformation case study in Sec.~\ref{sec:image-transformation}. 
+We also briefly provide several other examples in Sec.~\ref{sec:additional-examples}. 
+These case studies and examples have been implemented\todo{make this not a lie}{}
+in Hazel, a browser-based live programming environment for a pure typed functional language in the
+ML family, with which we assume basic familiarity. 
+We will describe the implementation of livelits in Hazel, and more generally, in Sec.~\ref{sec:implementation}.
 
-\subsection{Live Grading}\label{sec:live-grading}
+\subsection{Case Study: Grading with Livelits}\label{sec:live-grading}
+\begin{figure}
+\begin{lstlisting}
+let grades = $dataframe in 
+            A1          A2          A3          A4       Midterm      Final  +
+  Cyrus
+  Nick
+  David
+    + 
+let averages = compute_averages(grades) in 
+let cutoffs = $cutoffs(averages) in
+
+  |---------------------------------------------------|
+
+assign_grades(averages, cutoffs)
+\end{lstlisting}
+\caption{Grading Example}
+\label{fig:grading}
+\end{figure}
+
 \subsubsection{Overview}\label{sec:live-grading-overview}
@@
 
-\subsection{Image Transformation Pipeline}\label{sec:transformation}
+\subsection{Case Study: Image Transformations}\label{sec:image-transformation}
 Show example of an image transformation pipeline going through livelits

### 2020-05-25 case-studies.tex
added initial text for grading case study

@@
 two domain-specific case studies:
-a course grading case study in Sec.~\ref{sec:live-grading} 
+a course grade assignment case study in Sec.~\ref{sec:live-grading} 
 and an image transformation case study in Sec.~\ref{sec:image-transformation}. 
-We also briefly provide several other examples in Sec.~\ref{sec:additional-examples}. 
+We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}. 
 These case studies and examples have been implemented\todo{make this not a lie}{}
 in Hazel, a browser-based live programming environment for a pure typed functional language in the
-ML family, with which we assume basic familiarity. 
-We will describe the implementation of livelits in Hazel, and more generally, in Sec.~\ref{sec:implementation}.
+ML family, with which we assume basic familiarity.\todo{cite Elm tutorial for readers who are not familiar?}{}
 
@@
 
-assign_grades(averages, cutoffs)
+format_for_university(assign_grades(averages, cutoffs))
 \end{lstlisting}
@@
 \end{figure}
+
+Consider this commonplace scenario: an instructor needs 
+(1) to record numeric grades for various assignments and exams, and 
+(2) to visualize and perform various computations with these numeric grades 
+in order ultimately to assign final letter grades (\textbf{A} through \textbf{F}) to the students in the course.
+
+Perhaps the most common end-user application in this domain is the spreadsheet.
+A spreadsheet allows the instructor to record student grades using a natural tabular interface,
+visualize this data in one of a finite number of styles, and perform basic computations on this data,
+with results updated live, i.e. automatically after each cell has been edited. 
+However, these affordances are limited. It is difficult to package up common operations and workflows 
+into reusable libraries, interact with the data using domain-specific visualizations, 
+and perform operations that have not been anticipated by the designers of the spreadsheet system 
+(e.g. operations that prepare the data in an idiosyncratic format suitable for automated submission 
+to the university's grading system). 
+General-purpose programming languages equipped with suitable libraries for working with numeric data
+are flexible enough to handle these scenarios, but users of these languages 
+sacrifice the ability to receive live feedback and directly manipulate the data and visualizations thereof.
+
+Livelits are able to address this tension. 
+Fig.~\ref{fig:grading} shows a Hazel program where the user has:\todo{label each of these 1-5 in figure?}{}
+\begin{enumerate}
+\item directly recorded numeric \li{grades} for each student using a livelit, \li{\$dataframe}, that implements a tabular spreadsheet-like interface suitable for this purpose, 
+\item programmatically computed the overall \li{averages} for each student by calling a function, \li{compute_averages}, defined in a library (not shown), 
+\item determined the \li{cutoffs} between letter grades by directly manipulating a livelit that provides a drag-and-drop interface superimposed on a visualization of the grade distribution, 
+\item programmatically assigned letter grades to each student based on the cutoffs determined by the livelit in (3) by calling a library function, \li{assign_grades}, and 
+\item formatted the final grade assignments in a manner suitable for submission to a university system by calling another library function, \li{format_for_university}.
+\end{enumerate}
 

### 2020-05-25 case-studies.tex
tweaked text for grading case study

@@
 \begin{enumerate}
-\item directly recorded numeric \li{grades} for each student using a livelit, \li{\$dataframe}, that implements a tabular spreadsheet-like interface suitable for this purpose, 
-\item programmatically computed the overall \li{averages} for each student by calling a function, \li{compute_averages}, defined in a library (not shown), 
-\item determined the \li{cutoffs} between letter grades by directly manipulating a livelit that provides a drag-and-drop interface superimposed on a visualization of the grade distribution, 
-\item programmatically assigned letter grades to each student based on the cutoffs determined by the livelit in (3) by calling a library function, \li{assign_grades}, and 
-\item formatted the final grade assignments in a manner suitable for submission to a university system by calling another library function, \li{format_for_university}.
+\item recorded numeric \li{grades} for each student using a livelit, \li{\$dataframe}, that implements a tabular user interface with support for live evaluation of the Hazel expressions in cells, 
+\item programmatically computed the overall \li{averages} for each student by applying a function, \li{compute_averages}, defined in a library  (not shown) shared between multiple courses, 
+\item determined reasonable \li{cutoffs} between letter grades by directly manipulating a livelit that offers a domain-specific drag-and-drop interface superimposed on a live visualization of the grade distribution
+(for the sake of space, we plot dots rather than a full histogram)\todo{should we do a histogram?}{}, 
+\item programmatically assigned letter grades to each student based on the \li{cutoffs} determined in step (3) by calling another shared library function, \li{assign_grades}, 
+\item formatted the final grade assignments in a manner suitable for submission to the university system by calling another shared library function, \li{format_for_university}.
 \end{enumerate}
 
-\subsubsection{Overview}\label{sec:live-grading-overview}
-Talk about the scenario
+\subsubsection{Livelit Expansion} abc
 
-Live matrix: what is a matrix structurally? (data frame? list of lists?) + how does liveness work
-+ overview of hygiene guarantees
+\subsubsection{Typed Hygienic Splicing} abc
 
-Grade cutoffs: how does it work wrt liveness? (do we use a parameter here instead of a splice?) -- 
-focus on domain-specific benefits here.
+\subsubsection{Live Evaluation} abc
 

### 2020-05-26 case-studies.tex
first draft of livelit expansion section

@@
 
-\subsubsection{Livelit Expansion} abc
+\subsubsection{Livelit Expansion} 
+Livelit expressions are given dynamic meaning by expansion to standard expressions, i.e. expressions without livelits.
+The expansion of Fig.~\ref{fig:grading} is:\todo{change code once figure is final}{}
 
-\subsubsection{Typed Hygienic Splicing} abc
+\begin{lstlisting}
+let grades = (
+  ["A1", "A2", "A3", "A4", "Midterm", "Final"],
+  ("Cyrus", [0.0, 0.0, 0.0, 0.0, 0.0]),
+  ("Nick", [0.0, 0.0, 0.0, 0.0, 0.0]),
+  ("David", [0.0, 0.0, 0.0, 0.0, 0.0])) in 
+let averages = compute_averages(grades) in 
+let cutoffs = (.A 90.0, .B 80.0, .C 70.0, .D 60.0) in 
+format_for_university(assign_grades(averages, cutoffs))
+\end{lstlisting}
+
+The programmer can request to see this expansion in Hazel. 
+Ideally, however, it would not be necessary to inspect the underlying expansion 
+(or the livelit implementation, discussed in Sec.~\ref{sec:livelit-definitions}) 
+to reason about types and binding. 
+After all, one does not need to look inside 
+function bodies to reason about types and binding.
+Instead, in the words of Reynolds\todo{cite}{}, 
+``type structure is a syntactic discipline for maintaining levels of abstraction''.
+The livelits mechanism supports reasoning abstractly about types and binding by 
+several means.
+
+First, to support abstract reasoning about the type of the expansion, 
+livelit definitions must specify an \emph{expansion type} in their declarations.
+The declarations of the two livelits in Fig.~\ref{fig:grading}, 
+eliding their implementations for now, are:
+\begin{lstlisting}[numbers=none]
+livelit $dataframe at Dataframe { ... }
+livelit $cutoffs(averages: List(Float)) 
+  at (.A Float, .B Float, .C Float, .D Float) { ... }
+\end{lstlisting}
+The expansion type of \li{\$dataframe} is \li{Dataframe} 
+(a type for tabular floating point data together with string row and column keys, consistent with the expression in the expansion above)
+and the expansion type of \li{\$cutoffs} is a labeled product of grade cutoffs (Hazel notates labels \li{.label}).
+
+\subsubsection{Typed Hygienic Splicing and Parameterization} abc
+
+unlike arguments, not a fixed number of splices (e.g. adding a new row or column by pressing those buttons)
 

### 2020-05-26 livelits-paper.tex
first draft of livelit expansion section

@@
 numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML, showspaces=false,showstringspaces=false,xleftmargin=15pt, 
-morekeywords={string, float, int, Int, Float, String, livelit},
+morekeywords={string, float, int, Int, Float, String, livelit, at, List},
 classoffset=0,belowskip=\smallskipamount, aboveskip=\smallskipamount,

### 2020-05-28 case-studies.tex
added note about cursor inspector
tweaked note about cursor inspector
added section on parameters and partial application

--- 2020-05-25 case-studies.tex
@@
 
-\subsubsection{Livelit Expansion} abc
+\subsubsection{Livelit Expansion} 
+Livelit expressions are given dynamic meaning by expansion to core expressions, i.e. expressions that do not contain livelits.
+For example, the expansion of Fig.~\ref{fig:grading} is:\todo{change code once figure is final}{}
 
-\subsubsection{Typed Hygienic Splicing} abc
+\begin{lstlisting}
+let grades = Dataframe (
+  ["A1", "A2", "A3", "A4", "Midterm", "Final"],
+  [
+    ("Cyrus", [0.0, 0.0, 0.0, 0.0, 0.0]),
+    ("Nick", [0.0, 0.0, 0.0, 0.0, 0.0]),
+    ("David", [0.0, 0.0, 0.0, 0.0, 0.0])
+  ]) in 
+let averages = compute_averages(grades) in 
+let cutoffs = (.A 90.0, .B 80.0, .C 70.0, .D 60.0) in 
+format_for_university(assign_grades(averages, cutoffs))
+\end{lstlisting}
+
+The programmer can inspect this expansion in Hazel. 
+Ideally, however, it would not be necessary to inspect the expansion 
+(or the livelit implementation, which specifies the expansion logic and is discussed in Sec.~\ref{sec:livelit-definitions}) 
+to reason about types and binding. 
+After all, one does not need to look inside 
+function bodies to reason about types and binding.
+Instead, in the words of Reynolds\todo{cite}{}, 
+``type structure is a syntactic discipline for maintaining levels of abstraction''.
+The livelits mechanism supports reasoning abstractly about types and binding by 
+several means.
+
+To support abstract reasoning about the type of the expansion, 
+livelit definitions declare an \emph{expansion type}.
+The declarations of the two livelits in Fig.~\ref{fig:grading}, 
+eliding their implementations, are:
+\begin{lstlisting}[numbers=none]
+livelit $dataframe at Dataframe { ... }
+livelit $cutoffs(avgs: List(Float)) 
+  at (.A Float, .B Float, .C Float, .D Float) { ... }
+\end{lstlisting}
+The expansion type of \li{\$dataframe} is \li{Dataframe} 
+(a type classifying tabular floating point data together with string row and column keys, consistent with the expansion above)
+and the expansion type of \li{\$cutoffs} is a labeled product of grade cutoffs (Hazel notates labels \li{.label}).
+Hazel summarizes the typing information from the livelit definition for the programmer when the cursor is on the livelit's name, 
+just as it displays a function's type when its name is under the cursor (not shown).\todo{could show if Hannah et al are finished}{}
+
+\subsubsection{Typed Hygienic Splicing and Parameterization} 
+Livelit expressions can have sub-expressions provided by the client in one of two ways:
+as parameters or as splices.
+
+\paragraph{Parameters} Parameters appear immediately after the livelit name. 
+For example, \li{averages} is provided as a parameter to \li{\$cutoffs} in Fig.~\ref{fig:grading}.
+Livelits can declare a fixed number of parameters and specify their types in their declarations. 
+For example, the declaration of \li{\$cutoffs} above specifies one parameter of type \li{List(Float)}.
+Parameters can be partially applied in livelit abbreviations. For example, consider the slider livelit
+from Fig.~\ref{fig:color}:
+\begin{lstlisting}[numbers=none]
+livelit $slider (min: Float) (max: Float) at Float { ... }
+\end{lstlisting}
+We can partially apply the first parameter as follows to create a non-negative slider with one remaining parameter:
+\begin{lstlisting}[numbers=none]
+livelit $nnslider = $slider 0.0 in ...
+\end{lstlisting}
+The client must provide all remaining parameters to use a livelit to fill a hole in the program.
+So writing \li{\$nnslider} in expression position will not display the slider GUI. Instead, it will display as a livelit parameterization error.%
+\footnote{In Hazel, erroneous expressions are placed inside holes so that they do not prevent other parts of the program from executing
+\cite{HazelnutLive}.}
+
+We can apply both parameters in an abbreviation to create an unparameterized livelit:
+\begin{lstlisting}[numbers=none]
+livelit $percentage = $slider 0.0 1.0 in ...
+\end{lstlisting}
+
+\paragraph{Splices} Splices appear directly inside the livelit GUI.
+For example, the column headings and the cells in the \li{\$}
+
+unlike arguments, not a fixed number of splices (e.g. adding a new row or column by pressing those buttons)
 
@@
 many lines of code it took.
+
+Ideas:
+\begin{itemize}
+  \item derivation trees like Joomy's system (\url{https://joom.github.io/proof-tree-builder/src/})
+\end{itemize}

### 2020-05-31 case-studies.tex
hygiene section

@@
 
-\subsubsection{Typed Hygienic Splicing and Parameterization} 
-Livelit expressions can have sub-expressions provided by the client in one of two ways:
+\subsubsection{Typed Hygienic Splicing and Parameterization}\label{sec:splicing-and-parameterization}
+Livelit expressions can include sub-expressions in one of two ways:
 as parameters or as splices.
 
-\paragraph{Parameters} Parameters appear immediately after the livelit name. 
+\paragraph{Parameters}\label{sec:parameterization} Parameters appear immediately after the name of the applied livelit. 
 For example, \li{averages} is provided as a parameter to \li{\$cutoffs} in Fig.~\ref{fig:grading}.
-Livelits can declare a fixed number of parameters and specify their types in their declarations. 
-For example, the declaration of \li{\$cutoffs} above specifies one parameter of type \li{List(Float)}.
-Parameters can be partially applied in livelit abbreviations. For example, consider the slider livelit
+The livelit declares a fixed number of parameters and specifies their types in its declaration. 
+For example, the declaration of \li{\$cutoffs} shown above specifies one parameter of type \li{List(Float)}
+(the grades data to be plotted, see below).
+
+Livelit abbreviations can partially apply parameters. Consider the slider livelit
 from Fig.~\ref{fig:color}:
@@
 \end{lstlisting}
-We can partially apply the first parameter as follows to create a non-negative slider with one remaining parameter:
+We can partially apply the first parameter as follows to define a parameterized non-negative slider livelit with one remaining parameter:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-The client must provide all remaining parameters to use a livelit to fill a hole in the program.
-So writing \li{\$nnslider} in expression position will not display the slider GUI. Instead, it will display as a livelit parameterization error.%
-\footnote{In Hazel, erroneous expressions are placed inside holes so that they do not prevent other parts of the program from executing
+Only livelits with no remaining parameters can be used to fill a hole.
+So writing \li{\$nnslider} in expression position will not display the slider GUI. Instead, it will display as a ``missing livelit parameter'' error.%
+\footnote{\label{footnote:typing}In Hazel, erroneous expressions are automatically placed inside holes so that they do not prevent other parts of the program from executing
 \cite{HazelnutLive}.}
 
-We can apply both parameters in an abbreviation to create an unparameterized livelit:
+\paragraph{Splices}\label{sec:splices}
+Spliced expressions (or \emph{splices}) appear directly inside the livelit GUI.
+Splices can be filled with Hazel expressions of any form, including other livelit expressions, 
+and all of Hazel's editing affordances are available when editing these expressions.
+For example, each cell in the \li{\$dataframe} GUI in Fig.~\ref{fig:grading} 
+has a corresponding splice, which appears in the formula bar at the top of the livelit 
+when the user has selected that cell. 
+The cell itself displays the live value of the spliced expression, 
+following the example of spreadsheet interfaces 
+(we return to live evaluation in Sec.~\ref{sec:live-evaluation} below).
+
+The livelit provides an expected type for each splice.
+For example, the splices for the row and column keys in Fig.~\ref{fig:grading} are expected to be expressions
+of type \li{String}, and the remaining cells are expected to be expressions of type \li{Float}. 
+Hazel surfaces this typing information for the programmer when the cursor is in a splice, 
+so it is not necessary to inspect the expansion or the livelit implementation.
+If an expression of invalid type is entered, it will display in an error hole as usual,
+and in Hazel this will not prevent evaluation of other expressions (see Footnote \ref{footnote:typing}).
+
+Unlike parameters, the number of splices is not fixed in the livelit declaration. Splices can be created, 
+deleted, and filled through user interaction with the livelit. For example, clicking the \li{+} buttons
+in Fig.~\ref{fig:grading} will create new rows or columns, which will in turn generate new splices.
+
+\paragraph{Hygiene}\label{sec:hygiene}
+Parameters and splices can both appear in the underlying expansion. For example,
+the expansion of the \li{\$dataframe} livelit in Fig.~\ref{fig:grading} includes the 
+spliced expressions that the client entered the cell contents and the row and column keys.
+(Not every parameter or splice needs to appear in the expansion. For example,
+the bounds provided to the \li{\$slider} livelit are used only to constrain the user 
+interface, and do not appear directly in the expansion, which is simply a number literal.)
+
+The client cannot know, without looking at the expansion or the livelit implementation, 
+where in the expansion each parameter or splice will appear. This becomes relevant when 
+the expansion places the parameter or splice under a binder, e.g. in the body of a function or let binding.
+Na\"ively, this could cause inadvertent capture of the bound variable by a free variable 
+in the parameter or splice. For example, consider a livelit that generates an expansion 
+of the following form:
 \begin{lstlisting}[numbers=none]
-livelit $percentage = $slider 0.0 1.0 in ...
+let len = strlen <splice1> in 
+(<splice2>, <splice2> + 1)
 \end{lstlisting}
+Here, \li{<splice2>} appears under the binding of \li{len}. If the client enters an expression
+for \li{<splice2>} that itself refers to a client-side binding of the same variable, \li{len},
+it would na\"ively be captured. This sort of bug would difficult to debug, both for the provider
+and the client.
 
-\paragraph{Splices} Splices appear directly inside the livelit GUI.
-For example, the column headings and the cells in the \li{\$}
-
-unlike arguments, not a fixed number of splices (e.g. adding a new row or column by pressing those buttons)
-
-\subsubsection{Live Evaluation} abc
+To avoid this situation, the livelit expansion mechanism 
+maintains parameter and splice hygiene: parameters and splices are included in the expansion 
+in a capture-avoiding manner.
+Consequently, clients can reason abstractly about binding in splices: variables in splices
+always refer to bindings visible to the client, rather than bindings that are hidden inside expansions.
+(We discuss potentially relaxed variations of this discipline in Sec.~\ref{sec:discussion}).
+\subsubsection{Live Evaluation}\label{sec:live-evaluation}
 

### 2020-05-31 case-studies.tex
tweaked hygiene section

@@
 \paragraph{Hygiene}\label{sec:hygiene}
-Parameters and splices can both appear in the underlying expansion. For example,
+Parameters and splices can both appear in the expansion. For example,
 the expansion of the \li{\$dataframe} livelit in Fig.~\ref{fig:grading} includes the 
-spliced expressions that the client entered the cell contents and the row and column keys.
-(Not every parameter or splice needs to appear in the expansion. For example,
-the bounds provided to the \li{\$slider} livelit are used only to constrain the user 
-interface, and do not appear directly in the expansion, which is simply a number literal.)
+entered cell contents and the row and column keys.
 
@@
 \end{lstlisting}
-Here, \li{<splice2>} appears under the binding of \li{len}. If the client enters an expression
-for \li{<splice2>} that itself refers to a client-side binding of the same variable, \li{len},
-it would na\"ively be captured. This sort of bug would difficult to debug, both for the provider
-and the client.
+Here, \li{<splice2>} appears under the binding of \li{len}. If the client's choice for 
+\li{<splice2>} itself refers to a client-side binding of the same identifier, \li{len},
+it would na\"ively be captured. This sort of bug would difficult to debug, 
+both for the livelit provider and the client.
 
@@
 always refer to bindings visible to the client, rather than bindings that are hidden inside expansions.
-(We discuss potentially relaxed variations of this discipline in Sec.~\ref{sec:discussion}).
+This is implemented by alpha-renaming bindings internal to the expansion as necessary.
+(We discuss potentially relaxed variations of this hygiene discipline in Sec.~\ref{sec:discussion}).
 \subsubsection{Live Evaluation}\label{sec:live-evaluation}

### 2020-06-01 case-studies.tex
beginning of live evaluation section

@@
 \begin{lstlisting}
+let weights = [0.1, 0.1, 0.1, 0.1, 0.25, 0.35] in 
 let grades = $dataframe in 
@@
     + 
-let averages = compute_averages(grades) in 
+let averages = compute_averages(grades, weights) in 
 let cutoffs = $cutoffs(averages) in
@@
 This is implemented by alpha-renaming bindings internal to the expansion as necessary.
-(We discuss potentially relaxed variations of this hygiene discipline in Sec.~\ref{sec:discussion}).
+(We discuss potentially relaxed variations of this hygiene discipline in Sec.~\ref{sec:discussion}.)
+
 \subsubsection{Live Evaluation}\label{sec:live-evaluation}
+Livelits can request the result of evaluating a splice or parameter.
+However, evaluation is, as usual, defined only for closed expressions, 
+whereas parameters and splices can both refer to variables in the surrounding
+context. In order to provide evaluation results for these free variables, 
+Hazel performs \textbf{closure collection}. This proceeds in two phases.
+
+In the first phase, Hazel evaluates the program with holes appearing in place of each livelit 
+application and collects closures around each of these holes \cite{HazelnutLive}. 
+For example, the closure around the application of \li{\$dataframe} in Fig.~\ref{fig:grading}
+contains the value of the \li{weights} variable from Line 1 and any other bound variables, e.g. 
+from imported libraries, not shown. Consequently, when the livelit requests the evaluation result for
+splices that use these variables, such as the cell shown selected in Fig.~\ref{fig:grading}, their values are available.
+
+The closure for the \li{\$cutoffs} livelit in Fig.~\ref{fig:grading} includes a binding for 
+the \li{grades} variable. However, this variable's evaluation result, along with the evaluation 
+results for other variables that depend on \li{grades}, such as \li{averages}, ultimately depend on the 
+expansion of the \li{\$dataframe} livelit. If we stop after the first phase of closure collection,
+then no useful result will be available here because that livelit has been replaced with a hole.
+For this reason, there is a second phase of closure collection where any livelit holes that appear 
+in the collected closures are \emph{resumed}, i.e. the expansion is generated 
+(in this case, for the \li{\$dataframe} livelit) and is used to fill 
+the hole before evaluation resumes, following the semantics for resumption developed by \citet{HazelnutLive}.
+Livelit expansions do not themselves contain livelits, so no subsequent resumption phases are necessary.
 

### 2020-06-01 case-studies.tex
finished first draft of live evaluation section

--- 2020-05-31 case-studies.tex
@@
 This is implemented by alpha-renaming bindings internal to the expansion as necessary.
-(We discuss potentially relaxed variations of this hygiene discipline in Sec.~\ref{sec:discussion}).
+(We discuss potentially relaxed variations of this hygiene discipline in Sec.~\ref{sec:discussion}.)
+
 \subsubsection{Live Evaluation}\label{sec:live-evaluation}
+Livelits can request the result of evaluating a splice or a parameter. 
+For example, the \li{\$dataframe} livelit uses this facility to display
+the evaluation result for each cell when the cursor is not on it, like a spreadsheet.
+The \li{\$cutoffs} livelit uses this facility to plot the grades list, provided
+as a parameter, on the number line.
+
+\paragraph{Closure Collection} Evaluation in Hazel is, as usual, defined only for closed expressions, 
+but parameters and splices can both refer to variables in the surrounding
+context. In order to provide evaluation results for parameters or splices containing free variables, 
+Hazel performs \textbf{closure collection} in two phases.
+
+In the first phase, Hazel evaluates the program after placing holes in place of each livelit, 
+following the evaluation semantics for incomplete programs developed by \citet{HazelnutLive}. 
+The result is an expression with holes in it, each of which is equipped with a closure. 
+For example, there is a single closure associated with the application of \li{\$dataframe} in Fig.~\ref{fig:grading}.
+This closure contains the value of the \li{weights} variable from Line 1 and any other variables in the context 
+around that livelit, e.g. 
+from imported libraries, not shown. Consequently, when the livelit requests the evaluation result for
+splices that use these variables, such as the cell shown selected in Fig.~\ref{fig:grading}, 
+an evaluation result is available. (We consider the situation where more than one closure is available
+in the Sec.~\ref{sec:image-transformation}.)
+
+Similarly, the closure for the \li{\$cutoffs} livelit in Fig.~\ref{fig:grading} includes a result for 
+the \li{grades} variable. However, this variable's result, along with the 
+results for other variables that depend on \li{grades}, such as \li{averages}, ultimately depend on the 
+expansion of the \li{\$dataframe} livelit. If we stop after the first phase of closure collection,
+where these expansions have not yet been included, 
+then no useful result will be available here.
+For this reason, there is a second phase of closure collection where any livelit holes that appear 
+in the collected livelit closures are \emph{resumed}, i.e. the expansion is generated 
+(in this case, for the \li{\$dataframe} livelit) and is used to fill 
+the hole before evaluation resumes, following the semantics for resumption developed by \citet{HazelnutLive}.
+Livelit expansions do not themselves contain livelits, so no subsequent resumption phases are necessary.
+
+The final evaluation result for the program as a whole can also proceed by resumption, because 
+closure collection performs all of the computations that do not ultimately depend on the livelit expansion.
+Consequently, Hazel does not need to evaluate the program multiple times in order to support edit-time 
+(i.e. live) evaluation in this way.
+
+\paragraph{Indeterminate Results}
+We used the phrase \emph{evaluation result} rather than \emph{value} above purposefully: 
+splices and parameters can be incomplete, i.e. they can contain holes, so 
+evaluation does not always result in a value \cite{HazelnutLive}.
+In particular, when a hole appears in elimination position, e.g. in function position of a function application,
+ the result is instead an indeterminate expression. 
+When a livelit requests an evaluation result, it must be able to handle indeterminate results.
+For example, if the parameter to \li{\$cutoffs} were a hole or some other indeterminate expression\todo{small screenshot?}{}, 
+then it would have degraded functionality 
+(in this case, it would display the list elements that are values on the timeline, skipping indeterminate elements\todo{implement this}{}). 
+We will return to how livelit implementations handle this situation in Sec.~\ref{sec:implementation}.
+\todo{make sure we cover this}{}
 

### 2020-07-03 abstract.tex
substantial clarity pass over abstract, sec1, sec 2.1

@@
     Although textual program manipulation is powerful, 
-    some types of expressions are more naturally manipulated
+    some types of expressions are more naturally represented and manipulated
     graphically.     
@@
     GUI widgets, 
-    and many other domain-specific data structures.
+    and various other domain-specific data structures.
     This paper introduces \emph{live literals}, or \emph{livelits}, which allow
-    a programmer to fill holes of types like these by directly manipulating a     
+    the programmer to fill holes of types like these by directly manipulating a     
     GUI embedded persistently into the code. 
     Livelits are compositional: the GUI can itself contain
-    spliced expressions, which in turn can contain other livelits. 
+    spliced expressions, which can in turn contain other livelits. 
     These splices are typed, and the system 
@@
     This paper introduces livelits with several non-trivial case studies that exercise these capabilities. 
-    Our implementation extends Hazel, a web-based live programming environment able to assign
+    Our implementation extends Hazel, a browser-based live programming environment able to assign
     static and dynamic meaning to programs with holes. 
     We then define a typed lambda calculus that captures the essence of livelits, which
-    operate as graphical macros.     
+    operate as live, graphical macros.     
     The formalism is mechanized in Agda.

### 2020-07-03 case-studies.tex
substantial clarity pass over abstract, sec1, sec 2.1

@@
 \section{Case Studies}\label{sec:case-studies}
-This section introduces livelits from the client programmer's perspective by way of 
+Let us begin by considering livelits from the client programmer's perspective by way of 
 two domain-specific case studies:
@@
 in Hazel, a browser-based live programming environment for a pure typed functional language in the
-ML family, with which we assume basic familiarity.\todo{cite Elm tutorial for readers who are not familiar?}{}
+ML family. We assume basic familiarity with ML.\todo{cite Elm tutorial for readers who are not familiar?}{}
 
@@
 
-Consider this commonplace scenario: an instructor needs 
+Consider this familiar scenario: an instructor needs 
 (1) to record numeric grades for various assignments and exams, and 
@@
 \begin{enumerate}
-\item recorded numeric \li{grades} for each student using a livelit, \li{\$dataframe}, that implements a tabular user interface with support for live evaluation of the Hazel expressions in cells, 
+\item recorded numeric \li{grades} for each student using a livelit, \li{\$dataframe}, that implements a tabular user interface with support for live evaluation of the Hazel expression in each cell, 
 \item programmatically computed the overall \li{averages} for each student by applying a function, \li{compute_averages}, defined in a library  (not shown) shared between multiple courses, 

### 2020-07-03 intro.tex
added initial text for grading case study
added section on parameters and partial application
substantial clarity pass over abstract, sec1, sec 2.1

@@
 Textual program editors are powerful and expressive user interfaces, 
-so it is little wonder that they remain dominant decades after the teletype. 
+so it is perhaps little wonder that they remain dominant decades after the teletype. 
 However, textual user interfaces are not the best tool for every job. 
@@
 data types for which a non-textual  
-user interface may situationally be more appropriate \cite{Graphite}.
+user interface for expression construction may situationally be more appropriate \cite{Graphite}.
 
-For example, consider the following record type%
-\footnote{We will assume that we are in an ML-family language, 
-e.g. Elm \cite{Elm,ElmArchitecture,czaplicki2012elm}, 
-for the remainder of this paper, though the ideas are general.} 
-classifying colors using the RGBA encoding:
+As a simple example, consider a record type 
+classifying colors using an RGBA encoding:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-The problem is that this \emph{de facto} user interface for color selection 
+The problem is that this user interface for color selection 
 offers the user no immediate feedback 
@@
 and to interactively tweak that color.
-Better support for activities like these would be particularly helpful for users engaging in 
+
+Better support for manipulating data of types like these would be particularly helpful for users engaging in 
 live and exploratory programming in domains like web design, media production, 
-and data analysis.
-
-Indeed, practitioners in domains like these commonly eschew general-purpose programming 
+and data analysis. Indeed, practitioners in domains like these commonly eschew general-purpose programming 
 in favor of graphical end-user applications, like %
-image and video editors, spreadsheets, and music composition software, 
+image and video editors and music composition software, 
 in part because these applications provide domain-specific forms of live feedback and 
@@
 color to a variable for use in multiple locations, 
-or to computationally darken a color by passing it to a function, 
-or to organize and distribute color schemes using a package manager.
-Moreover, it is difficult to add new feedback and affordances and to compose 
-affordances in ways that the application designer has not anticipated, 
-e.g. to use a slider to manipulate the alpha channel 
-if the native color palette provides only a text box\todo{better example?}{}.
+or to transform a sound by passing it through a series of functions, 
+or to define a function that computes a vector graphic given a set
+of parameters (some of them perhaps themselves computed), 
+or to organize these constructions into modules.
+Moreover, it is difficult to add new affordances or to compose 
+affordances in ways that the application designer did not anticipate, 
+so users cannot easily make even simple changes like replacing a numeric text box in a color chooser with a slider, 
+much less more ambitious changes like a new visual interface for expressing geospatial data queries.
 
-This paper aims to resolve this apparent tension between 
-programmatic and direct manipulation interfaces by designing a programming system that 
-is able to surface GUIs when working with the data types for which 
-they are useful, while retaining support for textual program manipulation  
+This paper aims to resolve the apparent tension between 
+programmatic and direct manipulation user interfaces by designing a programming system that 
+is able to surface GUIs for the data types for which 
+they are useful, while retaining full support for symbolic program manipulation  
 and the abstraction and composition mechanisms
@@
 
-We are not, of course, the first to integrate direct manipulation user interfaces
+We are not, of course, the first to seek to integrate direct manipulation user interfaces
 into general-purpose programming systems.
 Prior work on projectional editing
-and active code completion, summarized in Sec. \ref{sec:related-work}, 
+and active code completion, detailed in Sec. \ref{sec:related-work}, 
 has also considered the problem of entering expressions 
 of certain types, like \li{Color}, 
-using specialized GUIs integrated into an editor. 
-The prior work most closely related to this paper is the {Graphite} system for Java \cite{Graphite}.
+using specialized GUIs integrated into a program editor. 
+The prior work most closely related to this paper is the {Graphite} system for Java \cite{Graphite},
+demonstrated in Fig.~\ref{fig:color}(a).
 Graphite allows library providers to associate GUIs, called \emph{palettes}, with types (using Java's class annotations). 
@@
 appropriate Java expression to fill the hole. 
-Figure~\ref{fig:color}(a)\todo{use subfigure?}{}, adapted from this prior work, demonstrates a simple color palette invoked using Graphite.
-When the user presses the \li{Enter} key, the Java code \li{new Color(0, 0, 128)}\todo{correct code}{} is inserted at the cursor (not shown) and the palette disappears.
+Figure~\ref{fig:color}(a)\todo{use subfigure?}{}, adapted from this prior work, demonstrates a color palette invoked using Graphite.
+When the user presses the \li{Enter} key, the Java expression \li{new Color(0, 0, 128)}\todo{correct code}{} is inserted at the cursor (not shown) and the palette disappears.
 
@@
 general-purpose languages, e.g. Typed CSS in Reason. TODO: CITE} 
-
 This survey also solicited a large number of use cases from the survey participants, 
@@
   \item \textbf{data analysis primitives}, such as plot setups and database queries; 
-  \item \textbf{data structures} such as matrices, maps, queues, and others for which specialized notation (e.g. JSON) and affordances may be more suitable than general-purpose mechanisms; and
+  \item \textbf{composite data structures} such as matrices, maps, queues, and others for which specialized notation and affordances may be more suitable than general-purpose mechanisms; and
   \item \textbf{data transformations} for which sensory feedback could aid in understanding, 
@@
 
-We take this extensive prior empirical evaluation as sufficient to establish 
-the utility and broad applicability of a mechanism of this sort.
+We take the extensive empirical evaluation performed by \citet{Graphite} as sufficient to establish 
+the utility and broad applicability of a mechanism of this class.
 
@@
   We use the word ``literal'' rather than ``palette'' because, with persistence, livelits
-  operate more like graphical literal notation, e.g. list literals.
+  operate as graphical literal notation, e.g. list literals.
 
@@
   For example, \li{\$slider} in Fig.~\ref{fig:color}(b) is parameterized by the lower and upper bound. 
-  Parameters behave like splices, differing in that livelit applications can be abbreviated, and abbreviations
-  support partial application. For example, it is possible to define a livelit named \li{\$percentage} 
+  Parameters behave like splices, differing in that livelit parameters can be partially applied in 
+  livelit abbreviations. For example, it is possible to define a livelit named \li{\$percentage} 
   that operates as a \li{\$slider} from \li{0} to \li{100} by scoped abbreviation:
@@
   leaving the hole being filled by the livelit temporarily unfilled and then evaluating
-  using a two-phased variation on the semantics for incomplete programs developed by \citet{HazelnutLive}. 
+  using a two-phased variation on the semantics for incomplete programs developed for Hazel \cite{HazelnutLive}. 
   We will see more sophisticated examples later in the paper, including a livelit 
   that appears inside a function that is called multiple times, leading to multiple closures that the client can 
-  interactively select between.
+   select between.
 \end{enumerate}
 
-The remainder of this paper is organized as follows. We begin in Sec.~\ref{sec:case-studies} with two case studies 
+\noindent
+\textbf{Outline.} The remainder of this paper is organized as follows. We begin in Sec.~\ref{sec:case-studies} with case studies 
 involving several non-trivial livelits, focusing on the livelit client's perspective. 

### 2020-07-09 livelit-calculus.tex
substantial clarity pass over abstract, sec1, sec 2.1
most of sec 4.... still going in circles with the last bit of closure collection and parameterization, but ready to start looking at

@@
-\section{LiveLittle: A Typed Livelit Calculus}\label{sec:livelit-calculus}
-Should we talk about anything other than expansion + hygiene, i.e. parameterization, closure collection, closure resumption?
-None of that is mechanized, but that may be okay. Closure collection and resumption seem to be worth formalizing.
+\section{A Simply Typed Livelit Calculus}\label{sec:livelit-calculus}
+In order to precisely capture the semantics of livelits  
+independently 
+of the specifics of the Hazel programming environment and GUI
+framework, we will now specify a simply typed \emph{livelit calculus}.
 
-Note: probably just going to omit the whole livelit function thing
+Fig.~\ref{fig:syntax} specifies the syntax of the livelit calculus. 
+Programs are written as \emph{unexpanded expressions}, $\hat e$, which are \emph{expanded} to 
+\emph{external expressions}, $e$, before being \emph{elaborated} 
+to internal expressions, $d$, for evaluation. All three stages are governed 
+by the same types, $\tau$. We include numbers, partial functions, products, sums, and recursive
+types, all in their standard form \cite{pfpl}, but this specific type structure is not critical. 
+Any language expressive
+enough to encode its own syntax (here as values of a recursive sum type)
+would be a suitable basis. We begin with a terse
+overview of the external and internal languages, 
+which are adapted straightforwardly from Hazelnut Live \cite{HazelnutLive} and therefore 
+serve as background material, in Sec.~\ref{sec:external-and-internal-lang}.
+Only unexpanded expressions contain 
+livelits, so they are the main focus of the remaining sections,
+which cover livelit expansion (Sec.~\ref{sec:calculus-expansion}), 
+livelit update actions (Sec.~\ref{sec:calculus-updates}), 
+liveness via closure collection (Sec.~\ref{sec:calculus-closure-collection}), 
+and parameterization (Sec.~\ref{sec:calculus-parameterization}).
 
-Subsection about Agda mechanization (Ian + Nick?)
+\subsection{Background: External and Internal Language}\label{sec:external-and-internal-lang}
+The external and internal languages are straightforward adaptations of the 
+external and internal languages of \emph{Hazelnut Live}, 
+a typed lambda calculus that assigns static and dynamic meaning to programs with holes,
+notated externally $\hehole{u}$ where $u$ is a \emph{hole name} \cite{HazelnutLive}.
+We omit non-empty holes (which allow reasoning about type inconsistencies \cite{Hazelnut}) and type holes
+(which operate like the unknown type from gradual type theory \cite{Siek06a,Hazelnut})
+because these mechanisms are orthogonal to livelits.
+
+The external language is governed by a typing judgement of the standard form, $\hasType{\Gamma}{e}{\tau}$, 
+where the typing context, $\Gamma$,
+is a finite set of typing assumptions of the form $x : \tau$ \cite{pfpl}.
+
+The internal language is a contextual type theory \cite{Nanevski2008}, i.e. the typing judgement is 
+of the form $\hasTypeD{\Delta}{\Gamma}{d}{\tau}$ where $\Delta$ is a finite set of hole typing
+assumptions of the form $u :: \tau[\Gamma]$.
+We need this hole typing context only for the internal language because, although hole names are assumed 
+unique in the external language, they can be duplicated during evaluation of internal expressions.
+Consequently, we need a context to ensure that each closure associated with hole $u$, written $\dehole{u}{\sigma}$, 
+has the same type and can be filled 
+with expressions valid under the same context.
+
+External expressions, $e$, are given dynamic meaning by typed elaboration to internal expressions, $d$, 
+according to the typed elaboration judgement, $\elabs{\Gamma}{e}{d}{\tau}{\Delta}$.
+The main purpose of this elaboration step is to initialize the substitution $\sigma$ on each hole closure, 
+which operates to capture 
+the substitutions that have occurred around that hole during evaluation. The key rule is:
+\begin{mathpar}
+\inferrule[Elab-Hole]{ }{
+    \elabs{\Gamma}{\hehole{u}}{\dehole{u}{\idof{\Gamma}}}{\tau}{u :: \tau[\Gamma]}
+}
+\end{mathpar}
+The substitution is initially the identity substitution, $\idof{\Gamma}$, i.e. the
+substitution that maps each variable in $\Gamma$ to itself, because no substitutions have yet occurred. For example, 
+\[ \elabs{ }{\helet{x}{\hnum{5}}{\hehole{u}} + \hnum{6}}{\helet{x}{\hnum{5}}{\dehole{u}{[x/x]} + \hnum{6}}}{\tnum}{u :: \tnum[x : \tnum]} \]
+During evaluation, $\evalsTo{d}{d'}$, the closure's substitution accumulates the substitutions that occur. For example,
+the internal expression above evalues as follows:
+\[
+  \evalsTo{\helet{x}{\hnum{5}}{\dehole{u}{[x/x]} + \hnum{6}}}{
+      \dehole{u}{[\hnum{5}/x]} + \hnum{6}
+  }
+\]
+We will use hole closures to support live programming with livelits in Sec.~\ref{sec:calculus-closure-collection}.
+
+All of these judgements are adapted directly from those of Hazelnut Live, 
+differing only in that we use a simpler declarative formulation rather than an algorithmic (bidirectional)
+formulation for the sake of simplicity and generality. Rather than restating the rules, we will simply state the governing
+metatheorems and refer the reader to the prior work for the full detalis \cite{HazelnutLive}.
+
+First, we have that an elaboration always exists and it preserves typing.
+\begin{theorem}[Typed Elaboration]
+    If $\hasType{\Gamma}{e}{\tau}$ then $\elabs{\Gamma}{e}{d}{\tau}{\Delta}$ for some $d$ and $\Delta$ such 
+    that $\hasTypeD{\Delta}{\Gamma}{d}{\tau}$. 
+\end{theorem}
+
+Next, we have that evaluation of an expression with holes leads to a final (i.e. irreducible) result of the same type (this is a corollary
+of the type safety properties established in the prior work.)
+\begin{theorem}[Preservation]
+    If $\hasTypeD{\Delta}{\cdot}{d}{\tau}$ and $\evalsTo{d}{d'}$ then $\isFinal{d'}$ and $\hasTypeD{\Delta}{\cdot}{d'}{\tau}$.
+\end{theorem}
+
+\subsection{Expansion}
+
+\begin{figure}
+    \[
+    \arraycolsep=4pt\begin{array}{llcl}
+        \mathsf{Typ} & \tau & ::= & 
+                                    \tnum ~\vert~
+                                    \tarr{\tau_1}{\tau_2} ~\vert~ 
+                                    \tprod{\tau_1}{\tau_2} ~\vert~ 
+                                    \tunit ~\vert~ 
+                                    \tsum{\tau_1}{\tau_2} ~\vert~
+                                    t ~\vert~
+                                    \trec{t}{\tau}\\
+        \mathsf{UExp} & \hat{e} & ::= & \hnum{n} ~\vert~ 
+                                 x ~\vert~
+                                 \hlam{x}{\hat e} ~\vert~
+                                 \hap{\hat e_1}{\hat e_2} ~\vert~
+                                 \cdots ~\vert~
+                                %  \hpair{\hat e_1}{\hat e_2} ~\vert~
+                                %  \hprl{\hat e} ~\vert~
+                                %  \hprr{\hat e} ~\vert~
+                                %  \htriv ~\vert~
+                                %  \hinL{\hat e} ~\vert~
+                                %  \hinR{\hat e} ~\vert~
+                                %  \hroll{\hat e} ~\vert~
+                                %  \hunroll{\hat e} \\
+                                 \hehole{u} ~\vert~
+                                 \haplivelit{u}{a}{\dtxt{model}}{\splices}\\
+        \mathsf{EExp} & e & ::= & \hnum{n} ~\vert~ x ~\vert~ \hlam{x}{e} ~\vert~ \hap{e_1}{e_2} ~\vert~ \cdots ~\vert~ \hehole{u}\\
+        \mathsf{IExp} & d & ::= & \hnum{n} ~\vert~ x ~\vert~ \hlam{x}{d} ~\vert~ \hap{d_1}{d_2} ~\vert~ \cdots ~\vert~ \dehole{u}{\sigma}\\
+        \mathsf{Splice} & \psi & ::= & \hsplice{\hat e}{\tau}
+    \end{array}
+    \]
+    \caption{Syntax of types, $\tau$, unexpanded expressions, $\hat{e}$, expanded expressions, $e$, and internal expressions, $d$. 
+    Metavariable $x$ ranges over variables, $u$ over hole names, and $\livelitname{a}$ over livelit names.
+    We write $\splat{\psi_i}$ for a finite sequence of $n \geq 0$ splices,
+    and $\sigma$ for finite substitutions of $n \geq 0$ internal expressions for variables, $[d_1/x_1, \cdots, d_n/x_n]$. 
+    We elide standard expression forms 
+    related to product, sum, and recursive types.
+    }
+    \label{fig:syntax}
+    \end{figure}
+
+    \begin{figure}
+    \begin{mathpar}
+        \inferrule[EVar]{
+            x : \tau \in \Gamma
+        }{
+            \expands{\Phi}{\Gamma}{x}{x}{\tau}
+        }
+
+        \inferrule[ELam]{
+            \expands{\Phi}{\Gamma, x : \taut{in}}{\hat e}{e}{\taut{out}}
+        }{
+            \expands{\Phi}{\Gamma}{\hlam{x}{\hat e}}{\hlam{x}{e}}{\tarr{\taut{in}}{\taut{out}}}
+        }
+
+        \inferrule[EAp]{
+            \expands{\Phi}{\Gamma}{\hat e_1}{e_1}{\tarr{\taut{in}}{\taut{out}}}\\\\
+            \expands{\Phi}{\Gamma}{\hat e_2}{e_2}{\taut{in}}
+        }{
+            \expands{\Phi}{\Gamma}{\hap{\hat e_1}{\hat e_2}}{\hap{e_1}{e_2}}{\taut{out}}
+        }
+
+        \cdots
+
+        \inferrule[EApLivelit]{
+            \livelitCtxEntry{a}{\taut{expand}}{\taut{model}}{\dtxt{expand}} \in \Phi\\\\
+            \hasType{ }{\dtxt{model}}{\taut{model}}\\
+            \evalsTo{\hap{\dtxt{expand}}{\dtxt{model}}}{\dtxt{encoded}}\\
+            \decodeExp{\dtxt{encoded}}{\etxt{pexpansion}}\\\\
+            \hasType{ }{\etxt{pexpansion}}{\tarr{\{\tau_i\}_{i < n}}{\taut{expand}}}\\
+            \{ \expands{\Phi}{\Gamma}{\hat e_i}{e_i}{\tau_i} \}_{i < n}
+        }{
+            \expands{\Phi}{\Gamma}{\haplivelit{u}{a}{\dtxt{model}}{\splat{\hsplice{\hat e_i}{\tau_i}}}}{
+                \hap{\etxt{pexpansion}}{\{e_i\}_{i < n}}
+            }{\taut{expand}}
+        }
+    \end{mathpar}
+    \caption{Expansion}
+    \label{fig:expansion}
+    \end{figure}
+The novelty of the livelit calculus lies entirely in its handling of unexpanded expressions,
+$\hat e$, which are given meaning by typed expansion to external expressions,
+$e$, according to the judgement $\expands{\Phi}{\Gamma}{\hat e}{e}{\tau}$ 
+defined in Fig.~\ref{fig:expansion}. 
+Unexpanded expressions mirror external expressions but for the presence of livelits, discussed below.
+The rules for standard
+forms like variables, functions and function application, shown in Fig.~\ref{fig:expansion},
+are straightforward.
+
+\subsubsection{Livelit Contexts}
+Livelit definitions are collected in the livelit context, $\Phi$, which
+maps livelit names to livelit definitions of the following form:
+\[ \livelitCtxEntry{a}{\taut{expand}}{\taut{model}}{\dtxt{expand}} \]
+Here, $\taut{expand}$ is the expansion type, $\taut{model}$ is the model type,
+and $\dtxt{expand}$ is the expansion function, which generates an expansion given a model.\todo{add view and update}
+
+For a livelit context to be well-formed, we require that each livelit definition be well-formed. 
+This in turn requires that the expansion function be hole-free and have the correct type.
+\begin{definition}[Livelit Context Well-Formedness]
+    A livelit context $\Phi$ is well-formed if and only if for each livelit definition, 
+    $\livelitCtxEntry{a}{\taut{expand}}{\taut{model}}{\dtxt{expand}} \in \Phi$,  we have
+    $\hasType{}{\dtxt{expand}}{\tarr{\taut{model}}{\expType}}$.
+\end{definition}
+Here, $\expType$ stands for a type whose values isomorphically encode 
+external expressions. The isomorphism is mediated in one direction by 
+the encoding judgement $\encodeExp{e}{d}$ and the other by the decoding judgement $\decodeExp{d}{e}$.
+Any isomorphism is sufficient, so we leave the details as a matter of implementation.
+The simplest approach would be to define $\expType$ as a recursive sum type,
+with one arm for each form of external expression (cf. \cite{TSLs} for an example).
+
+For simplicity, we assume that the livelit context is provided \emph{a priori} and therefore 
+that functions like $\dtxt{expand}$ are already closed and fully elaborated.
+In practice, the livelit context would be controlled by a definition form in the language
+that allows the expansion function to itself use livelits.
+This would require a staging mechanism,
+because we need to execute expansion functions in prior definitions to be able to 
+expand subsequent definitions.
+There are a number of ways to support the necessary staging in practice, e.g. 
+via explicit staging primitives \todo{cite ocaml macros paper, Racket stuff}{},
+by requiring that these definitions appear in separately compiled packages \cite{TLMs}, 
+or by using live
+programming mechanisms such as those available in Hazel 
+to evaluate ``up to'' each definition before proceeding.
+This choice is orthogonal to the mechanisms of interest in this section.
+
+\subsubsection{Hygienic Livelit Expansion}
+Unexpanded expressions are unique in that they include livelits:
+ \[\haplivelit{u}{a}{\dtxt{model}}{\splices}\] 
+Here, $\livelitname{a}$ names the livelit 
+ being applied. Livelits can be understood as filling holes, so $u$ identifies the hole
+ that is, conceptually, being filled. 
+The current state of the livelit is determined by the current model value, $\dtxt{model}$, 
+together with a collection of splices, $\splices$. Each splice $\psi_i$ 
+is of the form $\hsplice{\hat e_i}{\tau_i}$, where $\hat e_i$ is the spliced expression 
+itself (unexpanded, so it may contain other livelits) and $\tau_i$ is the type of that splice,
+as determined when the livelit definition first requested the splice (discussed in Sec.~\ref{sec:calculus-updates}).
+
+Rule \rulename{EApLivelit} performs livelit expansion. Its premises, in order, operate as follows:
+\begin{enumerate}
+    \item The first premise looks up the livelit name in the livelit context.
+    \item The second premise serves to ensure that the model value, $\dtxt{model}$, is of the 
+    specified model type, $\taut{model}$.
+    \item The third premise applies the expansion function, $\dtxt{expand}$, to the model value, $\dtxt{model}$, 
+    producing the encoded expansion, $\dtxt{encoded}$, which, by the definitions given above, is of type $\mathsf{Exp}$.
+    \item The fourth premise decodes $\dtxt{encoded}$, producing the external expression it encodes, $\etxt{pexpansion}$.
+    \item We call $\etxt{expansion}$ the \emph{parameterized expansion} because, according to the fifth premise,
+    it must be a function that returns a value of the expansion type, $\taut{expand}$, when applied (in curried fashion, though this is not critical)
+    to the splices, whose types, $\splat{\tau_i}$, are explicitly recorded in the livelit.
+    We must take care to ensure that the parameterized expansion is \emph{context independent}, 
+    i.e. that it cannot depend on the particular bindings available in the call site typing context, $\Gamma$. 
+    Context independence is one facet of what is colloquially known as \emph{hygiene} \cite{TLMs}.
+    In this simple formulation of the system, we maintain context independence simply by 
+    requiring that the parameterized expansion be 
+    closed. Consequently, any necessary helper functions used in the expansion must be provided explicitly by the client 
+    via a splice. In Sec.~\ref{sec:calculus-parameterization}, we will see how this context independence 
+    restriction can be made 
+    less onerous by passing definition-site bindings along as partially applied parameters.
+    \item Finally, the sixth premise inductively expands each of the spliced expressions in the same context as the livelit
+    expression itself.
+\end{enumerate}
+The conclusion of the rule then applies the parameterized expansion to the expanded splices. 
+By applying the splices as arguments, we maintain \emph{capture avoidance} -- splices cannot capture variables
+bound internally to the expansion. Capture avoidance is the other facet of \emph{hygiene} \cite{TLMs}.
+
+The typed expansion process is governed by the following metatheorem, which establishes that the expansion 
+is indeed an external expression of the indicated type.\todo{change to proof sketch}{}
+
+\begin{theorem}[Typed Expansion]
+    If $\expands{\Phi}{\Gamma}{\hat e}{e}{\tau}$ then $\hasType{\Gamma}{e}{\tau}$.
+\end{theorem}
+\begin{proof}
+    We proceed by rule induction on the assumption. 
+    The cases involving the standard forms follow by straightforward induction.
+    The only interesting case is the \rulename{EApLivelit} case.
+    In this case, we have that $e = \hap{\etxt{pexpansion}}{\splat{e_i}}$. 
+    By assumption, we have $\{ \expands{\Phi}{\Gamma}{\hat e_i}{e_i}{\tau_i} \}_{i < n}$.
+    By induction, we therefore have $\{ \hasType{\Gamma}{e_i}{\tau_i} \}_{i < n}$.
+    In addition, by assumption we have that $\hasType{}{\etxt{pexpansion}}{\tarr{\splat{\tau_i}}{\taut{expand}}}$.
+    By weakening, we therefore have $\hasType{\Gamma}{\etxt{pexpansion}}{\tarr{\splat{\tau_i}}{\taut{expand}}}$.
+    Finally, by iterated application of the typing rule for function application over the splices, 
+    using the judgements just established, 
+    we have $\hasType{\Gamma}{\hap{\etxt{pexpansion}}{\splat{e_i}}}{\taut{expand}}$ as desired.
+\end{proof}
+
+When composed with the Typed Elaboration theorem and the Type Safety properties established for the internal 
+language, we achieve end-to-end type safety for unexpanded expressions (every well-typed unexpanded 
+expression expands to a well-typed external expression, which in turn expands to a well-typed internal 
+expression, which evaluates in a type safe manner.)
+
+\subsubsection{Agda Mechanization}
+We have mechanized the typed expansion mechanism and proven the Typed Expansion theorem 
+using the Agda proof assistant.\todo{say more}{}
+
+\subsection{Live Feedback via Closure Collection}
+In order to support live feedback, a livelit needs to be able to ask the system
+to evaluate expressions under a selected closure for the hole that the livelit is filling.
+This mechanism was described by example in Sec.~\ref{sec:image-transformation}.
+
+In this section, we will formalize the process of efficiently collecting closures 
+for the livelits that appear in a program. The key idea is that we generate a temporary expansion,
+called the \emph{cc-expansion}, 
+where each livelit expands to an empty hole applied to each of its splices. 
+The key rule for the cc-expansion judgement, $\ccexpands{\Phi}{\Gamma}{\hat e}{e}{\tau}$, is:
+\begin{mathpar}
+\inferrule[CCApLivelit]{
+    \livelitCtxEntry{a}{\taut{expand}}{\taut{model}}{\dtxt{expand}} \in \Phi\\\\
+    \{ \ccexpands{\Phi}{\Gamma}{\hat e_i}{e_i}{\tau_i} \}_{i < n}
+}{
+    \ccexpands{\Phi}{\Gamma}{\haplivelit{u}{a}{\dtxt{model}}{\splat{\hsplice{\hat e_i}{\tau_i}}}}{\hap{\hehole{u}}{\splat{e_i}}}{\taut{expand}}
+}
+\end{mathpar}
+We then elaborate and evaluate the cc-expansion 
+following Hazelnut Live's dynamic semantics for programs with holes, which were summarized 
+in Sec.~\ref{sec:external-and-internal-lang}. The result will contain some number of hole closures
+for each livelit hole. We call these the proto-closures and their associated environments the proto-environments for that livelit hole, formally defined as follows.
+\begin{definition}[Proto-Environment Collection]
+If $\ccexpands{\Phi}{\cdot}{\hat e}{e}{\tau}$ and $\elabs{}{e}{d}{\tau}{\Delta}$
+and $\evalsTo{d}{d'}$ then $\protoEnvsOf{\hat e}{u} = \{ \sigma \mid \dehole{u}{\sigma} \in d' \}$.
+    % protoclosures(\hat e, u) = { sigma | hole^u_sigma in d' } where \hat e cc-expands to e and e elaborates to d and d evaluates to d' 
+\end{definition}
+A proto-environment for a livelit hole might itself contain a proto-closure for another livelit hole.
+For example, in Fig.~\ref{fig:color}, the proto-environment for the \li{\$color} livelit contained the 
+proto-closure for the \li{\$slider} livelit because it was used in the value of a bound variable, \li{grayLevel}.
+This is unsatisfying, because it means that the actual value of that variable, which requires computing the expansion,
+ is not available to the \li{\$color}
+livelit as expected.
+Consequently, we must fill any livelit holes that appear in the proto-environments for other livelit holes.
+We do so by computing the parameterized expansions of the corresponding livelits,  
+as specified in Sec.~\ref{sec:calculus-expansion} (we can skip the final premise of Rule \rulename{EApLivelit} because the 
+splice arguments have already been applied in the cc-expansion). We then perform a hole filling operation, 
+$\instantiate{\dtxt{pexpansion}}{u}{d}$ over each of the proto-environments associated with livelit holes. 
+Hole filling was defined by Hazelnut Live and operates to 
+fill every closure for hole $u$ in $d'$ with $\dtxt{pexpansion}$ and derives from the metavariable
+instantiation operation of contextual modal type theory (CMTT) \cite{HazelnutLive,Nanevski2008}. 
+Unlike substitution, hole filling is
+not capture-avoiding. Instead, the environment on each of these closures is applied to $\dtxt{pexpansion}$
+as a substitution, i.e. the delayed substitutions captured in the environment are realized.
+In this case, however, the proto-expansion is guaranteed to be closed due to 
+the context independence discipline we maintain in Rule \rulename{EApLivelit}, 
+so hole filling acts simply as a syntactic replacement.
+
+Let $\livelitsIn{\hat e}$ by the set of hole names for the livelits that appear in $\hat{e}$.
+
+\begin{definition}[Environment Collection]
+    If $u \in \livelitsIn{\hat e}$ and $\protoEnvsOf{\hat e}{u} = \splat{\sigma_i}$ 
+    % then $\EnvsOf{\hat e}{u} = \splat{fillem{\hat e\sigma_i}}$.
+\end{definition}
+
+This is not sufficient, however, because some of these hole closures might themselves 
+contain livelit holes. However, a livelit that uses this closure needs the final result, 
+not a hole. Consequently, we need to \emph{fill and resume} livelit closures that appear inside 
+other livelit closures. Fortunately, for a pure functional language can do so efficiently 
+by computing the expansion for the corresponding livelit expression, elaborating it in isolation,
+filling the closure with the resulting elaboration, and then resuming evaluation. A theorem from
+the prior paper establishes the correctness of this process, i.e. that the resulting program will 
+produce the same result as if it had been evaluated from scratch.
+
+In fact, even the cc-expansion evaluation process does not go to waste -- we can resume the 
+result of cc-expansion evaluation in the same way to arrive at the same final result 
+as the standard expansion and evaluation process.
+
+\subsection{Parameterization}
+
+Add livelit expressions, livelit abbreviations, 
+partial application support. Talk about how to work around the strict context
+independence discipline we have imposed.
+
+\subsection{Edit Action Semantics}
+So far, we have considered only the semantics of an individual editor state.
+However, livelits are useful because they evolve as the user interacts
+with them. In particular, the livelit's model, $\dtxt{model}$ can evolve
+due to livelit-specific user interactions. Furthermore, a livelit can also 
+create and delete splices when a user interaction demands it (unlike functions, which have a fixed 
+number of arguments).

### 2020-07-10 case-studies.tex
finally figured out how to formalize closure collection properly... subtle

@@
 
+We assume that the user has selected at most one closure for each livelit in the unexpanded expression.
+The corresponding environment is applied as a substitution before attempting to evaluate the requested expression.
+To maintain the context independence principle, this request must take the form of a closed function 
+of the splices. 
+As discussed in Sec.~\ref{sec:live-evaluation}, evaluation can fail if the free variables in the splices 
+overlap with the free variables in selected closure.
+
 \subsection{Additional Examples}\label{sec:additional-examples}

### 2020-07-10 livelit-calculus.tex
finally figured out how to formalize closure collection properly... subtle

@@
 In order to support live feedback, a livelit needs to be able to ask the system
-to evaluate expressions under a selected closure for the hole that the livelit is filling.
-This mechanism was described by example in Sec.~\ref{sec:image-transformation}.
+to evaluate expressions under one of the closures associated with the livelit.
+This mechanism was introduced by example in Sec.~\ref{sec:image-transformation}.
+In this section, we will formalize the process of efficiently collecting closures 
+for the livelits that appear in a program.
 
-In this section, we will formalize the process of efficiently collecting closures 
-for the livelits that appear in a program. The key idea is that we generate a temporary expansion,
+The key idea is that we generate an alternative expansion,
 called the \emph{cc-expansion}, 
-where each livelit expands to an empty hole applied to each of its splices. 
-The key rule for the cc-expansion judgement, $\ccexpands{\Phi}{\Gamma}{\hat e}{e}{\tau}$, is:
+where each livelit expands to an empty hole applied to the splices. In other words,
+we leave a hole in place of the parameterized expansion, which we expand and record in the \emph{cc-context}, $\Omega$, on the side.
+The key rule for the cc-expansion judgement, $\ccexpands{\Phi}{\Gamma}{\hat e}{e}{\tau}{\Omega}$, is:
 \begin{mathpar}
 \inferrule[CCApLivelit]{
-    \livelitCtxEntry{a}{\taut{expand}}{\taut{model}}{\dtxt{expand}} \in \Phi\\\\
-    \{ \ccexpands{\Phi}{\Gamma}{\hat e_i}{e_i}{\tau_i} \}_{i < n}
+    \{ \ccexpands{\Phi}{\Gamma}{\hat e_i}{e_i}{\tau_i}{\Omega_i} \}_{i < n}\\
+    \expands{\Phi}{\Gamma}{\haplivelit{u}{a}{\dtxt{model}}{\splat{\hsplice{\hat e_i}{\tau_i}}}}{
+        \hap{\etxt{pexpansion}}{\splat{e_i}}
+    }{\taut{expand}}\\
+    \elabs{ }{\etxt{pexpansion}}{\dtxt{pexpansion}}{\tarr{\splat{\tau_i}}{\taut{expand}}}{\Delta}
 }{
     \ccexpands{\Phi}{\Gamma}{\haplivelit{u}{a}{\dtxt{model}}{\splat{\hsplice{\hat e_i}{\tau_i}}}}{\hap{\hehole{u}}{\splat{e_i}}}{\taut{expand}}
+    {\cup_{i<n} \Omega_i \cup \{\Omegaitem{u}{\dtxt{pexpansion}}\}}
 }
 \end{mathpar}
+
 We then elaborate and evaluate the cc-expansion 
@@
 \begin{definition}[Proto-Environment Collection]
-If $\ccexpands{\Phi}{\cdot}{\hat e}{e}{\tau}$ and $\elabs{}{e}{d}{\tau}{\Delta}$
-and $\evalsTo{d}{d'}$ then $\protoEnvsOf{\hat e}{u} = \{ \sigma \mid \dehole{u}{\sigma} \in d' \}$.
+If $\ccexpands{\Phi}{\cdot}{\hat e}{e}{\tau}{\Omega}$ and $\elabs{}{e}{d}{\tau}{\Delta}$
+and $\evalsTo{d}{d'}$ and $u \in \domof{\Omega}$ then $\protoEnvsOf{\Phi}{\hat e}{u} = \{ \sigma \mid \dehole{u}{\sigma} \in d' \}$.
     % protoclosures(\hat e, u) = { sigma | hole^u_sigma in d' } where \hat e cc-expands to e and e elaborates to d and d evaluates to d' 
@@
 For example, in Fig.~\ref{fig:color}, the proto-environment for the \li{\$color} livelit contained the 
-proto-closure for the \li{\$slider} livelit because it was used in the value of a bound variable, \li{grayLevel}.
-This is unsatisfying, because it means that the actual value of that variable, which requires computing the expansion,
- is not available to the \li{\$color}
-livelit as expected.
-Consequently, we must fill any livelit holes that appear in the proto-environments for other livelit holes.
-We do so by computing the parameterized expansions of the corresponding livelits,  
-as specified in Sec.~\ref{sec:calculus-expansion} (we can skip the final premise of Rule \rulename{EApLivelit} because the 
-splice arguments have already been applied in the cc-expansion). We then perform a hole filling operation, 
-$\instantiate{\dtxt{pexpansion}}{u}{d}$ over each of the proto-environments associated with livelit holes. 
-Hole filling was defined by Hazelnut Live and operates to 
-fill every closure for hole $u$ in $d'$ with $\dtxt{pexpansion}$ and derives from the metavariable
-instantiation operation of contextual modal type theory (CMTT) \cite{HazelnutLive,Nanevski2008}. 
+proto-closure for the \li{\$slider} livelit, via the \li{grayLevel} variable.
+If we use proto-environments for live evaluation, then the actual value of that variable
+would not be available (here, the color could not be displayed).
+
+Consequently, the second step of closure selection is to fill any livelit holes that appear in the proto-environments for other livelit holes.
+We do so by filling them using the parameterized expansions gathered in $\Omega$ and then resuming
+evaluation where appropriate.
+Formally, this involves the hole filling operation $\instantiate{d_1}{u}{d_2}$ for Hazelnut Live
+(which derives from the metavariable instantation operation of contextual modal type theory \cite{HazelnutLive,Nanevski2008}).
+This operation 
+fills every closure for hole $u$ in $d_2$ with $d_1$. 
 Unlike substitution, hole filling is
@@
 as a substitution, i.e. the delayed substitutions captured in the environment are realized.
-In this case, however, the proto-expansion is guaranteed to be closed due to 
+In this case, however, the parameterized expansion is guaranteed to be closed due to 
 the context independence discipline we maintain in Rule \rulename{EApLivelit}, 
@@
 
-Let $\livelitsIn{\hat e}$ by the set of hole names for the livelits that appear in $\hat{e}$.
-
-\begin{definition}[Environment Collection]
-    If $u \in \livelitsIn{\hat e}$ and $\protoEnvsOf{\hat e}{u} = \splat{\sigma_i}$ 
-    % then $\EnvsOf{\hat e}{u} = \splat{fillem{\hat e\sigma_i}}$.
+Formally, we begin by defining an operation $\fillof{\Omega}{\sigma}$ which acts on proto-environments 
+to fill any livelit holes that appear.
+\begin{definition}[Livelit Hole Filling] ~
+    \begin{enumerate}
+        \item $\fillof{\Omega}{[d_1/x_1, \ldots, d_n/x_n]} = [\fillof{\Omega}{d_1}/x_1, \ldots, \fillof{\Omega}{d_n}/x_n]$
+        \item $\fillof{\Omega}{d} = \instantiateB{\dtxt{pexpansion}}{u}_{\Omegaitem{u}{\dtxt{pexpansion}} \in \Omega}{d}$
+    \end{enumerate}
 \end{definition}
 
-This is not sufficient, however, because some of these hole closures might themselves 
-contain livelit holes. However, a livelit that uses this closure needs the final result, 
-not a hole. Consequently, we need to \emph{fill and resume} livelit closures that appear inside 
-other livelit closures. Fortunately, for a pure functional language can do so efficiently 
-by computing the expansion for the corresponding livelit expression, elaborating it in isolation,
-filling the closure with the resulting elaboration, and then resuming evaluation. A theorem from
-the prior paper establishes the correctness of this process, i.e. that the resulting program will 
-produce the same result as if it had been evaluated from scratch.
+This first step may cause certain expressions to become non-final, because the filled hole is no longer
+blocking evaluation. We therefore define an operation $\resumeof{\sigma}$ that resumes evalution for all closed expressions in $\sigma$.
+(Open expressions can remain because some closures appear under binders in the final result, so a substitution has not yet been recorded
+for the bound variables.)
+\begin{definition}[Environment Resumption] ~
+    \begin{enumerate}
+        \item $\resumeof{[d_1/x_1, \ldots, d_n/x_n]} = [\resumeof{d_1}/x_1, \ldots, \resumeof{d_n}/x_n]$
+        \item $\resumeof{d} = d'$ if $\fvof{d} = \emptyset$ and $\evalsTo{d}{d'}$
+        \item $\resumeof{d} = d$ if $\fvof{d} \neq \emptyset$
+    \end{enumerate}
+\end{definition}
 
-In fact, even the cc-expansion evaluation process does not go to waste -- we can resume the 
-result of cc-expansion evaluation in the same way to arrive at the same final result 
-as the standard expansion and evaluation process.
+Finally, we can produce the final set of environments by filling and resuming the proto-environments.
+\begin{definition}[Environment Collection]
+    If $\ccexpands{\Phi}{\cdot}{\hat e}{e}{\tau}{\Omega}$
+    % and $\elabs{}{e}{d}{\tau}{\Delta}$
+    % and $\evalsTo{d}{d'}$ and $u \in \domof{\Omega}$
+    then \[\envsOf{\Phi}{\hat e}{u} = \{\resumeof{\fillof{\Omega}{\sigma}} \mid \sigma \in \protoEnvsOf{\Phi}{\hat e}{u}\}\]
+\end{definition}
+
+This same fill and resume operation can be used to avoid recomputation when evaluating the fully expanded version of the user's program.
+If the editor has already performed environment collection, then it can simply continue from where it left off
+by filling and resuming
+the remaining top-level livelit holes (those that do not appear in a proto-environment).
+
+The correctness of the mechanisms described in this section rest on the fact that evaluation commutes with hole filling. 
+This only holds when the language is pure: evaluation order does not matter 
+so the system is free to evaluate around holes initially, 
+and then return to finish the job once they have been filled.
+
+\begin{theorem}[Post-Collection Resumption]
+    If $\ccexpands{\Phi}{\cdot}{\hat e}{\etxt{cc}}{\tau}{\Omega}$ and $\elabs{}{\etxt{cc}}{\dtxt{cc}}{\tau}{\Delta}$
+    and $\evalsTo{\dtxt{cc}}{\dtxt{cc}'}$ and $\resumeof{\fillof{\Omega}{\dtxt{cc}'}} = d_1$ 
+    and $\expands{\Phi}{\cdot}{\hat e}{\etxt{full}}{\tau}$
+    and $\elabs{}{\etxt{full}}{\dtxt{full}}{\tau}{\Delta}$ 
+    and $\evalsTo{\dtxt{full}}{d_2}$ then $d_1 = d_2$.
+\end{theorem}
+\begin{proof}
+    The key observation is that filling the livelit holes in the cc-expansion gives the full expansion,
+    i.e. $\fillof{\Omega}{\dtxt{cc}} = \dtxt{full}$. Resumption is simply evaluation for closed expressions.
+    By commutativity of hole filling, established in the prior work \cite{HazelnutLive}, 
+    we can delay hole filling until $\dtxt{cc}$ has first been evaluated to
+    $\dtxt{cc}'$.
+\end{proof}
+
+In a language with side effects, one would instead need to modify evaluation of  
+cc-expansions such that the expansion of each livelit application is evaluated alongside the 
+temporary livelit expansions,
+with the result stored for subsequent hole filling. This would ensure that
+side effects happen in the same order and only once. The details are beyond the scope of this work.
 

### 2020-07-10 livelit-calculus.tex
finished first pass through sec 3

@@
 \begin{enumerate}
-    \item The first premise looks up the livelit name in the livelit context.
-    \item The second premise serves to ensure that the model value, $\dtxt{model}$, is of the 
+    \item \textbf{Lookup.} The first premise looks up the livelit name in the livelit context.
+    \item \textbf{Model Validation.} The second premise serves to ensure that the model value, $\dtxt{model}$, is of the 
     specified model type, $\taut{model}$.
-    \item The third premise applies the expansion function, $\dtxt{expand}$, to the model value, $\dtxt{model}$, 
+    \item \textbf{Expansion.} The third premise applies the expansion function, $\dtxt{expand}$, to the model value, $\dtxt{model}$, 
     producing the encoded expansion, $\dtxt{encoded}$, which, by the definitions given above, is of type $\mathsf{Exp}$.
-    \item The fourth premise decodes $\dtxt{encoded}$, producing the external expression it encodes, $\etxt{pexpansion}$.
-    \item We call $\etxt{expansion}$ the \emph{parameterized expansion} because, according to the fifth premise,
+    \item \textbf{Decoding.} The fourth premise decodes $\dtxt{encoded}$, producing the external expression it encodes, $\etxt{pexpansion}$.
+    \item \textbf{Expansion Validation.} We call $\etxt{expansion}$ the \emph{parameterized expansion} because, according to the fifth premise,
     it must be a function that returns a value of the expansion type, $\taut{expand}$, when applied (in curried fashion, though this is not critical)
@@
     restriction can be made 
-    less onerous by passing definition-site bindings along as partially applied parameters.
-    \item Finally, the sixth premise inductively expands each of the spliced expressions in the same context as the livelit
+    less onerous by passing definition-site bindings along as partially applied parameters.\todo{remove?}{}
+    \item \textbf{Splice Expansion.} Finally, the sixth premise inductively expands each of the spliced expressions in the same context as the livelit
     expression itself.
@@
 In a language with side effects, one would instead need to modify evaluation of  
-cc-expansions such that the expansion of each livelit application is evaluated alongside the 
-temporary livelit expansions,
+cc-expansions such that the full expansion of each livelit application is evaluated at the same time as the 
+temporary expansion,
 with the result stored for subsequent hole filling. This would ensure that
-side effects happen in the same order and only once. The details are beyond the scope of this work.
+side effects happen in the same order and only once. The full details are beyond the scope of this work.
 

### 2020-07-10 livelits-paper.tex
substantial clarity pass over abstract, sec1, sec 2.1
finished first pass through sec 3

--- 2020-05-24 livelits-paper.tex
@@
 %% Cyrus packages
-\usepackage{microtype}
-\usepackage{mdframed}
-\usepackage{colortab}
+% \usepackage{microtype}
+% \usepackage{mdframed}
+% \usepackage{colortab}
 \usepackage{mathpartir}
-\usepackage{enumitem}
-\usepackage{bbm}
+% \usepackage{enumitem}
+% \usepackage{bbm}
 \usepackage{stmaryrd}
-\usepackage{mathtools}
-\usepackage{leftidx}
+% \usepackage{mathtools}
+% \usepackage{leftidx}
 \usepackage{todonotes}
 \usepackage{xspace}
-\usepackage{wrapfig}
+% \usepackage{wrapfig}
 
@@
 numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML, showspaces=false,showstringspaces=false,xleftmargin=15pt, 
-morekeywords={string, float, int, Int, Float, String, livelit},
+morekeywords={string, float, int, Int, Float, String, livelit, at, SpliceRef, UpdateMonad, ViewMonad, Html, return,
+capture, Typ, Exp, Option, List, Result, Dimensions, Unit},
 classoffset=0,belowskip=\smallskipamount, aboveskip=\smallskipamount,
@@
 %% Can remove this eventually
-\usepackage{blindtext}
+% \usepackage{blindtext}
 
-\usepackage{enumitem}
+% \usepackage{enumitem}
 
@@
 %% management, these would otherwise names clash).
-\let\clipbox\relax
-\usepackage[export]{adjustbox}% http://ctan.org/pkg/adjustbox
+% \let\clipbox\relax
+% \usepackage[export]{adjustbox}% http://ctan.org/pkg/adjustbox
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
+% \citestyle{acmauthoryear}   %% For author/year citations
 
@@
 \clearpage
-\input{formal-system}
+% \input{formal-system}
 

### 2020-07-10 implementation.tex
implementation sec

@@
 \section{Implementation}\label{sec:implementation}
-What should we talk about here? Layout of livelits?
+Implementing livelits requires tight integration between a rich editor, 
+a type checker, and a live evaluator. The evaluator is used
+both for running the model-view-update-expand system that controls 
+each livelit and for evaluating splices. Consequently, 
+the evaluator must be capable of evaluating incomplete programs and gathering hole closures
+as specified by Hazelnut Live \cite{HazelnutLive}.
+
+\subsection{Hazel}
+Hazelnut Live is the formal foundation for the Hazel programming environment, so 
+it was natural to choose the the Hazel programming environment for our primary implementation\todo{cite?}{}. 
+ Hazel is a browser-based live functional programming environment 
+organized from the ground up around typed holes. Uniquely, every editor state in Hazel is 
+semantically meaningful: it has a type, it can be evaluated, and it can be transformed 
+in a type-aware manner. This implies that livelits remain fully functional at all times, 
+even when the program is incomplete or erroneous.
+
+Hazel is implemented in OCaml and compiled to JavaScript using the \li{js_of_ocaml} compiler\todo{cite}.
+Hazel is written in a pure functional style, so our implementation is able to closely 
+follow the formalism. The Hazel language has been evolving in parallel with our work, 
+and has only recently become expressive enough for it to be feasible to implement livelits within Hazel itself.
+The examples in this paper are instead implemented in OCaml as modules satisfying the signature 
+implied by Fig.~\ref{fig:color-impl}. A functor then adapts these OCaml-based livelit implementations 
+to livelit definitions suitable for inclusion in the livelit context used in the semantics.
+
+Hazel achieves its ``gap-free'' liveness guarantee by automatically inserting explicit holes as necessary 
+while the user edits 
+the program. Formally, Hazel is a type-aware structure editor \cite{Hazelnut}, rather than a text editor, 
+although the developers are aiming to maintain a text-like programming experience when editing 
+symbolic code (this effort is orthogonal to our own). 
+This means that it is conceptually straightforward to hide the model and splice data and instead display 
+the computed view. However, there are a number of non-trivial layout challenges that come up, which we discuss
+in Sec.~\ref{sec:layout} below.
+
+\subsection{Text Editor Integration}
+Before delving into these issues, let us note that livelits do not require the use of a structure editor. 
+We have also developed a 
+proof-of-concept textual program editor, based on Sketch-n-Sketch \cite{sns-pldi,sns-uist}, with support for livelits.
+This proof-of-concept is not at feature parity with 
+our main implementation, but it demonstrates that a text editor with support for 
+parsing out livelit applications and replacing them with interactive graphical inclusions is sufficient.
+
+\subsection{Layout}\label{sec:layout}
+Whether implemented in a structure editor or a text editor, livelits present 
+interesting layout challenges: it can be difficult to sensibly 
+interleave symbolic code with livelit views.
+
+Hazel uses an optimizing pretty printer to determine layout, an approach 
+based on the work of Bernardy et al \todo{cite bernardy}. This system relies 
+fundamentally on character counts. Consequently, our implementation asks each 
+livelit to specify dimensions in terms of character counts rather than pixels.
+Livelits can be laid out in one of two ways: as inline livelits, like \li{\$slider},
+which must be exactly one character high and appear inline with the code,
+or as \emph{multi-line livelits}, which occupy up to the full width of the editor 
+and a specified number of code lines. The pretty printer lays out multi-line livelits
+on their own line, as seen in the figures in this paper.
+
+We are currently exploring a more flexible system whereby 
+multi-line livelits can be named and applied within a larger expression, but
+the GUI is not displayed until the next line, even if there are intervening characters.
+This is inspired by the \li{~} placeholder for large textual literals in the 
+Wyvern programming language \cite{TSLs}. We are also considering a number of other layout 
+options, e.g. pop-up livelits, livelits pinned to sidebars, and livelits that are rendered  
+on a separate canvas or document while still formally being located within an underlying functional program.
+
+This latter option 
+would be particularly interesting for end-user programming scenarios: users with limited
+programming experience 
+could interact with a collection of livelits laid out separately, without necessarily
+even being aware that their interactions are actually edits to an underlying typed
+functional program. More experienced users on the team could then make programmatic use of the entered
+data and parameters in a seamless manner. We are also excited about the possibility of end users
+getting curious enough to ``View Source'' in this setting, and in so doing 
+being exposed to typed functional programming concepts gradually.
+
+The layout of splices within livelits is also an interesting problem, and our current approach, where 
+the provider must specify a fixed size for each splice, is not particularly flexible. 
+A more principled system for GUI layout that integrates with program pretty 
+printing could be of substantial use in some settings. Formula bars, like the one in the \li{\$dataframe}
+livelit, are a useful design pattern to avoid splice layout issues, and it may be possible to provide 
+a generalized formula bar in the editor user interface that multiple livelits could share.
+While splices can in principle contain other livelits, it is difficult for small splice editors to 
+accomodate large livelit views. Fortunately, variables can be used to support composition via substitution 
+rather than nesting. A linter might suggest such a refactoring.
+
+The Hazel pretty printer uses memoization and incremental diffs to maintain reasonable performance,
+and adding livelits to the system did not substantially interfere with the existing approach.
+For the relatively simple livelits that we have developed, performance has been adequate.
+However, more complex  livelits or livelits that update quickly 
+could easily cause problems with editor responsiveness.
+This could perhaps be resolved by performing livelit-related computations asynchronously. Care 
+would need to be taken to ensure that intervening edits do not cause synchronization issues.
+
+Many livelit-related computations can be intelligently memoized, e.g. view computations and unchanged splice evaluations,
+especially with the edit awareness that Hazel provides. We leave further optimizations along these lines to future work.

### 2020-07-10 livelits-paper.tex
fix

@@
 
-\clearpage
-\appendix
-\section{Misc. Material from Prior Draft}
-
-To address these limitations, we introduce live literals, or livelits.
-Figure~\ref{fig:cutoffs} shows a mockup of the definition and application of a livelit named \li{$grade_cutoffs} for adjusting grade cutoffs, represented as values of  record type \li{grade_cutoffs}.
-Like textual literal forms (e.g.
-list literals), livelits are alternative representations of expressions of the associated type \cite{DBLP:journals/pacmpl/OmarA18}.
-% From the perspective of the remainder of the program, \li{cutoffs} is a value of type \li{grade_cutoffs} like any other.
-We are implementing livelits in Hazel (\url{hazel.org}), a live functional programming environment with support for typed holes \cite{popl-paper}. We plan to perform a live demo.
-
-The definition of
-the livelit, outlined in Figure~\ref{fig:cutoffs}, follows the Elm architecture,
-i.e. there are types representing the abstract model and the messages that the 
-GUI generates.
-Livelits are \textbf{persistent} rather than ephemeral, i.e.
-the model is recorded in the underlying syntax tree.
-The \li{view} function generates the GUI, implemented using HTML, on demand (so the view is not persisted).
-Another function, \li{to_exp}, is responsible for generating the underlying expression, called the \emph{expansion}, from the \li{model}.
-While other projectional editors, e.g. those generated by Citrus \cite{DBLP:conf/uist/KoM05}, also support persistent GUIs in code, they are not user-extensible.
-
-The GUI can itself contain typed holes, represented by the \li{HtmlHole} constructor. In this case, there is a single hole
-for entering an expression of type \li{list(float)}, i.e. the list of weighted
-averages. In this example, the user has filled this hole with a variable,
-\li{weighted_averages} (the definition of which precedes the content of this figure and is not shown). In other words,
-livelits support \textbf{open expressions} and therefore interact cleanly with 
-standard abstraction mechanisms, i.e. they can appear under binders. We follow the reasoning principles for literals with spliced sub-expressions established  by \citet{DBLP:journals/pacmpl/OmarA18}.
-
-The main complication when dealing with open expressions relates to how live feedback
-is to be generated. Given just the symbolic expression in the hole, it would be 
-impossible to plot (as orange dots) the actual data from the list that the variable refers to. To resolve this issue, the system evaluates the program 
-as if it the livelits were empty holes, 
-relying on the support for evaluating incomplete programs described recently 
-by \citet{DBLP:journals/pacmpl/OmarVCH19}. The result of evaluation is an expression containing
-hole closures, i.e. holes equipped with environments. 
-%Hazel allows the user to interactively select which closure
-%interests them if multiple closures for a given hole appear in the result (e.g. if a livelit appears inside a function called multiple times). 
-The livelit 
-can then evaluate the expression in the hole against the closure selected by the user (not shown) using \li{Env.run}.
-
-%%%%%
-
-\todo{this is the WIP one from last year}{}
-Programs are most often represented as text.
-%
-This representation format provides expert programmers access to a variety of
-text editors and text edit actions---to affect fine-grained control over
-programming decisions---and a large ecosystem of other text-based
-utilities---for example, for line-based file differencing of source code
-versions.
-%
-On the other hand, the flexibility and low-level nature of text comes at some
-costs.
-%
-For experts, some tedious and manual edits could lead to
-inefficiency---especially because many routine code editing operations do not
-require the full flexibility of text~\citep{XXX}.
-%
-For novices, the large class of syntax errors that stem from text-edits presents
-a steep learning curve.
-
-Motivated to overcome the disadvantages of represent programs as text, a variety
-of alternative code editing interfaces have been investigated.
-%
-At the opposite end of text are \emph{visual programming languages}, which often
-completely represent a program with graphical elements~\citep{XXX}.
-%
-To forgo text completely, such approaches often target domain-specific
-languages, such as dataflow programing~\citep{XXX} and pedagogical languages
-that are, by design, restricted to relatively few building blocks.
-%
-Other \emph{structure}, or \emph{projectional} editors, still use a significant
-amount of text to render programs, but forgo text-edits in favor of
-\emph{structure edit actions} which transform the program (represented as a tree
-or some other structure), sidestepping the danger of invalid intermediate states
-of concrete syntax.
-
-On the spectrum somewhere between fully text- and fully structure-based are
-``hybrid'' editors, which augment text with additional ways to visualize and
-manipulate the structure of the program.
-%
-Victor Scrubbing, APX~\citep{APX}, Sketch-n-Sketch~\citep{sns-pldi}, and Carbide
-IDE~\citep{XXX}, among others, allow numeric values to be ``scrubbed'' by
-directly manipulating sliders rather than text-editing numeric literals.
-%
-Barista~\cite{Barista} is a hybrid Java editor where custom \emph{structure
-view} GUIs provide alternate representations of expressions instead of text.
-%
-For example, an arithmetic expression may be rendered with mathematical symbols,
-a method may be accompanied by interactive documentation with input-output
-examples, and structures may be selectively collapsed, expanded, or zoomed.
-%
-Graphite~\citep{Graphite} allows custom GUIs---called \emph{palettes}---to help
-the programmer fill missing expressions---``holes''---in the program.
-%
-For example, a color palette can provide visual previews of different candidate
-color values, and a regular expression palette can show input-output examples
-for different candidate regular expressions.
-
-The GUI representations and interactions enabled by the above hybrid editors are
-useful, but several limitations likely preclude wider utility:
-%
-the types of expressions that benefit from alternative GUI editors are limited
-to
-%
-privileged types that have baked-in interfaces~\citep{XXX}
-%
-or just of base type~\citep{XXX};
-%
-the GUI view is ephemeral, in that it disappears once it has been used to
-generate an expression~\citep{XXX}; or
-%
-expressions generated by the GUI are not deeply integrated with the static type
-system and interpreter of the language~\citep{XXX,XXX,XXX}.
-
-%% refactoring
-%% 
-%% DNDRefactoring~\citep{DNDRefactoring}
-%% Deuce~\citep{sns-deuce}
-
-\parahead{Persistent, Composable, and Live GUIs for Filling Holes}
-
-compared to the ``simple'' palettes
-
-extend palettes with: persistence, composition, and live feedback
-
-macro systems allow alternative syntactic (i.e. string) representations, to
-expand into underlying expressions
-
-palatable
-
-by analogy to these ``literal'' macro systems, palettes are ``graphical
-macros'': through interaction with the user, the GUI generates the underlying
-expression.
-
-TLMs~\cite{TLMs}
-
-\parahead{Contributions and Paper Outline}
-
-design for palettes. specifically, within the Hazelnut Live framework,
-which is a system to address the gap problem that arises in traditional editor
-frontends.
-
-demonstration of the expressiveness of the approach through a series of
-examples, many of which are drawn from the user study mentioned before.
-
-a prototype implementation of palettes within Hazel, which currently
-supports several examples using a core calculus with minimal features. the
-implementation provides a clear path for scaling up to larger, full-featured
-syntactic programming conveniences, as well as further UI design in future work.
-
-prototype UIs:
-
-  \begin{enumerate}
-    \item \Hazel{}: (palette macros;) structure edit actions; expression formula bar; inline layout
-    \item \sns{}: (palette functions;) text editor; pop-up menu
-  \end{enumerate}
-
-hazelnut~\citep{Hazelnut,HazelnutLive}
-
-\begin{figure*}[t!]
-  \begin{center}
-  \includegraphics[width=34pc]{cutoffs-mockup.png}\end{center}
-  \caption{A mockup of a livelit for adjusting grade cutoffs. Livelits are persistent and can access the live environment.}
-  \label{fig:cutoffs}
-  \end{figure*}
-
-%% \section{Defining Palettes}
-\input{overview}
-\section{Overview Examples: Defining Palettes}
-
-Palette definitions take the following general form:
-\begin{lstlisting}
-palette $name 
-  (arg1 : t1) 
-  ... 
-  (argn : tn) 
-  at t 
-  implementation P in package pkg;
-\end{lstlisting}
-The static semantics requires the following:
-\begin{enumerate}
-\item \li{arg1} ... \li{argn} for \li{n} $\geq 0$ are distinct labels
-\item \li{t1} ...\li{t1} are valid types
-\item \li{t} is a valid type
-\item and \li{P} identifies a module that can be loaded from package \li{pkg} such that \begin{lstlisting}
-P : IPalette
-    with type args = { 
-      arg1 : HoleRef.t<t1>, 
-      ..., 
-      argn : HoleRef.t<tn> 
-    } 
-    with type output = t
-\end{lstlisting}
-where \li{IPalette} is defined in Fig.~\ref{fig:IPalette} and \li{HoleRef} in Fig.~\ref{fig:HoleRefs}.
-\end{enumerate}
-
-\begin{figure}
-\begin{lstlisting}
-module type IPalette = {
-  type args
-  type output
-  type model
-  type msg
-  val init    : args -> HRG.t(model)
-  val update  : (model, msg) -> HRG.t(model)
-  val view    : model -> HRE.t(Html.t(msg))
-  val compute : model -> HRE.t(output)
-}
-\end{lstlisting}
-\caption{\li{IPalette} module type}
-\label{fig:IPalette}
-\end{figure}
-
-\begin{figure}
-\begin{lstlisting}
-(* Abstract type of hole refs *)
-module HoleRef : {
-  type t('a)
-}
-
-(* HRG is the hole ref generation monad *)
-module HRG : {
-  type t('a)
-  val fresh('a) : t(HoleRef.t('a))
-  val bind : t('a) -> 
-             ('a -> t('b)) -> 
-             t('b)
-  val return : 'a -> t('a)
-}
-
-(* HRE is the hole ref evaluation monad *)
-module HRE : {
-  type t('a) 
-  type result('a) = Value('a * Exp)
-                  | Indet(Exp)
-  val eval('a) : HoleRef.t('a) -> 
-                 t(result('a))
-  val bind : t('a) -> 
-             ('a -> t('b)) -> 
-             t('b)
-  val return : 'a -> t('a)
-}
-\end{lstlisting}
-\caption{Modules for working with hole refs}
-\label{fig:HoleRefs}
-\end{figure}
-
-\subsection{Background: Elm architecture}
-
-\subsection{GUIs with Typed Holes}
-
-\subsection{Palette-Specific Actions?}
-
-\subsection{Reasoning Principles}
-
-\subsection{Deriving Palettes from Type Definitions?}
-
-\clearpage
-% \input{formal-system}
-
-\clearpage
-\input{evaluation}
-
-\clearpage
-\section{Related Work}
-Graphite, Relit, projectional editors, notebooks / Mathematica,
-
-\url{https://twitter.com/johnregehr/status/1095018518737637376} -- ascii art in source code
-
-\url{http://idl.cs.washington.edu/papers/lyra/} - plotting UIs
-
-\begin{figure*}
-\begin{lstlisting}
-                  Graphite    Projectional Editors    TLMs     Live Palette Expressions
-interactive       Y           Y                       N        Y
-extensible        Y           N                       Y        Y
-persistent        N           Y                       Y        Y
-compositional     N           N                       Y        Y
-live              N           N                       N        Y
-formalized        N           N                       Y        Y
-\end{lstlisting}
-\caption{Related Work Table}
-\label{fig:related-work}
-\end{figure*}
-
-\section{Discussion}
-
-\parahead{User Interface Design}
-
-\begin{itemize}
-
-\item
-%
-our \Hazel{} and \sns{} prototypes demonstrated two simple approaches to palette
-layout: inline and floating, respectively.
-%
-might also want user-controlled layout and pinning.
-
-\end{itemize}
-
-\parahead{Deriving Palettes from Type Definitions}
-
-\begin{itemize}
-
-\item \cite{SSS}
-
-\end{itemize}
-
-\section{Conclusion}
-
 \begin{quote}

### 2020-07-10 intro.tex
Add color figure

@@
   \begin{center}
-  \includegraphics[width=35pc]{color-palettes.png}\end{center}
+    \begin{subfigure}[t]{0.5\textwidth}
+      \centering
+      \includegraphics[width=15pc]{slider-color-livelits.png}
+    \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
+      \centering
+      \includegraphics[width=15pc]{slider-color-livelits.png}
+    \end{subfigure}
+  \end{center}
   \caption{
@@
   computed color.
-  The value of \li{defaultColor} is determined by the livelit via a macro expansion step.}
+  The value of \li{default\_color} is determined by the livelit via a macro expansion step.}
   \label{fig:color}
@@
   the terminology of \citet{TLMs} who developed reasoning principles for splicing in a strictly textual setting).
-  Fig.~\ref{fig:color}(b) demonstrates splicing: the client is able to define a variable, \li{grayLevel}, 
+  Fig.~\ref{fig:color}(b) demonstrates splicing: the client is able to define a variable, \li{gray\_level},
   to capture the relationship between the red, green, and blue components.

### 2020-07-10 implementation.tex
fix
related work

@@
 or as \emph{multi-line livelits}, which occupy up to the full width of the editor 
-and a specified number of code lines. The pretty printer lays out multi-line livelits
+and a specified number of code lines (not shown in Fig.~\ref{fig:color-impl}). 
+The pretty printer lays out multi-line livelits
 on their own line, as seen in the figures in this paper.
@@
 the GUI is not displayed until the next line, even if there are intervening characters.
-This is inspired by the \li{~} placeholder for large textual literals in the 
+This is inspired by the \verb|~| placeholder for large textual literals in the 
 Wyvern programming language \cite{TSLs}. We are also considering a number of other layout 
@@
 programming experience 
-could interact with a collection of livelits laid out separately, without necessarily
+could interact with a collection of livelits laid out separately in the popular ``dashboard'' style, 
+without necessarily
 even being aware that their interactions are actually edits to an underlying typed
@@
 While splices can in principle contain other livelits, it is difficult for small splice editors to 
-accomodate large livelit views. Fortunately, variables can be used to support composition via substitution 
-rather than nesting. A linter might suggest such a refactoring.
+visually accomodate large livelit views in practice. 
+Fortunately, variables can be used to support composition via substitution 
+rather than nesting. A linter might suggest a refactoring to a variable in these scenarios.
 
@@
 Many livelit-related computations can be intelligently memoized, e.g. view computations and unchanged splice evaluations,
-especially with the edit awareness that Hazel provides. We leave further optimizations along these lines to future work.
+especially with the edit awareness that Hazel provides. 
+We leave further optimizations along these lines to future work.
+Prior work on incrementally updating UIs as code changes could be relevant to this effort \cite{burckhardt2013s}.

### 2020-07-10 livelit-calculus.tex

@@
 For example, in Fig.~\ref{fig:color}, the proto-environment for the \li{\$color} livelit contained the
-proto-closure for the \li{\$slider} livelit, via the \li{grayLevel} variable.
+proto-closure for the \li{\$slider} livelit, via the \li{gray\_level} variable.
 If we use proto-environments for live evaluation, then the actual value of that variable

### 2020-07-10 intro.tex
Add Graphite color palette
tweaks

@@
 data types for which a non-textual
-user interface for expression construction may situationally be more appropriate \cite{Graphite}.
+user interface for expression construction may situationally be more appropriate.
 
@@
 to activate the palette.
-Once the interaction is finished, it is the palette's job to generate an
-appropriate Java expression to fill the hole.
+Once the interaction is finished, the palette must generate a
+Java expression to fill the hole.
 Figure~\ref{fig:color}(a)\todo{use subfigure?}{}, adapted from this prior work, demonstrates a color palette invoked using Graphite.
-When the user presses the \li{Enter} key, the Java expression \li{new Color(0, 0, 128)}\todo{correct code}{} is inserted at the cursor (not shown) and the palette disappears.
+When the user presses the \li{Enter} key, the Java expression \li{new Color(173, 173, 173, 94)}\todo{correct code}{} is inserted at the cursor (not shown) and the palette disappears.
 
@@
 many participants remarked that they rarely entered colors directly into Java code,
-but rather into stylesheets or other files governed by a domain-specific language.
+but rather into stylesheets.
 Other palettes, e.g. a palette
-that supported regular expression construction and testing, were viewed as more
+that supported regular expression construction, were viewed as more
 suitable for Java code.
 The mechanisms being considered are suitable both for general-purpose languages
-and typed domain-specific languages, many of which can be embedded into modern
-general-purpose languages, e.g. Typed CSS in Reason. TODO: CITE}
+and typed domain-specific languages like typed stylesheets, which can often be embedded into modern
+general-purpose languages.}
 This survey also solicited a large number of use cases from the survey participants,
@@
     In Graphite, only one palette can be associated with a type, and this must
-    be done at the definition site.
+    be done at its definition site.
     By contrast, livelit definitions are orthogonal to type definitions.
@@
   to persist GUI state that is not included in the generated code.}
-  By contrast, livelits are persistent. We use a pure functional model-view-update architecture
-  (i.e. the Elm architecture\todo{what to cite?}{}) as our GUI framework,
+  By contrast, livelits are persistent. We use a pure functional model-view-update-expand architecture
+  (a variation on the Elm architecture \cite{ElmArchitecture}) as our GUI framework,
   so only the model needs to be persisted, rather than the full GUI state.
   We use the word ``literal'' rather than ``palette'' because, with persistence, livelits
-  operate as graphical literal notation, e.g. list literals.
+  operate as graphical literal notation.
 
@@
   At best, palettes can contain text boxes, but expressions entered into these text boxes
-  do not have any editor support (which creates many problems, one of which is that they cannot themselves be generated by palettes).
+  do not have any editor support (and thus cannot themselves be generated by palettes).
   This implies that palettes can only reasonably generate {closed expressions}.
@@
   \end{lstlisting}
-  Parameterization also underlies the hygiene mechanism, as we will discuss in Sec.~\ref{sec:livelit-definitions}.
+  % Parameterization also underlies the hygiene mechanism, as we will discuss in Sec.~\ref{sec:livelit-definitions}.
 

### 2020-07-10 case-studies.tex
finished first pass through sec 3
removing todos

@@
 and an image transformation case study in Sec.~\ref{sec:image-transformation}.
-We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}. 
-These case studies and examples have been implemented\todo{make this not a lie}{}
+We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}.\todo{do we?}{}
+These case studies and examples have been implemented
 in Hazel, a browser-based live programming environment for a pure typed functional language in the
-ML family. We assume basic familiarity with ML.\todo{cite Elm tutorial for readers who are not familiar?}{}
+ML family. We assume basic familiarity with ML.
 
@@
 \begin{figure}
-\begin{lstlisting}
-let weights = [0.1, 0.1, 0.1, 0.1, 0.25, 0.35] in 
-let grades = $dataframe in 
-            A1          A2          A3          A4       Midterm      Final  +
-  Cyrus
-  Nick
-  David
-    + 
-let averages = compute_averages(grades, weights) in 
-let cutoffs = $cutoffs(averages) in
-
-  |---------------------------------------------------|
-
-format_for_university(assign_grades(averages, cutoffs))
-\end{lstlisting}
-\caption{Grading Example}
+    \includegraphics[width=30pc]{grade-cutoff-livelit.png}
+\caption{Grading with Livelits}
 \label{fig:grading}
@@
 A spreadsheet allows the instructor to record student grades using a natural tabular interface,
-visualize this data in one of a finite number of styles, and perform basic computations on this data,
+visualize this data in one of a finite number of plot styles, and perform basic computations on this data,
 with results updated live, i.e. automatically after each cell has been edited.
@@
 are flexible enough to handle these scenarios, but users of these languages
-sacrifice the ability to receive live feedback and directly manipulate the data and visualizations thereof.
+sacrifice the ability to receive live feedback and directly manipulate data and visualizations thereof.
 
 Livelits are able to address this tension.
-Fig.~\ref{fig:grading} shows a Hazel program where the user has:\todo{label each of these 1-5 in figure?}{}
+Fig.~\ref{fig:grading} shows a Hazel program where the user has:
 \begin{enumerate}
@@
 \item determined reasonable \li{cutoffs} between letter grades by directly manipulating a livelit that offers a domain-specific drag-and-drop interface superimposed on a live visualization of the grade distribution
-(for the sake of space, we plot dots rather than a full histogram)\todo{should we do a histogram?}{}, 
+(for the sake of space, we plot dots rather than a full histogram),
 \item programmatically assigned letter grades to each student based on the \li{cutoffs} determined in step (3) by calling another shared library function, \li{assign_grades},
@@
 
-\subsubsection{Livelit Expansion} 
+\subsubsection{Livelit Expansion}\label{sec:livelit-expansion}
 Livelit expressions are given dynamic meaning by expansion to core expressions, i.e. expressions that do not contain livelits.
-For example, the expansion of Fig.~\ref{fig:grading} is:\todo{change code once figure is final}{}
+For example, the expansion of Fig.~\ref{fig:grading} is:
 
@@
   [
-    ("Cyrus", [0.0, 0.0, 0.0, 0.0, 0.0]),
-    ("Nick", [0.0, 0.0, 0.0, 0.0, 0.0]),
-    ("David", [0.0, 0.0, 0.0, 0.0, 0.0])
+    ("Alice", [24. +. 36. +. 33., 92., 83.5, 87.5, 95., 88.]),
+    ("Bob", [61., 64., 98., 89.5, 70., 85.]),
+    ("Carol", [75., 81., 73., 71., 82., 79.]),
+    (* ... *)
   ]) in
-let averages = compute_averages(grades) in 
+let averages = compute_averages grades weights in
 let cutoffs = (.A 90.0, .B 80.0, .C 70.0, .D 60.0) in
-format_for_university(assign_grades(averages, cutoffs))
+format_for_university (assign_grades averages cutoffs)
 \end{lstlisting}
@@
 function bodies to reason about types and binding.
-Instead, in the words of Reynolds\todo{cite}{}, 
+Instead, in the words of \citet{DBLP:conf/ifip/Reynolds83},
 ``type structure is a syntactic discipline for maintaining levels of abstraction''.
@@
 livelit $dataframe at Dataframe { ... }
-livelit $cutoffs(avgs: List(Float)) 
+livelit $grade_cutoffs(avgs: List(Float))
   at (.A Float, .B Float, .C Float, .D Float) { ... }
@@
 (a type classifying tabular floating point data together with string row and column keys, consistent with the expansion above)
-and the expansion type of \li{\$cutoffs} is a labeled product of grade cutoffs (Hazel notates labels \li{.label}).
-Hazel summarizes the typing information from the livelit definition for the programmer when the cursor is on the livelit's name, 
-just as it displays a function's type when its name is under the cursor (not shown).\todo{could show if Hannah et al are finished}{}
+and the expansion type of \li{\$grade_cutoffs} is a labeled product of grade cutoffs (Hazel notates labels \li{.label}).
+Hazel summarizes the typing information from the livelit definition when the cursor is on the livelit's name,
+just as it displays a function's type when its name is under the cursor (not shown).
 
@@
 \paragraph{Parameters}\label{sec:parameterization} Parameters appear immediately after the name of the applied livelit.
-For example, \li{averages} is provided as a parameter to \li{\$cutoffs} in Fig.~\ref{fig:grading}.
+For example, \li{averages} is provided as a parameter to \li{\$grade_cutoffs} in Fig.~\ref{fig:grading}.
 The livelit declares a fixed number of parameters and specifies their types in its declaration.
-For example, the declaration of \li{\$cutoffs} shown above specifies one parameter of type \li{List(Float)}
+For example, the declaration of \li{\$grade_cutoffs} shown above specifies one parameter of type \li{List(Float)}
 (the grades data to be plotted, see below).
@@
 always refer to bindings visible to the client, rather than bindings that are hidden inside expansions.
-This is implemented by alpha-renaming bindings internal to the expansion as necessary.
-(We discuss potentially relaxed variations of this hygiene discipline in Sec.~\ref{sec:discussion}.)
+% This is implemented by alpha-renaming bindings internal to the expansion as necessary.
+% (We discuss potentially relaxed variations of this hygiene discipline in Sec.~\ref{sec:discussion}.)
 
@@
 the evaluation result for each cell when the cursor is not on it, like a spreadsheet.
-The \li{\$cutoffs} livelit uses this facility to plot the grades list, provided
+The \li{\$grade_cutoffs} livelit uses this facility to plot the grades list, provided
 as a parameter, on the number line.
@@
 
-Similarly, the closure for the \li{\$cutoffs} livelit in Fig.~\ref{fig:grading} includes a result for 
+Similarly, the closure for the \li{\$grade_cutoffs} livelit in Fig.~\ref{fig:grading} includes a result for
 the \li{grades} variable. However, this variable's result, along with the
@@
 When a livelit requests an evaluation result, it must be able to handle indeterminate results.
-For example, if the parameter to \li{\$cutoffs} were a hole or some other indeterminate expression\todo{small screenshot?}{}, 
+For example, if the parameter to \li{\$grade_cutoffs} were a hole or some other indeterminate expression,
 then it would have degraded functionality
-(in this case, it would display the list elements that are values on the timeline, skipping indeterminate elements\todo{implement this}{}). 
-We will return to how livelit implementations handle this situation in Sec.~\ref{sec:implementation}.
-\todo{make sure we cover this}{}
+(in this case, it would display the list elements that are values on the timeline, skipping indeterminate elements).
+We will return to how livelit implementations handle this situation in Sec.~\ref{sec:livelit-definitions}.
 

### 2020-07-10 intro.tex
Restore subfigure labels
removing todos

@@
       \centering
-      \includegraphics[width=15pc]{slider-color-livelits.png}
+      \vspace{-3.4cm}
+      \includegraphics[width=15pc]{graphite-color-palette.png}
     \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
@@
 Figure~\ref{fig:color}(a)\todo{use subfigure?}{}, adapted from this prior work, demonstrates a color palette invoked using Graphite.
-When the user presses the \li{Enter} key, the Java expression \li{new Color(173, 173, 173, 94)}\todo{correct code}{} is inserted at the cursor (not shown) and the palette disappears.
+When the user presses the \li{Enter} key, the Java expression \li{new Color(173, 173, 173, 85)} is inserted at the cursor (not shown) and the palette disappears.
 
@@
 Our focus in this paper is on addressing several non-trivial technical
-deficiencies that limit both palette providers and palette clients in Graphite.
+deficiencies that limit both palette providers and palette clients in Graphite
+(and other projectional editing systems).
 The system of \emph{live literals}, or \emph{livelits} introduced in
@@
   \begin{lstlisting}[numbers=none]
-  let livelit $percentage = $slider(0, 100) in ...
+  let $percentage = $slider 0 100 in ...
   \end{lstlisting}
@@
   leaving the hole being filled by the livelit temporarily unfilled and then evaluating
-  using a two-phased variation on the semantics for incomplete programs developed for Hazel \cite{HazelnutLive}.
-  We will see more sophisticated examples later in the paper, including a livelit
+  using a two-phased variation on the semantics for incomplete programs developed for the \Hazel programming environment \cite{HazelnutLive}.
+  We will see more sophisticated examples later on, including a livelit
   that appears inside a function that is called multiple times, leading to multiple closures that the client can
@@
 This calculus and the key metatheoretic properties has been mechanized in Agda.
-In Sec.~\ref{sec:extensions}, we describe some extensions of the core calculus\todo{what are the extensions?}{}.
 In Sec.~\ref{sec:implementation}, we provide a more detailed account of our implementation of livelits within Hazel,

### 2020-07-10 livelit-definitions.tex
tweaked text for grading case study
removing todos

@@
 \section{Livelit Definitions}\label{sec:livelit-definitions}
-Which livelit implementation do we want to show? Maybe grade cutoffs? Slider as simple intro + color chooser?
 
-Talk about each component of a livelit definition:
-\begin{itemize}
-  \item livelit name (talk about binding structure)
-  \item expansion type
-  \item model type (must be serializable)
-  \item action type (mention integration into Hazel's undo system)
-  \item initial model
-  \item update function (monadic actions; talk about color palette update; SpliceRefs)
-  \item view function (monadic actions; talk about GUI stuff? tab order? dimensions? mostly defer to Elm for details)
-  \item expand function (should we call it fill?)
-\end{itemize}
+\begin{figure}
+\begin{lstlisting}
+  type Color = (.r Int, .g Int, .b Int, .a Int)
 
-Hygiene from a provider's perspective -- have to use parameters to explicitly capture definition site bindings.
+  livelit $color at Color {
+    capture { }
+
+    type Model = (.r SpliceRef, .g SpliceRef, .b SpliceRef, .a SpliceRef)
+    let init : UpdateMonad(Model) = do 
+      r <- new_splice(`Int`, Some(`0`))
+      g <- new_splice(`Int`, Some(`0`))
+      b <- new_splice(`Int`, Some(`0`))
+      a <- new_splice(`Int`, Some(`100`))
+      return (r, g, b, a)
+
+    type Action = 
+    | ClickOnColor(Color)
+
+    let view : Model -> ViewMonad(Html(Action)) = 
+      fun model -> do 
+        (* determine a color to display to the client, if possible *)
+        r_res <- eval_splice(model.r)
+        g_res <- eval_splice(model.g)
+        b_res <- eval_splice(model.b)
+        a_res <- eval_splice(model.a)
+        let cur_color : Color = 
+          case (r_res, g_res, b_res, a_res) 
+          | (Some(Val(IntLit(r))), 
+             Some(Val(IntLit(g))), 
+             Some(Val(IntLit(b))), 
+             Some(Val(IntLit(a)))) -> 
+            Some((r, g, b, a))
+          | _ -> None
+        in
+
+        r_editor <- editor(model.r, FixedWidth(20))
+        g_editor <- editor(model.g, FixedWidth(20))
+        b_editor <- editor(model.b, FixedWidth(20))
+        a_editor <- editor(model.a, FixedWidth(20))
+
+        (* ...render the UI using these splice editors... *)
+
+    let update : Model -> Action -> UpdateMonad(Model) =
+      fun model (ClickOnColor(r, g, b, a)) -> do 
+        set_splice(model.r, IntLit(r))
+        set_splice(model.g, IntLit(g))
+        set_splice(model.b, IntLit(b))
+        set_splice(model.a, IntLit(a))
+        return (r, g, b, a)
+
+    let expand : Model -> (Exp, List(SpliceRef)) = 
+      fun model -> 
+        (`fun r g b a -> (r, g, b, a)`, [model.r, model.g, model.b, model.a])
+  }
+\end{lstlisting}
+\caption{Livelit Implementation: \li{\$color} (simplified for presentation)}
+\label{fig:color-impl}
+\end{figure}
+
+\noindent
+Let us now look in more detail at a livelit definition. Fig.~\ref{fig:color-impl}
+outlines the \li{\$color} livelit from Fig.~\ref{fig:color}. (We omit certain 
+incidental implementation details and use unimplemented syntactic sugar, including
+Haskell-style \li{do} notation and simplified quasiquotation, for the sake of presentation.
+We say more about the implementation in Sec.~\ref{sec:implementation}.)
+
+Line 3 of Fig.~\ref{fig:color-impl} specifies the livelit's name, \li{\$color}, and its \emph{expansion type},
+\li{Color}, which was discussed in Sec.~\ref{sec:livelit-expansion}.
+The remainder of the definition is the livelit's implementation.
+Livelits are implemented using a variation on the functional model-view-update
+framework popularized by the Elm programming language \cite{ElmArchitecture}. We add a fourth component,
+expansion generation. In addition, we add a simple monadic framework to provide the necessary  
+interface between the livelit and the programming environment, discussed below, while retaining
+a pure functional programming style.
+
+\subsection{Model}\label{sec:model}
+Line 6 of Fig.~\ref{fig:color-impl} specifies the livelit's \emph{model type},
+here a labeled 4-tuple of \emph{splice references}, one for each of the four splices
+that appear in the GUI in Fig.~\ref{fig:color}.
+Models are persisted in the syntax tree of the program, so in practice,
+ the system would check that the model type supports automatic serialization. 
+ (Hazel is able to serialize all types because it is implemented as a 
+ simple interpreter.)
+
+The initial value of the model is specified starting on Line 7, 
+which defines a command in the \li{UpdateMonad} that returns the 
+initial model value after generating four new splice references
+using the \li{new_splice} command:
+\begin{lstlisting}[numbers=none]
+new_splice : (Typ, Option(Exp)) -> UpdateMonad(SpliceRef)
+\end{lstlisting}
+This command creates a fresh splice
+of the given type and, optionally, with the given initial expression,
+and returns a unique splice reference, which uniquely and abstractly identifies that splice.
+Bolded types like \li{Typ}, \li{Exp}, \li{UpdateMonad}, and \li{SpliceRef}
+are provided by the standard library. The \li{Typ} and \li{Exp} types 
+encode the syntax of Hazel's types and expressions.
+These two arguments are provided here in quasiquoted form, e.g. \li{`0`} \cite{bawden1999quasiquotation}. 
+If no initial expression is provided, the splice is initialized with 
+an empty hole.
+
+The system checks that the provided type and initial content is valid 
+assuming only the explicitly specified set of captured bindings on Line 4.
+Here, we do not need to capture any bindings because \li{Int} is a built in type
+and the initial content is closed. However, 
+for splices of user-defined types, those types need to be explicitly captured.
+Similarly, helper functions that are used in initial splices (and in the expansion,
+discussed below) must be explicitly captured. 
+This serves to ensure \emph{context independence} -- the livelit needs to make no 
+assumptions about the typing context at use sites.
+We use an explicit capture set, rather than implicitly capture of all bindings at the definition site, 
+both to prevent inadvertent capture 
+and to ensure that the capture set is part
+of the interface of the livelit definition, which simplifies matters related to 
+module export and determining paths to the captured bindings at use sites \cite{TLMs}.
+Formally, captured variables can be understood as parameters that have been 
+immediately partially applied.
+
+\subsection{Action}
+Line 14 defines the \li{Action} type for the \li{\$color} livelit, which 
+specifies a single user-initiated action: clicking on a color using the user 
+interface on the right side of Fig.~\ref{fig:color}. Actions are emitted
+from event handlers (e.g. click handlers) defined in the computed \li{view}, 
+and actions are consumed by the \li{update} function, causing a change in the model. 
+Let us discuss each of these functions in turn.
+
+\subsection{View}
+The \li{view} function computes the view given a model and access to the commands in 
+the \li{ViewMonad}. The computed view is a value of type \li{Html(Action)}.  
+The type family \li{Html(a)} provides a simple immutable
+encoding of an HTML tree, where the type parameter \li{a} is the type of actions that 
+are emitted by event handlers that can be attached to elements of the tree, e.g.
+\li{on_click} and so on. 
+We elide the details of the particular user interface in Fig.~\ref{fig:color}
+to focus on two key mechanisms exposed by \li{ViewMonad}: live evaluation 
+and splice rendering.
+
+\subsubsection{Live Evaluation}
+In order to provide the client with live feedback about the dynamic implications 
+of the programmer's choices, 
+the view function can ask the system to evaluate a splice, or an encoded function of 
+multiple splices as long as the function is well-typed using only the capture set, 
+under the closure that the client has selected using one of the following
+commands:
+\begin{lstlisting}[numbers=none]
+eval_splice : SpliceRef -> ViewMonad(Option(Result))
+eval_fun    : (Exp, List(SpliceRef)) -> ViewMonad(Option(Result))
+\end{lstlisting}
+
+The \li{None} case 
+arises when evaluation is not possible, either because no closures are available
+or because the free variables in the selected closure overlap with the free 
+variables in the provided expression. This occurs only when the user has selected
+a closure that appears under a binder in the evaluation result (e.g. because 
+the result is a lambda). 
+We discuss this further in Sec.~\ref{sec:calculus-closure-collection}.
+
+If a result is available, the \li{Result} type distinguishes two possibilities:
+\begin{lstlisting}[numbers=none]
+type Result = 
+| Val(Exp)
+| Indet(Exp)
+\end{lstlisting}
+The \li{Val} case arises when evaluation produces a true value, whereas the 
+\li{Indet} case arises when evaluation results in an indeterminate expression,
+i.e. an expression that cannot be further evaluated due to one or more holes 
+in critical elimination positions \cite{HazelnutLive}.
+
+In our example, the livelit determines a color to display to the client 
+if all four splices evaluate to integer literals. Otherwise, there is not 
+enough information to determine a color, and the livelit can decide how to 
+communicate this information to the client, e.g. with an "X" over the color 
+preview.
+
+Livelits can attempt to offer feedback even when the result is indeterminate,
+because indeterminate expressions might nevertheless contain useful information.
+For example, a livelit that previews a sequence of notes as audio might be able 
+to handle a list of notes where certain notes are missing, i.e. holes, by 
+simply playing silence or some other default sound when it encounters them.
+This behavior is highly domain-specific, so each livelit provider must decide 
+whether and how indeterminate results are supported.
+
+\subsubsection{Splice Rendering}
+The remainder of the \li{view} function is responsible for computing the view.
+This is straightforward but for the question of splices. In Fig.~\ref{fig:color},
+we want the splices to render as expression editors, with all of Hazel's editor 
+services available to the user. In order to support this, the \li{view} function
+can request an editor for a given splice, with given dimensions:
+\begin{lstlisting}[numbers=none]
+editor : (SpliceRef, Dimensions) -> ViewMonad(Html(a))
+\end{lstlisting}
+The result is an opaque \li{Html(a)} value that the remainder of the function 
+can place as needed. When the livelit is rendered, this part of the tree is 
+under the control of Hazel. The \li{Dimensions} parameter currently supports only a fixed
+character width, with overflow causing scrolling, but in the future we plan to offer 
+to offer more flexible layout options.
+
+Another situation that can arise is when the UI needs to include not an editor but 
+a rendered evaluation result. For example, each of the cells in the \li{\$dataframe}
+livelit in Fig.~\ref{fig:grading} shows the evaluation result for the corresponding 
+cell. Only the formula bar at the top is an editor. To support this, the \li{view}
+function can use the \li{result_view} commands, which mirror the \li{eval} commands:
+\begin{lstlisting}[numbers=none]
+result_view_splice : (SpliceRef, Dimensions) -> ViewMonad(Html(a))
+result_view_fun    : (Exp, List(SpliceRef), Dimensions) -> ViewMonad(Html(a))
+\end{lstlisting}
+
+\subsection{Update}
+When the user triggers an event in a livelit view, it emits an \li{Action}.
+The system responds by calling the \li{update} function to determine how 
+this action should affect the model and the splices.
+
+We discussed in Sec.~\ref{sec:model} that the \li{UpdateMonad}
+allows new splices to be generated in response to user actions. 
+For example, the \li{\$dataframe} livelit uses this when new rows
+or columns are added.
+
+In Fig.~\ref{fig:color-impl}, we see that the \li{\$color} livelit responds to 
+the \li{ClickOnColor} action by invoking the \li{set_splice} command to overwrite 
+the current splices with integer literals determined based on which color the user
+clicked on:
+\begin{lstlisting}[numbers=none]
+  set_splice : (SpliceRef, Exp) -> UpdateMonad(Unit)
+\end{lstlisting}
+The provided expression must satisfy the splice type and only make use of the capture set.
+
+When the model is updated, a new view is 
+computed. The system then performs a diff between the old and new view in order to 
+efficiently perform the necessary imperative updates to the editor's visual state.
+Changes to splices can also cause the view to be recomputed, because the view might 
+need to evaluate the splices. The \li{update} function does not itself 
+have the ability to request evaluation, because the model should not depend directly  
+which closure the user has selected. Of course, the \li{view} might emit 
+result-dependent actions when appropriate.
+
+\subsection{Expansion}
+The ultimate purpose of a livelit is to fill the hole where it appears by generating an expansion,
+i.e. a standard symbolic expression of the expansion type, here \li{Color}.
+The \li{expand} function, shown in Fig.~\ref{fig:color-impl}, is responsible for generating 
+an expansion given the current model.
+
+The expansion can include splices, but we the system does not make the contents of a splice 
+available directly. Instead, the \li{expand} must return a pair consisting of an encoded expression, of type 
+\li{Exp}, with a list of \li{SpliceRef}s that the expansion depends on. 
+We call the \li{Exp} the \emph{parameterized expansion}
+ because it must take an argument for each listed \li{SpliceRef}s. 
+ That argument will be the corresponding splice 
+type, which was provided when the splice was initialized. 
+ The return type of the parameterized expansion is the expansion type, here \li{Color}.
+
+This parameterization strategy makes it simple to enforce hygiene: the parameterized expansion 
+can depend only on the capture set, whereas the splices are entered by the client and so they can 
+depend only on the client site typing context. The use of standard function application ensures
+that splices are capture avoiding. We consider this more formally in the next section.

### 2020-07-10 case-studies.tex
Add image transformation figure
sketchy conclusion, removed examples sec

@@
 and an image transformation case study in Sec.~\ref{sec:image-transformation}.
-We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}.\todo{do we?}{}
-These case studies and examples have been implemented
+% We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}.\todo{do we?}{}
+These case studies have been implemented
 in Hazel, a browser-based live programming environment for a pure typed functional language in the
@@
 
-\subsection{Additional Examples}\label{sec:additional-examples}
-It would be nice to have a gallery-style figure and a brief overview of some other case studies
-and how they exercise the novel features of the livelits mechanism. Maybe some statistics on how
-many lines of code it took.
+% \subsection{Additional Examples}\label{sec:additional-examples}
+% It would be nice to have a gallery-style figure and a brief overview of some other case studies
+% and how they exercise the novel features of the livelits mechanism. Maybe some statistics on how
+% many lines of code it took.
 
-Ideas:
-\begin{itemize}
-  \item derivation trees like Joomy's system (\url{https://joom.github.io/proof-tree-builder/src/})
-\end{itemize}
+% Ideas:
+% \begin{itemize}
+%   \item derivation trees like Joomy's system (\url{https://joom.github.io/proof-tree-builder/src/})
+% \end{itemize}

### 2020-07-10 discussion.tex
sketchy conclusion, removed examples sec

@@
 \section{Discussion and Conclusion}\label{sec:discussion}\label{sec:conclusion}
-Future work:
-\begin{itemize}
-  \item pattern matching
-  \item type splices
-  \item module splices
-  \item explicitly make bindings available in splices
-  \item better UI for closure provenance / connect with control flow / call stack better
-  \item deriving livelits from type definitions
-  \item bidirectional evaluation ala SnS
-  \item integration with structure editing more cleanly
-  \item \dots
-  \item collaborative editing?
-  \item side effects?
-  \item full screening livelits (or compositions thereof) as a way to create end-user workflows
-  \item big vision: authoring environment based on livelits
-\end{itemize}
+\begin{quote}
+  %
+  \textit{
+  %
+  The arithmetical symbols are written diagrams and the geometrical figures are graphic formulas.''
+  %
+  }
+
+  \vspace{3pt}
+
+  \hfill{}--- David Hilbert~\cite{hilbert1902mathematical}
+  \end{quote}
+
+  Diagrams have played a pivotal role in mathematical thought since antiquity,
+  indeed long predating symbolic mathematics \cite{cajori1993history}. 
+  Popular computing and creative tooling, too, has embraced visual representation and direct manipulation 
+  interfaces for decades.
+  Programming, however, has remained stubbornly mired in textual user interfaces. 
+  Certainly, one must acknowledge that textual and symbolic notation
+  is an indispensable tool for abstract thought. 
+  Indeed, it is now widely recognized that variables and functions are at the 
+  very foundation of computing. 
+  Our hope with this paper is to demonstrate that principled, mathematically structured
+  programming is not only compatible with live graphical user interfaces, but that the 
+  combination of these two holds tremendous promise for the future of programming.
+
+  There remain a number of avenues for future work. First, we have only just started
+  to apply the livelits mechanism in various application domains. As the implementation
+  matures, we plan to introduce enthusiasts in a wide variety of problem domains
+  to livelits and continue the empirical evaluations started by \citet{Graphite}.
+
+  In particular, we would like to better understand the conceptual and practical 
+  difficulties that livelit providers face, with the aim of making livelit 
+  implementation an extremely low cost activity. Livelits could themselves be 
+  useful for this task, if we develop a GUI widget library with support for livelits.
+In addition, recent approaches to deriving type-specific structure editors automatically from pretty printing logic \cite{hempeltiny} could perhaps 
+be adapted to generate livelit implementations. Mechanisms for deriving simple 
+livelit definitions from type definitions, perhaps similar to Haskell's \li{deriving} directive, 
+may also prove helpful \cite{magalhaes2010generic}. The GEC toolkit has also explored  
+the automatic generation of simple user interfaces from type definitions \cite{DBLP:conf/afp/AchtenEPW04}.
+
+  The livelits mechanism as described in this paper operates only on expressions,
+  but livelits might be useful for generating other sorts of terms, such as types
+  and entire modules. Support for type and module splices, too, would likely be useful.
+  With these, better support for reflecting on the provided type might allow for 
+  the development of more general livelits, e.g. a variant of \li{\$datatype} 
+  with individually typed columns, rather than only floating point data.
+
+  The strict hygiene discipline that we maintain has, we believe, substantial 
+  benefits---programmers will inevitably encounter unfamiliar livelits, and 
+  the reasoning principles that we enforce are likely to help them ``reason around''
+  the situation. However, it may be useful in certain circumstances to carefully
+  relax these restrictions. For example, some livelits might benefit from the 
+  ability to introduce bindings into splices, which is currently prevented by 
+  the capture avoidance discipline. Perhaps this could be done in an 
+  editor capable of highlighting the unusual situation.
+
+  Another direction for future work has to do with pushing edits from computed results
+  back into livelits. For example, a slider expands to a number, which may 
+  then flow through a computation. Bidirectional evaluation techniques may allow
+  the user to edit a number in the result of a computation and see the necessary
+  change to a slider in the program.
+
+  Programming and authoring have much in common. Documents often contain structured
+  information, and programs are written to manipulate structured information.
+  Another future direction for livelits is as the basis for a programmable authoring 
+  system, where each element on a page is actually an expression in a programming language
+  being rendered by a livelit. Taking this further, a networked collection of these
+  documents could form a powerful computational wiki. This would 
+  require addressing the difficult problem of supporting collaborative interactions 
+  involving edits to arbitrary user interfaces. 
+  We believe that these problems are surmountable.
+
+% \begin{itemize}
+%   \item pattern matching
+%   \item type splices
+%   \item module splices
+%   \item explicitly make bindings available in splices
+%   \item better UI for closure provenance / connect with control flow / call stack better
+%   \item deriving livelits from type definitions
+%   \item bidirectional evaluation ala SnS
+%   \item integration with structure editing more cleanly
+%   \item \dots
+%   \item collaborative editing?
+%   \item side effects?
+%   \item full screening livelits (or compositions thereof) as a way to create end-user workflows
+%   \item big vision: authoring environment based on livelits
+% \end{itemize}

### 2020-07-10 case-studies.tex
image transformations case study text

@@
 \subsection{Case Study: Image Transformations}\label{sec:image-transformation}
-Show example of an image transformation pipeline going through livelits
 
-This might also benefit from Parameterization
+\begin{figure*}
+  \begin{center}
+    \begin{subfigure}[t]{0.5\textwidth}
+      \centering
+      \includegraphics[width=15pc]{img-filter-1.png}
+      \caption{}
+    \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
+      \centering
+      \includegraphics[width=15pc]{img-filter-2.png}
+      \caption{}
+    \end{subfigure}
+  \end{center}
+  \caption{Case Study: Image Transformation. The image shown is determined based on the selected closure.}
+  \label{fig:img-transformation}
+\end{figure*}
 
-Show multiple calls with different example images + closure selector UIs
+Our next case study is a simple image transformation livelit, \li{\$img_filter},
+demonstrated in Fig.~\ref{fig:img-transformation}. This livelit takes 
+a URL to an image as a parameter, and contains two splices of type \li{Int},
+one to adjust the contrast, and the other to adjust the brightness.
+In this example, we have filled those splices with a \li{\$slider}, but 
+as above we could enter any expression of type \li{Int}.
+The livelit shows a live preview of the transformed image.
+The expansion generates the necessary calls to image processing functions, 
+not shown.
 
-Go into more detail about how evaluation works + fill-and-resume mechanics (and efficiency nod)
-
-Talk about probes?
-
-We assume that the user has selected at most one closure for each livelit in the unexpanded expression.
-The corresponding environment is applied as a substitution before attempting to evaluate the requested expression.
-To maintain the context independence principle, this request must take the form of a closed function
-of the splices.
-As discussed in Sec.~\ref{sec:live-evaluation}, evaluation can fail if the free variables in the splices
-overlap with the free variables in selected closure.
+The novelty in this example is that the livelit appears inside a function, 
+\li{classic}, which takes an image URL as input. At the bottom 
+of the figure, \li{classic} is mapped over a list (here, containing only 
+two elements for the sake of presentation). Consequently, there are now 
+two closures associated with the application of \li{\$img_filter} in the 
+body of \li{classic}. Rather than disabling live evaluation in situations like 
+this, Hazel instead allows the programmer to select between the closures 
+as shown (inset). The live feedback is derived based on the selected closure.
+This makes it easy to see how the image filter being designed here will affect a
+number of example images by quickly toggling between closures. The underlying expansion remains abstract, i.e. it refers to the image via the \li{url} variable.
 

### 2020-09-06 case-studies.tex
popl21 reviews

--- 2020-07-10 case-studies.tex
@@
 and an image transformation case study in Sec.~\ref{sec:image-transformation}.
-We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}.\todo{do we?}{}
-These case studies and examples have been implemented
+We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}.
+These case studies and examples have been implemented\todo{make this not a lie}{}
 in Hazel, a browser-based live programming environment for a pure typed functional language in the
-ML family. We assume basic familiarity with ML.
+ML family. We assume basic familiarity with ML.\todo{cite Elm tutorial for readers who are not familiar?}{}
 
@@
     \includegraphics[width=30pc]{grade-cutoff-livelit.png}
-\caption{Grading with Livelits}
+\caption{Grading Example}
 \label{fig:grading}
@@
 A spreadsheet allows the instructor to record student grades using a natural tabular interface,
-visualize this data in one of a finite number of plot styles, and perform basic computations on this data,
+visualize this data in one of a finite number of styles, and perform basic computations on this data,
 with results updated live, i.e. automatically after each cell has been edited.
@@
 are flexible enough to handle these scenarios, but users of these languages
-sacrifice the ability to receive live feedback and directly manipulate data and visualizations thereof.
+sacrifice the ability to receive live feedback and directly manipulate the data and visualizations thereof.
 
 Livelits are able to address this tension.
-Fig.~\ref{fig:grading} shows a Hazel program where the user has:
+Fig.~\ref{fig:grading} shows a Hazel program where the user has:\todo{label each of these 1-5 in figure?}{}
 \begin{enumerate}
@@
 \item determined reasonable \li{cutoffs} between letter grades by directly manipulating a livelit that offers a domain-specific drag-and-drop interface superimposed on a live visualization of the grade distribution
-(for the sake of space, we plot dots rather than a full histogram),
+(for the sake of space, we plot dots rather than a full histogram)\todo{should we do a histogram?}{},
 \item programmatically assigned letter grades to each student based on the \li{cutoffs} determined in step (3) by calling another shared library function, \li{assign_grades},
@@
 Livelit expressions are given dynamic meaning by expansion to core expressions, i.e. expressions that do not contain livelits.
-For example, the expansion of Fig.~\ref{fig:grading} is:
+For example, the expansion of Fig.~\ref{fig:grading} is:\todo{change code once figure is final}{}
 
@@
   [
-    ("Alice", [24. +. 36. +. 33., 92., 83.5, 87.5, 95., 88.]),
-    ("Bob", [61., 64., 98., 89.5, 70., 85.]),
-    ("Carol", [75., 81., 73., 71., 82., 79.]),
-    (* ... *)
+    ("Cyrus", [0.0, 0.0, 0.0, 0.0, 0.0]),
+    ("Nick", [0.0, 0.0, 0.0, 0.0, 0.0]),
+    ("David", [0.0, 0.0, 0.0, 0.0, 0.0])
   ]) in
-let averages = compute_averages grades weights in
+let averages = compute_averages(grades) in
 let cutoffs = (.A 90.0, .B 80.0, .C 70.0, .D 60.0) in
-format_for_university (assign_grades averages cutoffs)
+format_for_university(assign_grades(averages, cutoffs))
 \end{lstlisting}
@@
 function bodies to reason about types and binding.
-Instead, in the words of \citet{DBLP:conf/ifip/Reynolds83},
+Instead, in the words of Reynolds\todo{cite}{},
 ``type structure is a syntactic discipline for maintaining levels of abstraction''.
@@
 livelit $dataframe at Dataframe { ... }
-livelit $grade_cutoffs(avgs: List(Float))
+livelit $cutoffs(avgs: List(Float))
   at (.A Float, .B Float, .C Float, .D Float) { ... }
@@
 (a type classifying tabular floating point data together with string row and column keys, consistent with the expansion above)
-and the expansion type of \li{\$grade_cutoffs} is a labeled product of grade cutoffs (Hazel notates labels \li{.label}).
-Hazel summarizes the typing information from the livelit definition when the cursor is on the livelit's name,
-just as it displays a function's type when its name is under the cursor (not shown).
+and the expansion type of \li{\$cutoffs} is a labeled product of grade cutoffs (Hazel notates labels \li{.label}).
+Hazel summarizes the typing information from the livelit definition for the programmer when the cursor is on the livelit's name,
+just as it displays a function's type when its name is under the cursor (not shown).\todo{could show if Hannah et al are finished}{}
 
@@
 \paragraph{Parameters}\label{sec:parameterization} Parameters appear immediately after the name of the applied livelit.
-For example, \li{averages} is provided as a parameter to \li{\$grade_cutoffs} in Fig.~\ref{fig:grading}.
+For example, \li{averages} is provided as a parameter to \li{\$cutoffs} in Fig.~\ref{fig:grading}.
 The livelit declares a fixed number of parameters and specifies their types in its declaration.
-For example, the declaration of \li{\$grade_cutoffs} shown above specifies one parameter of type \li{List(Float)}
+For example, the declaration of \li{\$cutoffs} shown above specifies one parameter of type \li{List(Float)}
 (the grades data to be plotted, see below).
@@
 always refer to bindings visible to the client, rather than bindings that are hidden inside expansions.
-% This is implemented by alpha-renaming bindings internal to the expansion as necessary.
-% (We discuss potentially relaxed variations of this hygiene discipline in Sec.~\ref{sec:discussion}.)
+This is implemented by alpha-renaming bindings internal to the expansion as necessary.
+(We discuss potentially relaxed variations of this hygiene discipline in Sec.~\ref{sec:discussion}.)
 
@@
 the evaluation result for each cell when the cursor is not on it, like a spreadsheet.
-The \li{\$grade_cutoffs} livelit uses this facility to plot the grades list, provided
+The \li{\$cutoffs} livelit uses this facility to plot the grades list, provided
 as a parameter, on the number line.
@@
 
-Similarly, the closure for the \li{\$grade_cutoffs} livelit in Fig.~\ref{fig:grading} includes a result for
+Similarly, the closure for the \li{\$cutoffs} livelit in Fig.~\ref{fig:grading} includes a result for
 the \li{grades} variable. However, this variable's result, along with the
@@
 When a livelit requests an evaluation result, it must be able to handle indeterminate results.
-For example, if the parameter to \li{\$grade_cutoffs} were a hole or some other indeterminate expression,
+For example, if the parameter to \li{\$cutoffs} were a hole or some other indeterminate expression\todo{small screenshot?}{},
 then it would have degraded functionality
-(in this case, it would display the list elements that are values on the timeline, skipping indeterminate elements).
-We will return to how livelit implementations handle this situation in Sec.~\ref{sec:livelit-definitions}.
+(in this case, it would display the list elements that are values on the timeline, skipping indeterminate elements\todo{implement this}{}).
+We will return to how livelit implementations handle this situation in Sec.~\ref{sec:implementation}.
+\todo{make sure we cover this}{}
 

### 2020-09-06 discussion.tex
popl21 reviews

--- 2020-05-24 discussion.tex
@@
 \section{Discussion and Conclusion}\label{sec:discussion}\label{sec:conclusion}
+\begin{quote}
+  %
+  \textit{
+  %
+  ``The arithmetical symbols are written diagrams and the geometrical figures are graphic formulas.''
+  %
+  }
+
+  \vspace{3pt}
+
+  \hfill{}--- David Hilbert~\cite{XXX}
+  \end{quote}
+
 Future work:
+Recent approaches to deriving type-specific structure editors automatically from pretty printing logic \cite{hempeltiny} could perhaps 
+be adapted to generate livelit implementations. Mechanisms for deriving simple 
+livelit definitions from type definitions, perhaps similar to Haskell's \li{deriving} directive, 
+may also prove helpful \cite{magalhaes2010generic}. The GEC toolkit has explored  
+the automatic generation of simple user interfaces from type definitions \cite{DBLP:conf/afp/AchtenEPW04}.
+
 \begin{itemize}

### 2020-09-06 livelit-definitions.tex
popl21 reviews

@@
 
-Line 3 of Fig.~\ref{fig:color-impl} specifies the livelit's name, \li{\$color}, and its \emph{expansion type},
+Line 1 of Fig.~\ref{fig:color-impl} specifies the livelit's name, \li{\$color}, and its \emph{expansion type},
 \li{Color}, which was discussed in Sec.~\ref{sec:livelit-expansion}.
@@
 Livelits are implemented using a variation on the functional model-view-update
-framework popularized by the Elm programming language \cite{ElmArchitecture}. We add a fourth component,
+framework popularized by the Elm programming language\todo{cites}. We add a fourth component,
 expansion generation. In addition, we add a simple monadic framework to provide the necessary  
 interface between the livelit and the programming environment, discussed below, while retaining
-a pure functional programming style.
+a pure functional programming model.\todo{cite monads in Haskell}
 
 \subsection{Model}\label{sec:model}
-Line 6 of Fig.~\ref{fig:color-impl} specifies the livelit's \emph{model type},
+\todo{final check of line numbers}{}Line 2 of Fig.~\ref{fig:color-impl} specifies the livelit's \emph{model type},
 here a labeled 4-tuple of \emph{splice references}, one for each of the four splices
@@
 
-The initial value of the model is specified starting on Line 7, 
+The initial value of the model is specified starting on Line 4, 
 which defines a command in the \li{UpdateMonad} that returns the 
@@
 encode the syntax of Hazel's types and expressions.
-These two arguments are provided here in quasiquoted form, e.g. \li{`0`} \cite{bawden1999quasiquotation}. 
+These two arguments are provided here in quasiquoted form, e.g. \li{`0`}\todo{cite something}. 
 If no initial expression is provided, the splice is initialized with 
@@
 The system checks that the provided type and initial content is valid 
-assuming only the explicitly specified set of captured bindings on Line 4.
-Here, we do not need to capture any bindings because \li{Int} is a built in type
+assuming only the explicitly specified set of captured bindings on Line 3.
+Here, we do not need to capture anything because \li{Int} is a built in type
 and the initial content is closed. However, 
@@
 We use an explicit capture set, rather than implicitly capture of all bindings at the definition site, 
-both to prevent inadvertent capture 
+both to prevent inadvertent capture\todo{cite heathers paper}{} 
 and to ensure that the capture set is part
@@
 module export and determining paths to the captured bindings at use sites \cite{TLMs}.
-Formally, captured variables can be understood as parameters that have been 
-immediately partially applied.
 
 \subsection{Action}
-Line 14 defines the \li{Action} type for the \li{\$color} livelit, which 
+Line 11 defines the \li{Action} type for the \li{\$color} livelit, which 
 specifies a single user-initiated action: clicking on a color using the user 
@@
 Changes to splices can also cause the view to be recomputed, because the view might 
-need to evaluate the splices. The \li{update} function does not itself 
+need to evaluate the splices\todo{optimizaiton?}. The \li{update} function does not itself 
 have the ability to request evaluation, because the model should not depend directly  

### 2020-09-06 related-work.tex
popl21 reviews

@@
 \section{Related Work}\label{sec:related-work}
-Cyrus has a list somewhere, ask if you want to work on this section
+As detailed in Sec.~\ref{sec:intro},
+the \emph{Graphite} system introduced the idea of filling 
+typed holes using a type-specific user interface \cite{Graphite}.
+This prior work engaged in substantial qualitative evaluation, 
+which due to the fundamental similarities between the two systems 
+is as relevant to our design as theirs. However, the prior work 
+left a number of core technical issues unresolved, 
+as summarized in Sec.~\ref{sec:intro}. 
+Livelits resolve these issues in large part by bringing together 
+seemingly disparate ideas from several other pieces of recent work.
+
+In particular, recent work on type-specific languages (TSLs) \cite{TSLs} 
+and typed literal macros (TLMs) \cite{TLMs}
+explored similar ideas of user-defined
+literal forms with support for hygienic splicing, with the former 
+using type-directed dispatch similar to Graphite and the latter supporting 
+decentralized extensibility via explicit naming as in our approach. 
+However, these systems operate in a purely 
+textual setting. There is no support for graphical notation or 
+interaction, nor live feedback, which are the primary contributions of this paper.
+The typed expansion judgement central to the typed livelit calculus in this paper 
+is structured similarly 
+to the corresponding judgements in the formal systems describing TLMs and TSLs,
+and the reasoning principles (e.g. hygiene properties, the focus on reasoning abstractly 
+about typing) was inspired directly by this work. However, the  
+approach to hygiene we take is cleaner by its use of function application 
+rather than direct insertion of splices (and could perhaps be ported to those formalisms).
+Splices also operate quite differently in our work,
+because they are placed automatically and structurally delimited in the user interface, rather than 
+placed by the client and then parsed out of the text by a custom parser charged 
+with retaining provenance information.
+This substantially complicates the design of splices in the prior work.
+
+This structural delimitation of different notations is reminiscent of work on 
+\emph{language boxes} \cite{DBLP:conf/sle/RenggliDN09}, which focused primarily on 
+combining different notations, primarily textual, using structural delimiters inserted 
+explicitly by the user using a special-purpose editor.,
+
+There has been a long line of research on \emph{projectional editing}, where
+the user edits graphical representations (projections) of code constructs \cite{DBLP:journals/ile/MillerPMV94,read1996generating,DBLP:conf/sde/Reiss84,DBLP:conf/uist/KoM05}. 
+Many of the oldest systems offer only a fixed set of projections and interaction techniques. 
+More recently, language workbenches with support for projectional editing like Citrus \cite{DBLP:conf/uist/KoM05}
+and MPS \cite{voelter2011language} have made it easier to define new projectional editors.
+However, these systems generate entire editors, whereas livelit definitions can be 
+included granularly in a livelit context and support decentralized extensibility.
+Furthermore, livelit definitions are governed by a type and binding discipline that 
+serves to ensure that a programmer unfamiliar with the semantics of an unfamiliar livelit
+can still be sure that it is not operating outside of the bounds of the type system.
+In many ways, an unfamiliar livelit can be approached much like an unfamiliar function
+can be approached: by inspecting its type signature and reasoning compositionally 
+about arguments, i.e. splices.
+Finally, livelits integrate cleanly with live evaluation of programs with holes \cite{HazelnutLive},
+whereas other projectional editors either do not support evaluation of holes or do 
+not use that information to provide feedback to the user.
+
+Related to projectional editors are ``syntax-recognizing editors'' \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84} and 
+a great variety of systems that generate visualizations from code \cite{DBLP:conf/chi/Lerner20,koschke2003software,urquiza2004survey}. Livelits differ 
+from these systems in directionality: visualization systems are given code and generate
+a visualization, whereas livelits are asked to create code where there would otherwise
+be a hole based on interactions that the programmer has with a visual interface.
+
+Conal Elliott's work on tangible functional programming \cite{DBLP:conf/icfp/Elliott07} similarly explored a
+system that provides visual feedback about the result of a computation, 
+but the computation itself could be constructed using only a fixed structural syntax
+manifest as a series of connected windows.
+
+The Vital programming environment for Haskell \cite{hanna2002interactive} supports type-specific stylesheets
+that can be used to create custom user interfaces for both editing and displaying
+values. The editors support splices, called \emph{cells}, that can contain 
+Haskell code. Results are computed in a live manner, though there is no support
+for evaluating incomplete programs. The visualizations themselves also cannot provide
+live feedback. Moreover, the system does not enforce any hygiene or typing principles:
+the user is entirely responsible for syntactic and semantic correctness, and the 
+underlying semantics have not been formally defined.
+
+% Hazelnut
+% Hazelnut Live
+% Its Alive! paper (GUI widgets)
+% Tangible FUnctional Programming 
+% Vital https://dl.acm.org/doi/10.1145/581478.581493 and https://link.springer.com/chapter/10.1007/11964681_12
+% Projection Boxes http://cseweb.ucsd.edu/~lerner/proj-boxes.html
+% Tiny Structure Editors for Low Low Prices
+% Typed Holes Work (synthesis stuff from Haskell, mention Haskell)
+% Pharo / Smalltalk 
+% GEC Toolkit: https://link.springer.com/chapter/10.1007/11546382_5
+% Previous use of live literals https://homepages.cwi.nl/~storm/livelit/livelit.html
+% Visual Haskell and related visual programming environments: https://pdfs.semanticscholar.org/6e43/5494ef1aaaa6ec0fcf47bb5b03d7c06ef937.pdf
+% widgets in dataflow languages like puredata and labview

### 2020-09-06 discussion.tex
final draft of popl

--- 2020-07-10 discussion.tex
@@
   %
-  The arithmetical symbols are written diagrams and the geometrical figures are graphic formulas.''
+  The arithmetical symbols are written diagrams and the geometrical figures are graphic formulas.
   %
@@
   the user to edit a number in the result of a computation and see the necessary
-  change to a slider in the program.
+  change to a slider in the program \cite{sns-pldi,sns-uist}.
 

### 2020-09-06 implementation.tex
final draft of popl

@@
 Hazelnut Live is the formal foundation for the Hazel programming environment, so 
-it was natural to choose the the Hazel programming environment for our primary implementation\todo{cite?}{}. 
+it was natural to choose the the Hazel programming environment for our primary implementation. 
  Hazel is a browser-based live functional programming environment 
@@
 
-Hazel is implemented in OCaml and compiled to JavaScript using the \li{js_of_ocaml} compiler\todo{cite}.
-Hazel is written in a pure functional style, so our implementation is able to closely 
+Hazel is implemented in OCaml and compiled to JavaScript using the \li{js_of_ocaml} compiler \cite{DBLP:conf/aplas/RadanneVB16}.
+The core of Hazel is written in a pure functional style, so our implementation is able to closely 
 follow the formalism. The Hazel language has been evolving in parallel with our work, 
 and has only recently become expressive enough for it to be feasible to implement livelits within Hazel itself.
-The examples in this paper are instead implemented in OCaml as modules satisfying the signature 
+The examples in this paper were instead implemented in OCaml as modules satisfying the signature 
 implied by Fig.~\ref{fig:color-impl}. A functor then adapts these OCaml-based livelit implementations 
@@
 \subsection{Text Editor Integration}
-Before delving into these issues, let us note that livelits do not require the use of a structure editor. 
+Before delving into layout issues, let us note that livelits do not require the use of a structure editor. 
 We have also developed a 
@@
 This proof-of-concept is not at feature parity with 
-our main implementation, but it demonstrates that a text editor with support for 
-parsing out livelit applications and replacing them with interactive graphical inclusions is sufficient.
+our main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84} 
+is sufficient to support livelits, albeit with some gaps in livelit UI availability when 
+there are critical errors.
 
@@
 Whether implemented in a structure editor or a text editor, livelits present 
-interesting layout challenges: it can be difficult to sensibly 
-interleave symbolic code with livelit views.
+interesting layout challenges--it can be difficult to sensibly 
+interleave symbolic code with livelit views in a clean, readable manner.
 
 Hazel uses an optimizing pretty printer to determine layout, an approach 
-based on the work of Bernardy et al \todo{cite bernardy}. This system relies 
+based on the work of \citet{DBLP:journals/pacmpl/Bernardy17}. This system relies 
 fundamentally on character counts. Consequently, our implementation asks each 
@@
 or as \emph{multi-line livelits}, which occupy up to the full width of the editor 
-and a specified number of code lines (not shown in Fig.~\ref{fig:color-impl}). 
+and a specified number of code lines (we omitted these specifications in Fig.~\ref{fig:color-impl}). 
 The pretty printer lays out multi-line livelits
-on their own line, as seen in the figures in this paper.
+on their own line, as seen in Fig.~\ref{fig:grading}.
 
@@
 We leave further optimizations along these lines to future work.
-Prior work on incrementally updating UIs as code changes could be relevant to this effort \cite{burckhardt2013s}.
+Prior work on incrementally updating UIs after code changes could be relevant to this effort \cite{burckhardt2013s}.

### 2020-11-18 popl21/popl21-rebuttal.txt
moved popl21 submission into a directory for archival

Thank you for your reviews!

## Review A

### Corrections

The quoted sentence was part of our discussion of _future_ work. It was not written as motivation for this paper.

Livelits can generate expressions, not just values. This includes expressions that have non-trivial computational behavior. 
For example, Fig 3 demonstrates image transformations.

## Review B

### Comments

Although the five examples in the paper are non-trivial, we agree that additional screenshots might help readers build 
better intuitions for the expressive power of livelits. Consequently, we would be happy to include 4-6 additional 
screenshots of novel livelits in the form of an "example gallery" in our final revision. None of these introduce
new technical content into the paper.

Our prior work on Graphite included several empirical studies which are just as relevant to livelits.
The present paper can be read as a technical complement to that qualitative/empirical paper [Omar et al, 2012].

The Hazel livelits implementation as described in the paper was functional at the time of submission
(in the livelits branch on Github). It will be sent for artifact evaluation and released publicly with the 
next major Hazel release before POPL. The public will be able to play with livelits in their browser. We are 
working on a number of ``quality of life'' improvements on an ongoing basis, but these do not change the 
content or contributions of the paper.

## Review C

### Corrections

MPS and Barista (via Citrus) were discussed in detail in the paper. These systems, as well as the others that 
the review mentions, fail to achieve multiple of the design goals from Sec. 1 (discussed on line 921). We
will include missing citations in the final revision, thank you.

Livelits could be defined using other livelits when using the definition mechanism defined in Sec. 3. It is not the case that ``the definitions for livelits are textual'' in this setting. Indeed, related work on literal macros supports using literal macros in literal macro definitions [Omar et al, 2018] and the same mechanisms could be deployed here (once Hazel has a proper module / package management system).

We believe pure functional programming in languages like Haskell and Elm is practical. 
That said, it is not true that livelits are limited to the pure functional setting. 
The issue is simply that running a program with side effects
during editing might not be _desirable_ in some settings, but that choice is up to the programmer.
For example, live programming in Jupyter lab notebooks often involves side effects 
occurring in arbitrary order depending on user interactions and programmers by and large accept this. 
Livelits could be added to Jupyter-like systems such that they always use this ambient state.
The issue could also be avoided by using algebraic effects, so that destructive effects could be ``shimmed'' 
during development. These considerations are not livelit specific -- they are inherited from the work 
in [Omar et al., POPL 2019] and indeed live programming itself is fundamentally in tension with 
side effects. We will include a more substantial discussion of this point in the final revision.

We do discuss livelit persistence in Sec. 3.1. Only the model is serialized. In our case, we use a generic S-expression
textual syntax for Hazel expressions, but this is an incidental implementation detail (Hazel programs are not
intended to be edited using text editors). We do discuss a proof-of-concept text editor with support for 
livelits in Sec. 5.2.

### Comments

Pattern livelits would be a straightforward extension of our system. The necessary theory for pattern macros
was developed in [Omar et al, 2018].

Declaration livelits can be subsumed by module livelits, which can in turn be subsumed by expression livelits
in a language that has first class modules. This was also briefly discussed in [Omar et al, 2018].

While our formalism of expansion does build on our prior macro system formalism, we see this as a strength
of our paper. Our treatment of splices is distinct, as is our treatment of hygiene. Closure collection is entirely
new. The purpose of the formalism was to make the mechanism precise and reduce it to its essential form.

Hazel functions that use GUI libraries must ultimately execute by extraction to OCaml, so the fact that 
we implemented our GUI code directly in OCaml was merely a matter of expediency. The Hazel language is still
evolving (in large part due to contributors who are not involved with this paper) and there is not yet a 
full-scale GUI widgets library. We plan to "self host" in the coming months before POPL.

The interface for selecting between closures is the same as the interface demonstrated in [Omar et al, POPL 2019]. We will include a screenshot in the final revision.

The capture list in Fig. 4 is empty because generated expansions do not use any libraries or definition site values
(they are simple tuples of numbers in this case). See [Omar et al, ICFP 2018] for a discussion on explicit capture.

## Review D

### Corrections

The ipywidgets library only allows simple non-compositional widgets (no splices inside) and these widgets do not have access to the surrounding 
closures or state so there is no support for liveness. 
None of the central examples in our paper could be implemented as ipywidgets. We will, however, make sure to cite this work in the final revisions.

### Comments

See our response to Review B regarding user studies.

Typed holes are central in that they enable closure collection, which is what enables liveness.

The core calculus includes sums, products, and recursive types because these are minimal type theoretic constructs sufficient to achieve
self-hosting of syntax. Alternative encoding schemes could work just as well (e.g. Church encodings in System F) but are more unwieldy.
Self-hosting is necessary because the expand function returns an encoding of syntax, which needs to be decoded (Sec. 4.2.1).

Splices are simply expressions that appear inside (i.e. are notionally spliced into) livelit GUIs.

We will cite Lerner 2020, which has related motivations, thank you.

We will use a different connective for expanding and elaborating in the final version, thank you.

### 2020-11-18 popl21/case-studies.tex
moved popl21 submission into a directory for archival

--- 2020-07-10 case-studies.tex
@@
 body of \li{classic}. Rather than disabling live evaluation in situations like 
-this, Hazel instead allows the programmer to select between the closures 
-as shown (inset). The live feedback is derived based on the selected closure.
+this, Hazel instead allows the programmer to select between the closures when 
+the cursor is on the livelit expression (via a simple sidebar toggle, not shown, which is shared 
+with the general closure selection UI \cite{HazelnutLive}). 
+The live feedback is based on the selected closure.
 This makes it easy to see how the image filter being designed here will affect a

### 2020-11-18 popl21/livelit-calculus.tex
moved popl21 submission into a directory for archival

--- 2020-07-10 livelit-calculus.tex
@@
 \section{A Simply Typed Livelit Calculus}\label{sec:livelit-calculus}
+
+\begin{figure}
+    \[
+    \arraycolsep=4pt\begin{array}{llcl}
+        \mathsf{Typ} & \tau & ::= &
+                                    \tnum ~\vert~
+                                    \tarr{\tau_1}{\tau_2} ~\vert~
+                                    \tprod{\tau_1}{\tau_2} ~\vert~
+                                    \tunit ~\vert~
+                                    \tsum{\tau_1}{\tau_2} ~\vert~
+                                    t ~\vert~
+                                    \trec{t}{\tau}\\
+        \mathsf{UExp} & \hat{e} & ::= & \hnum{n} ~\vert~
+                                 x ~\vert~
+                                 \hlam{x}{\hat e} ~\vert~
+                                 \hap{\hat e_1}{\hat e_2} ~\vert~
+                                 \cdots ~\vert~
+                                %  \hpair{\hat e_1}{\hat e_2} ~\vert~
+                                %  \hprl{\hat e} ~\vert~
+                                %  \hprr{\hat e} ~\vert~
+                                %  \htriv ~\vert~
+                                %  \hinL{\hat e} ~\vert~
+                                %  \hinR{\hat e} ~\vert~
+                                %  \hroll{\hat e} ~\vert~
+                                %  \hunroll{\hat e} \\
+                                 \hehole{u} ~\vert~
+                                 \haplivelit{u}{a}{\dtxt{model}}{\splices}\\
+        \mathsf{EExp} & e & ::= & \hnum{n} ~\vert~ x ~\vert~ \hlam{x}{e} ~\vert~ \hap{e_1}{e_2} ~\vert~ \cdots ~\vert~ \hehole{u}\\
+        \mathsf{IExp} & d & ::= & \hnum{n} ~\vert~ x ~\vert~ \hlam{x}{d} ~\vert~ \hap{d_1}{d_2} ~\vert~ \cdots ~\vert~ \dehole{u}{\sigma}\\
+        \mathsf{Splice} & \psi & ::= & \hsplice{\hat e}{\tau}
+    \end{array}
+    \]
+    \caption{Syntax of types, $\tau$, unexpanded expressions, $\hat{e}$, expanded expressions, $e$, and internal expressions, $d$.
+    Metavariable $x$ ranges over variables, $u$ over hole names, and $\livelitname{a}$ over livelit names.
+    We write $\splat{\psi_i}$ for a finite sequence of $n \geq 0$ splices,
+    and $\sigma$ for finite substitutions of $n \geq 0$ internal expressions for variables, $[d_1/x_1, \cdots, d_n/x_n]$.
+    We elide standard expression forms
+    related to product, sum, and recursive types.
+    }
+    \label{fig:syntax}
+    \end{figure}
+
 In order to precisely capture the semantics of livelits
@@
 livelits, so they are the main focus of the remaining sections,
-which cover livelit expansion (Sec.~\ref{sec:calculus-expansion}),
-livelit update actions (Sec.~\ref{sec:calculus-updates}),
-liveness via closure collection (Sec.~\ref{sec:calculus-closure-collection}),
-and parameterization (Sec.~\ref{sec:calculus-parameterization}).
+which cover livelit expansion (Sec.~\ref{sec:calculus-expansion}) and 
+liveness via closure collection (Sec.~\ref{sec:calculus-closure-collection}). 
+% and parameterization (Sec.~\ref{sec:calculus-parameterization}).
 
@@
 notated externally $\hehole{u}$ where $u$ is a \emph{hole name} \cite{HazelnutLive}.
-We omit non-empty holes (which allow reasoning about type inconsistencies \cite{Hazelnut}) and type holes
+We omit non-empty holes (which internalize type inconsistencies \cite{Hazelnut}) and type holes
 (which operate like the unknown type from gradual type theory \cite{Siek06a,Hazelnut})
@@
 
-\subsection{Expansion}
+\subsection{Expansion}\label{sec:calculus-expansion}
 
-\begin{figure}
-    \[
-    \arraycolsep=4pt\begin{array}{llcl}
-        \mathsf{Typ} & \tau & ::= &
-                                    \tnum ~\vert~
-                                    \tarr{\tau_1}{\tau_2} ~\vert~
-                                    \tprod{\tau_1}{\tau_2} ~\vert~
-                                    \tunit ~\vert~
-                                    \tsum{\tau_1}{\tau_2} ~\vert~
-                                    t ~\vert~
-                                    \trec{t}{\tau}\\
-        \mathsf{UExp} & \hat{e} & ::= & \hnum{n} ~\vert~
-                                 x ~\vert~
-                                 \hlam{x}{\hat e} ~\vert~
-                                 \hap{\hat e_1}{\hat e_2} ~\vert~
-                                 \cdots ~\vert~
-                                %  \hpair{\hat e_1}{\hat e_2} ~\vert~
-                                %  \hprl{\hat e} ~\vert~
-                                %  \hprr{\hat e} ~\vert~
-                                %  \htriv ~\vert~
-                                %  \hinL{\hat e} ~\vert~
-                                %  \hinR{\hat e} ~\vert~
-                                %  \hroll{\hat e} ~\vert~
-                                %  \hunroll{\hat e} \\
-                                 \hehole{u} ~\vert~
-                                 \haplivelit{u}{a}{\dtxt{model}}{\splices}\\
-        \mathsf{EExp} & e & ::= & \hnum{n} ~\vert~ x ~\vert~ \hlam{x}{e} ~\vert~ \hap{e_1}{e_2} ~\vert~ \cdots ~\vert~ \hehole{u}\\
-        \mathsf{IExp} & d & ::= & \hnum{n} ~\vert~ x ~\vert~ \hlam{x}{d} ~\vert~ \hap{d_1}{d_2} ~\vert~ \cdots ~\vert~ \dehole{u}{\sigma}\\
-        \mathsf{Splice} & \psi & ::= & \hsplice{\hat e}{\tau}
-    \end{array}
-    \]
-    \caption{Syntax of types, $\tau$, unexpanded expressions, $\hat{e}$, expanded expressions, $e$, and internal expressions, $d$.
-    Metavariable $x$ ranges over variables, $u$ over hole names, and $\livelitname{a}$ over livelit names.
-    We write $\splat{\psi_i}$ for a finite sequence of $n \geq 0$ splices,
-    and $\sigma$ for finite substitutions of $n \geq 0$ internal expressions for variables, $[d_1/x_1, \cdots, d_n/x_n]$.
-    We elide standard expression forms
-    related to product, sum, and recursive types.
-    }
-    \label{fig:syntax}
-    \end{figure}
+The novelty of the livelit calculus lies entirely in its handling of unexpanded expressions,
+$\hat e$, which are given meaning by typed expansion to external expressions,
+$e$, according to the judgement $\expands{\Phi}{\Gamma}{\hat e}{e}{\tau}$
+defined in Fig.~\ref{fig:expansion}.
+Unexpanded expressions mirror external expressions but for the presence of livelits, discussed below.
+The rules for standard
+forms like variables, functions and function application, shown in Fig.~\ref{fig:expansion},
+are straightforward.
 
+\subsubsection{Livelit Contexts}
+
+Livelit definitions are collected in the livelit context, $\Phi$, which
+maps livelit names to livelit definitions of the following form:
+\[ \livelitCtxEntry{a}{\taut{expand}}{\taut{model}}{\dtxt{expand}} \]
+Here, $\taut{expand}$ is the expansion type, $\taut{model}$ is the model type,
+and $\dtxt{expand}$ is the expansion function, which generates an expansion given a model.
+We omit logic related to view computations and actions, which are tied to a particular 
+UI framework.
+
+For a livelit context to be well-formed, we require that each livelit definition be well-formed.
+This in turn requires that the expansion function be hole-free and have the correct type.
+\begin{definition}[Livelit Context Well-Formedness]
+    A livelit context $\Phi$ is well-formed if and only if for each livelit definition,
+    $\livelitCtxEntry{a}{\taut{expand}}{\taut{model}}{\dtxt{expand}} \in \Phi$,  we have
+    $\hasType{}{\dtxt{expand}}{\tarr{\taut{model}}{\expType}}$.
+\end{definition}
+Here, $\expType$ stands for a type whose values isomorphically encode
+external expressions. The isomorphism is mediated in one direction by
+the encoding judgement $\encodeExp{e}{d}$ and the other by the decoding judgement $\decodeExp{d}{e}$.
+Any isomorphism is sufficient, so we leave the details as a matter of implementation.
+The simplest approach would be to define $\expType$ as a recursive sum type,
+with one arm for each form of external expression (cf. \cite{TSLs} for an example).
+
+For simplicity, we assume that the livelit context is provided \emph{a priori} and therefore
+that functions like $\dtxt{expand}$ are already closed and fully elaborated.
+In practice, the livelit context would be controlled by a definition form in the language
+that allows the expansion function to itself use livelits.
+This would require a staging mechanism,
+because we need to execute expansion functions in prior definitions to be able to
+expand subsequent definitions.
+There are a number of ways to support the necessary staging in practice, e.g.
+via explicit staging primitives \cite{DBLP:conf/icfp/Flatt02},
+by requiring that these definitions appear in separately compiled packages \cite{TLMs},
+or by using live
+programming mechanisms such as those available in Hazel
+to evaluate ``up to'' each definition before proceeding \cite{HazelnutLive}.
+This choice is orthogonal to the mechanisms of interest in this section.
+
+\subsubsection{Hygienic Livelit Expansion}
 \begin{figure}
@@
     \end{figure}
-The novelty of the livelit calculus lies entirely in its handling of unexpanded expressions,
-$\hat e$, which are given meaning by typed expansion to external expressions,
-$e$, according to the judgement $\expands{\Phi}{\Gamma}{\hat e}{e}{\tau}$
-defined in Fig.~\ref{fig:expansion}.
-Unexpanded expressions mirror external expressions but for the presence of livelits, discussed below.
-The rules for standard
-forms like variables, functions and function application, shown in Fig.~\ref{fig:expansion},
-are straightforward.
-
-\subsubsection{Livelit Contexts}
-Livelit definitions are collected in the livelit context, $\Phi$, which
-maps livelit names to livelit definitions of the following form:
-\[ \livelitCtxEntry{a}{\taut{expand}}{\taut{model}}{\dtxt{expand}} \]
-Here, $\taut{expand}$ is the expansion type, $\taut{model}$ is the model type,
-and $\dtxt{expand}$ is the expansion function, which generates an expansion given a model.\todo{add view and update}
-
-For a livelit context to be well-formed, we require that each livelit definition be well-formed.
-This in turn requires that the expansion function be hole-free and have the correct type.
-\begin{definition}[Livelit Context Well-Formedness]
-    A livelit context $\Phi$ is well-formed if and only if for each livelit definition,
-    $\livelitCtxEntry{a}{\taut{expand}}{\taut{model}}{\dtxt{expand}} \in \Phi$,  we have
-    $\hasType{}{\dtxt{expand}}{\tarr{\taut{model}}{\expType}}$.
-\end{definition}
-Here, $\expType$ stands for a type whose values isomorphically encode
-external expressions. The isomorphism is mediated in one direction by
-the encoding judgement $\encodeExp{e}{d}$ and the other by the decoding judgement $\decodeExp{d}{e}$.
-Any isomorphism is sufficient, so we leave the details as a matter of implementation.
-The simplest approach would be to define $\expType$ as a recursive sum type,
-with one arm for each form of external expression (cf. \cite{TSLs} for an example).
-
-For simplicity, we assume that the livelit context is provided \emph{a priori} and therefore
-that functions like $\dtxt{expand}$ are already closed and fully elaborated.
-In practice, the livelit context would be controlled by a definition form in the language
-that allows the expansion function to itself use livelits.
-This would require a staging mechanism,
-because we need to execute expansion functions in prior definitions to be able to
-expand subsequent definitions.
-There are a number of ways to support the necessary staging in practice, e.g.
-via explicit staging primitives \todo{cite ocaml macros paper, Racket stuff}{},
-by requiring that these definitions appear in separately compiled packages \cite{TLMs},
-or by using live
-programming mechanisms such as those available in Hazel
-to evaluate ``up to'' each definition before proceeding.
-This choice is orthogonal to the mechanisms of interest in this section.
-
-\subsubsection{Hygienic Livelit Expansion}
 Unexpanded expressions are unique in that they include livelits:
@@
 itself (unexpanded, so it may contain other livelits) and $\tau_i$ is the type of that splice,
-as determined when the livelit definition first requested the splice (discussed in Sec.~\ref{sec:calculus-updates}).
+as determined when the livelit definition first requested the splice (discussed in Sec.~\ref{sec:livelit-definitions}).
 
@@
     closed. Consequently, any necessary helper functions used in the expansion must be provided explicitly by the client
-    via a splice. In Sec.~\ref{sec:calculus-parameterization}, we will see how this context independence
-    restriction can be made
-    less onerous by passing definition-site bindings along as partially applied parameters.\todo{remove?}{}
+    via a splice. In Sec.~\ref{sec:livelit-definitions}, we discussed how the use of explicit capture sets, which can be understood as 
+    immediate partial application, can lower the cognitive burden of this strict context independence discipline.
+    Parameters formally operate in the same way as splices, and partial application is a well-understood technique, so we omit these details from the calculus.
     \item \textbf{Splice Expansion.} Finally, the sixth premise inductively expands each of the spliced expressions in the same context as the livelit
@@
 The typed expansion process is governed by the following metatheorem, which establishes that the expansion
-is indeed an external expression of the indicated type.\todo{change to proof sketch}{}
+is indeed an external expression of the indicated type.
 
@@
 expression expands to a well-typed external expression, which in turn expands to a well-typed internal
-expression, which evaluates in a type safe manner.)
+expression, which in turn evaluates in a type safe manner.)
 
@@
 We have mechanized the typed expansion mechanism and proven the Typed Expansion theorem
-using the Agda proof assistant.\todo{say more}{}
+using the Agda proof assistant, using the same techniques as were used in the Agda mechanization 
+of Hazelnut Live \cite{HazelnutLive}. This mechanization is available in the anonymous 
+supplemental material.
 
-\subsection{Live Feedback via Closure Collection}
+\subsection{Live Feedback via Closure Collection}\label{sec:calculus-closure-collection}
 In order to support live feedback, a livelit needs to be able to ask the system
@@
 Unlike substitution, hole filling is
-not capture-avoiding. Instead, the environment on each of these closures is applied to $\dtxt{pexpansion}$
+not capture-avoiding. Instead, the environment on each of these closures is applied to $d_1$
 as a substitution, i.e. the delayed substitutions captured in the environment are realized.
@@
 
-\subsection{Parameterization}
+% \subsection{Parameterization}
 
-Add livelit expressions, livelit abbreviations,
-partial application support. Talk about how to work around the strict context
-independence discipline we have imposed.
+% Add livelit expressions, livelit abbreviations,
+% partial application support. Talk about how to work around the strict context
+% independence discipline we have imposed.
 
-\subsection{Edit Action Semantics}
-So far, we have considered only the semantics of an individual editor state.
-However, livelits are useful because they evolve as the user interacts
-with them. In particular, the livelit's model, $\dtxt{model}$ can evolve
-due to livelit-specific user interactions. Furthermore, a livelit can also
-create and delete splices when a user interaction demands it (unlike functions, which have a fixed
-number of arguments).
+% \subsection{Edit Action Semantics}
+% So far, we have considered only the semantics of an individual editor state.
+% However, livelits are useful because they evolve as the user interacts
+% with them. In particular, the livelit's model, $\dtxt{model}$ can evolve
+% due to livelit-specific user interactions. Furthermore, a livelit can also
+% create and delete splices when a user interaction demands it (unlike functions, which have a fixed
+% number of arguments).

### 2020-11-18 pldi21/abstract.tex
ported to PLDI format and tightened up abstract

% !TEX root = livelits-paper.tex

%% no citations in Abstract
%%
%% \cite{popl-paper}

\begin{abstract}
    Although textual program manipulation is powerful, 
    some types of expressions are better represented and manipulated
    graphically.
    % Examples include expressions that compute 
    % colors,
    % animation parameters,
    % musical sequences,
    % tabular data, 
    % plot parameters,
    % GUI widgets, 
    % and various other domain-specific data structures.
    This paper introduces \emph{live literals}, or \emph{livelits}, 
    a mechanism that allows
    client programmers to fill holes of types like these 
    by directly manipulating a     
    provider-specified GUI embedded persistently into otherwise 
    symbolic code. 
    Uniquely, livelits are compositional: the GUI can itself contain
    spliced expressions, which can in turn contain other livelits. 
    Splices are typed and the system 
    ensures that the livelit treats each splice hygienically. 
    Livelits are also uniquely live: 
    they can offer {immediate feedback} about the run-time implications 
    of the programmer's choices 
    even when the spliced expressions mention bound variables, 
    because the system gathers closures associated with the 
    hole that the livelit is tasked with filling.
    This paper introduces livelits with case studies that exercise these novel capabilities. 
    We implement livelits in Hazel, 
    a live programming environment able to 
    typecheck and run programs with holes. 
    We then define a typed lambda calculus that captures the essence of livelits, which
    operate as live, graphical macros.     
    The macro expansion process has been mechanized in Agda.
\end{abstract}

### 2020-11-18 pldi21/related-work.tex
ported to PLDI format and tightened up abstract

--- 2020-09-06 related-work.tex
@@
 As detailed in Sec.~\ref{sec:intro},
-the \emph{Graphite} system introduced the idea of filling 
-typed holes using a type-specific user interface \cite{Graphite}.
+the \emph{Graphite} system substantially developed the idea of filling 
+typed holes using a type-specific user interface, and it was the starting 
+point for our work \cite{Graphite}.
 This prior work engaged in substantial qualitative evaluation, 

### 2020-11-18 pldi21/intro.tex
ported to PLDI format and tightened up abstract
tighten up 1.0

--- 2020-07-10 intro.tex
@@
 \section{Introduction}\label{sec:intro}
-Textual program editors are powerful and expressive user interfaces,
-so it is perhaps little wonder that they remain dominant decades after the teletype.
-However, textual user interfaces are not the best tool for every job.
-In particular, there are many
-data types for which a non-textual
-user interface for expression construction may situationally be more appropriate.
+Text-based program editors are flexible and expressive user interfaces,
+so it is little wonder that they remain dominant decades after the teletype.
+However, textual user interfaces are not the best tool for every programming job.
+% In particular, there are countless 
+% data types for which a non-textual
+% user interface may situationally be more appropriate.
 
 As a simple example, consider a record type
-classifying colors using an RGBA encoding:
-\begin{lstlisting}[numbers=none]
-type Color = { r: Int, g: Int, b: Int, a: Int }
-\end{lstlisting}
+classifying colors using the RGBA encoding. 
 It is possible to select a particular color by entering
-an expression of this type using a text editor:
-\begin{lstlisting}[numbers=none]
-let bgcolor : Color = { r: 255, g: 178, b: 45, a: 100 }
-\end{lstlisting}
-The problem is that this user interface for color selection
-offers the user no immediate feedback
-and limited editing affordances.
-It is difficult for the programmer, or anyone subsequently reading or modifying the code, to know which color is represented
-and to interactively tweak that color.
+an expression of this type using a text editor, e.g. \li{\{ r: 255, g: 178, b: 45, a: 100 \}}. 
+The problem with this textual user interface for color selection is that 
+it offers the programmer no live feedback about which color has been selected 
+and limited editing affordances for tweaking the chosen color.
+Analagous critiques apply to strictly textual user interfaces for 
+countless other data structures,
+such as animation parameters,
+vector graphics,
+musical sequences,
+audio filters,
+game states, 
+GUI primitives,
+tabular data, 
+plots,
+geospatial data, 
+neural network diagrams, 
+biological neuron models, 
+and the list goes on.
 
-Better support for manipulating data of types like these would be particularly helpful for users engaging in
-live and exploratory programming in domains like web design, media production,
-and data analysis. Indeed, practitioners in domains like these commonly eschew general-purpose programming
-in favor of graphical end-user applications, like %
-image and video editors and music composition software,
-in part because these applications provide domain-specific forms of live feedback and
-direct manipulation affordances, e.g. color palettes, visual timelines, and interactive plots.
-The trade-off is that these applications have limited support for abstraction and composition.
+% It is difficult for the programmer, or anyone subsequently reading or modifying the code, to know which color is represented
+% and to interactively tweak that color.
+
+Practitioners in domains where manipulating data of types like these is central 
+have largely eschewed general-purpose programming languages 
+in favor of more specialized graphical end-user applications, like %
+image and video editors, music composition software, level design tools, 
+and bespoke GUIs written by technicians, 
+in large part because these applications 
+take seriously the need for domain-specific forms of live feedback, graphical representations, and
+direct manipulation affordances, 
+e.g. color palettes, visual timelines, interactive plots, and maps.
+
+The tragic trade-off is that these applications have limited support for abstraction and composition.
 It is difficult, for example, to bind a
 color to a variable for use in multiple locations,
-or to transform a sound by passing it through a series of functions,
 or to define a function that computes a vector graphic given a set
 of parameters (some of them perhaps themselves computed),
+or to transform a sound by passing it through a series of functions,
 or to organize these constructions into modules.
 Moreover, it is difficult to add new affordances or to compose
-affordances in ways that the application designer did not anticipate,
+affordances in ways that the application developer did not anticipate,
 so users cannot easily make even simple changes like replacing a numeric text box in a color chooser with a slider,
 much less more ambitious changes like a new visual interface for expressing geospatial data queries.
+
+% Better support for manipulating data of types like these would be particularly helpful for users engaging in
+% live and exploratory programming in domains like web design, media production,
+% and data analysis. Indeed,
 
@@
 and the abstraction and composition mechanisms
-available in modern general-purpose languages.
+available in modern general-purpose languages as the ``glue'' between these 
+GUIs.
 
@@
        \includegraphics[width=15pc]{graphite-color-palette.png}
+      \vspace{0.1cm}
+      \caption{}
     \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
@@
       \includegraphics[width=15pc]{slider-color-livelits.png}
+      \vspace{-11px}
+      \caption{}
     \end{subfigure}
@@
 Java expression to fill the hole.
-Figure~\ref{fig:color}(a)\todo{use subfigure?}{}, adapted from this prior work, demonstrates a color palette invoked using Graphite.
+Figure~\ref{fig:color}(a), adapted from this prior work, demonstrates a color palette invoked using Graphite.
 When the user presses the \li{Enter} key, the Java expression \li{new Color(173, 173, 173, 85)} is inserted at the cursor (not shown) and the palette disappears.

### 2020-11-19 pldi21/abstract.tex
further tightening of abstract  typo fix

@@
 \begin{abstract}
-    Although textual program manipulation is powerful, 
-    some types of expressions are better represented and manipulated
+    Although text editing is powerful, 
+    some types of expressions are more naturally represented and manipulated
     graphically.
@@
     This paper introduces \emph{live literals}, or \emph{livelits}, 
-    a mechanism that allows
-    client programmers to fill holes of types like these 
+    which allow
+    clients to fill holes of these types  
     by directly manipulating a     
-    provider-specified GUI embedded persistently into otherwise 
+    provider-defined GUI embedded persistently into otherwise 
     symbolic code. 
     Uniquely, livelits are compositional: the GUI can itself contain
-    spliced expressions, which can in turn contain other livelits. 
+    spliced expressions, which have full editor support and can in turn contain other livelits. 
     Splices are typed and the system 
-    ensures that the livelit treats each splice hygienically. 
+    ensures that livelits treat splices hygienically. 
     Livelits are also uniquely live: 
     they can offer {immediate feedback} about the run-time implications 
-    of the programmer's choices 
-    even when the spliced expressions mention bound variables, 
-    because the system gathers closures associated with the 
+    of the client's choices 
+    even when splices mention bound variables, 
+    because the system continuously gathers closures associated with the 
     hole that the livelit is tasked with filling.
@@
     typecheck and run programs with holes. 
-    We then define a typed lambda calculus that captures the essence of livelits, which
-    operate as live, graphical macros.     
+    We then define a typed lambda calculus that captures the essence of livelits as 
+    live graphical macros.     
     The macro expansion process has been mechanized in Agda.

### 2020-11-19 pldi21/intro.tex
reworking intro

@@
 \section{Introduction}\label{sec:intro}
-Text-based program editors are flexible and expressive user interfaces,
+Text-based program editors are flexible and expressive user interfaces
 so it is little wonder that they remain dominant decades after the teletype.
-However, textual user interfaces are not the best tool for every programming job.
+However, textual user interfaces are not the best tool for every computational job.
 % In particular, there are countless 
@@
 As a simple example, consider a record type
-classifying colors using the RGBA encoding. 
+classifying RGBA-encoded colors. 
 It is possible to select a particular color by entering
-an expression of this type using a text editor, e.g. \li{\{ r: 255, g: 178, b: 45, a: 100 \}}. 
+an expression of this type in a text editor, e.g. \li{\{ r: 255, g: 178, b: 45, a: 100 \}}. 
 The problem with this textual user interface for color selection is that 
-it offers the programmer no live feedback about which color has been selected 
-and limited editing affordances for tweaking the chosen color.
+it offers no live feedback about which color has been selected 
+and limited editing affordances for tweaking the selected color.
 Analagous critiques apply to strictly textual user interfaces for 
 countless other data structures,
-such as animation parameters,
-vector graphics,
+such as vector graphics,
+animation parameters,
 musical sequences,
 audio filters,
-game states, 
-GUI primitives,
+board game states, 
+GUI widgets and layouts,
 tabular data, 
@@
 biological neuron models, 
-and the list goes on.
+mathematical diagrams, 
+and so on.
 
@@
 
-Practitioners in domains where manipulating data of types like these is central 
-have largely eschewed general-purpose programming languages 
+Practitioners in domains where manipulating data of types like these is 
+a central activity 
+have largely eschewed general-purpose programming environments 
 in favor of more specialized graphical end-user applications, like %
 image and video editors, music composition software, level design tools, 
-and bespoke GUIs written by technicians, 
+and bespoke GUIs written by students or lab technicians, 
 in large part because these applications 
-take seriously the need for domain-specific forms of live feedback, graphical representations, and
+take seriously the need for domain-specific forms of live feedback, 
+graphical data representations, and 
 direct manipulation affordances, 
@@
 
-The tragic trade-off is that these applications have limited support for abstraction and composition.
+The tragedy is that these applications have 
+limited support for abstraction and composition.
 It is difficult, for example, to bind a
-color to a variable for use in multiple locations,
-or to define a function that computes a vector graphic given a set
-of parameters (some of them perhaps themselves computed),
-or to transform a sound by passing it through a series of functions,
-or to organize these constructions into modules.
+color to a variable for use in multiple locations in an
+otherwise directly constructed game map,
+or to define a function that computes portions of an 
+otherwise directly constructed vector graphic,
+or to transform a directly constructed musical sequence 
+by passing it through a series of symbolically specified functions.
 Moreover, it is difficult to add new affordances or to compose
-affordances in ways that the application developer did not anticipate,
-so users cannot easily make even simple changes like replacing a numeric text box in a color chooser with a slider,
-much less more ambitious changes like a new visual interface for expressing geospatial data queries.
+affordances in ways that the application developer did not anticipate.
+Users cannot easily make even simple changes like replacing a numeric text box in a color palette with a slider,
+much less more ambitious changes like installing an alternative visual interface for expressing geospatial data queries in a workflow.
 
@@
 
-This paper aims to resolve the apparent tension between
-programmatic and direct manipulation user interfaces by designing a programming system that
-is able to surface GUIs for the data types for which
+This paper aims to resolve this tension between
+programmatic and direct manipulation user interfaces by designing 
+a programming environment that
+is able to surface GUIs when working with types for which
 they are useful, while retaining full support for symbolic program manipulation
 and the abstraction and composition mechanisms
-available in modern general-purpose languages as the ``glue'' between these 
-GUIs.
+available in modern general-purpose programming languages.
 
@@
 \begin{figure*}
-  \begin{center}
-    \begin{subfigure}[t]{0.5\textwidth}
-      \centering
-       \vspace{-3.4cm}
-       \includegraphics[width=15pc]{graphite-color-palette.png}
-      \vspace{0.1cm}
-      \caption{}
-    \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
-      \centering
-      \includegraphics[width=15pc]{slider-color-livelits.png}
-      \vspace{-11px}
-      \caption{}
-    \end{subfigure}
-  \end{center}
-  \caption{
-  (a) This figure, adapted from the prior work of \citet{Graphite},
-  shows a Graphite palette associated with the \texttt{Color} class.
-  The client can specify the RGBA components only using literal numbers.
-  (b) A similar Hazel livelit for colors uses live splices for the RGBA components,
-  so the client can enter any Hazel expression.
-  In this case, the client has entered a variable, itself defined using a slider livelit, into the RGB
-  splices. The A splice is filled directly with a slider livelit.
-  The color livelit evaluates splices under its closure to display the
-  computed color.
-  The value of \li{default\_color} is determined by the livelit via a macro expansion step.}
+  \includegraphics[width=29pc]{grade-cutoff-livelit.png}
+\caption{Grading with Livelits}
+\label{fig:grading}
 \label{fig:color}
@@
 
-We are not, of course, the first to seek to integrate direct manipulation user interfaces
-into general-purpose programming systems.
-Prior work on projectional editing
-and active code completion, detailed in Sec. \ref{sec:related-work},
-has also considered the problem of entering expressions
-of certain types, like \li{Color},
-using specialized GUIs integrated into a program editor.
-The prior work most closely related to this paper is the {Graphite} system for Java \cite{Graphite},
-demonstrated in Fig.~\ref{fig:color}(a).
-Graphite allows library providers to associate GUIs, called \emph{palettes}, with types (using Java's class annotations).
-Wherever an expression of a type equipped with a palette is needed,
-i.e. wherever there is a \emph{hole} of that type in the program,
-the programming environment offers the programmer the option, via the code completion menu,
-to activate the palette.
-Once the interaction is finished, the palette must generate a
+% \begin{figure*}
+%   \begin{center}
+%     \begin{subfigure}[t]{0.5\textwidth}
+%       \centering
+%        \vspace{-3.4cm}
+%        \includegraphics[width=15pc]{graphite-color-palette.png}
+%       \vspace{0.1cm}
+%       \caption{}
+%     \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
+%       \centering
+%       \includegraphics[width=15pc]{slider-color-livelits.png}
+%       \vspace{-11px}
+%       \caption{}
+%     \end{subfigure}
+%   \end{center}
+%   \caption{
+%   (a) This figure, adapted from the prior work of \citet{Graphite},
+%   shows a Graphite palette associated with the \texttt{Color} class.
+%   The client can specify the RGBA components only using literal numbers.
+%   (b) A similar Hazel livelit for colors uses live splices for the RGBA components,
+%   so the client can enter any Hazel expression.
+%   In this case, the client has entered a variable, itself defined using a slider livelit, into the RGB
+%   splices. The A splice is filled directly with a slider livelit.
+%   The color livelit evaluates splices under its closure to display the
+%   computed color.
+%   The value of \li{default\_color} is determined by the livelit via a macro expansion step.}
+%   \label{fig:color}
+% \end{figure*}
+
+Of course, we are not the first to integrate direct manipulation interfaces
+into symbolic programming environments.
+% Prior work on projectional editing
+% and active code completion, detailed in Sec. \ref{sec:related-work},
+% has also considered the problem of entering expressions
+% of certain types, like \li{Color},
+% using specialized GUIs integrated into a program editor.
+% We detail prior work in Sec.~\ref{sec:related-work}, but 
+The prior work most relevant to this paper is the {Graphite} system for Eclipse for Java,
+demonstrated in Fig.~\ref{fig:color}(a) \cite{Graphite}.
+Graphite allows a library provider to associate a GUI, called a \emph{palette}, with a type 
+(via a Java class annotation).
+Wherever an expression of this type is needed,
+i.e. wherever there is a \emph{hole} of that type in the program
+(as determined by Eclipse's online parser and typechecker),
+the environment offers the client the option, via the code completion menu,
+to interact with the palette.
+Once the interaction is finished, the palette generates a
 Java expression to fill the hole.
 Figure~\ref{fig:color}(a), adapted from this prior work, demonstrates a color palette invoked using Graphite.
-When the user presses the \li{Enter} key, the Java expression \li{new Color(173, 173, 173, 85)} is inserted at the cursor (not shown) and the palette disappears.
+When the user presses the \li{Enter} key, the Java expression \li{new Color(173, 173, 173, 85)} is inserted at the cursor and the palette disappears.
+Several related systems, such as the  
+\textbf{mage} system for the Jupyter notebook environment \cite{DBLP:conf/uist/KeryRHMWP20}
+and the interactive visual syntax system for Racket \cite{interactive-visual-syntax}, behave fundamentally similarly.
+asdfasdfa
+fasdfasdf
+asdfasd\todo{placeholder for future text that might be added}
 
-\citet{Graphite} extensively evaluated this mechanism.
-Most notably, a survey of approximately 450 developers found that
-most of these developers viewed the proposed mechanism favorably and stated that they
-would use a suitable palette some or all of the time if one was available.%
-\footnote{When presented with a color palette,
-many participants remarked that they rarely entered colors directly into Java code,
-but rather into stylesheets.
-Other palettes, e.g. a palette
-that supported regular expression construction, were viewed as more
-suitable for Java code.
-The mechanisms being considered are suitable both for general-purpose languages
-and typed domain-specific languages like typed stylesheets, which can often be embedded into modern
-general-purpose languages.}
-This survey also solicited a large number of use cases from the survey participants,
-which the authors grouped into several broad categories. Notable examples include:
-\begin{itemize}
-  \item \textbf{graphical elements}, such as brushes, fonts, polygon editors, user interface widgets, and 3D primitives;
-  \item \textbf{data analysis primitives}, such as plot setups and database queries;
-  \item \textbf{composite data structures} such as matrices, maps, queues, and others for which specialized notation and affordances may be more suitable than general-purpose mechanisms; and
-  \item \textbf{data transformations} for which sensory feedback could aid in understanding,
-  e.g. audio filters, image transformations, and animations with parameters such as speed and shape.
-\end{itemize}
-
-We take the extensive empirical evaluation performed by \citet{Graphite} as sufficient to establish
-the utility and broad applicability of a mechanism of this class.
+\citet{Graphite} evaluated Graphite by surveying 473 developers 
+and \citet{DBLP:conf/uist/KeryRHMWP20} evaluated \textbf{mage} by interviewing 9 developers.
+Both studies found that
+participants viewed the proposed mechanism favorably and 
+would use a suitable GUI some or all of the time.
+% \footnote{When presented with a color palette,
+% many participants remarked that they rarely entered colors directly into Java code,
+% but rather into stylesheets.
+% Other palettes, e.g. a palette
+% that supported regular expression construction, were viewed as more
+% suitable for Java code.
+% The mechanisms being considered are suitable both for general-purpose languages
+% and typed domain-specific languages like typed stylesheets, which can often be embedded into modern
+% general-purpose languages.}
+This and other prior work also collectively showcase a wide variety of use cases \cite{Graphite,DBLP:conf/uist/KeryRHMWP20,interactive-visual-syntax}\todo{more cites}, 
+and the Graphite survey solicited dozens of additional use cases from participants,
+which the authors systematically taxonomize \cite{Graphite}. 
+We take these extensive empirical findings   
+as evidence for, and a showcase of, 
+the value of this class of mechanisms for integrating GUIs with code.
 
@@
 
-Our focus in this paper is on addressing several non-trivial technical
-deficiencies that limit both palette providers and palette clients in Graphite
-(and other projectional editing systems).
-The system of \emph{live literals}, or \emph{livelits} introduced in
-this paper and demonstrated in Fig.~\ref{fig:color}(b) improves upon Graphite
-in each of the following ways.
-\begin{enumerate}
-  \setlength\itemsep{0.5em}
-  \item \textbf{Decentralized Extensibility}:
-    In Graphite, only one palette can be associated with a type, and this must
-    be done at its definition site.
-    By contrast, livelit definitions are orthogonal to type definitions.
-    Clients invoke livelits explicitly via their names, which are prefixed by the \li{\$} character,
-    e.g. \li{\$color} and \li{\$slider} in Fig.~\ref{fig:color}(b).
-    We call this \emph{decentralized extensibility}.
-  \item \textbf{Persistence}: Graphite's palettes are {ephemeral},
-  i.e. they disappear after the initial interaction,
-  leaving behind only the generated textual code.
-  Consequently, only the programmer that initially enters the expression
-  benefits from the feedback and affordances that the palette provides.%
-  \footnote{Graphite does include an \emph{ad hoc} mechanism that
-  allows palettes to parse the code that is selected in the editor
-  when the palette loads, but this requires that each palette implement
-  a parser for the subset of Java used in the code that it generates,
-  and therefore this mechanism is quite brittle. It is also difficult
-  to persist GUI state that is not included in the generated code.}
-  By contrast, livelits are persistent. We use a pure functional model-view-update-expand architecture
-  (a variation on the Elm architecture \cite{ElmArchitecture}) as our GUI framework,
-  so only the model needs to be persisted, rather than the full GUI state.
-  We use the word ``literal'' rather than ``palette'' because, with persistence, livelits
-  operate as graphical literal notation.
+We turn our attention in this paper to fundamental technical
+deficiencies that limit GUI providers and clients using these prior systems.
+To address these, 
+we introduce a system of \emph{live literals}, or \emph{livelits}, 
+demonstrated in Fig.~\ref{fig:color}(b). 
+This system is unique in achieving all of the following properties.
+(We describe the subset of these properties 
+achieved by prior systems,
+including those just mentioned, 
+in Sec.~\ref{sec:related-work}.)
 
-  \item \textbf{Compositionality}:
-  Graphite does not provide any way to {enter subexpressions within a palette}.
-  At best, palettes can contain text boxes, but expressions entered into these text boxes
-  do not have any editor support (and thus cannot themselves be generated by palettes).
-  This implies that palettes can only reasonably generate {closed expressions}.
-  For example, in Figure~\ref{fig:color}(a), the color palette
-  is useful only for generating simple color constants.
-  By contrast, livelits have first-class support for subexpressions, which we call \emph{splices} (following
-  the terminology of \citet{TLMs} who developed reasoning principles for splicing in a strictly textual setting).
-  Fig.~\ref{fig:color}(b) demonstrates splicing: the client is able to define a variable, \li{gray\_level},
-  to capture the relationship between the red, green, and blue components.
-  The client is also able to use a slider inline in the splice governing
-  the alpha (A) component.
+\newcommand{\llproperty}[1]{\vspace{5px}\noindent\textbf{#1}.}
 
-  \item \textbf{Parameterization}: It is possible to define parameterized families of livelits.
-  For example, \li{\$slider} in Fig.~\ref{fig:color}(b) is parameterized by the lower and upper bound.
-  Parameters behave like splices, differing in that livelit parameters can be partially applied in
-  livelit abbreviations. For example, it is possible to define a livelit named \li{\$percentage}
-  that operates as a \li{\$slider} from \li{0} to \li{100} by scoped abbreviation:
-  \begin{lstlisting}[numbers=none]
-  let $percentage = $slider 0 100 in ...
-  \end{lstlisting}
+\llproperty{Decentralized Extensibility}
+    Providers define livelits in libraries, and 
+    clients invoke livelits by name. Livelit names, e.g. \li{\$color} and \li{\$slider} in Fig.~\ref{fig:color}(b),  are prefixed by \li{\$},    
+    to distinguish them from variables.
+    % We call this \emph{decentralized extensibility}.
+    % Both mage and Racket's visual syntax system are similarly extensible, 
+    % but 
+    % In Graphite, palettes are associated with class definitions, 
+    % so there is a tighter association between 
+    % Other systems, discussed in Sec.~\ref{sec:related-work}, 
+    % are either non-extensible, or cannot be extended by library providers 
+    % (only by, for example, separately installed editor extensions, or by 
+    % regenerating the editor using a language workbench.)
+
+\llproperty{Persistence}
+  % In Graphite and \textbf{mage}, GUIs are {ephemeral},
+  % i.e. they disappear after the initial interaction,
+  % leaving behind only the generated textual code.
+  % Only the programmer that initially enters the expression
+  % benefits from the feedback and affordances that the GUI provides.
+  % %
+  % \footnote{Graphite does include an \emph{ad hoc} mechanism that
+  % allows palettes to parse the code that is selected in the editor
+  % when the palette loads, but this requires that each palette implement
+  % a parser for the subset of Java used in the code that it generates,
+  % and therefore this mechanism is quite brittle. It is also difficult
+  % to persist GUI state that is not included in the generated code.}
+  Livelits are persistent elements of the syntax tree. They operate as  
+  graphical literals, rather than as the ephemeral code generation GUIs of Graphite and \textbf{mage}. 
+  We define a pure model-view-update-expand architecture
+  (a variation on Elm's model-view-update architecture \cite{ElmArchitecture}) 
+  where only the model needs to be persisted.
+  The dynamic meaning of a livelit is determined by macro expansion.
+  % We chose the word ``literal'' rather than ``palette'' because,  persistence, livelits
+  % operate as graphical literals.
+
+\llproperty{Hygienic Composition}
+  Prior systems have limited or no support for {entering sub-expressions within the GUI}, 
+  so they are useful mainly for generating expressions composed of constants,
+  e.g. color constants in Fig.~\ref{fig:color}(a).
+
+  Livelits have first-class support for sub-expressions, which we call \emph{splices} (after
+   \citet{TLMs}).
+  Fig.~\ref{fig:color}(b) demonstrates splicing: the RGBA entry boxes in the GUI 
+  are splices, so the client can define a variable, \li{baseline},
+  to create a relationship between the red, green, and blue components.\todo{tweak figure}{}
+  The client can also use a slider to specify the alpha (A) component, 
+  without the provider of the \li{\$color} slider 
+  needing to anticipate this composition. 
+  Similarly, Fig.~\ref{fig:grading}(b)\todo{subfigure labels}{} demonstrates a 
+  compositional data table livelit where each entry is a splice.
+
+  Crucially, composition is strictly 
+  governed by a hygiene discipline that ensures
+  (1) \textbf{capture avoidance}, i.e. that variables that the client uses in splices 
+  will not capture expansion-internal bindings; and 
+  (2) \textbf{context independence}, i.e. that the livelit 
+  can be invoked in any program context without pre-condition or conflict.
+
+\llproperty{Parameterization} It is possible to define parameterized families of livelits.
+  For example, \li{\$slider} in Fig.~\ref{fig:color}(b) is parameterized by the slider's bounds.
+  Parameters operate like splices, differing only in that they can be partially applied in
+  livelit abbreviations. For example, Fig.~\ref{fig:color}(b)\todo{add slider}{} 
+  partially applies \li{\$slider} to \li{0} to \li{100} to define a \li{\$percentage} slider.
+  % \begin{lstlisting}[numbers=none]
+  % let $percentage = $slider 0 100 in ...
+  % \end{lstlisting}
   % Parameterization also underlies the hygiene mechanism, as we will discuss in Sec.~\ref{sec:livelit-definitions}.
 
-  \item \textbf{Liveness}: Graphite's palettes do not have
+\llproperty{Typing} Each livelit specifies the type of expansions it generates, 
+and parameters and splices also specify types, 
+so livelits are compatible with type-driven methodologies and tools.
+Together with the hygiene discipline, this allow clients to reason abstractly about expansions, i.e.  
+without inspecting the expansions or livelit implementations directly.
+
+\llproperty{Liveness} Graphite's palettes do not have
   access to the run-time environment. By contrast, livelits can evaluate splices
@@
    select between.
-\end{enumerate}
 
-\noindent
-\textbf{Outline.} The remainder of this paper is organized as follows. We begin in Sec.~\ref{sec:case-studies} with case studies
-involving several non-trivial livelits, focusing on the livelit client's perspective.
+\paragraph{Outline.} We begin in Sec.~\ref{sec:case-studies} from the livelit client perspective 
+with case studies involving several non-trivial livelits chosen to demonstrate the novel contributions of this paper.
 In Sec.~\ref{sec:livelit-definitions}, we consider the livelit provider's perspective by detailing the livelit
 definition mechanism by example.
-In Sec.~\ref{sec:livelit-calculus}, we provide a formal account of the livelit mechanism as a typed lambda calculus.
-This calculus formally specifies the essential aspects of livelit definition, representation, expansion,
-and run-time closure collection independent of the particularities of GUI framework and other implementation details.
-This calculus and the key metatheoretic properties has been mechanized in Agda.
-In Sec.~\ref{sec:implementation}, we provide a more detailed account of our implementation of livelits within Hazel,
-and discuss several implementation details that would need to be considered by other implementations.
-In Sec.~\ref{sec:related-work}, we compare livelits to related work.
+In Sec.~\ref{sec:livelit-calculus}, we provide a formal account of the livelit mechanism, including structural representation, expansion, and live closure collection, as a typed lambda calculus. 
+We have mechanically specified the central mechanism, livelit expansion, and proven the associated metatheorems in Agda.
+The livelit calculus 
+serves to capture the essential nature of the mechanism independent of the particularities of syntax, a GUI framework, and other irrelevant design and implementation details,
+because we believe livelits can be integrated into a wide variety of programming systems. 
+In Sec.~\ref{sec:implementation}, we provide a more detailed account of our two implementations of livelits.
+Our primary implementation is integrated into Hazel, a live programming environment designed from the ground up 
+around live hole-driven development. 
+We have also prototyped livelits within a more conventional text editor 
+and discuss several considerations unique to this setting.
+In Sec.~\ref{sec:related-work}, we compare livelits to related work using the design criteria outlined above.
 Finally, we conclude in Sec.~\ref{sec:discussion} with a discussion of some present limitations and future work.

### 2020-11-20 pldi21/case-studies.tex
ported to PLDI format and tightened up abstract
reworking intro
tighten up sec 2

--- 2020-11-18 popl21/case-studies.tex
@@
-\section{Case Studies}\label{sec:case-studies}
-Let us begin by considering livelits from the client programmer's perspective by way of
+\section{Livelits by Example}\label{sec:case-studies}
+
+In this section, we will detail the livelits mechanism by way of  
 two domain-specific case studies:
@@
 These case studies have been implemented
-in Hazel, a browser-based live programming environment for a pure typed functional language in the
-ML family. We assume basic familiarity with ML.
+in Hazel, a browser-based live programming environment for a dialect of Elm. 
+Elm is an industrial pure typed functional language in the
+ML family used for client-side web development. 
+We assume basic familiarity with ML.
 
 \subsection{Case Study: Grading with Livelits}\label{sec:live-grading}
-\begin{figure}
-    \includegraphics[width=30pc]{grade-cutoff-livelit.png}
-\caption{Grading with Livelits}
-\label{fig:grading}
-\end{figure}
-
 Consider this familiar scenario: an instructor needs
@@
 (2) to visualize and perform various computations with these numeric grades
-in order ultimately to assign final letter grades (\textbf{A} through \textbf{F}) to the students in the course.
+in order ultimately to assign final letter grades.
+(In fact, this case study is not contrived: one author is using Hazel to compute grades this semester.)
 
-Perhaps the most common end-user application in this domain is the spreadsheet.
-A spreadsheet allows the instructor to record student grades using a natural tabular interface,
-visualize this data in one of a finite number of plot styles, and perform basic computations on this data,
-with results updated live, i.e. automatically after each cell has been edited.
-However, these affordances are limited. It is difficult to package up common operations and workflows
+The most common end-user application for this task is the spreadsheet, because 
+it allows the instructor to record grades using a natural tabular interface,
+visualize this data in one of a finite number of plot styles, 
+and perform basic computations,
+with results updated live.
+However, these affordances are limited. 
+It is difficult to package up common operations 
 into reusable libraries, interact with the data using domain-specific visualizations,
-and perform operations that have not been anticipated by the designers of the spreadsheet system
-(e.g. operations that prepare the data in an idiosyncratic format suitable for automated submission
-to the university's grading system).
-General-purpose programming languages equipped with suitable libraries for working with numeric data
-are flexible enough to handle these scenarios, but users of these languages
-sacrifice the ability to receive live feedback and directly manipulate data and visualizations thereof.
+and perform complex, unanticipated operations 
+(e.g. preparing the data in an idiosyncratic format demanded by the university's grading system).
+
+General-purpose programming languages 
+can handle these scenarios, but users 
+lose the ability to receive live feedback and  
+directly manipulate data and visualizations in the editor.
 
 Livelits are able to address this tension.
-Fig.~\ref{fig:grading} shows a Hazel program where the user has:
-\begin{enumerate}
-\item recorded numeric \li{grades} for each student using a livelit, \li{\$dataframe}, that implements a tabular user interface with support for live evaluation of the Hazel expression in each cell,
-\item programmatically computed the overall \li{averages} for each student by applying a function, \li{compute_averages}, defined in a library  (not shown) shared between multiple courses,
-\item determined reasonable \li{cutoffs} between letter grades by directly manipulating a livelit that offers a domain-specific drag-and-drop interface superimposed on a live visualization of the grade distribution
-(for the sake of space, we plot dots rather than a full histogram),
-\item programmatically assigned letter grades to each student based on the \li{cutoffs} determined in step (3) by calling another shared library function, \li{assign_grades},
-\item formatted the final grade assignments in a manner suitable for submission to the university system by calling another shared library function, \li{format_for_university}.
-\end{enumerate}
+Fig.~\ref{fig:grading}(c)\todo{sublabels} shows a Hazel program where 
+the instructor alternates between programmatic and direct manipulation in several situations.
+
+First, the instructor defines a value \li{grades} 
+that records the grades for each student using a livelit, \li{\$dataframe}, 
+that implements a tabular user interface. The formula bar 
+allows the selected cell to be filled with an arbitrary Hazel expression, 
+with all of Hazel's editor services available. 
+For the sake of demonstration, we show a cell that has been filled using another livelit, 
+\li{\$slider}, in combination with symbolic manipulation.\todo{do this?}{}
+The table itself displays not the expression itself but rather its value, just as in a spreadsheet.
+
+Next, the instructor programmatically computes the overall \li{averages} 
+for each student by applying \li{compute_averages}, a helper function 
+defined in the usual way in a library  (not shown) shared between multiple courses.
+
+Next, the instructor wants to ``eyeball'' reasonable \li{cutoffs} between letter grades 
+by directly manipulating a highly domain-specific livelit, \li{\$cutoffs}, that provides draggable ``paddles'' 
+superimposed on a live visualization of the distribution of the \li{averages} value provided as a parameter 
+(for the sake of space, we plot dots rather than a histogram).
+% The value of \li{cutoffs} is a labeled 4-tuple containing each cut-off.
+
+Finally, 
+the instructor programmatically assigns letter grades to each student 
+based on the selected \li{cutoffs} 
+by calling \li{assign_grades} 
+and formats these for submission with \li{format_for_university},
+again shared functions.
 
 \subsubsection{Livelit Expansion}\label{sec:livelit-expansion}
-Livelit expressions are given dynamic meaning by expansion to core expressions, i.e. expressions that do not contain livelits.
+Livelit expressions are given dynamic meaning by expansion to core expressions, which do not contain livelits.
 For example, the expansion of Fig.~\ref{fig:grading} is:
 
-\begin{lstlisting}
+\begin{lstlisting}[xleftmargin=0.2cm]
 let grades = Dataframe (
-  ["A1", "A2", "A3", "A4", "Midterm", "Final"],
-  [
-    ("Alice", [24. +. 36. +. 33., 92., 83.5, 87.5, 95., 88.]),
-    ("Bob", [61., 64., 98., 89.5, 70., 85.]),
-    ("Carol", [75., 81., 73., 71., 82., 79.]),
-    (* ... *)
-  ]) in
+  ["A1", "A2", "A3", "Midterm", "Final"],
+  [("Alice", [24. +. 36. +. 33., 
+               92., 83.5, 95., 88.]),
+   ("Bob", [61., 64., 98., 70., 85.]),
+   ("Ciri", [75., 81., 73., 82., 79.]),
+   (* ... *) ]) in
 let averages = compute_averages grades weights in
-let cutoffs = (.A 90.0, .B 80.0, .C 70.0, .D 60.0) in
-format_for_university (assign_grades averages cutoffs)
+let cutoffs = (.A 90., .B 80., .C 70., .D 60.) in
+format_for_university 
+  (assign_grades averages cutoffs)
 \end{lstlisting}
@@
 \cite{HazelnutLive}.}
+
+\begin{figure*}
+  \begin{center}
+    \begin{subfigure}[t]{0.5\textwidth}
+      \centering
+      \includegraphics[width=15pc]{img-filter-1.png}
+      \caption{}
+    \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
+      \centering
+      \includegraphics[width=15pc]{img-filter-2.png}
+      \caption{}
+    \end{subfigure}
+  \end{center}
+  \caption{Case Study: Image Transformation. The image shown is determined based on the selected closure.}
+  \label{fig:img-transformation}
+\end{figure*}
 
@@
 
-\begin{figure*}
-  \begin{center}
-    \begin{subfigure}[t]{0.5\textwidth}
-      \centering
-      \includegraphics[width=15pc]{img-filter-1.png}
-      \caption{}
-    \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
-      \centering
-      \includegraphics[width=15pc]{img-filter-2.png}
-      \caption{}
-    \end{subfigure}
-  \end{center}
-  \caption{Case Study: Image Transformation. The image shown is determined based on the selected closure.}
-  \label{fig:img-transformation}
-\end{figure*}
-
 Our next case study is a simple image transformation livelit, \li{\$img_filter},

### 2020-11-20 pldi21/intro.tex
tighten up sec 2

@@
 
-\llproperty{Liveness} Graphite's palettes do not have
-  access to the run-time environment. By contrast, livelits can evaluate splices
-  and therefore provide feedback related to the run-time behavior of the expression being generated.
-  For example, in Fig.~\ref{fig:color}(b), the color that is displayed is determined by evaluating the RGBA
+\llproperty{Liveness} Uniquely, livelits can evaluate splices
+  and so provide feedback related to the run-time behavior of the expression being generated.
+  For example, in Fig.~\ref{fig:color}(b), 
+  the color that is displayed is determined by evaluating the RGBA
   component splices to numeric values.
@@
   leaving the hole being filled by the livelit temporarily unfilled and then evaluating
-  using a two-phased variation on the semantics for incomplete programs developed for the \Hazel programming environment \cite{HazelnutLive}.
-  We will see more sophisticated examples later on, including a livelit
-  that appears inside a function that is called multiple times, leading to multiple closures that the client can
-   select between.
+  using a two-phased variation on the semantics for 
+  live programming with typed holes developed by \citet{HazelnutLive}.
+  We support live evaluation even for livelits 
+  that appears inside a function that is called multiple times: this leads 
+  to multiple closures that the client 
+   selects between.
 
-\paragraph{Outline.} We begin in Sec.~\ref{sec:case-studies} from the livelit client perspective 
-with case studies involving several non-trivial livelits chosen to demonstrate the novel contributions of this paper.
-In Sec.~\ref{sec:livelit-definitions}, we consider the livelit provider's perspective by detailing the livelit
-definition mechanism by example.
-In Sec.~\ref{sec:livelit-calculus}, we provide a formal account of the livelit mechanism, including structural representation, expansion, and live closure collection, as a typed lambda calculus. 
+\paragraph{Outline.} We begin in Sec.~\ref{sec:case-studies} by introducing
+livelits from the perspective of client programmers. 
+Our examples are organized into case studies 
+and are chosen to specifically demonstrate the novel contributions of this paper.
+In Sec.~\ref{sec:livelit-definitions}, 
+we consider the livelit provider's perspective by detailing the livelit
+definition mechanism with an example.
+In Sec.~\ref{sec:livelit-calculus}, we define the \emph{typed livelit calculus}. 
 We have mechanically specified the central mechanism, livelit expansion, and proven the associated metatheorems in Agda.
-The livelit calculus 
-serves to capture the essential nature of the mechanism independent of the particularities of syntax, a GUI framework, and other irrelevant design and implementation details,
+This calculus 
+serves to capture the essential nature of livelits 
+independent of the particularities of syntax, GUI frameworks, 
+and other orthogonal design details,
 because we believe livelits can be integrated into a wide variety of programming systems. 
 In Sec.~\ref{sec:implementation}, we provide a more detailed account of our two implementations of livelits.
-Our primary implementation is integrated into Hazel, a live programming environment designed from the ground up 
-around live hole-driven development. 
-We have also prototyped livelits within a more conventional text editor 
-and discuss several considerations unique to this setting.
-In Sec.~\ref{sec:related-work}, we compare livelits to related work using the design criteria outlined above.
-Finally, we conclude in Sec.~\ref{sec:discussion} with a discussion of some present limitations and future work.
+Our primary implementation, used in the screenshots in the paper,
+is within Hazel, a live programming environment designed 
+around hole-driven development. 
+We have also prototyped livelits within a standard text editor. 
+We also discuss factors to consider when adding livelits to languages with side effects.
+In Sec.~\ref{sec:related-work}, we compare livelits to related work using the design properties outlined above 
+as a rubric.
+Finally, we conclude in Sec.~\ref{sec:discussion} after a discussion of some present limitations and future work.

### 2020-11-20 pldi21/case-studies.tex
nearly have sec 2 into 5 pages

@@
 that implements a tabular user interface. The formula bar 
-allows the selected cell to be filled with an arbitrary Hazel expression, 
-with all of Hazel's editor services available. 
+allows the selected cell to be filled with an arbitrary Hazel expression. 
 For the sake of demonstration, we show a cell that has been filled using another livelit, 
@@
 
-Next, the instructor programmatically computes the overall \li{averages} 
+Next, the instructor computes \li{averages} 
 for each student by applying \li{compute_averages}, a helper function 
-defined in the usual way in a library  (not shown) shared between multiple courses.
+defined in a library  (not shown) shared between multiple courses.
 
 Next, the instructor wants to ``eyeball'' reasonable \li{cutoffs} between letter grades 
-by directly manipulating a highly domain-specific livelit, \li{\$cutoffs}, that provides draggable ``paddles'' 
+by directly manipulating a highly domain-specific livelit, \li{\$grade_cutoffs}, that provides draggable ``paddles'' 
 superimposed on a live visualization of the distribution of the \li{averages} value provided as a parameter 
@@
 Finally, 
-the instructor programmatically assigns letter grades to each student 
-based on the selected \li{cutoffs} 
+the instructor programmatically assigns grades to students 
+based on these \li{cutoffs} 
 by calling \li{assign_grades} 
-and formats these for submission with \li{format_for_university},
+and \li{format_for_university},
 again shared functions.
 
-\subsubsection{Livelit Expansion}\label{sec:livelit-expansion}
-Livelit expressions are given dynamic meaning by expansion to core expressions, which do not contain livelits.
-For example, the expansion of Fig.~\ref{fig:grading} is:
+\subsection{Livelit Expansion}\label{sec:livelit-expansion}
+Livelit invocations 
+expand to expressions.
+For example, the expansion of Fig.~\ref{fig:grading}(c) is:
 
@@
 
-The programmer can inspect this expansion in Hazel.
-Ideally, however, it would not be necessary to inspect the expansion
-(or the livelit implementation, which specifies the expansion logic and is discussed in Sec.~\ref{sec:livelit-definitions})
-to reason about types and binding.
-After all, one does not need to look inside
+The client can inspect this expansion in Hazel via a toggle (not shown).
+Ideally, however, reasoning about types and binding
+should not require the client to inspect the expansion 
+nor the livelit implementation (which specifies the expansion logic 
+as we will describe in Sec.~\ref{sec:livelit-definitions}).
+After all, function clients do not need to look inside
 function bodies to reason about types and binding.
@@
 ``type structure is a syntactic discipline for maintaining levels of abstraction''.
-The livelits mechanism supports reasoning abstractly about types and binding by
-several means.
+Livelits maintain this discipline by
+several means, described next in Sec.~\ref{sec:expansion-typing}-\ref{sec:hygiene}.
 
+\subsubsection{Expansion Typing} 
+\label{sec:expansion-typing}
 To support abstract reasoning about the type of the expansion,
@@
 eliding their implementations, are:
-\begin{lstlisting}[numbers=none]
-livelit $dataframe at Dataframe { ... }
-livelit $grade_cutoffs(avgs: List(Float))
-  at (.A Float, .B Float, .C Float, .D Float) { ... }
+\begin{lstlisting}[numbers=none,xleftmargin=0cm]
+livelit $dataframe at Dataframe {...}
+livelit $grade_cutoffs(averages: List(Float)) at 
+  (.A Float, .B Float, .C Float, .D Float) {...}
 \end{lstlisting}
-The expansion type of \li{\$dataframe} is \li{Dataframe}
-(a type classifying tabular floating point data together with string row and column keys, consistent with the expansion above)
-and the expansion type of \li{\$grade_cutoffs} is a labeled product of grade cutoffs (Hazel notates labels \li{.label}).
-Hazel summarizes the typing information from the livelit definition when the cursor is on the livelit's name,
-just as it displays a function's type when its name is under the cursor (not shown).
+The expansion type of \li{\$dataframe} is \li{Dataframe},
+which classifies tabular floating point data together with string row and column names (see the expansion above).
+The expansion type of \li{\$grade_cutoffs} is a labeled product of grade cutoffs (field labels are written \li{.label}
+rather than \li{label:} in Hazel).
+Hazel displays the information in the livelit declaration when the cursor is on the livelit's name,
+just as it displays typing information in other situations (not shown).\todo{cite HATRA}{}
 
-\subsubsection{Typed Hygienic Splicing and Parameterization}\label{sec:splicing-and-parameterization}
-Livelit expressions can include sub-expressions in one of two ways:
-as parameters or as splices.
+\subsection{Compositionality}\label{sec:splicing-and-parameterization}
+Livelits are compositional: they can work with sub-expressions  
+in the form of parameters and splices.
 
-\paragraph{Parameters}\label{sec:parameterization} Parameters appear immediately after the name of the applied livelit.
-For example, \li{averages} is provided as a parameter to \li{\$grade_cutoffs} in Fig.~\ref{fig:grading}.
-The livelit declares a fixed number of parameters and specifies their types in its declaration.
-For example, the declaration of \li{\$grade_cutoffs} shown above specifies one parameter of type \li{List(Float)}
-(the grades data to be plotted, see below).
+\subsubsection{Parameters}\label{sec:parameterization} 
+Livelit can declare a finite number of parameters of specified types. 
+For example, \li{\$grade_cutoffs} above declares one parameter,
+the averages to be plotted, of type \li{List(Float)}.
+Parameters are applied  
+using function application notation 
+as seen in Fig.~\ref{fig:grading}(c) or 
+using the pipelining (i.e. reverse function application) operators, \li{<|} and \li{|>}, 
+which allow multiple livelits to form dataflows (not shown).
 
-Livelit abbreviations can partially apply parameters. Consider the slider livelit
-from Fig.~\ref{fig:color}:
-\begin{lstlisting}[numbers=none]
-livelit $slider (min: Float) (max: Float) at Float { ... }
+Livelit abbreviations can partially apply parameters. For example, consider the slider livelit
+from Fig.~\ref{fig:color}\todo{where?}{}\todo{int or float?}:
+\begin{lstlisting}[numbers=none,xleftmargin=0cm]
+livelit $slider (min: Int) (max: Int) at Int {...}
 \end{lstlisting}
-We can partially apply the first parameter as follows to define a parameterized non-negative slider livelit with one remaining parameter:
-\begin{lstlisting}[numbers=none]
-livelit $nnslider = $slider 0.0 in ...
+We can partially apply the first parameter to define a parameterized non-negative slider livelit:
+\begin{lstlisting}[numbers=none,xleftmargin=0cm]
+let $nnslider = $slider 0 in ...
 \end{lstlisting}
-Only livelits with no remaining parameters can be used to fill a hole.
-So writing \li{\$nnslider} in expression position will not display the slider GUI. Instead, it will display as a ``missing livelit parameter'' error.%
-\footnote{\label{footnote:typing}In Hazel, erroneous expressions are automatically placed inside holes so that they do not prevent other parts of the program from executing
+Only livelits with no remaining parameters can be invoked, 
+so writing \li{\$nnslider} in expression position will display as a ``missing livelit parameter'' error.%
+\footnote{\label{footnote:typing}In Hazel, erroneous expressions 
+are automatically placed inside holes so that they do not prevent other parts of the program from evaluating
 \cite{HazelnutLive}.}
@@
 
-\paragraph{Splices}\label{sec:splices}
-Spliced expressions (or \emph{splices}) appear directly inside the livelit GUI.
-Splices can be filled with Hazel expressions of any form, including other livelit expressions,
-and all of Hazel's editing affordances are available when editing these expressions.
+\subsubsection{Splices}\label{sec:splices}
+Spliced expressions, or \emph{splices}, appear directly inside the livelit GUI.
+Splices can be filled with Hazel expressions of any form, including other livelit invocations.
 For example, each cell in the \li{\$dataframe} GUI in Fig.~\ref{fig:grading}
-has a corresponding splice, which appears in the formula bar at the top of the livelit
-when the user has selected that cell.
-The cell itself displays the live value of the spliced expression,
-following the example of spreadsheet interfaces
-(we return to live evaluation in Sec.~\ref{sec:live-evaluation} below).
+has a corresponding splice. The formula bar at the top 
+allows the user to edit the splice corresponding to the selected cell,
+and all of Hazel's editing affordances are available when the client does so.
+% Not all splices necessarily appear in the GUI.
+Unlike parameters, the number of splices can change 
+as the user interacts with the livelit, e.g. when changing the number of rows or columns in a \li{\$dataframe}.
+% The cell itself displays the live value of the spliced expression---
+% (we return to live evaluation in Sec.~\ref{sec:live-evaluation} below.
 
-The livelit provides an expected type for each splice.
-For example, the splices for the row and column keys in Fig.~\ref{fig:grading} are expected to be expressions
-of type \li{String}, and the remaining cells are expected to be expressions of type \li{Float}.
-Hazel surfaces this typing information for the programmer when the cursor is in a splice,
-so it is not necessary to inspect the expansion or the livelit implementation.
-If an expression of invalid type is entered, it will display in an error hole as usual,
-and in Hazel this will not prevent evaluation of other expressions (see Footnote \ref{footnote:typing}).
+The livelit provides an expected type for each splice when it is created.
+For example, the splices for the row and column keys in Fig.~\ref{fig:grading}(c)\todo{go through subfigures}{} have expected type \li{String}, 
+and the remaining cells have expected type \li{Float}.
+Hazel surfaces and uses the expected type when the cursor is in the splice.\todo{cite HATRA paper}{}
+% If an expression of invalid type is entered, it will display in an error hole as usual,
+% and in Hazel this will not prevent evaluation of other expressions (see Footnote \ref{footnote:typing}).
 
-Unlike parameters, the number of splices is not fixed in the livelit declaration. Splices can be created,
-deleted, and filled through user interaction with the livelit. For example, clicking the \li{+} buttons
-in Fig.~\ref{fig:grading} will create new rows or columns, which will in turn generate new splices.
+% Unlike parameters, the number of splices is not fixed in the livelit declaration. Splices can be created,
+% deleted, and filled through user interaction with the livelit. For example, clicking the \li{+} buttons
+% in Fig.~\ref{fig:grading} will create new rows or columns, which will in turn generate new splices.
 
-\paragraph{Hygiene}\label{sec:hygiene}
-Parameters and splices can both appear in the expansion. For example,
-the expansion of the \li{\$dataframe} livelit in Fig.~\ref{fig:grading} includes the
-entered cell contents and the row and column keys.
+\subsubsection{Hygienic Composition}\label{sec:hygiene}
 
-The client cannot know, without looking at the expansion or the livelit implementation,
-where in the expansion each parameter or splice will appear. This becomes relevant when
-the expansion places the parameter or splice under a binder, e.g. in the body of a function or let binding.
+Ensuring that clients can reason about binding while leaving expansions
+ invisible 
+ requires a hygiene discipline that enforces \emph{capture avoidance}
+and \emph{context independence} \cite{TLMs}\todo{cite michael hygiene paper}.
+
+\paragraph{Capture Avoidance}
+Splices and parameters can appear anywhere in the expansion. 
+This becomes potentially problematic when
+the expansion places the parameter or splice under a binder, e.g. in the body of a function or \li{let} binding.
 Na\"ively, this could cause inadvertent capture of the bound variable by a free variable
 in the parameter or splice. For example, consider a livelit that generates an expansion
-of the following form:
+of the following seemingly innocuous form:
 \begin{lstlisting}[numbers=none]
 let len = strlen <splice1> in
-(<splice2>, <splice2> + 1)
+Some (<splice2> + len)
 \end{lstlisting}
-Here, \li{<splice2>} appears under the binding of \li{len}. If the client's choice for
-\li{<splice2>} itself refers to a client-side binding of the same identifier, \li{len},
-it would na\"ively be captured. This sort of bug would difficult to debug,
-both for the livelit provider and the client.
+Here, \li{<splice2>} appears under the binding of \li{len}. If the client has filled
+\li{<splice2>} with an expression that refers to a client-side binding of \li{len},
+these references would na\"ively be captured. This would not occur in \li{<splice1>}, 
+because the \li{let} is not recursive.
+This breaks abstraction and is notoriously difficult to debug,
+both for the livelit provider, who has no way to predict which variables a client will use,
+ and the client, who does not know which variables the provider used.
 
-To avoid this situation, the livelit expansion mechanism
-maintains parameter and splice hygiene: parameters and splices are included in the expansion
-in a capture-avoiding manner.
-Consequently, clients can reason abstractly about binding in splices: variables in splices
-always refer to bindings visible to the client, rather than bindings that are hidden inside expansions.
+To avoid this situation, parameters and splices are placed in the expansion
+in a capture-avoiding manner: variables in splices
+always refer to the bindings visible to the client, 
+rather than bindings that are hidden inside the expansion.
+We discuss how this is implemented in Sec.~\ref{sec:expansion}.
 % This is implemented by alpha-renaming bindings internal to the expansion as necessary.
@@
 
-\subsubsection{Live Evaluation}\label{sec:live-evaluation}
-Livelits can request the result of evaluating a splice or a parameter.
-For example, the \li{\$dataframe} livelit uses this facility to display
-the evaluation result for each cell when the cursor is not on it, like a spreadsheet.
-The \li{\$grade_cutoffs} livelit uses this facility to plot the grades list, provided
-as a parameter, on the number line.
+\paragraph{Context Independence}
+The example expansion above used a library function, \li{strlen}.
+Na\"ively, this expansion would break if placed 
+in client contexts where \li{strlen} is not bound, or bound to 
+an unexpected value.
+To avoid requiring clients to determine and satisfy these invisible 
+dependencies, the livelits mechanism enforces \emph{context independence}:
+generated expansions are valid in any context. Dependencies are bound 
+relative to the livelit definition site (see Sec.~\ref{sec:expansion}).
 
-\paragraph{Closure Collection} Evaluation in Hazel is, as usual, defined only for closed expressions,
-but parameters and splices can both refer to variables in the surrounding
-context. In order to provide evaluation results for parameters or splices containing free variables,
+\subsection{Live Evaluation}\label{sec:live-evaluation}
+Livelits have the ability to evaluate a splice or a parameter 
+in order to provide better feedback about run-time behavior to the client.
+The \li{\$dataframe} livelit uses this facility to display
+the evaluation result for each cell, like a spreadsheet.
+The \li{\$grade_cutoffs} livelit uses this facility to plot the grades, which were 
+passed in as a parameter, on the number line.
+
+\subsubsection{Closure Collection} The subtlety here is that 
+evaluation in Hazel is defined for closed expressions as is typical,
+but parameters and splices can be open, i.e. refer to variables in the surrounding
+context. In order to provide a suitable run-time environment that binds these variables, 
 Hazel performs \textbf{closure collection} in two phases.
 
-In the first phase, Hazel evaluates the program after placing holes in place of each livelit,
-following the evaluation semantics for incomplete programs developed by \citet{HazelnutLive}.
-The result is an expression with holes in it, each of which is equipped with a closure.
-For example, there is a single closure associated with the application of \li{\$dataframe} in Fig.~\ref{fig:grading}.
-This closure contains the value of the \li{weights} variable from Line 1 and any other variables in the context
-around that livelit, e.g.
-from imported libraries, not shown. Consequently, when the livelit requests the evaluation result for
-splices that use these variables, such as the cell shown selected in Fig.~\ref{fig:grading},
-an evaluation result is available. (We consider the situation where more than one closure is available
-in the Sec.~\ref{sec:image-transformation}.)
+In the first phase, \emph{proto-closure collection}, 
+Hazel replaces each livelit with a uniquely numbered hole and then evaluates the program 
+using the semantics for evaluating programs with holes developed by \citet{HazelnutLive}.
+Evaluation proceeds around these holes, producing a result with hole closures, i.e. holes with environments.
 
-Similarly, the closure for the \li{\$grade_cutoffs} livelit in Fig.~\ref{fig:grading} includes a result for
-the \li{grades} variable. However, this variable's result, along with the
-results for other variables that depend on \li{grades}, such as \li{averages}, ultimately depend on the
+For example, there is one closure for \li{\$dataframe} in Fig.~\ref{fig:grading}.
+It contains the value of the \li{q1max}\todo{what?}{} variable from Line 1 and any other variables in the context
+around that livelit. 
+These values can be used to evaluate 
+splices that use the corresponding variables, such as the cell selected in Fig.~\ref{fig:grading}(c).
+
+Similarly, the closure for \li{\$grade_cutoffs} in Fig.~\ref{fig:grading} includes a result for
+the \li{grades} variable. However, there is a problem: this variable's result depends on the
 expansion of the \li{\$dataframe} livelit. If we stop after the first phase of closure collection,
-where these expansions have not yet been included,
-then no useful result will be available here.
-For this reason, there is a second phase of closure collection where any livelit holes that appear
+where livelits have not yet been expanded,
+then no useful value for the \li{grades} variable will be available after proto-closure collection:
+it will just be a \li{\$dataframe} livelit hole closure.
+For this reason, there is a second phase of closure collection, called \emph{closure resumption}, 
+where any livelit holes that appear
 in the collected livelit closures are \emph{resumed}, i.e. the expansion is generated
-(in this case, for the \li{\$dataframe} livelit) and is used to fill
-the hole before evaluation resumes, following the semantics for resumption developed by \citet{HazelnutLive}.
-Livelit expansions do not themselves contain livelits, so no subsequent resumption phases are necessary.
+(in this case, for the \li{\$dataframe} livelit), the hole is filled, 
+and evaluation resumes.
+Livelit expansions do not contain livelits, so no subsequent resumption phases are necessary.
 
-The final evaluation result for the program as a whole can also proceed by resumption, because
-closure collection performs all of the computations that do not ultimately depend on the livelit expansion.
-Consequently, Hazel does not need to evaluate the program multiple times in order to support edit-time
-(i.e. live) evaluation in this way.
+Hazel does not need to evaluate the program multiple times in order to support live closure collection
+as well as the usual live evaluation services it offers, because the final evaluation result 
+can also be determined by resumption.
+(We discuss how to address subtleties that arise 
+in the presence of non-commutative side effects in Sec.~\ref{sec:calculus-closure-collection}.)
 
-\paragraph{Indeterminate Results}
+\subsubsection{Indeterminate Results}
 We used the phrase \emph{evaluation result} rather than \emph{value} above purposefully:
@@
 (in this case, it would display the list elements that are values on the timeline, skipping indeterminate elements).
-We will return to how livelit implementations handle this situation in Sec.~\ref{sec:livelit-definitions}.
+We will return to how livelit implementations handle this situation in Sec.~\ref{sec:live-evaluation-def}.
 
-\subsection{Case Study: Image Transformations}\label{sec:image-transformation}
-
+\subsubsection{Case Study: Live Image Transformations}\label{sec:image-transformation}
+\todo{revise this once we have final figure}{}
 Our next case study is a simple image transformation livelit, \li{\$img_filter},

### 2020-11-20 pldi21/intro.tex
nearly have sec 2 into 5 pages

@@
 
-We turn our attention in this paper to fundamental technical
+We turn our attention in this paper to several fundamental technical
 deficiencies that limit GUI providers and clients using these prior systems.
@@
 demonstrated in Fig.~\ref{fig:color}(b). 
-This system is unique in achieving all of the following properties.
-(We describe the subset of these properties 
-achieved by prior systems,
+Livelits are unique in achieving all of the following properties.
+(We describe which subset of these properties 
+are achieved by prior systems,
 including those just mentioned, 
@@
   livelit abbreviations. For example, Fig.~\ref{fig:color}(b)\todo{add slider}{} 
-  partially applies \li{\$slider} to \li{0} to \li{100} to define a \li{\$percentage} slider.
+  partially applies \li{\$slider} to \li{0} and \li{100} to define a \li{\$percentage} slider\todo{todo}{}.
   % \begin{lstlisting}[numbers=none]
@@
 \llproperty{Liveness} Uniquely, livelits can evaluate splices 
-  and so provide feedback related to the run-time behavior of the expression being generated.
+  throughout the editing process 
+  (i.e. in a \emph{live} manner \cite{DBLP:conf/icse/Tanimoto13}) 
+  to provide feedback related to run-time behavior.
   For example, in Fig.~\ref{fig:color}(b), 
-  the color that is displayed is determined by evaluating the RGBA
+  the selected color can be displayed by evaluating the RGBA
   component splices to numeric values.
@@
   leaving the hole being filled by the livelit temporarily unfilled and then evaluating
-  using a two-phased variation on the semantics for 
+  using a two-phased variant of the semantics for 
   live programming with typed holes developed by \citet{HazelnutLive}.
   We support live evaluation even for livelits 
-  that appears inside a function that is called multiple times: this leads 
+  that appear inside a function. Multiple function calls lead 
   to multiple closures that the client 
-   selects between.
+   can select between.
 
@@
 Our examples are organized into case studies 
-and are chosen to specifically demonstrate the novel contributions of this paper.
+and are chosen to demonstrate the novel contributions of this paper.
 In Sec.~\ref{sec:livelit-definitions}, 
-we consider the livelit provider's perspective by detailing the livelit
-definition mechanism with an example.
+we consider the livelit provider's perspective by introducing livelit
+definitions with a detailed example.
 In Sec.~\ref{sec:livelit-calculus}, we define the \emph{typed livelit calculus}. 
@@
 Our primary implementation, used in the screenshots in the paper,
-is within Hazel, a live programming environment designed 
+is integrated into Hazel, a live programming environment designed 
 around hole-driven development. 
 We have also prototyped livelits within a standard text editor. 
-We also discuss factors to consider when adding livelits to languages with side effects.
+Additionally, we discuss factors that must be considered when integrating livelits into languages with side effects.
 In Sec.~\ref{sec:related-work}, we compare livelits to related work using the design properties outlined above 
 as a rubric.
-Finally, we conclude in Sec.~\ref{sec:discussion} after a discussion of some present limitations and future work.
+Finally, we conclude in Sec.~\ref{sec:discussion} after a discussion of present limitations and future work.

### 2020-11-20 pldi21/livelit-definitions.tex
further tightening of abstract  typo fix
nearly have sec 2 into 5 pages

--- 2020-07-10 livelit-definitions.tex
@@
 \subsubsection{Live Evaluation}
+\label{sec:live-evaluation-def}
 In order to provide the client with live feedback about the dynamic implications 
@@
 \subsection{Expansion}
+\label{sec:expansion}
 The ultimate purpose of a livelit is to fill the hole where it appears by generating an expansion,
@@
 
-The expansion can include splices, but we the system does not make the contents of a splice 
+The expansion can include splices, but the system does not make the contents of a splice 
 available directly. Instead, the \li{expand} must return a pair consisting of an encoded expression, of type 

### 2020-11-20 pldi21/intro.tex
fix and make fig 3 fit

@@
 or to transform a directly constructed musical sequence 
-by passing it through a series of symbolically specified functions.
+by passing it through a series of symbolically defined functions.
 Moreover, it is difficult to add new affordances or to compose
 affordances in ways that the application developer did not anticipate.
-Users cannot easily make even simple changes like replacing a numeric text box in a color palette with a slider,
-much less more ambitious changes like installing an alternative visual interface for expressing geospatial data queries in a workflow.
+Users cannot easily make even simple changes like replacing a numeric text box in a dialog with a slider,
+much less more ambitious changes like installing an alternative visual interface for expressing geospatial data queries 
+into a database frontend.
 

### 2020-11-20 pldi21/livelit-definitions.tex
fix and make fig 3 fit

@@
 \begin{figure}
-\begin{lstlisting}
+\begin{lstlisting}[xleftmargin=0.2cm,morekeywords={expansions,require}]
 type Color = (.r Int, .g Int, .b Int, .a Int)
@@
 livelit $color at Color {
-    capture { }
+  type Model = (.r SpliceRef, .g SpliceRef, 
+                .b SpliceRef, .a SpliceRef)
 
-    type Model = (.r SpliceRef, .g SpliceRef, .b SpliceRef, .a SpliceRef)
   let init : UpdateMonad(Model) = do 
@@
   type Action = 
-    | ClickOnColor(Color)
+  | ClickOn(Color)
 
@@
     fun model -> do 
-        (* determine a color to display to the client, if possible *)
+      (* determine a color to display *)
       r_res <- eval_splice(model.r)
@@
           Some((r, g, b, a))
-          | _ -> None
+        | _ -> 
+          (* indetermine color shown as X *)
+          None 
       in
 
-        r_editor <- editor(model.r, FixedWidth(20))
-        g_editor <- editor(model.g, FixedWidth(20))
-        b_editor <- editor(model.b, FixedWidth(20))
-        a_editor <- editor(model.a, FixedWidth(20))
+      (* generate splice editors *)
+      let size = FixedWidth(20) in 
+      r_editor <- editor(model.r, size)
+      g_editor <- editor(model.g, size)
+      b_editor <- editor(model.b, size)
+      a_editor <- editor(model.a, size)
 
-        (* ...render the UI using these splice editors... *)
+      (* ... now we can render the UI ... *)
 
-    let update : Model -> Action -> UpdateMonad(Model) =
-      fun model (ClickOnColor(r, g, b, a)) -> do 
-        set_splice(model.r, IntLit(r))
-        set_splice(model.g, IntLit(g))
-        set_splice(model.b, IntLit(b))
-        set_splice(model.a, IntLit(a))
-        return (r, g, b, a)
+  let update : 
+      Model -> Action -> UpdateMonad(Model) =
+    fun model (ClickOn c) -> do 
+      set_splice(model.r, IntLit(c.r))
+      set_splice(model.g, IntLit(c.g))
+      set_splice(model.b, IntLit(c.b))
+      set_splice(model.a, IntLit(c.a))
+      return model
 
+  expansions require { }
   let expand : Model -> (Exp, List(SpliceRef)) = 
-      fun model -> 
-        (`fun r g b a -> (r, g, b, a)`, [model.r, model.g, model.b, model.a])
+    fun model -> (`fun r g b a -> (r, g, b, a)`, 
+      [model.r, model.g, model.b, model.a])
 }
 \end{lstlisting}
-\caption{Livelit Implementation: \li{\$color} (simplified for presentation)}
+\vspace{-8px}
+\caption{Livelit Implementation}
 \label{fig:color-impl}

### 2020-11-21 pldi21/case-studies.tex
finished tweaking sec 2 and expanded filter case study substantially

@@
 Next, the instructor wants to ``eyeball'' reasonable \li{cutoffs} between letter grades 
-by directly manipulating a highly domain-specific livelit, \li{\$grade_cutoffs}, that provides draggable ``paddles'' 
-superimposed on a live visualization of the distribution of the \li{averages} value provided as a parameter 
-(for the sake of space, we plot dots rather than a histogram).
+by directly manipulating a domain-specific livelit, \li{\$grade_cutoffs}, that provides draggable ``paddles'' 
+superimposed on a live visualization of the distribution of \li{averages}, which is provided as a livelit parameter.
 % The value of \li{cutoffs} is a labeled 4-tuple containing each cut-off.
@@
 livelit definitions declare an \emph{expansion type}.
-The declarations of the two livelits in Fig.~\ref{fig:grading},
+The declarations of the livelits in Fig.~\ref{fig:grading},
 eliding their implementations, are:
@@
   (.A Float, .B Float, .C Float, .D Float) {...}
+livelit $slider (min: Int) (max: Int) at Int {...}
 \end{lstlisting}
@@
 
-Livelit abbreviations can partially apply parameters. For example, consider the slider livelit
-from Fig.~\ref{fig:color}\todo{where?}{}\todo{int or float?}:
+Livelit abbreviations can partially apply parameters. For example, 
+we can partially apply the first parameter of \li{\$slider} to define a parameterized unsigned slider livelit:
 \begin{lstlisting}[numbers=none,xleftmargin=0cm]
-livelit $slider (min: Int) (max: Int) at Int {...}
-\end{lstlisting}
-We can partially apply the first parameter to define a parameterized non-negative slider livelit:
-\begin{lstlisting}[numbers=none,xleftmargin=0cm]
-let $nnslider = $slider 0 in ...
+let $uslider = $slider 0 in ...
 \end{lstlisting}
 Only livelits with no remaining parameters can be invoked, 
-so writing \li{\$nnslider} in expression position will display as a ``missing livelit parameter'' error.%
+so writing \li{\$uslider} in expression position will display as a ``missing livelit parameter'' error.%
 \footnote{\label{footnote:typing}In Hazel, erroneous expressions 
@@
 
-\paragraph{Capture Avoidance}
+\textbf{Capture Avoidance.}
 Splices and parameters can appear anywhere in the expansion. 
-This becomes potentially problematic when
-the expansion places the parameter or splice under a binder, e.g. in the body of a function or \li{let} binding.
+This becomes problematic when
+they appear under a binder, e.g. in the body of a function or \li{let} binding.
 Na\"ively, this could cause inadvertent capture of the bound variable by a free variable
@@
 
-\paragraph{Context Independence}
+\textbf{Context Independence.}
 The example expansion above used a library function, \li{strlen}.
@@
 
-\subsubsection{Closure Collection} The subtlety here is that 
-evaluation in Hazel is defined for closed expressions as is typical,
-but parameters and splices can be open, i.e. refer to variables in the surrounding
-context. In order to provide a suitable run-time environment that binds these variables, 
+\subsubsection{Closure Collection} The subtlety is that 
+evaluation in Hazel is defined for closed expressions as usual,
+but parameters and splices can be open, i.e. refer to surrounding variables. 
+To provide a environment that binds these variables, 
 Hazel performs \textbf{closure collection} in two phases.
@@
 using the semantics for evaluating programs with holes developed by \citet{HazelnutLive}.
-Evaluation proceeds around these holes, producing a result with hole closures, i.e. holes with environments.
+Evaluation proceeds around these holes, producing a result containing 
+corresponding hole closures, i.e. holes with environments.
 
 For example, there is one closure for \li{\$dataframe} in Fig.~\ref{fig:grading}.
-It contains the value of the \li{q1max}\todo{what?}{} variable from Line 1 and any other variables in the context
-around that livelit. 
+It contains the value of \li{q1max}\todo{what?}{} and the other variables in scope. 
 These values can be used to evaluate 
@@
 
-Similarly, the closure for \li{\$grade_cutoffs} in Fig.~\ref{fig:grading} includes a result for
-the \li{grades} variable. However, there is a problem: this variable's result depends on the
-expansion of the \li{\$dataframe} livelit. If we stop after the first phase of closure collection,
-where livelits have not yet been expanded,
-then no useful value for the \li{grades} variable will be available after proto-closure collection:
-it will just be a \li{\$dataframe} livelit hole closure.
+Similarly, the closure for \li{\$grade_cutoffs} in Fig.~\ref{fig:grading} includes 
+the necessary \li{averages} variable, but 
+its value depends on \li{grades}, which is determined by \li{\$dataframe}. 
+If we stop after proto-closure collection,
+then no useful value will be available:
+the result will be \emph{indeterminate}, because \li{\$dataframe} has been replaced with a hole \cite{HazelnutLive}.
 For this reason, there is a second phase of closure collection, called \emph{closure resumption}, 
@@
 in the collected livelit closures are \emph{resumed}, i.e. the expansion is generated
-(in this case, for the \li{\$dataframe} livelit), the hole is filled, 
-and evaluation resumes.
-Livelit expansions do not contain livelits, so no subsequent resumption phases are necessary.
+(in this case, for the \li{\$dataframe} livelit) and evaluation resumes.
+% Livelit expansions do not contain livelits, so no subsequent expansion / resumption phases are necessary.
 
-Hazel does not need to evaluate the program multiple times in order to support live closure collection
-as well as the usual live evaluation services it offers, because the final evaluation result 
-can also be determined by resumption.
-(We discuss how to address subtleties that arise 
-in the presence of non-commutative side effects in Sec.~\ref{sec:calculus-closure-collection}.)
+% Hazel does not need to evaluate the program multiple times in order to support live closure collection
+% as well as the usual live evaluation services it offers, because the final evaluation result 
+% can also be determined by resumption.
+% (We discuss how to address subtleties that arise 
+% in the presence of non-commutative side effects in Sec.~\ref{sec:calculus-closure-collection}.)
 
 \subsubsection{Indeterminate Results}
-We used the phrase \emph{evaluation result} rather than \emph{value} above purposefully:
-splices and parameters can be incomplete, i.e. they can contain holes, so
-evaluation does not always result in a value \cite{HazelnutLive}.
-In particular, when a hole appears in elimination position, e.g. in function position of a function application,
- the result is instead an indeterminate expression.
-When a livelit requests an evaluation result, it must be able to handle indeterminate results.
-For example, if the parameter to \li{\$grade_cutoffs} were a hole or some other indeterminate expression,
-then it would have degraded functionality
-(in this case, it would display the list elements that are values on the timeline, skipping indeterminate elements).
-We will return to how livelit implementations handle this situation in Sec.~\ref{sec:live-evaluation-def}.
+Even after resumption, some elements of the closure may remain indeterminate, e.g. due to holes that 
+are not filled with livelits. 
+When a livelit requests an evaluation result, it must be able to handle these indeterminate results.
+For example, if there were missing grades,
+then \li{\$grade_cutoffs} would have degraded functionality: 
+it would display only the list elements that are values on the timeline, skipping indeterminate elements.
+We will return to how this occurs in Sec.~\ref{sec:live-evaluation-def}.
 
-\subsubsection{Case Study: Live Image Transformations}\label{sec:image-transformation}
+\subsubsection{Case Study: Live Photo Filters}\label{sec:image-transformation}
 \todo{revise this once we have final figure}{}
-Our next case study is a simple image transformation livelit, \li{\$img_filter},
-demonstrated in Fig.~\ref{fig:img-transformation}. This livelit takes 
+Closure collection can produce multiple closures when, for example, 
+a livelit appears inside a function applied multiple times. 
+Our next case study, in the photography domain,
+demonstrates how Hazel handle this situation, and 
+the workflow it enables.
+
+We interviewed a photographer 
+who described a typical workflow: 
+they use the Lightroom application to 
+apply a set of adjustments and filters 
+across all photos in a collection before making 
+individual adjustments. 
+Many photographers do this one photo at a time,
+though this photographer had recently learned how to
+use Lightroom's saved presets to 
+apply adjustments to multiple photos at once.
+However, they remained dissatisfied by the workflow.
+They wanted to be able to see how the shared settings affected 
+multiple photos as they tweaked them, without having to 
+save and apply the preset repeatedly.
+They also wanted to be able to change the shared settings
+even after making certain individual adjustments.
+They also expressed interest in defining parameterized 
+presets, and in automating the process of 
+posting sets to a website and social media.
+
+% For our next case study, we 
+% developed a library of 
+% image transformation livelits, 
+% motivated by a desire to automate
+
+% The livelits we have considered so far generate simple values.
+
+Motivated by this interview, 
+we prototyped a collection of image transformation livelits.
+One of these, \li{\$something} is demonstrated in Fig.~\ref{fig:img-transformation},
+within a function, \li{classic_look}, that serveas as a ``preset'' that is 
+mapped over a list of images (loaded by URL) at the bottom of the figure. 
+% We show two for space reasons.
+
+...\todo{finalize?}{}
+This livelit takes 
 a URL to an image as a parameter, and contains two splices of type \li{Int},
@@
 In this example, we have filled those splices with a \li{\$slider}, but 
-as above we could enter any expression of type \li{Int}.
+as above we could enter any expression of type \li{Int}, e.g. 
+an arithmetic expression or a variable. 
 The livelit shows a live preview of the transformed image.
-The expansion generates the necessary calls to image processing functions, 
+The expansion generates the necessary calls to an image processing library, 
 not shown.
 
-The novelty in this example is that the livelit appears inside a function, 
-\li{classic}, which takes an image URL as input. At the bottom 
-of the figure, \li{classic} is mapped over a list (here, containing only 
-two elements for the sake of presentation). Consequently, there are now 
-two closures associated with the application of \li{\$img_filter} in the 
-body of \li{classic}. Rather than disabling live evaluation in situations like 
+Because the livelit appears inside a function applied (by \li{map}) twice, 
+there are now two closures associated with the livelit. 
+Rather than disabling live evaluation in situations like 
 this, Hazel instead allows the programmer to select between the closures when 
-the cursor is on the livelit expression (via a simple sidebar toggle, not shown, which is shared 
-with the general closure selection UI \cite{HazelnutLive}). 
+the cursor is on the livelit expression via a simple sidebar toggle. 
+This causes the view to change as shown in ...\todo{todo}{}
 The live feedback is based on the selected closure.
 This makes it easy to see how the image filter being designed here will affect a
-number of example images by quickly toggling between closures. The underlying expansion remains abstract, i.e. it refers to the image via the \li{url} variable.
+number of example images by quickly toggling between closures. 
+The underlying expansion remains abstract, i.e. it refers to the image via the \li{url} variable.
+
+We showed this and similar examples to the photographer we had interviewed. 
+They expressed substantial interest in this approach despite having only 
+limited programming experience (with Python). They made the fair point that it 
+would take substantial effort to match Lightroom's breadth of filters and effects, 
+but with sustained effort, they agreed that this approach could be more powerful than 
+Lightroom's point-and-click interface while retaining many of its benefits (specifically
+mentioning the graphical sliders). Although this was only a single interview,
+it is consistent with the body of evidence summarized in Sec.~\ref{sec:background}.
 

### 2020-11-21 pldi21/case-studies.tex
got sec 2 into 5 pages

@@
 Only livelits with no remaining parameters can be invoked, 
-so writing \li{\$uslider} in expression position will display as a ``missing livelit parameter'' error.%
-\footnote{\label{footnote:typing}In Hazel, erroneous expressions 
-are automatically placed inside holes so that they do not prevent other parts of the program from evaluating
-\cite{HazelnutLive}.}
+so writing \li{\$uslider} in expression position will display as a ``missing livelit parameter'' error. 
+(In Hazel, erroneous expressions 
+are automatically placed inside holes and do not prevent other parts of the program from evaluating
+\cite{HazelnutLive}.)
+% \footnote{\label{footnote:typing}}
 
@@
 If we stop after proto-closure collection,
-then no useful value will be available:
-the result will be \emph{indeterminate}, because \li{\$dataframe} has been replaced with a hole \cite{HazelnutLive}.
-For this reason, there is a second phase of closure collection, called \emph{closure resumption}, 
-where any livelit holes that appear
-in the collected livelit closures are \emph{resumed}, i.e. the expansion is generated
-(in this case, for the \li{\$dataframe} livelit) and evaluation resumes.
+no useful value will be available:
+\li{averages} will be \emph{indeterminate}, because \li{\$dataframe} has been replaced with a hole \cite{HazelnutLive}.
+For this reason, there is a second phase of closure collection, \emph{closure resumption}, 
+where any livelit holes 
+in the collected livelit closures are \emph{resumed}, i.e. the hole is filled with the expansion 
+ and evaluation resumes.
 % Livelit expansions do not contain livelits, so no subsequent expansion / resumption phases are necessary.
@@
 \subsubsection{Indeterminate Results}
-Even after resumption, some elements of the closure may remain indeterminate, e.g. due to holes that 
+Even after closure resumption, some elements of the closure may remain indeterminate, e.g. due to holes that 
 are not filled with livelits. 
@@
 \todo{revise this once we have final figure}{}
-Closure collection can produce multiple closures when, for example, 
-a livelit appears inside a function applied multiple times. 
-Our next case study, in the photography domain,
-demonstrates how Hazel handle this situation, and 
-the workflow it enables.
+Our next case study 
+considers the situation where multiple closures are collected, and 
+the workflows it enables.
 
@@
 they use the Lightroom application to 
-apply a set of adjustments and filters 
+apply a set of adjustments 
 across all photos in a collection before making 
@@
 multiple photos as they tweaked them, without having to 
-save and apply the preset repeatedly.
-They also wanted to be able to change the shared settings
-even after making certain individual adjustments.
-They also expressed interest in defining parameterized 
-presets, and in automating the process of 
-posting sets to a website and social media.
+save and reapply the preset.
+They also wanted to be able to change the applied preset
+even after making individual adjustments.
+Finally, they 
+expressed interest in parameterizing  
+presets, and in automating parts of the post-production process.
 
@@
 Motivated by this interview, 
-we prototyped a collection of image transformation livelits.
+we prototyped a collection of photo filter livelits.
 One of these, \li{\$something} is demonstrated in Fig.~\ref{fig:img-transformation},
@@
 In this example, we have filled those splices with a \li{\$slider}, but 
-as above we could enter any expression of type \li{Int}, e.g. 
-an arithmetic expression or a variable. 
+as above we could enter any expression of type \li{Int}, e.g. a variable. 
 The livelit shows a live preview of the transformed image.
-The expansion generates the necessary calls to an image processing library, 
+The expansion generates calls to a browser image processing library, 
 not shown.
@@
 We showed this and similar examples to the photographer we had interviewed. 
-They expressed substantial interest in this approach despite having only 
+They expressed enthusiasm for this approach despite having only 
 limited programming experience (with Python). They made the fair point that it 
-would take substantial effort to match Lightroom's breadth of filters and effects, 
-but with sustained effort, they agreed that this approach could be more powerful than 
+would take substantial effort to match Lightroom's breadth of filters, 
+but stated that this approach could be more powerful than 
 Lightroom's point-and-click interface while retaining many of its benefits (specifically
-mentioning the graphical sliders). Although this was only a single interview,
+mentioning sliders). Although this was only a single interview,
 it is consistent with the body of evidence summarized in Sec.~\ref{sec:background}.

### 2020-11-21 pldi21/intro.tex
finished tweaking sec 2 and expanded filter case study substantially
got sec 2 into 5 pages

@@
 
-\subsection{Background}
+\subsection{Background}\label{sec:background}
 \begin{figure*}
@@
 as evidence for, and a showcase of, 
-the value of this class of mechanisms for integrating GUIs with code.
+the value of this class of mechanisms for integrating GUIs into code.
 
@@
     Providers define livelits in libraries, and 
-    clients invoke livelits by name. Livelit names, e.g. \li{\$color} and \li{\$slider} in Fig.~\ref{fig:color}(b),  are prefixed by \li{\$},    
+    clients invoke livelits by name. Livelit names, e.g. \li{\$color},  are prefixed by \li{\$},    
     to distinguish them from variables.
@@
 \llproperty{Hygienic Composition}
-  Prior systems have limited or no support for {entering sub-expressions within the GUI}, 
-  so they are useful mainly for generating expressions composed of constants,
-  e.g. color constants in Fig.~\ref{fig:color}(a).
-
-  Livelits have first-class support for sub-expressions, which we call \emph{splices} (after
+% Prior systems have limited or no support for {entering sub-expressions within the GUI}, 
+% so they are useful mainly for generating expressions composed of constants,
+% e.g. color constants in Fig.~\ref{fig:color}(a).
+%
+  Livelits support sub-expressions directly in the GUI, which we call \emph{splices} (after
    \citet{TLMs}).
-  Fig.~\ref{fig:color}(b) demonstrates splicing: the RGBA entry boxes in the GUI 
-  are splices, so the client can define a variable, \li{baseline},
-  to create a relationship between the red, green, and blue components.\todo{tweak figure}{}
-  The client can also use a slider to specify the alpha (A) component, 
-  without the provider of the \li{\$color} slider 
-  needing to anticipate this composition. 
-  Similarly, Fig.~\ref{fig:grading}(b)\todo{subfigure labels}{} demonstrates a 
-  compositional data table livelit where each entry is a splice.
+  Fig.~\ref{fig:color}(b) demonstrates splicing: the RGBA components  
+  are splice editors, so the client can define a variable, \li{baseline},
+  to relate the color components\todo{tweak figure}{} 
+  and use a slider livelit inline to specify the alpha component.
+  % Similarly, Fig.~\ref{fig:grading}(b)\todo{subfigure labels}{} demonstrates a 
+  %  data table livelit where each entry is a splice.
 
@@
   (2) \textbf{context independence}, i.e. that the livelit 
-  can be invoked in any program context without pre-condition or conflict.
+  can be invoked in any program context.%without pre-condition or conflict.
 
-\llproperty{Parameterization} It is possible to define parameterized families of livelits.
+\llproperty{Parameterization} Livelits can form parameterized families.
   For example, \li{\$slider} in Fig.~\ref{fig:color}(b) is parameterized by the slider's bounds.
-  Parameters operate like splices, differing only in that they can be partially applied in
+  Parameters operate like splices, differing in that they can be partially applied in
   livelit abbreviations. For example, Fig.~\ref{fig:color}(b)\todo{add slider}{} 
@@
   For example, in Fig.~\ref{fig:color}(b), 
-  the selected color can be displayed by evaluating the RGBA
+  displaying the selected color requires evaluating the RGBA
   component splices to numeric values.

### 2020-11-21 pldi21/livelit-definitions.tex
got sec 2 into 5 pages

@@
 \noindent
-Let us now look in more detail at a livelit definition. Fig.~\ref{fig:color-impl}
-outlines the \li{\$color} livelit from Fig.~\ref{fig:color}. (We omit certain 
-incidental implementation details and use unimplemented syntactic sugar, including
-Haskell-style \li{do} notation and simplified quasiquotation, for the sake of presentation.
-We say more about the implementation in Sec.~\ref{sec:implementation}.)
+We now take the perspective of a livelit provider.
+Fig.~\ref{fig:color-impl}
+defines \li{\$color} from Fig.~\ref{fig:color}(b). We omit certain 
+incidental details and use unimplemented syntactic sugar, including
+Haskell-style \li{do} notation and quasiquotation, for presentation.
 
-Line 3 of Fig.~\ref{fig:color-impl} specifies the livelit's name, \li{\$color}, and its \emph{expansion type},
-\li{Color}, which was discussed in Sec.~\ref{sec:livelit-expansion}.
-The remainder of the definition is the livelit's implementation.
+Livelit definitions are scoped and packaged like 
+any other definition. They consist of a declaration and an implementation. 
+Line 3 of Fig.~\ref{fig:color-impl} is \li{\$color}'s declaration,
+which defines its name and its \emph{expansion type}. 
+Livelit parameters would also appear here. 
+The declaration is part of the client interface, 
+as discussed in Sec.~\ref{sec:expansion-typing}-\ref{sec:parameterization}.
+
+\subsection{Livelit Implementations}
+The curly braces delimit the livelit's implementation. 
 Livelits are implemented using a variation on the functional model-view-update
@@
 interface between the livelit and the programming environment, discussed below, while retaining
-a pure functional programming style.
+a pure functional programming model.
 
-\subsection{Model}\label{sec:model}
+\subsubsection{Model}\label{sec:model}
 Line 6 of Fig.~\ref{fig:color-impl} specifies the livelit's \emph{model type},
@@
 
-\subsection{Action}
+\subsubsection{Action}
 Line 14 defines the \li{Action} type for the \li{\$color} livelit, which 
@@
 
-\subsection{View}
+\subsubsection{View}
 The \li{view} function computes the view given a model and access to the commands in 
@@
 
-\subsubsection{Live Evaluation}
+\paragraph{Live Evaluation}
 \label{sec:live-evaluation-def}
@@
 
-\subsubsection{Splice Rendering}
+\paragraph{Splice Rendering}
 The remainder of the \li{view} function is responsible for computing the view.
@@
 
-\subsection{Update}
+\subsubsection{Update}
 When the user triggers an event in a livelit view, it emits an \li{Action}.
@@
 
-\subsection{Expansion}
+\subsubsection{Expansion}
 \label{sec:expansion}

### 2020-11-21 pldi21/implementation.tex
fit sec 3 into 2 pages

--- 2020-09-06 implementation.tex
@@
 Prior work on incrementally updating UIs after code changes could be relevant to this effort \cite{burckhardt2013s}.
+
+\subsection{Integration into Imperative Languages}\todo{write me}{}
+\label{sec:imperative-langs}
+The Elm architecture has proven practical for sophisticated user interfaces, 
+e.g. the Sketch-n-Sketch system \cite{sns-uist}, 
+and its purity addresses means there are no issues 
+related to initialization, maintenance, and persistence of mutable state
+by the editor.
+Imperative languages can express the same architecture (indeed, a close cousin 
+is at the heart of the popular React framework for Javascript).
+The system would only need to specify that any transient mutable state will be reset between updates.

### 2020-11-21 pldi21/livelit-definitions.tex
fit sec 3 into 2 pages

@@
 \begin{figure}
-\begin{lstlisting}[xleftmargin=0.2cm,morekeywords={expansions,require}]
+\begin{lstlisting}[xleftmargin=0.2cm]
 type Color = (.r Int, .g Int, .b Int, .a Int)
-
 livelit $color at Color {
@@
                 .b SpliceRef, .a SpliceRef)
+
+  captures { }
 
@@
 
-  expansions require { }
   let expand : Model -> (Exp, List(SpliceRef)) = 
@@
 \noindent
-We now take the perspective of a livelit provider.
+We now take the perspective of a livelit provider.\todo{line num pass}{}
 Fig.~\ref{fig:color-impl}
@@
 Line 3 of Fig.~\ref{fig:color-impl} is \li{\$color}'s declaration,
-which defines its name and its \emph{expansion type}. 
+which defines its name and its {expansion type}. 
 Livelit parameters would also appear here. 
@@
 Livelits are implemented using a variation on the functional model-view-update
-framework popularized by the Elm programming language \cite{ElmArchitecture}. We add a fourth component,
-expansion generation. In addition, we add a simple monadic framework to provide the necessary  
-interface between the livelit and the programming environment, discussed below, while retaining
-a pure functional programming model.
+architecture popularized by the Elm language, upon which Hazel is based \cite{ElmArchitecture}. We add a fourth component,
+expansion generation. In addition, we add a simple monadic framework (\emph{a la} Haskell\todo{citation}{}) to provide the necessary  
+interface between the livelit and the programming environment, while retaining
+a pure functional programming model. (We discuss imperative languages in Sec.~\ref{sec:imperative-langs}.)
 
 \subsubsection{Model}\label{sec:model}
-Line 6 of Fig.~\ref{fig:color-impl} specifies the livelit's \emph{model type},
+The state of a livelit invocation is encoded in its model value. 
+Line 4 of Fig.~\ref{fig:color-impl} specifies the livelit's \emph{model type},
 here a labeled 4-tuple of \emph{splice references}, one for each of the four splices
 that appear in the GUI in Fig.~\ref{fig:color}.
-Models are persisted in the syntax tree of the program, so in practice,
- the system would check that the model type supports automatic serialization. 
- (Hazel is able to serialize all types because it is implemented as a 
- simple interpreter.)
+The model is how the livelit state is persisted in the syntax tree, so 
+ the system requires that the model type supports automatic serialization
+ (so, no functions in models).
 
-The initial value of the model is specified starting on Line 7, 
-which defines a command in the \li{UpdateMonad} that returns the 
-initial model value after generating four new splice references
+The \li{init} value on Line 9 determines the value of the model
+when the livelit is first invoked in the editor.
+It is a command in the \li{UpdateMonad}, discussed below, that returns the 
+initial model value after generating four new splices 
 using the \li{new_splice} command:
 \begin{lstlisting}[numbers=none]
-new_splice : (Typ, Option(Exp)) -> UpdateMonad(SpliceRef)
+new_splice : (Typ, Maybe(Exp)) 
+             -> UpdateMonad(SpliceRef)
 \end{lstlisting}
-This command creates a fresh splice
-of the given type and, optionally, with the given initial expression,
-and returns a unique splice reference, which uniquely and abstractly identifies that splice.
-Bolded types like \li{Typ}, \li{Exp}, \li{UpdateMonad}, and \li{SpliceRef}
-are provided by the standard library. The \li{Typ} and \li{Exp} types 
+This command creates a splice
+of the given type and, optionally, specifies its initial contents. 
+It returns a splice reference, which uniquely identifies that splice.
+Bolded types  
+are defined in the standard library. The \li{Typ} and \li{Exp} types 
 encode the syntax of Hazel's types and expressions.
-These two arguments are provided here in quasiquoted form, e.g. \li{`0`} \cite{bawden1999quasiquotation}. 
-If no initial expression is provided, the splice is initialized with 
-an empty hole.
+These two arguments are expressed here using quasiquotes, e.g. \li{`0`} \cite{bawden1999quasiquotation}. 
+% If no initial expression is provided, the splice starts as  
+% an empty hole.
 
-The system checks that the provided type and initial content is valid 
-assuming only the explicitly specified set of captured bindings on Line 4.
-Here, we do not need to capture any bindings because \li{Int} is a built in type
-and the initial content is closed. However, 
-for splices of user-defined types, those types need to be explicitly captured.
-Similarly, helper functions that are used in initial splices (and in the expansion,
-discussed below) must be explicitly captured. 
-This serves to ensure \emph{context independence} -- the livelit needs to make no 
-assumptions about the typing context at use sites.
-We use an explicit capture set, rather than implicitly capture of all bindings at the definition site, 
-both to prevent inadvertent capture 
-and to ensure that the capture set is part
-of the interface of the livelit definition, which simplifies matters related to 
-module export and determining paths to the captured bindings at use sites \cite{TLMs}.
-Formally, captured variables can be understood as parameters that have been 
-immediately partially applied.
+The system checks that the splice type and initial content are valid 
+assuming only the parameters and explicitly specified set of captured bindings on Line 7.
+Here, the capture set is empty because \li{Int} is a built in type
+and the initial content is closed. 
+This serves to ensure \emph{context independence}: the livelit needs to make no 
+assumptions about the typing context at invocation sites.
+We use an explicit capture set, rather than implicitly capturing all bindings 
+at the definition site, 
+to ensure that private bindings are not unintentionally  
+leaked to clients as detailed by \citet{TLMs}.
+% Formally, captured variables can be understood as parameters that have been 
+% immediately partially applied.
 
@@
 Line 14 defines the \li{Action} type for the \li{\$color} livelit, which 
-specifies a single user-initiated action: clicking on a color using the user 
-interface on the right side of Fig.~\ref{fig:color}. Actions are emitted
+specifies a single user-initiated action: clicking on a color using
+the right side of Fig.~\ref{fig:color}(b). Actions are emitted
 from event handlers (e.g. click handlers) defined in the computed \li{view}, 
-and actions are consumed by the \li{update} function, causing a change in the model. 
-Let us discuss each of these functions in turn.
+and actions are consumed by the \li{update} function, causing a change to the model. 
+Let us discuss each of these functions in turn in Sec.~\ref{sec:def-view}-\ref{sec:def-update}.
 
-\subsubsection{View}
+\subsubsection{View}\label{sec:def-view}
 The \li{view} function computes the view given a model and access to the commands in 
@@
 We elide the details of the particular user interface in Fig.~\ref{fig:color}
-to focus on two key mechanisms exposed by \li{ViewMonad}: live evaluation 
-and splice rendering.
+to focus on three key editor mechanisms exposed by \li{ViewMonad}: live evaluation,
+splice editors, and result rendering.
 
@@
 \label{sec:live-evaluation-def}
-In order to provide the client with live feedback about the dynamic implications 
-of the programmer's choices, 
-the view function can ask the system to evaluate a splice, or an encoded function of 
-multiple splices as long as the function is well-typed using only the capture set, 
-under the closure that the client has selected using one of the following
-commands:
-\begin{lstlisting}[numbers=none]
-eval_splice : SpliceRef -> ViewMonad(Option(Result))
-eval_fun    : (Exp, List(SpliceRef)) -> ViewMonad(Option(Result))
+As discussed in Sec.~\ref{sec:live-evaluation}, 
+the view can depend on the result of evaluating a splice 
+under the closure the client has selected.
+The interface between the view and the live evaluator 
+is via the 
+following command:
+% Actually, it can depend on an encoded function of 
+% multiple splices as long as the function is well-typed using only the capture set, 
+% under the closure that the client has selected using one of the following
+% commands:
+\begin{lstlisting}[numbers=none,xleftmargin=0cm]
+eval_splice : SpliceRef -> ViewMonad(Maybe(Result))
 \end{lstlisting}
 
+%eval_fun : (Exp, List(SpliceRef)) -> ViewMonad(Maybe(Result))
+
 The \li{None} case 
-arises when evaluation is not possible, either because no closures are available
-or because the free variables in the selected closure overlap with the free 
-variables in the provided expression. This occurs only when the user has selected
-a closure that appears under a binder in the evaluation result (e.g. because 
-the result is a lambda). 
-We discuss this further in Sec.~\ref{sec:calculus-closure-collection}.
-
+arises when evaluation is not possible, e.g. because no closures are collected.
+% because the free variables in the selected closure overlap with the free 
+% variables in the provided expression. This occurs only when the user has selected
+% a closure that appears under a binder in the evaluation result (e.g. because 
+% the result is a lambda). 
+% We discuss this further in Sec.~\ref{sec:calculus-closure-collection}.
 If a result is available, the \li{Result} type distinguishes two possibilities:
 \begin{lstlisting}[numbers=none]
-type Result = 
-| Val(Exp)
-| Indet(Exp)
+type Result = Val(Exp) | Indet(Exp)
 \end{lstlisting}
-The \li{Val} case arises when evaluation produces a true value, whereas the 
+The \li{Val} case arises when evaluation produces a value, whereas the 
 \li{Indet} case arises when evaluation results in an indeterminate expression,
-i.e. an expression that cannot be further evaluated due to one or more holes 
-in critical elimination positions \cite{HazelnutLive}.
+i.e. an expression that cannot be fully evaluated due to holes 
+in critical positions \cite{HazelnutLive}.
 
-In our example, the livelit determines a color to display to the client 
-if all four splices evaluate to integer literals. Otherwise, there is not 
-enough information to determine a color, and the livelit can decide how to 
-communicate this information to the client, e.g. with an "X" over the color 
-preview.
+Lines X-Y determine a color to display in the color preview 
+if all four splices evaluate to integers. Otherwise, there is not 
+enough information to determine a color. The livelit 
+communicates this indeterminacy with an ``X'' over the preview.
 
@@
 to handle a list of notes where certain notes are missing, i.e. holes, by 
-simply playing silence or some other default sound when it encounters them.
+simply playing silence or some default sound when it encounters them.
 This behavior is highly domain-specific, so each livelit provider must decide 
@@
 
-\paragraph{Splice Rendering}
-The remainder of the \li{view} function is responsible for computing the view.
-This is straightforward but for the question of splices. In Fig.~\ref{fig:color},
-we want the splices to render as expression editors, with all of Hazel's editor 
-services available to the user. In order to support this, the \li{view} function
-can request an editor for a given splice, with given dimensions:
+\paragraph{Splice Editors}
+The view includes an editing area for each splice. These editors must support all of 
+Hazel's editing services. To support this, the \li{view} function
+can request an editor with a given dimension (in character units) for a given splice:
 \begin{lstlisting}[numbers=none]
-editor : (SpliceRef, Dimensions) -> ViewMonad(Html(a))
+editor : (SpliceRef, Dim) -> ViewMonad(Html(a))
 \end{lstlisting}
 The result is an opaque \li{Html(a)} value that the remainder of the function 
-can place as needed. When the livelit is rendered, this part of the tree is 
-under the control of Hazel. The \li{Dimensions} parameter currently supports only a fixed
+can place where needed. When the livelit is rendered, this part of the tree is 
+under the control of Hazel. The \li{Dim} parameter currently supports only a fixed
 character width, with overflow causing scrolling, but in the future we plan to offer 
@@
 
-Another situation that can arise is when the UI needs to include not an editor but 
+\paragraph{Result Rendering}
+Some livelit views needs to include 
 a rendered evaluation result. For example, each of the cells in the \li{\$dataframe}
@@
 cell. Only the formula bar at the top is an editor. To support this, the \li{view}
-function can use the \li{result_view} commands, which mirror the \li{eval} commands:
+function can use the \li{result_view} command, which mirrors the \li{eval} command:
 \begin{lstlisting}[numbers=none]
-result_view_splice : (SpliceRef, Dimensions) -> ViewMonad(Html(a))
-result_view_fun    : (Exp, List(SpliceRef), Dimensions) -> ViewMonad(Html(a))
+result_view : 
+  (SpliceRef, Dim) -> ViewMonad(Maybe(Html(a)))
 \end{lstlisting}
 
-\subsubsection{Update}
+\subsubsection{Update}\label{sec:def-update}
 When the user triggers an event in a livelit view, it emits an \li{Action}.
 The system responds by calling the \li{update} function to determine how 
-this action should affect the model and the splices.
+this action should affect the model and, in some cases, the splices.
 
-We discussed in Sec.~\ref{sec:model} that the \li{UpdateMonad}
-allows new splices to be generated in response to user actions. 
-For example, the \li{\$dataframe} livelit uses this when new rows
-or columns are added.
+% We discussed in Sec.~\ref{sec:model} that the \li{UpdateMonad}
+% allows new splices to be generated in response to user actions. 
+% For example, the \li{\$dataframe} livelit uses this when new rows
+% or columns are added.
 
@@
 \begin{lstlisting}[numbers=none]
-  set_splice : (SpliceRef, Exp) -> UpdateMonad(Unit)
+  set_splice : (SpliceRef, Exp) -> UpdateMonad ()
 \end{lstlisting}
-The provided expression must satisfy the splice type and only make use of the capture set.
+The system checks that the expression must satisfy the splice type 
+and onlys make use of the capture set, as in Sec.~\ref{sec:model}.
 
@@
 Changes to splices can also cause the view to be recomputed, because the view might 
-need to evaluate the splices. The \li{update} function does not itself 
+evaluate the splices. The \li{update} function does not itself 
 have the ability to request evaluation, because the model should not depend directly  
-which closure the user has selected. Of course, the \li{view} might emit 
-result-dependent actions when appropriate.
+on which closure the user has selected. 
+%Of course, the \li{view} might emit 
+%result-dependent actions when appropriate.
 
@@
 The ultimate purpose of a livelit is to fill the hole where it appears by generating an expansion,
-i.e. a standard symbolic expression of the expansion type, here \li{Color}.
-The \li{expand} function, shown in Fig.~\ref{fig:color-impl}, is responsible for generating 
-an expansion given the current model.
+i.e. an expression of the expansion type, here \li{Color}.
+The \li{expand} function determines this expansion based on the model.
 
 The expansion can include splices, but the system does not make the contents of a splice 
-available directly. Instead, the \li{expand} must return a pair consisting of an encoded expression, of type 
-\li{Exp}, with a list of \li{SpliceRef}s that the expansion depends on. 
-We call the \li{Exp} the \emph{parameterized expansion}
- because it must take an argument for each listed \li{SpliceRef}s. 
- That argument will be the corresponding splice 
+available directly as a value of type \li{Exp}. Instead, \li{expand} returns a \emph{parameterized expansion}
+paired with a list of \li{SpliceRef}s. 
+The {parameterized expansion}
+ is an encoding (of type \li{Exp}) of a function 
+ that takes an argument for each listed \li{SpliceRef}s. 
+ That argument is of the corresponding splice 
 type, which was provided when the splice was initialized. 
- The return type of the parameterized expansion is the expansion type, here \li{Color}.
+ The return type of the parameterized expansion is the expansion type.
+ Here, the parameterized expansion for \li{\$color} encodes a function of type 
+ \li{Int -> Int -> Int -> Int -> Color}.
 
 This parameterization strategy makes it simple to enforce hygiene: the parameterized expansion 
-can depend only on the capture set, whereas the splices are entered by the client and so they can 
-depend only on the client site typing context. The use of standard function application ensures
+can depend only on the livelit parameters and capture set, whereas the splices are entered by the client and so they can 
+depend only on the client site typing context. The use of function application ensures
 that splices are capture avoiding. We consider this more formally in the next section.

### 2020-11-21 pldi21/intro.tex

@@
 \subsection{Background}\label{sec:background}
-\begin{figure*}
-  \includegraphics[width=29pc]{grade-cutoff-livelit.png}
+
+ \begin{figure*}[htb]
+   \begin{minipage}[b]{0.3\textwidth}
+     \begin{subfigure}[b]{0.3\textwidth}
+       \includegraphics[width=15pc]{graphite-color-palette.png}
+       \caption{}
+     \end{subfigure}
+     \begin{subfigure}[b]{0.3\textwidth}
+       \includegraphics[width=15pc]{slider-color-livelits.png}
+       \caption{}
+      \end{subfigure}
+  \end{minipage}
+
+  \begin{subfigure}[b]{0.7\textwidth}
+    \includegraphics[width=15pc]{grade-cutoff-livelit.png}
     \caption{Grading with Livelits}
@@
     \label{fig:color}
+   \end{subfigure}
+
+ \caption{
+   (a) This figure, adapted from the prior work of \citet{Graphite},
+   shows a Graphite palette associated with the \texttt{Color} class.
+   The client can specify the RGBA components only using literal numbers.
+   (b) A similar Hazel livelit for colors uses live splices for the RGBA components,
+   so the client can enter any Hazel expression.
+   In this case, the client has entered a variable, itself defined using a slider livelit, into the RGB
+   splices. The A splice is filled directly with a slider livelit.
+   The color livelit evaluates splices under its closure to display the
+   computed color.
+   The value of \li{default\_color} is determined by the livelit via a macro expansion step.}
+   \label{fig:color}
  \end{figure*}
-
-% \begin{figure*}
-%   \begin{center}
-%     \begin{subfigure}[t]{0.5\textwidth}
-%       \centering
-%        \vspace{-3.4cm}
-%        \includegraphics[width=15pc]{graphite-color-palette.png}
-%       \vspace{0.1cm}
-%       \caption{}
-%     \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
-%       \centering
-%       \includegraphics[width=15pc]{slider-color-livelits.png}
-%       \vspace{-11px}
-%       \caption{}
-%     \end{subfigure}
-%   \end{center}
-%   \caption{
-%   (a) This figure, adapted from the prior work of \citet{Graphite},
-%   shows a Graphite palette associated with the \texttt{Color} class.
-%   The client can specify the RGBA components only using literal numbers.
-%   (b) A similar Hazel livelit for colors uses live splices for the RGBA components,
-%   so the client can enter any Hazel expression.
-%   In this case, the client has entered a variable, itself defined using a slider livelit, into the RGB
-%   splices. The A splice is filled directly with a slider livelit.
-%   The color livelit evaluates splices under its closure to display the
-%   computed color.
-%   The value of \li{default\_color} is determined by the livelit via a macro expansion step.}
-%   \label{fig:color}
-% \end{figure*}
 

### 2020-11-21 pldi21/case-studies.tex
fix sec 4 at page 10

@@
 
-\subsubsection{Closure Collection} The subtlety is that 
+\subsubsection{Closure Collection}\label{sec:closure-collection-example} The subtlety is that 
 evaluation in Hazel is defined for closed expressions as usual,

### 2020-11-21 pldi21/livelit-calculus.tex
got sec 2 into 5 pages
fit sec 3 into 2 pages
fix sec 4 at page 10

--- 2020-11-18 popl21/livelit-calculus.tex
@@
     \[
-    \arraycolsep=4pt\begin{array}{llcl}
+    \arraycolsep=3pt\begin{array}{rlcl}
         \mathsf{Typ} & \tau & ::= &
-                                    \tnum ~\vert~
+                                    % \tnum ~\vert~
                                     \tarr{\tau_1}{\tau_2} ~\vert~
@@
                                     \trec{t}{\tau}\\
-        \mathsf{UExp} & \hat{e} & ::= & \hnum{n} ~\vert~
+        \mathsf{UExp} & \hat{e} & ::= & 
                                  x ~\vert~
@@
                                  \hap{\hat e_1}{\hat e_2} ~\vert~
-                                 \cdots ~\vert~
+                                 ... ~\vert~
                                 %  \hpair{\hat e_1}{\hat e_2} ~\vert~
@@
                                  \haplivelit{u}{a}{\dtxt{model}}{\splices}\\
-        \mathsf{EExp} & e & ::= & \hnum{n} ~\vert~ x ~\vert~ \hlam{x}{e} ~\vert~ \hap{e_1}{e_2} ~\vert~ \cdots ~\vert~ \hehole{u}\\
-        \mathsf{IExp} & d & ::= & \hnum{n} ~\vert~ x ~\vert~ \hlam{x}{d} ~\vert~ \hap{d_1}{d_2} ~\vert~ \cdots ~\vert~ \dehole{u}{\sigma}\\
+        \mathsf{EExp} & e & ::= & x ~\vert~ \hlam{x}{e} ~\vert~ \hap{e_1}{e_2} ~\vert~ ... ~\vert~ \hehole{u}\\
+        \mathsf{IExp} & d & ::= & x ~\vert~ \hlam{x}{d} ~\vert~ \hap{d_1}{d_2} ~\vert~ ... ~\vert~ \dehole{u}{\sigma}\\
         \mathsf{Splice} & \psi & ::= & \hsplice{\hat e}{\tau}
@@
     \caption{Syntax of types, $\tau$, unexpanded expressions, $\hat{e}$, expanded expressions, $e$, and internal expressions, $d$.
-    Metavariable $x$ ranges over variables, $u$ over hole names, and $\livelitname{a}$ over livelit names.
+    Here, $x$ ranges over variables, $u$ over hole names, and $\livelitname{a}$ over livelit names.
     We write $\splat{\psi_i}$ for a finite sequence of $n \geq 0$ splices,
     and $\sigma$ for finite substitutions of $n \geq 0$ internal expressions for variables, $[d_1/x_1, \cdots, d_n/x_n]$.
-    We elide standard expression forms
+    We elide standard forms
     related to product, sum, and recursive types.
@@
 
-In order to precisely capture the semantics of livelits
+In order to communicate the semantics of livelits
 independently
-of the specifics of the Hazel programming environment and GUI
-framework, we will now specify a simply typed \emph{livelit calculus}.
+of the specifics of the Hazel environment and the web platform, 
+we now specify a simply typed \emph{livelit calculus}.
 
@@
 \emph{external expressions}, $e$, before being \emph{elaborated}
-to internal expressions, $d$, for evaluation. All three stages are governed
-by the same types, $\tau$. We include numbers, partial functions, products, sums, and recursive
+to \emph{internal expressions}, $d$, for evaluation. All three sorts are classified
+by the same types, $\tau$. We include partial functions, products, sums, and recursive
 types, all in their standard form \cite{pfpl}, but this specific type structure is not critical.
 Any language expressive
-enough to encode its own syntax (here as values of a recursive sum type)
-would be a suitable basis. We begin with a terse
+enough to encode its own abstract syntax 
+would be a suitable basis for a livelit calculus.
+
+We begin in Sec.~\ref{sec:external-and-internal-lang} with a terse
 overview of the external and internal languages,
-which are adapted straightforwardly from Hazelnut Live \cite{HazelnutLive} and therefore
-serve as background material, in Sec.~\ref{sec:external-and-internal-lang}.
-Only unexpanded expressions contain
-livelits, so they are the main focus of the remaining sections,
-which cover livelit expansion (Sec.~\ref{sec:calculus-expansion}) and 
+which are adapted from Hazelnut Live \cite{HazelnutLive}.
+We then detail livelit expansion (Sec.~\ref{sec:calculus-expansion}) and 
 liveness via closure collection (Sec.~\ref{sec:calculus-closure-collection}). 
@@
 a typed lambda calculus that assigns static and dynamic meaning to programs with holes,
-notated externally $\hehole{u}$ where $u$ is a \emph{hole name} \cite{HazelnutLive}.
+notated $\hehole{u}$ where $u$ is a \emph{hole name} \cite{HazelnutLive}.
 We omit non-empty holes (which internalize type inconsistencies \cite{Hazelnut}) and type holes
-(which operate like the unknown type from gradual type theory \cite{Siek06a,Hazelnut})
-because these mechanisms are orthogonal to livelits.
+(which operate like the unknown type from gradual type theory \cite{Siek06a,Hazelnut}).
+These mechanisms are orthogonal to livelits and are included in our implementation.
 
-The external language is governed by a typing judgement of the standard form, $\hasType{\Gamma}{e}{\tau}$,
+External expressions, e, are governed by a typing judgement of the form, $\hasType{\Gamma}{e}{\tau}$,
 where the typing context, $\Gamma$,
@@
 unique in the external language, they can be duplicated during evaluation of internal expressions.
-Consequently, we need a context to ensure that each closure associated with hole $u$, written $\dehole{u}{\sigma}$,
-has the same type and can be filled
-with expressions valid under the same context.
+The hole context ensure that each closure associated with hole $u$, written $\dehole{u}{\sigma}$,
+shares a type and can be filled
+with expressions valid under a shared context.
 
-External expressions, $e$, are given dynamic meaning by typed elaboration to internal expressions, $d$,
-according to the typed elaboration judgement, $\elabs{\Gamma}{e}{d}{\tau}{\Delta}$.
-The main purpose of this elaboration step is to initialize the substitution $\sigma$ on each hole closure,
+\newcommand{\tnat}{\mathsf{nat}}
+
+External expressions are given dynamic meaning by typed elaboration to internal expressions, $d$,
+according to the judgement $\elabs{\Gamma}{e}{d}{\tau}{\Delta}$.
+The main purpose of elaboration is to initialize the substitution $\sigma$ on each hole closure,
 which operates to capture
@@
 substitution that maps each variable in $\Gamma$ to itself, because no substitutions have yet occurred. For example,
-\[ \elabs{ }{\helet{x}{\hnum{5}}{\hehole{u}} + \hnum{6}}{\helet{x}{\hnum{5}}{\dehole{u}{[x/x]} + \hnum{6}}}{\tnum}{u :: \tnum[x : \tnum]} \]
+\[ \elabs{ }{(\hlam{x}{\hehole{u}})~\hnum{5}}{(\hlam{x}{\dehole{u}{[x/x]}})~{\hnum{5}}}{\tnat}{u :: \tnum[x : \tnat]} \]
+
 During evaluation, $\evalsTo{d}{d'}$, the closure's substitution accumulates the substitutions that occur. For example,
-the internal expression above evalues as follows:
+the internal expression above evaluates as follows:
 \[
-  \evalsTo{\helet{x}{\hnum{5}}{\dehole{u}{[x/x]} + \hnum{6}}}{
-      \dehole{u}{[\hnum{5}/x]} + \hnum{6}
+  \evalsTo{(\hlam{x}{\dehole{u}{[x/x]}})~{\hnum{5}}}{
+      \dehole{u}{[\hnum{5}/x]}
   }
 \]
-We will use hole closures to support live programming with livelits in Sec.~\ref{sec:calculus-closure-collection}.
+% We will use hole closures to support live programming with livelits in Sec.~\ref{sec:calculus-closure-collection}.
 
-All of these judgements are adapted directly from those of Hazelnut Live,
-differing only in that we use a simpler declarative formulation rather than an algorithmic (bidirectional)
-formulation for the sake of simplicity and generality. Rather than restating the rules, we will simply state the governing
-metatheorems and refer the reader to the prior work for the full detalis \cite{HazelnutLive}.
+All of these judgements are adapted directly from Hazelnut Live,
+differing only in that we use a declarative rather than an algorithmic (bidirectional)
+formulation for simplicity. Rather than restating the rules, we simply state the key governing
+metatheorems and refer the reader to the prior work and our Agda mechanization (Sec.~\ref{sec:agda}) for the full details \cite{HazelnutLive}.
 
-First, we have that an elaboration always exists and it preserves typing.
+First, we have that elaboration preserves typing.
 \begin{theorem}[Typed Elaboration]
@@
 
-Next, we have that evaluation of an expression with holes leads to a final (i.e. irreducible) result of the same type (this is a corollary
-of the type safety properties established in the prior work.)
+Next, we have that evaluation of a closed expression with holes produces a final (i.e. irreducible) preserves typing.
 \begin{theorem}[Preservation]
@@
 
-The novelty of the livelit calculus lies entirely in its handling of unexpanded expressions,
+The novelty of the livelit calculus is entirely in its handling of unexpanded expressions,
 $\hat e$, which are given meaning by typed expansion to external expressions,
@@
 defined in Fig.~\ref{fig:expansion}.
-Unexpanded expressions mirror external expressions but for the presence of livelits, discussed below.
-The rules for standard
-forms like variables, functions and function application, shown in Fig.~\ref{fig:expansion},
-are straightforward.
+Unexpanded expressions mirror external expressions but for the presence of livelit invocations.
+The rules for the mirrored 
+forms like variables and functions, both shown in Fig.~\ref{fig:expansion},
+are simple.
 
-\subsubsection{Livelit Contexts}
+\subsubsection{Livelit Contexts}\label{sec:livelit-contexts}
 
@@
 and $\dtxt{expand}$ is the expansion function, which generates an expansion given a model.
-We omit logic related to view computations and actions, which are tied to a particular 
-UI framework.
+We omit the logic related to view computations and actions, which are tied to a particular 
+UI framework and have only indirect semantic significance.
 
-For a livelit context to be well-formed, we require that each livelit definition be well-formed.
-This in turn requires that the expansion function be hole-free and have the correct type.
+% For a livelit context to be well-formed, we require that each livelit definition be well-formed.
+% This in turn requires that the expansion function be hole-free and have the correct type.
 \begin{definition}[Livelit Context Well-Formedness]
@@
 the encoding judgement $\encodeExp{e}{d}$ and the other by the decoding judgement $\decodeExp{d}{e}$.
-Any isomorphism is sufficient, so we leave the details as a matter of implementation.
-The simplest approach would be to define $\expType$ as a recursive sum type,
-with one arm for each form of external expression (cf. \cite{TSLs} for an example).
+Any scheme is sufficient, so we leave it as a matter of implementation.
+The simplest approach is to define $\expType$ as a recursive sum type,
+with one arm for each form of external expression (cf. \cite{TSLs}).
 
 For simplicity, we assume that the livelit context is provided \emph{a priori} and therefore
-that functions like $\dtxt{expand}$ are already closed and fully elaborated.
+that the expansion function is already closed and fully elaborated.
 In practice, the livelit context would be controlled by a definition form in the language
-that allows the expansion function to itself use livelits.
+that allows the definition to itself invoke livelits.
 This would require a staging mechanism,
-because we need to execute expansion functions in prior definitions to be able to
+because we need to evaluate expansion functions in prior definitions to be able to
 expand subsequent definitions.
@@
 to evaluate ``up to'' each definition before proceeding \cite{HazelnutLive}.
-This choice is orthogonal to the mechanisms of interest in this section.
+% This choice is orthogonal to the mechanisms of interest in this section.
 
@@
             \expands{\Phi}{\Gamma}{\hlam{x}{\hat e}}{\hlam{x}{e}}{\tarr{\taut{in}}{\taut{out}}}
-        }
+        }~~~~~~~~\cdots
 
-        \inferrule[EAp]{
-            \expands{\Phi}{\Gamma}{\hat e_1}{e_1}{\tarr{\taut{in}}{\taut{out}}}\\\\
-            \expands{\Phi}{\Gamma}{\hat e_2}{e_2}{\taut{in}}
-        }{
-            \expands{\Phi}{\Gamma}{\hap{\hat e_1}{\hat e_2}}{\hap{e_1}{e_2}}{\taut{out}}
-        }
-
-        \cdots
-
+        % \inferrule[EAp]{
+        %     \expands{\Phi}{\Gamma}{\hat e_1}{e_1}{\tarr{\taut{in}}{\taut{out}}}\\\\
+        %     \expands{\Phi}{\Gamma}{\hat e_2}{e_2}{\taut{in}}
+        % }{
+        %     \expands{\Phi}{\Gamma}{\hap{\hat e_1}{\hat e_2}}{\hap{e_1}{e_2}}{\taut{out}}
+        % }
         \inferrule[EApLivelit]{
@@
     \end{figure}
-Unexpanded expressions are unique in that they include livelits:
+Unexpanded expressions are unique in that they include livelit invocations:
  \[\haplivelit{u}{a}{\dtxt{model}}{\splices}\]
@@
 The current state of the livelit is determined by the current model value, $\dtxt{model}$,
-together with a collection of splices, $\splices$. Each splice $\psi_i$
+together with the splice list, $\splices$. Each splice $\psi_i$
 is of the form $\hsplice{\hat e_i}{\tau_i}$, where $\hat e_i$ is the spliced expression
 itself (unexpanded, so it may contain other livelits) and $\tau_i$ is the type of that splice,
-as determined when the livelit definition first requested the splice (discussed in Sec.~\ref{sec:livelit-definitions}).
+as determined when the livelit definition first requested the splice (discussed in Sec.~\ref{sec:model}).
 
 Rule \rulename{EApLivelit} performs livelit expansion. Its premises, in order, operate as follows:
-\begin{enumerate}
+\begin{enumerate}[itemsep=3px,leftmargin=*]
     \item \textbf{Lookup.} The first premise looks up the livelit name in the livelit context.
@@
     \item \textbf{Expansion.} The third premise applies the expansion function, $\dtxt{expand}$, to the model value, $\dtxt{model}$,
-    producing the encoded expansion, $\dtxt{encoded}$, which, by the definitions given above, is of type $\mathsf{Exp}$.
-    \item \textbf{Decoding.} The fourth premise decodes $\dtxt{encoded}$, producing the external expression it encodes, $\etxt{pexpansion}$.
-    \item \textbf{Expansion Validation.} We call $\etxt{expansion}$ the \emph{parameterized expansion} because, according to the fifth premise,
-    it must be a function that returns a value of the expansion type, $\taut{expand}$, when applied (in curried fashion, though this is not critical)
+    producing the encoded parameterized expansion, $\dtxt{encoded}$, which, by the definitions given above, is of type $\mathsf{Exp}$.
+    \item \textbf{Decoding.} The fourth premise decodes $\dtxt{encoded}$, producing the \emph{parameterized expansion}, $\etxt{pexpansion}$.
+    \item \textbf{Expansion Validation.} The fifth premise checks that {parameterized expansion} is a function that returns a value of the expansion type, $\taut{expand}$, when applied (in curried fashion, though this is not critical)
     to the splices, whose types, $\splat{\tau_i}$, are explicitly recorded in the livelit.
@@
     i.e. that it cannot depend on the particular bindings available in the call site typing context, $\Gamma$.
-    Context independence is one facet of what is colloquially known as \emph{hygiene} \cite{TLMs}.
-    In this simple formulation of the system, we maintain context independence simply by
+    % Context independence is one facet of what is colloquially known as \emph{hygiene} \cite{TLMs}.
+    In this simple formulation of the system, we maintain context independence by
     requiring that the parameterized expansion be
     closed. Consequently, any necessary helper functions used in the expansion must be provided explicitly by the client
-    via a splice. In Sec.~\ref{sec:livelit-definitions}, we discussed how the use of explicit capture sets, which can be understood as 
-    immediate partial application, can lower the cognitive burden of this strict context independence discipline.
-    Parameters formally operate in the same way as splices, and partial application is a well-understood technique, so we omit these details from the calculus.
-    \item \textbf{Splice Expansion.} Finally, the sixth premise inductively expands each of the spliced expressions in the same context as the livelit
-    expression itself.
+    via a splice. 
+    In Sec.~\ref{sec:livelit-definitions}, we discussed how the use of explicit capture sets eliminates this client burden.
+    For simplicity, we omit capture sets from the calculus, but see \cite{TLMs} for a formal treatment.
+    \item \textbf{Splice Expansion.} The sixth premise inductively expands each of the spliced expressions in the same context as the livelit
+    invocation itself.
 \end{enumerate}
@@
 By applying the splices as arguments, we maintain \emph{capture avoidance} -- splices cannot capture variables
-bound internally to the expansion. Capture avoidance is the other facet of \emph{hygiene} \cite{TLMs}.
+bound internally to the expansion because beta reduction performs capture-avoiding substitution. 
+% Capture avoidance is the other facet of \emph{hygiene} \cite{TLMs}.
 
@@
 \end{theorem}
-\begin{proof}
-    We proceed by rule induction on the assumption.
-    The cases involving the standard forms follow by straightforward induction.
-    The only interesting case is the \rulename{EApLivelit} case.
-    In this case, we have that $e = \hap{\etxt{pexpansion}}{\splat{e_i}}$.
-    By assumption, we have $\{ \expands{\Phi}{\Gamma}{\hat e_i}{e_i}{\tau_i} \}_{i < n}$.
-    By induction, we therefore have $\{ \hasType{\Gamma}{e_i}{\tau_i} \}_{i < n}$.
-    In addition, by assumption we have that $\hasType{}{\etxt{pexpansion}}{\tarr{\splat{\tau_i}}{\taut{expand}}}$.
-    By weakening, we therefore have $\hasType{\Gamma}{\etxt{pexpansion}}{\tarr{\splat{\tau_i}}{\taut{expand}}}$.
-    Finally, by iterated application of the typing rule for function application over the splices,
-    using the judgements just established,
-    we have $\hasType{\Gamma}{\hap{\etxt{pexpansion}}{\splat{e_i}}}{\taut{expand}}$ as desired.
-\end{proof}
+% \begin{proof}
+%     We proceed by rule induction on the assumption.
+%     The cases involving the standard forms follow by straightforward induction.
+%     The only interesting case is the \rulename{EApLivelit} case.
+%     In this case, we have that $e = \hap{\etxt{pexpansion}}{\splat{e_i}}$.
+%     By assumption, we have $\{ \expands{\Phi}{\Gamma}{\hat e_i}{e_i}{\tau_i} \}_{i < n}$.
+%     By induction, we therefore have $\{ \hasType{\Gamma}{e_i}{\tau_i} \}_{i < n}$.
+%     In addition, by assumption we have that $\hasType{}{\etxt{pexpansion}}{\tarr{\splat{\tau_i}}{\taut{expand}}}$.
+%     By weakening, we therefore have $\hasType{\Gamma}{\etxt{pexpansion}}{\tarr{\splat{\tau_i}}{\taut{expand}}}$.
+%     Finally, by iterated application of the typing rule for function application over the splices,
+%     using the judgements just established,
+%     we have $\hasType{\Gamma}{\hap{\etxt{pexpansion}}{\splat{e_i}}}{\taut{expand}}$ as desired.
+% \end{proof}
 
 When composed with the Typed Elaboration theorem and the Type Safety properties established for the internal
-language, we achieve end-to-end type safety for unexpanded expressions (every well-typed unexpanded
-expression expands to a well-typed external expression, which in turn expands to a well-typed internal
-expression, which in turn evaluates in a type safe manner.)
+language, we achieve end-to-end type safety: every well-typed unexpanded
+expression expands to a well-typed external expression, which in turn elaborates to a well-typed internal
+expression, which in turn evaluates in a type safe manner.
 
 \subsubsection{Agda Mechanization}
-We have mechanized the typed expansion mechanism and proven the Typed Expansion theorem
-using the Agda proof assistant, using the same techniques as were used in the Agda mechanization 
-of Hazelnut Live \cite{HazelnutLive}. This mechanization is available in the anonymous 
-supplemental material.
+\label{sec:agda}
+We have mechanized the external and internal language and 
+typed expansion and proven the aforementioned theorems
+using the Agda proof assistant, building on the Agda mechanization 
+of Hazelnut Live \cite{HazelnutLive}. This mechanization is available in the supplement.
 
@@
 to evaluate expressions under one of the closures associated with the livelit.
-This mechanism was introduced by example in Sec.~\ref{sec:image-transformation}.
-In this section, we will formalize the process of efficiently collecting closures
-for the livelits that appear in a program.
+This mechanism was introduced by example in Sec.~\ref{sec:closure-collection-example}.
+In this section, we will formalize the process of efficiently collecting closures.
 
-The key idea is that we generate an alternative expansion,
+\subsubsection{Proto-Environment Collection}
+We begin by generating an alternative expansion,
 called the \emph{cc-expansion},
-where each livelit expands to an empty hole applied to the splices. In other words,
-we leave a hole in place of the parameterized expansion, which we expand and record in the \emph{cc-context}, $\Omega$, on the side.
+where each livelit invocation expands to an empty hole applied to its splices. In other words,
+the hole is in place of the parameterized expansion, which we instead record in a \emph{cc-context}, $\Omega$, on the side.
 The key rule for the cc-expansion judgement, $\ccexpands{\Phi}{\Gamma}{\hat e}{e}{\tau}{\Omega}$, is:
@@
     }{\taut{expand}}\\
-    \elabs{ }{\etxt{pexpansion}}{\dtxt{pexpansion}}{\tarr{\splat{\tau_i}}{\taut{expand}}}{\Delta}
+    \elabs{ }{\etxt{pexpansion}}{\dtxt{pexpansion}}{\tarr{\splat{\tau_i}}{\taut{expand}}}{\Delta}\\
+    \Omega = \cup_{i<n} \Omega_i \cup \{\Omegaitem{u}{\dtxt{pexpansion}}\}
 }{
     \ccexpands{\Phi}{\Gamma}{\haplivelit{u}{a}{\dtxt{model}}{\splat{\hsplice{\hat e_i}{\tau_i}}}}{\hap{\hehole{u}}{\splat{e_i}}}{\taut{expand}}
-    {\cup_{i<n} \Omega_i \cup \{\Omegaitem{u}{\dtxt{pexpansion}}\}}
+    {\Omega}
 }
@@
 
-We then elaborate and evaluate the cc-expansion
-following Hazelnut Live's dynamic semantics for programs with holes, which were summarized
-in Sec.~\ref{sec:external-and-internal-lang}. The result will contain some number of hole closures
-for each livelit hole. We call these the proto-closures and their associated environments the proto-environments for that livelit hole, formally defined as follows.
-\begin{definition}[Proto-Environment Collection]
+We then elaborate and evaluate the cc-expansion. 
+The result will contain some number of hole closures
+for each livelit hole. We call these the proto-closures and their environments the proto-environments for that livelit hole.
+\begin{definition}[Proto-Closure Collection]
 If $\ccexpands{\Phi}{\cdot}{\hat e}{e}{\tau}{\Omega}$ and $\elabs{}{e}{d}{\tau}{\Delta}$
@@
 \end{definition}
-A proto-environment for a livelit hole might itself contain a proto-closure for another livelit hole.
-For example, in Fig.~\ref{fig:color}, the proto-environment for the \li{\$color} livelit contained the
-proto-closure for the \li{\$slider} livelit, via the \li{gray\_level} variable.
-If we use proto-environments for live evaluation, then the actual value of that variable
-would not be available (here, the color could not be displayed).
 
-Consequently, the second step of closure selection is to fill any livelit holes that appear in the proto-environments for other livelit holes.
+\subsubsection{Closure Resumption}
+A proto-environment for a livelit hole might itself contain a proto-closure for another livelit hole,
+which is problematic for the reasons detailed in Sec.~\ref{sec:closure-collection-example}.
+% For example, in Fig.~\ref{fig:color}, the proto-environment for the \li{\$color} livelit contained the
+% proto-closure for the \li{\$slider} livelit, via the \li{baseline} variable.
+% If we use proto-environments for live evaluation, then the actual value of that variable
+% would not be available (here, the color could not be displayed).
+Consequently, the second step of closure collection, called \emph{closure resumption}, is to fill any livelit holes that appear in the proto-environments for other livelit holes.
 We do so by filling them using the parameterized expansions gathered in $\Omega$ and then resuming
@@
 as a substitution, i.e. the delayed substitutions captured in the environment are realized.
-In this case, however, the parameterized expansion is guaranteed to be closed due to
+In this case, however, the parameterized expansion is necessarily closed due to
 the context independence discipline we maintain in Rule \rulename{EApLivelit},
-so hole filling acts simply as a syntactic replacement.
+so hole filling amounts to syntactic replacement.
 
 Formally, we begin by defining an operation $\fillof{\Omega}{\sigma}$ which acts on proto-environments
-to fill any livelit holes that appear.
+to fill the livelit holes.
+
 \begin{definition}[Livelit Hole Filling] ~
     \begin{enumerate}
-        \item $\fillof{\Omega}{[d_1/x_1, \ldots, d_n/x_n]} = [\fillof{\Omega}{d_1}/x_1, \ldots, \fillof{\Omega}{d_n}/x_n]$
+        \item $\fillof{\Omega}{[d_1/x_1, \ldots, d_n/x_n]} = $\\
+        $\hspace{10px}[\fillof{\Omega}{d_1}/x_1, \ldots, \fillof{\Omega}{d_n}/x_n]$
         \item $\fillof{\Omega}{d} = \instantiateB{\dtxt{pexpansion}}{u}_{\Omegaitem{u}{\dtxt{pexpansion}} \in \Omega}{d}$
@@
 blocking evaluation. We therefore define an operation $\resumeof{\sigma}$ that resumes evalution for all closed expressions in $\sigma$.
-(Open expressions can remain because some closures appear under binders in the final result, so a substitution has not yet been recorded
-for the bound variables.)
+(The only open expressions that might remain are the initial variables from the identity substitution generated by elaboration.
+Some closures appear under binders in the final result, so these variables will not have yet recorded a substitution.)
 \begin{definition}[Environment Resumption] ~
     \begin{enumerate}
-        \item $\resumeof{[d_1/x_1, \ldots, d_n/x_n]} = [\resumeof{d_1}/x_1, \ldots, \resumeof{d_n}/x_n]$
+        \item $\resumeof{[d_1/x_1, \ldots, d_n/x_n]} = $\\
+        $[\resumeof{d_1}/x_1, \ldots, \resumeof{d_n}/x_n]$
         \item $\resumeof{d} = d'$ if $\fvof{d} = \emptyset$ and $\evalsTo{d}{d'}$
@@
 
-The correctness of the mechanisms described in this section rest on the fact that evaluation commutes with hole filling.
-This only holds when the language is pure: evaluation order does not matter
-so the system is free to evaluate around holes initially,
-and then return to finish the job once they have been filled.
+The correctness of the mechanisms described in this section rest on the fact that evaluation commutes with hole filling
+in the pure setting.
+% This only holds when the language is pure: evaluation order does not matter
+% so the system is free to evaluate around holes initially,
+% and then return to finish the job once they have been filled.
 
@@
 
-In a language with side effects, one would instead need to modify evaluation of
-cc-expansions such that the full expansion of each livelit application is evaluated at the same time as the
-temporary expansion,
-with the result stored for subsequent hole filling. This would ensure that
-side effects happen in the same order and only once. The full details are beyond the scope of this work.
+In a language with side effects, one could evaluate 
+cc-expansions in an alternative mode where the full expansion of each livelit invocation is evaluated 
+in the order that corresponding livelit hole, and any expressions blocked on it are resumed immediately. 
+The results would need to be recorded in memory for use in the filling phase. This would ensure that
+side effects happen in the same order and only once. %A formal specification is left as future work.
+Alternatively, an imperative language might use a different mechanism to make
+environments available to livelits, e.g. by environment logging \emph{a la} Lamdu \cite{lamdu}.
+This is more limited in situations where a livelit is never evaluated, as sometimes 
+occurs during development of a livelit in a function which has yet to be applied, 
+but does not require cc-expansion. We leave further exploration of this design space 
+to future work.
 

### 2020-11-21 pldi21/case-studies.tex
Add basic_adjustments figure
updates

@@
   \end{center}
-  \caption{Case Study: Image Transformation. The image shown is determined based on the selected closure.}
+  \caption{Case Study II: Image Transformation. The image shown is determined based on the selected closure.}
   \label{fig:img-transformation}
@@
 we prototyped a collection of photo filter livelits.
-One of these, \li{\$something} is demonstrated in Fig.~\ref{fig:img-transformation},
-within a function, \li{classic_look}, that serveas as a ``preset'' that is 
-mapped over a list of images (loaded by URL) at the bottom of the figure. 
-% We show two for space reasons.
-
-...\todo{finalize?}{}
-This livelit takes 
-a URL to an image as a parameter, and contains two splices of type \li{Int},
+One of these, \li{\$basic_adjustments}, is demonstrated in Fig.~\ref{fig:img-transformation}. 
+This livelit contains two splices of type \li{Int},
 one to adjust the contrast, and the other to adjust the brightness.
@@
 The livelit shows a live preview of the transformed image.
-The expansion generates calls to a browser image processing library, 
+The expansion generates calls to a browser image processing framework, 
 not shown.
+
+This livelit is used within a function, \li{classic_look}, that creates a ``preset'' filter. 
+This function is mapped over a list of images (loaded by URL) at the bottom of the figure. 
+% We show two for space reasons.
+The provided URL is passed into to livelit as a parameter.
 
@@
 this, Hazel instead allows the programmer to select between the closures when 
-the cursor is on the livelit expression via a simple sidebar toggle. 
-This causes the view to change as shown in ...\todo{todo}{}
-The live feedback is based on the selected closure.
+the cursor is on the livelit expression via a simple sidebar toggle,
+shown in the middle of Fig.~\ref{fig:img-transformation}.
+This causes the view to change as shown. 
 This makes it easy to see how the image filter being designed here will affect a

### 2020-11-21 pldi21/discussion.tex
updates

\section{Discussion and Conclusion}\label{sec:discussion}\label{sec:conclusion}
\begin{quote}
  %
  \textit{
  %
  The arithmetical symbols are written diagrams and the geometrical figures are graphic formulas.
  %
  }

  \vspace{3pt}

  \hfill{}--- David Hilbert~\cite{hilbert1902mathematical}
  \end{quote}

  \noindent
  Diagrams have played a pivotal role in mathematical thought since antiquity,
  indeed predating symbolic mathematics \cite{cajori1993history}. 
  Popular computing and creative tooling, too, has embraced visual representation and direct manipulation 
  interfaces for decades.
  Programming, however, has remained stubbornly mired in textual user interfaces. 
  % Certainly, one must acknowledge that textual and symbolic notation
  % is an indispensable tool for abstract thought. 
  % Indeed, it is now widely recognized that variables and functions are at the 
  % very foundation of computing. 
  Our hope with this paper is to demonstrate that principled, mathematically structured
  programming is not only compatible with live graphical user interfaces, but that the 
  combination of these two holds promise for the future of programming.

  This paper's contributions are 
  in advancing the expressive power of the mechanism in several directions, 
  most notably in terms of compositionality and liveness. These 
  technical contributions are a complement to the empirical findings by \citet{Graphite} and others.
  That said, the two case studies we considered in this paper were motivated by 
  real world problems. 
  As the implementation
  matures, we plan to introduce enthusiasts in a wide variety of problem domains
  to livelits and continue these empirical evaluations.

  % There remain a number of avenues for future work. First, we have only just started
  % to apply the livelits mechanism in various application domains.  
  % started .

%   In particular, we would like to better understand the conceptual and practical 
%   difficulties that livelit providers face, with the aim of making livelit 
%   implementation an extremely low cost activity. Livelits could themselves be 
%   useful for this task, if we develop a GUI widget library with support for livelits.
% In addition, recent approaches to deriving type-specific structure editors automatically from pretty printing logic \cite{hempeltiny} could perhaps 
% be adapted to generate livelit implementations.

Mechanisms for deriving simple 
livelit definitions from type definitions, perhaps similar to Haskell's \li{deriving} directive
or the GEC toolkit \cite{DBLP:conf/afp/AchtenEPW04}, or from \li{to_string} functions \cite{DBLP:conf/vl/HempelC20}, 
may prove fruitful in the future.

  The livelits mechanism as described in this paper operates only on expressions,
  but livelits might be useful for generating other sorts of terms, such as types,
  patterns, and entire modules. Prior work on literal macros has explored this \cite{TLMs}.
  % Support for type and module splices, too, would likely be useful.
  % With these, better support for reflecting on the provided type might allow for 
  % the development of more general livelits, e.g. a variant of \li{\$datatype} 
  % with individually typed columns, rather than only floating point data.

  % The strict hygiene discipline that we maintain has, we believe, substantial 
  % benefits---programmers will inevitably encounter unfamiliar livelits, and 
  % the reasoning principles that we enforce are likely to help them ``reason around''
  % the situation. However, it may be useful in certain circumstances to carefully
  % relax these restrictions. For example, some livelits might benefit from the 
  % ability to introduce bindings into splices, which is currently prevented by 
  % the capture avoidance discipline. Perhaps the 
  % editor could highlight the unusual situation.

  Another direction for future work has to do with pushing edits from computed results
  back into livelits. For example, a slider expands to a number, which may 
  then flow through a computation. Bidirectional evaluation techniques may allow
  the user to edit a number in the result of a computation and see the necessary
  change to a slider in the program \cite{sns-pldi,sns-uist}.

  Programming and authoring have much in common. Documents often contain structured
  information, and programs are written to manipulate structured information.
  Another future direction for livelits is as the basis for a programmable authoring 
  system, where the non-symbolic elements on the page are revealed to be code after all,
  albeit code generated by a livelit invocation that presents a more natural editing experience. 
  Taking this further, a networked collection of these
  documents could form a powerful computational wiki.
  We present this paper as a foundation for such exploration.
  %  This would 
  % require addressing the difficult problem of supporting collaborative interactions 
  % involving edits to arbitrary user interfaces. 
  % We believe that these problems are surmountable.

% \begin{itemize}
%   \item pattern matching
%   \item type splices
%   \item module splices
%   \item explicitly make bindings available in splices
%   \item better UI for closure provenance / connect with control flow / call stack better
%   \item deriving livelits from type definitions
%   \item bidirectional evaluation ala SnS
%   \item integration with structure editing more cleanly
%   \item \dots
%   \item collaborative editing?
%   \item side effects?
%   \item full screening livelits (or compositions thereof) as a way to create end-user workflows
%   \item big vision: authoring environment based on livelits
% \end{itemize}

### 2020-11-21 pldi21/implementation.tex
updates

--- 2020-09-06 implementation.tex
@@
 Implementing livelits requires tight integration between a rich editor, 
-a type checker, and a live evaluator. The evaluator is used
-both for running the model-view-update-expand system that controls 
-each livelit and for evaluating splices. Consequently, 
-the evaluator must be capable of evaluating incomplete programs and gathering hole closures
-as specified by Hazelnut Live \cite{HazelnutLive}.
+a type checker, and a live evaluator 
+capable of evaluating incomplete programs and gathering hole closures.
 
 \subsection{Hazel}
-Hazelnut Live is the formal foundation for the Hazel programming environment, so 
-it was natural to choose the the Hazel programming environment for our primary implementation. 
- Hazel is a browser-based live functional programming environment 
-organized from the ground up around typed holes. Uniquely, every editor state in Hazel is 
+Hazelnut Live is the foundation of the Hazel programming environment, 
+and Hazel has support for all of the necessary mechanisms, so 
+it was natural to choose Hazel for our primary implementation.
+Hazel is implemented in OCaml and compiled to JavaScript using the \li{js_of_ocaml} compiler \cite{DBLP:conf/aplas/RadanneVB16}.
+% The core of Hazel is written in a pure functional style, so our implementation closely 
+% follows the formalism.  
+%  Hazel is a browser-based live functional programming environment 
+% organized from the ground up around typed holes, with full .
+
+The livelit definition mechanism described in Sec.~\ref{sec:livelit-definitions}
+is implemented in Hazel, albeit without some of the syntactic sugar in Fig.~\ref{fig:color-impl}.
+This, together with the fact that Hazel lacks a mature GUI widget libraries as of this writing, 
+makes complex examples tedious to implement within Hazel, so we also 
+added the ability to define livelits using JavaScript or OCaml. These are loaded when 
+Hazel is compiled. As Hazel evolves, we expect to need to define such ``primitive'' livelits less 
+frequently, using them mainly for livelits that would benefit from access to established JavaScript 
+or OCaml libraries.
+
+Uniquely, every editor state in Hazel is 
 semantically meaningful: it has a type, it can be evaluated, and it can be transformed 
 in a type-aware manner. This implies that livelits remain fully functional at all times, 
-even when the program is incomplete or erroneous.
-
-Hazel is implemented in OCaml and compiled to JavaScript using the \li{js_of_ocaml} compiler \cite{DBLP:conf/aplas/RadanneVB16}.
-The core of Hazel is written in a pure functional style, so our implementation is able to closely 
-follow the formalism. The Hazel language has been evolving in parallel with our work, 
-and has only recently become expressive enough for it to be feasible to implement livelits within Hazel itself.
-The examples in this paper were instead implemented in OCaml as modules satisfying the signature 
-implied by Fig.~\ref{fig:color-impl}. A functor then adapts these OCaml-based livelit implementations 
-to livelit definitions suitable for inclusion in the livelit context used in the semantics.
-
-Hazel achieves its ``gap-free'' liveness guarantee by automatically inserting explicit holes as necessary 
+even when the program is incomplete or erroneous. Hazel achieves this ``gap-free'' liveness guarantee by automatically inserting explicit holes as necessary 
 while the user edits 
 the program. Formally, Hazel is a type-aware structure editor \cite{Hazelnut}, rather than a text editor, 
-although the developers are aiming to maintain a text-like programming experience when editing 
-symbolic code (this effort is orthogonal to our own). 
-This means that it is conceptually straightforward to hide the model and splice data and instead display 
-the computed view. However, there are a number of non-trivial layout challenges that come up, which we discuss
-in Sec.~\ref{sec:layout} below.
+although the developers are aiming to maintain a text-like experience (this effort is orthogonal to our own).
+
+% The examples in this paper were instead implemented in OCaml as modules satisfying the signature 
+% implied by Fig.~\ref{fig:color-impl}. A functor then adapts these OCaml-based livelit implementations 
+% to livelit definitions suitable for inclusion in the livelit context used in the semantics.
+
+% This means that it is conceptually straightforward to hide the model and splice data and instead display 
+% the computed view. However, there are a number of non-trivial layout challenges that come up, which we discuss
+% in Sec.~\ref{sec:layout} below.
 
 \subsection{Text Editor Integration}
-Before delving into layout issues, let us note that livelits do not require the use of a structure editor. 
+Livelits do not require the use of a structure editor. 
 We have also developed a 
-proof-of-concept textual program editor, based on Sketch-n-Sketch \cite{sns-pldi,sns-uist}, with support for livelits.
+proof-of-concept implementation of livelits in a textual program editor, Sketch-n-Sketch \cite{sns-pldi,sns-uist},
+which recently added support for the necessary mechanisms \cite{DBLP:journals/pacmpl/LubinCOC20}.
+Livelit interaction causes the serialized model in the text buffer to be changed, which updates the view.
 This proof-of-concept is not at feature parity with 
-our main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84} 
+our main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84,interactive-visual-notation} 
 is sufficient to support livelits, albeit with some gaps in livelit UI availability when 
-there are critical errors.
+there are syntax errors.
 
@@
 Whether implemented in a structure editor or a text editor, livelits present 
-interesting layout challenges--it can be difficult to sensibly 
-interleave symbolic code with livelit views in a clean, readable manner.
-
-Hazel uses an optimizing pretty printer to determine layout, an approach 
-based on the work of \citet{DBLP:journals/pacmpl/Bernardy17}. This system relies 
+interesting layout challenges. 
+Hazel uses an optimizing pretty printer based on the work of \citet{DBLP:journals/pacmpl/Bernardy17} to determine layout. This system relies 
 fundamentally on character counts. Consequently, our implementation asks each 
 livelit to specify dimensions in terms of character counts rather than pixels.
-Livelits can be laid out in one of two ways: as inline livelits, like \li{\$slider},
-which must be exactly one character high and appear inline with the code,
-or as \emph{multi-line livelits}, which occupy up to the full width of the editor 
-and a specified number of code lines (we omitted these specifications in Fig.~\ref{fig:color-impl}). 
-The pretty printer lays out multi-line livelits
-on their own line, as seen in Fig.~\ref{fig:grading}.
+Livelits can be laid out either as inline livelits, like \li{\$slider},
+which are one character high and appear inline with the code,
+or as \emph{multi-line livelits}, which occupy up to the full width 
+and a specified number of lines. % We omitted these specifications in Fig.~\ref{fig:color-impl}. 
+% The pretty printer lays out multi-line livelits
+% on their own line, as seen in Fig.~\ref{fig:grading}.
 
-We are currently exploring a more flexible system whereby 
-multi-line livelits can be named and applied within a larger expression, but
-the GUI is not displayed until the next line, even if there are intervening characters.
-This is inspired by the \verb|~| placeholder for large textual literals in the 
-Wyvern programming language \cite{TSLs}. We are also considering a number of other layout 
-options, e.g. pop-up livelits, livelits pinned to sidebars, and livelits that are rendered  
+% We are currently exploring a more flexible system whereby 
+% multi-line livelits can be named and applied within a larger expression, but
+% the GUI is not displayed until the next line, even if there are intervening characters.
+% This is inspired by the \verb|~| placeholder for large textual literals in the 
+% Wyvern programming language \cite{TSLs}.
+
+One might also considering a number of other layout 
+options, e.g. inline-block literals \emph{a la} Wyvern \cite{TSLs}, pop-up livelits, livelits pinned to sidebars, and livelits that are rendered  
 on a separate canvas or document while still formally being located within an underlying functional program.
@@
 even being aware that their interactions are actually edits to an underlying typed
-functional program. More experienced users on the team could then make programmatic use of the entered
-data and parameters in a seamless manner. We are also excited about the possibility of end users
-getting curious enough to ``View Source'' in this setting, and in so doing 
-being exposed to typed functional programming concepts gradually.
+functional program. 
+% More experienced users on the team could then make programmatic use of the entered
+% data and parameters in a seamless manner. We are also excited about the possibility of end users
+% getting curious enough to ``View Source'' in this setting, and in so doing 
+% being exposed to typed functional programming concepts gradually.
 
-The layout of splices within livelits is also an interesting problem, and our current approach, where 
-the provider must specify a fixed size for each splice, is not particularly flexible. 
-A more principled system for GUI layout that integrates with program pretty 
-printing could be of substantial use in some settings. Formula bars, like the one in the \li{\$dataframe}
-livelit, are a useful design pattern to avoid splice layout issues, and it may be possible to provide 
-a generalized formula bar in the editor user interface that multiple livelits could share.
-While splices can in principle contain other livelits, it is difficult for small splice editors to 
-visually accomodate large livelit views in practice. 
-Fortunately, variables can be used to support composition via substitution 
-rather than nesting. A linter might suggest a refactoring to a variable in these scenarios.
+% The layout of splices within livelits is also an interesting problem, and our current approach, where 
+% the provider must specify a fixed size for each splice, is not particularly flexible. 
+% A more principled system for GUI layout that integrates with program pretty 
+% printing could be of substantial use in some settings. Formula bars, like the one in the \li{\$dataframe}
+% livelit, are a useful design pattern to avoid splice layout issues, and it may be possible to provide 
+% a generalized formula bar in the editor user interface that multiple livelits could share.
+% While splices can in principle contain other livelits, it is difficult for small splice editors to 
+% visually accomodate large livelit views in practice. 
+% Fortunately, variables can be used to support composition via substitution 
+% rather than nesting. A linter might suggest a refactoring to a variable in these scenarios.
 
-The Hazel pretty printer uses memoization and incremental diffs to maintain reasonable performance,
-and adding livelits to the system did not substantially interfere with the existing approach.
-For the relatively simple livelits that we have developed, performance has been adequate.
-However, more complex  livelits or livelits that update quickly 
-could easily cause problems with editor responsiveness.
-This could perhaps be resolved by performing livelit-related computations asynchronously. Care 
-would need to be taken to ensure that intervening edits do not cause synchronization issues.
+% The Hazel pretty printer uses memoization and incremental diffs to maintain reasonable performance,
+% and adding livelits to the system did not substantially interfere with the existing approach.
+% For the relatively simple livelits that we have developed, performance has been adequate.
+% However, more complex  livelits or livelits that update quickly 
+% could easily cause problems with editor responsiveness.
+% This could perhaps be resolved by performing livelit-related computations asynchronously. Care 
+% would need to be taken to ensure that intervening edits do not cause synchronization issues.
 
-Many livelit-related computations can be intelligently memoized, e.g. view computations and unchanged splice evaluations,
-especially with the edit awareness that Hazel provides. 
-We leave further optimizations along these lines to future work.
-Prior work on incrementally updating UIs after code changes could be relevant to this effort \cite{burckhardt2013s}.
+% Many livelit-related computations can be intelligently memoized, e.g. view computations and unchanged splice evaluations,
+% especially with the edit awareness that Hazel provides. 
+% We leave further optimizations along these lines to future work.
+% Prior work on incrementally updating UIs after code changes could be relevant to this effort \cite{burckhardt2013s}.
+
+\subsection{Integration into Imperative Languages}
+\label{sec:imperative-langs}
+Our focus in this paper was on pure languages.
+Side effects pose a challenge on two fronts.
+If they occur in a livelit implementation,
+then the state exists at edit-time. 
+The most viable approach would be to retain the model-view-update
+architecture and specify that any transient mutable state will be reset between updates.
+If side effects occur in livelit expansions, 
+then the closure collection mechanism requires more 
+care to avoid unsoundness, as described in Sec.~\ref{sec:calculus-closure-collection}.
+However, these problems are surmountable, and we look forward to 
+efforts to integrate livelits into imperative languages.
+
+% The Elm architecture has proven practical for sophisticated user interfaces, 
+% e.g. the Sketch-n-Sketch system \cite{sns-uist}, 
+% and its purity addresses means there are no issues 
+% related to initialization, maintenance, and persistence of mutable state
+% by the editor.
+% Imperative languages can express the same architecture (indeed, a close cousin 
+% is at the heart of the popular React framework for Javascript).

### 2020-11-21 pldi21/intro.tex
Updates from Overleaf
updates

--- 2020-11-21 pldi21/intro.tex
@@
 \subsection{Contributions}
-
+\label{sec:contributions}
 We turn our attention in this paper to several fundamental technical

### 2020-11-21 pldi21/related-work.tex
updates

--- 2020-09-06 related-work.tex
@@
 \section{Related Work}\label{sec:related-work}
-As detailed in Sec.~\ref{sec:intro},
-the \emph{Graphite} system introduced the idea of filling 
-typed holes using a type-specific user interface \cite{Graphite}.
-This prior work engaged in substantial qualitative evaluation, 
+As detailed in Sec.~\ref{sec:background},
+the Graphite system developed the idea of filling 
+typed holes using a type-specific user interface, and it was the starting 
+point for our work \cite{Graphite}.
+Subsequent work on \textbf{mage} further explores this design space \cite{DBLP:conf/uist/KeryRHMWP20}.
+This prior work engaged in substantial qualitative evaluations, 
 which due to the fundamental similarities between the two systems 
@@
 left a number of core technical issues unresolved, 
-as summarized in Sec.~\ref{sec:intro}. 
+as summarized in Sec.~\ref{sec:contributions}. 
 Livelits resolve these issues in large part by bringing together 
@@
 
-This structural delimitation of different notations is reminiscent of work on 
-\emph{language boxes} \cite{DBLP:conf/sle/RenggliDN09}, which focused primarily on 
+Recent work on a interactive visual macro system in Racket 
+is quite similar in spirit to our work \cite{interactive-visual-syntax}.
+However, it supports only a limited form of splicing via a pop up text editor
+that does not support full compositionality as described here, where livelits 
+can appear within other livelits. Splicing is not hygienic, nor is there any 
+consideration of typing. Parameterization and partial application is also impossible, because invocation 
+is via an editor command rather than a language-integrated mechanism. Finally,
+and most critically, there is no support for liveness.
+
+Our structural delimitation of splices is reminiscent of work on 
+\emph{language boxes} \cite{DBLP:conf/sle/RenggliDN09}, which focused on 
 combining different notations, primarily textual, using structural delimiters inserted 
-explicitly by the user using a special-purpose editor.,
+explicitly using a special-purpose editor.
 
 There has been a long line of research on \emph{projectional editing}, where
-the user edits graphical representations (projections) of code constructs \cite{DBLP:journals/ile/MillerPMV94,read1996generating,DBLP:conf/sde/Reiss84,DBLP:conf/uist/KoM05}. 
+the user edits graphical representations (projections) of code constructs 
+\cite{DBLP:journals/ile/MillerPMV94,read1996generating,DBLP:conf/sde/Reiss84,DBLP:conf/uist/KoM05}. 
 Many of the oldest systems offer only a fixed set of projections and interaction techniques. 
@@
 In many ways, an unfamiliar livelit can be approached much like an unfamiliar function
-can be approached: by inspecting its type signature and reasoning compositionally 
-about arguments, i.e. splices.
+can be approached: by inspecting its declarations and reasoning compositionally 
+about arguments, i.e. splices and parameters.
 Finally, livelits integrate cleanly with live evaluation of programs with holes \cite{HazelnutLive},
-whereas other projectional editors either do not support evaluation of holes or do 
-not use that information to provide feedback to the user.
+whereas many projectional editors do not support live evaluation or do 
+not integrate liveness into projections.
+
+A number of notebook systems, including Mathematica, Jupyter \cite{Guo13}, and others, 
+support the insertion of simple widgets like sliders that respond live to changes.
 
 Related to projectional editors are ``syntax-recognizing editors'' \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84} and 
-a great variety of systems that generate visualizations from code \cite{DBLP:conf/chi/Lerner20,koschke2003software,urquiza2004survey}. Livelits differ 
+a variety of systems that generate visualizations from code \cite{DBLP:conf/chi/Lerner20,koschke2003software,urquiza2004survey}. Livelits differ 
 from these systems in directionality: visualization systems are given code and generate
 a visualization, whereas livelits are asked to create code where there would otherwise
-be a hole based on interactions that the programmer has with a visual interface.
+be a hole.
 
@@
 
-The Vital programming environment for Haskell \cite{hanna2002interactive} supports type-specific stylesheets
+The Vital programming environment for Haskell \cite{hanna2002interactive} 
+supports type-specific stylesheets
 that can be used to create custom user interfaces for both editing and displaying
@@
 Haskell code. Results are computed in a live manner, though there is no support
-for evaluating incomplete programs. The visualizations themselves also cannot provide
+for evaluating incomplete programs. The visualizations themselves cannot provide
 live feedback. Moreover, the system does not enforce any hygiene or typing principles:

### 2020-11-21 pldi21/intro.tex

--- 2020-11-21 pldi21/intro.tex
@@
 \subsection{Background}\label{sec:background}
+\definecolor{mygray}{rgb}{0.93, 0.93, 0.93}
+\definecolor{shadecolor}{named}{mygray}
 
- \begin{figure*}[htb]
-   \begin{minipage}[b]{0.3\textwidth}
-     \begin{subfigure}[b]{0.3\textwidth}
-       \includegraphics[width=15pc]{graphite-color-palette.png}
-       \caption{}
+\begin{figure*}
+  \begin{minipage}[t]{0.38\textwidth}
+    \begin{subfigure}[t]{\linewidth}
+    \begin{snugshade}
+      \vspace*{-2mm}
+      \caption{\textbf{Prior Work:} Graphite}
+      \vspace*{1mm}
+     \end{snugshade}
+      \vspace*{1.5mm}
+      \includegraphics[width=\linewidth]{graphite-color-palette.png}
+      \vspace*{2mm}
     \end{subfigure}
-     \begin{subfigure}[b]{0.3\textwidth}
-       \includegraphics[width=15pc]{slider-color-livelits.png}
-       \caption{}
+
+    \begin{subfigure}[t]{\linewidth}
+     \begin{snugshade}
+      \vspace*{-2mm}
+      \caption{\textbf{This Paper:} Livelits are live and compositional}
+      \vspace*{1mm}
+     \end{snugshade}
+      \vspace*{1.5mm}
+      \includegraphics[width=\linewidth]{slider-color-livelits.png}
     \end{subfigure}
   \end{minipage}
-
-  \begin{subfigure}[b]{0.7\textwidth}
-    \includegraphics[width=15pc]{grade-cutoff-livelit.png}
-    \caption{Grading with Livelits}
-    \label{fig:grading}
-    \label{fig:color}
+  \hspace{3mm}
+  \begin{subfigure}[t]{0.585\textwidth}
+  \begin{snugshade}
+   \vspace*{-2mm}
+    \caption{\textbf{Case Study}: Grading with Livelits}
+    \vspace*{1mm}
+     \end{snugshade}
+    \vspace*{1.5mm}
+    \includegraphics[width=\linewidth]{grade-cutoff-livelit.png}
   \end{subfigure}
-
    \caption{

### 2020-11-21 pldi21/intro.tex
Updates from Overleaf

@@
 \begin{figure*}
-  \begin{minipage}[t]{0.38\textwidth}
+  \begin{minipage}[t]{0.37\textwidth}
     \begin{subfigure}[t]{\linewidth}
@@
       \vspace*{-2mm}
-      \caption{\textbf{Prior Work:} Graphite}
+      \caption{\textbf{Prior Work:} Graphite \cite{Graphite}}
       \vspace*{1mm}
      \end{snugshade}
-      \vspace*{1.5mm}
-      \includegraphics[width=\linewidth]{graphite-color-palette.png}
-      \vspace*{2mm}
+      \vspace*{-1mm}
+      \includegraphics[width=\linewidth]{graphite-color-palette-green.png}
+      \vspace*{-5mm}
     \end{subfigure}
-
+    \hspace{8mm}
     \begin{subfigure}[t]{\linewidth}
@@
      \end{snugshade}
-      \vspace*{1.5mm}
+      \vspace*{-1mm}
       \includegraphics[width=\linewidth]{slider-color-livelits.png}
@@
   \hspace{3mm}
-  \begin{subfigure}[t]{0.585\textwidth}
+  \begin{subfigure}[t]{0.50\textwidth}
   \begin{snugshade}
@@
      \end{snugshade}
-    \vspace*{1.5mm}
+    \vspace*{-1mm}
     \includegraphics[width=\linewidth]{grade-cutoff-livelit.png}
   \end{subfigure}
-   \caption{
-   (a) This figure, adapted from the prior work of \citet{Graphite},
-   shows a Graphite palette associated with the \texttt{Color} class.
-   The client can specify the RGBA components only using literal numbers.
-   (b) A similar Hazel livelit for colors uses live splices for the RGBA components,
-   so the client can enter any Hazel expression.
-   In this case, the client has entered a variable, itself defined using a slider livelit, into the RGB
-   splices. The A splice is filled directly with a slider livelit.
-   The color livelit evaluates splices under its closure to display the
-   computed color.
-   The value of \li{default\_color} is determined by the livelit via a macro expansion step.}
+   \vspace*{-7mm}
+   \caption{}
+   \vspace*{-2mm}
    \label{fig:color}
@@
 \subsection{Contributions}
-\label{sec:contributions}
+
 We turn our attention in this paper to several fundamental technical

### 2020-11-21 pldi21/intro.tex
merge

--- 2020-11-21 pldi21/intro.tex
@@
   \end{subfigure}
-   \caption{
-   (a) This figure, adapted from the prior work of \citet{Graphite},
-   shows a Graphite palette associated with the \texttt{Color} class.
-   The client can specify the RGBA components only using literal numbers.
-   (b) A similar Hazel livelit for colors uses live splices for the RGBA components,
-   so the client can enter any Hazel expression.
-   In this case, the client has entered a variable, itself defined using a slider livelit, into the RGB
-   splices. The A splice is filled directly with a slider livelit.
-   The color livelit evaluates splices under its closure to display the
-   computed color.
-   The value of \li{default\_color} is determined by the livelit via a macro expansion step.}
+   \caption{}
    \label{fig:color}

### 2020-11-21 pldi21/case-studies.tex
merge
final pldi submission

@@
 Livelits are able to address this tension.
-Fig.~\ref{fig:grading}(c)\todo{sublabels} shows a Hazel program where 
+Fig.~\ref{fig:grading} shows a Hazel program where
 the instructor alternates between programmatic and direct manipulation in several situations.
@@
 For the sake of demonstration, we show a cell that has been filled using another livelit,
-\li{\$slider}, in combination with symbolic manipulation.\todo{do this?}{}
+\li{\$slider}, in combination with symbolic manipulation.
 The table itself displays not the expression itself but rather its value, just as in a spreadsheet.
@@
 expand to expressions.
-For example, the expansion of Fig.~\ref{fig:grading}(c) is:
+For example, the expansion of Fig.~\ref{fig:grading} is:
 
@@
 Hazel displays the information in the livelit declaration when the cursor is on the livelit's name,
-just as it displays typing information in other situations (not shown).\todo{cite HATRA}{}
+just as it displays typing information in other situations (not shown) \cite{hazeltutor}.
 
@@
 using function application notation
-as seen in Fig.~\ref{fig:grading}(c) or 
+as seen in Fig.~\ref{fig:grading} or
 using the pipelining (i.e. reverse function application) operators, \li{<|} and \li{|>},
@@
   \begin{center}
-    \begin{subfigure}[t]{0.5\textwidth}
-      \centering
-      \includegraphics[width=15pc]{img-filter-1.png}
-      \caption{}
-    \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
-      \centering
-      \includegraphics[width=15pc]{img-filter-2.png}
-      \caption{}
-    \end{subfigure}
+    \includegraphics[width=40pc]{img-filter.png}
+    % \begin{subfigure}[t]{0.5\textwidth}
+    %   \centering
+    %   \includegraphics[width=15pc]{img-filter-1.png}
+    %   \caption{}
+    % \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
+    %   \centering
+    %   \includegraphics[width=15pc]{img-filter-2.png}
+    %   \caption{}
+    % \end{subfigure}
   \end{center}
-  \caption{Case Study II: Image Transformation. The image shown is determined based on the selected closure.}
+  \caption{Case Study: Image Transformation. The image shown is determined based on the selected closure.}
   \label{fig:img-transformation}
@@
 The livelit provides an expected type for each splice when it is created.
-For example, the splices for the row and column keys in Fig.~\ref{fig:grading}(c)\todo{go through subfigures}{} have expected type \li{String}, 
+For example, the splices for the row and column keys in Fig.~\ref{fig:grading} have expected type \li{String},
 and the remaining cells have expected type \li{Float}.
-Hazel surfaces and uses the expected type when the cursor is in the splice.\todo{cite HATRA paper}{}
+Hazel surfaces and uses the expected type when the cursor is in the splice \cite{hazeltutor}.
 % If an expression of invalid type is entered, it will display in an error hole as usual,
@@
  requires a hygiene discipline that enforces \emph{capture avoidance}
-and \emph{context independence} \cite{TLMs}\todo{cite michael hygiene paper}.
+and \emph{context independence} \cite{TLMs,adamsHygiene}.
 
@@
 For example, there is one closure for \li{\$dataframe} in Fig.~\ref{fig:grading}.
-It contains the value of \li{q1max}\todo{what?}{} and the other variables in scope. 
+It contains the value of \li{q1\_max} and the other variables in scope.
 These values can be used to evaluate
-splices that use the corresponding variables, such as the cell selected in Fig.~\ref{fig:grading}(c).
+splices that use the corresponding variables, such as the cell selected in Fig.~\ref{fig:grading}.
 
@@
 
-\subsubsection{Case Study: Live Photo Filters}\label{sec:image-transformation}
-\todo{revise this once we have final figure}{}
+\subsubsection{Case Study: Live Image Filters}\label{sec:image-transformation}
 Our next case study
@@
 one to adjust the contrast, and the other to adjust the brightness.
-In this example, we have filled those splices with a \li{\$slider}, but 
+In this example, we have filled those splices with \li{\$percent}, but
 as above we could enter any expression of type \li{Int}, e.g. a variable.
@@
 there are now two closures associated with the livelit. 
-Rather than disabling live evaluation in situations like 
-this, Hazel instead allows the programmer to select between the closures when 
-the cursor is on the livelit expression via a simple sidebar toggle,
+Hazel allows the programmer to select between the closures when 
+the cursor is on the livelit expression via the sidebar toggle,
 shown in the middle of Fig.~\ref{fig:img-transformation}.
-This causes the view to change as shown. 
-This makes it easy to see how the image filter being designed here will affect a
+This allows the client to see how the filter being designed will affect a
 number of example images by quickly toggling between closures.

### 2020-11-21 pldi21/discussion.tex
final pldi submission

@@
 
-  % The strict hygiene discipline that we maintain has, we believe, substantial 
-  % benefits---programmers will inevitably encounter unfamiliar livelits, and 
-  % the reasoning principles that we enforce are likely to help them ``reason around''
-  % the situation. However, it may be useful in certain circumstances to carefully
-  % relax these restrictions. For example, some livelits might benefit from the 
-  % ability to introduce bindings into splices, which is currently prevented by 
-  % the capture avoidance discipline. Perhaps the 
-  % editor could highlight the unusual situation.
+   The strict hygiene discipline has, we believe, substantial 
+   benefits---programmers will inevitably encounter unfamiliar livelits, and 
+   the reasoning principles that we enforce are likely to help them ``reason around''
+   the situation. However, it may be useful in certain circumstances to 
+   relax these, with the editor alerting the user to the unusual situation.
 
@@
   documents could form a powerful computational wiki.
-  We present this paper as a foundation for such exploration.
+  We present this paper as a foundation for such explorations.
   %  This would 
@@
 
+\newpage
 % \begin{itemize}

### 2020-11-21 pldi21/implementation.tex
final pldi submission

@@
 This proof-of-concept is not at feature parity with 
-our main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84,interactive-visual-notation} 
-is sufficient to support livelits, albeit with some gaps in livelit UI availability when 
+our main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84,interactive-visual-syntax} 
+is sufficient to support livelits, albeit with some gaps in availability when 
 there are syntax errors.
@@
 care to avoid unsoundness, as described in Sec.~\ref{sec:calculus-closure-collection}.
-However, these problems are surmountable, and we look forward to 
-efforts to integrate livelits into imperative languages.
+% However, these problems are surmountable, and we look forward to 
+% efforts to integrate livelits into imperative languages.
 

### 2020-11-21 pldi21/intro.tex
final pldi submission

@@
       \caption{\textbf{Prior Work:} Graphite \cite{Graphite}}
+       \label{fig:graphite}
       \vspace*{1mm}
@@
       \caption{\textbf{This Paper:} Livelits are live and compositional}
+    \label{fig:color}
       \vspace*{1mm}
@@
   \end{minipage}
-  \hspace{3mm}
+  \hspace{10.5mm}
   \begin{subfigure}[t]{0.50\textwidth}
@@
     \caption{\textbf{Case Study}: Grading with Livelits}
+    \label{fig:grading}
     \vspace*{1mm}
@@
   \end{subfigure}
-   \vspace*{-7mm}
-   \caption{}
-   \vspace*{-2mm}
-   \label{fig:color}
+  \vspace{-4mm}
+   \caption{Introductory Examples}
+%   \vspace*{-2mm}
 \end{figure*}
@@
 The prior work most relevant to this paper is the {Graphite} system for Eclipse for Java,
-demonstrated in Fig.~\ref{fig:color}(a) \cite{Graphite}.
+demonstrated in Fig.~\ref{fig:graphite} \cite{Graphite}.
 Graphite allows a library provider to associate a GUI, called a \emph{palette}, with a type 
@@
 Java expression to fill the hole.
-Figure~\ref{fig:color}(a), adapted from this prior work, demonstrates a color palette invoked using Graphite.
+Figure~\ref{fig:graphite}, adapted from this prior work, demonstrates a color palette invoked using Graphite.
 When the user presses the \li{Enter} key, the Java expression \li{new Color(173, 173, 173, 85)} is inserted at the cursor and the palette disappears.
@@
 \textbf{mage} system for the Jupyter notebook environment \cite{DBLP:conf/uist/KeryRHMWP20}
-and the interactive visual syntax system for Racket \cite{interactive-visual-syntax}, behave fundamentally similarly.
-asdfasdfa
-fasdfasdf
-asdfasd\todo{placeholder for future text that might be added}
+behave fundamentally similarly.
+Projectional editing, the visual macro system in Racket \cite{interactive-visual-syntax}, and a number of other designs
+also confront this general problem of integrating GUIs and code.
 
+\newpage
 \citet{Graphite} evaluated Graphite by surveying 473 developers 
@@
 % general-purpose languages.}
-This and other prior work also collectively showcase a wide variety of use cases \cite{Graphite,DBLP:conf/uist/KeryRHMWP20,interactive-visual-syntax}\todo{more cites}, 
+This and other prior work also collectively showcase a wide variety of use cases \cite{Graphite,DBLP:conf/uist/KeryRHMWP20,interactive-visual-syntax}, 
 and the Graphite survey solicited dozens of additional use cases from participants,
@@
 the value of this class of mechanisms for integrating GUIs into code.
-
+\vspace{-1mm}
 \subsection{Contributions}
@@
 we introduce a system of \emph{live literals}, or \emph{livelits}, 
-demonstrated in Fig.~\ref{fig:color}(b). 
+demonstrated in Fig.~\ref{fig:color}. 
 Livelits are unique in achieving all of the following properties.
@@
    \citet{TLMs}).
-  Fig.~\ref{fig:color}(b) demonstrates splicing: the RGBA components  
+  Fig.~\ref{fig:color} demonstrates splicing: the RGBA components  
   are splice editors, so the client can define a variable, \li{baseline},
-  to relate the color components\todo{tweak figure}{} 
+  to relate the color components 
   and use a slider livelit inline to specify the alpha component.
@@
 \llproperty{Parameterization} Livelits can form parameterized families.
-  For example, \li{\$slider} in Fig.~\ref{fig:color}(b) is parameterized by the slider's bounds.
+  For example, \li{\$slider} in Fig.~\ref{fig:color} is parameterized by the slider's bounds.
   Parameters operate like splices, differing in that they can be partially applied in
-  livelit abbreviations. For example, Fig.~\ref{fig:color}(b)\todo{add slider}{} 
-  partially applies \li{\$slider} to \li{0} and \li{100} to define a \li{\$percentage} slider\todo{todo}{}.
+  livelit abbreviations. For example, Fig.~\ref{fig:color}
+  partially applies \li{\$slider} to \li{0} and \li{100} to define a \li{\$percent} slider.
   % \begin{lstlisting}[numbers=none]
@@
   to provide feedback related to run-time behavior.
-  For example, in Fig.~\ref{fig:color}(b), 
+  For example, in Fig.~\ref{fig:color}, 
   displaying the selected color requires evaluating the RGBA

### 2020-11-21 pldi21/livelit-calculus.tex
final pldi submission

@@
 or by using live
-programming mechanisms such as those available in Hazel
+programming mechanisms such as those in Hazel
 to evaluate ``up to'' each definition before proceeding \cite{HazelnutLive}.

### 2020-11-21 pldi21/livelit-definitions.tex
fix sec 4 at page 10
final pldi submission

@@
 \noindent
-We now take the perspective of a livelit provider.\todo{line num pass}{}
+We now take the perspective of a livelit provider.
 Fig.~\ref{fig:color-impl}
-defines \li{\$color} from Fig.~\ref{fig:color}(b). We omit certain 
+defines \li{\$color} from Fig.~\ref{fig:color}. We omit certain 
 incidental details and use unimplemented syntactic sugar, including
@@
 any other definition. They consist of a declaration and an implementation. 
-Line 3 of Fig.~\ref{fig:color-impl} is \li{\$color}'s declaration,
+Line 2 of Fig.~\ref{fig:color-impl} is \li{\$color}'s declaration,
 which defines its name and its {expansion type}. 
@@
 architecture popularized by the Elm language, upon which Hazel is based \cite{ElmArchitecture}. We add a fourth component,
-expansion generation. In addition, we add a simple monadic framework (\emph{a la} Haskell\todo{citation}{}) to provide the necessary  
+expansion generation. In addition, we add a simple monadic framework (\emph{a la} Haskell \cite{marlow2010haskell}) to provide the necessary  
 interface between the livelit and the programming environment, while retaining
 a pure functional programming model. (We discuss imperative languages in Sec.~\ref{sec:imperative-langs}.)
+Livelit implementations can themselves invoke other livelits (see Sec.~\ref{sec:livelit-contexts}).
 
@@
 The state of a livelit invocation is encoded in its model value. 
-Line 4 of Fig.~\ref{fig:color-impl} specifies the livelit's \emph{model type},
+Line 3 of Fig.~\ref{fig:color-impl} specifies the livelit's \emph{model type},
 here a labeled 4-tuple of \emph{splice references}, one for each of the four splices
@@
 
-The \li{init} value on Line 9 determines the value of the model
+The \li{init} value on Line 8 determines the value of the model
 when the livelit is first invoked in the editor.
@@
 The system checks that the splice type and initial content are valid 
-assuming only the parameters and explicitly specified set of captured bindings on Line 7.
+assuming only the parameters and explicitly specified set of captured bindings on Line 6.
 Here, the capture set is empty because \li{Int} is a built in type
@@
 \subsubsection{Action}
-Line 14 defines the \li{Action} type for the \li{\$color} livelit, which 
+Line 15 defines the \li{Action} type for the \li{\$color} livelit, which 
 specifies a single user-initiated action: clicking on a color using
-the right side of Fig.~\ref{fig:color}(b). Actions are emitted
+the right side of Fig.~\ref{fig:color}. Actions are emitted
 from event handlers (e.g. click handlers) defined in the computed \li{view}, 
@@
 
-Lines X-Y determine a color to display in the color preview 
+Lines 26-34 determine a color to display in the color preview 
 if all four splices evaluate to integers. Otherwise, there is not 
 enough information to determine a color. The livelit 
-communicates this indeterminacy with an ``X'' over the preview.
+indicates this indeterminacy with an ``X'' over the preview.
 

### 2020-11-21 pldi21/related-work.tex
merge
final pldi submission

@@
 left a number of core technical issues unresolved,
-as summarized in Sec.~\ref{sec:contributions}. 
-Livelits resolve these issues in large part by bringing together 
-seemingly disparate ideas from several other pieces of recent work.
+as summarized Sec.~\ref{sec:contributions}.
+Livelits resolve these in large part by bringing together  ideas from other recent work.
 
@@
 However, these systems operate in a purely 
-textual setting. There is no support for graphical notation or 
-interaction, nor live feedback, which are the primary contributions of this paper.
+textual setting and have no support for live feedback.
 The typed expansion judgement central to the typed livelit calculus in this paper 
@@
 to the corresponding judgements in the formal systems describing TLMs and TSLs,
-and the reasoning principles (e.g. hygiene properties, the focus on reasoning abstractly 
-about typing) was inspired directly by this work. However, the  
+and the reasoning principles are closely related. However, the  
 approach to hygiene we take is cleaner by its use of function application 
@@
 with retaining provenance information.
-This substantially complicates the design of splices in the prior work.
+This substantially complicates the design of splices in that work.
 
@@
 consideration of typing. Parameterization and partial application is also impossible, because invocation 
-is via an editor command rather than a language-integrated mechanism. Finally,
-and most critically, there is no support for liveness.
+is via an editor command rather than a syntactic mechanism. Finally,
+there is no support for liveness.
 
@@
 the user edits graphical representations (projections) of code constructs 
-\cite{DBLP:journals/ile/MillerPMV94,read1996generating,DBLP:conf/sde/Reiss84,DBLP:conf/uist/KoM05}. 
+\cite{DBLP:journals/ile/MillerPMV94,read1996generating,DBLP:conf/sde/Reiss84,DBLP:conf/uist/KoM05}. Livelits are a form of projectional editing. 
 Many of the oldest systems offer only a fixed set of projections and interaction techniques. 
@@
 and MPS \cite{voelter2011language} have made it easier to define new projectional editors.
-However, these systems generate entire editors, whereas livelit definitions can be 
-included granularly in a livelit context and support decentralized extensibility.
-Furthermore, livelit definitions are governed by a type and binding discipline that 
-serves to ensure that a programmer unfamiliar with the semantics of an unfamiliar livelit
-can still be sure that it is not operating outside of the bounds of the type system.
-In many ways, an unfamiliar livelit can be approached much like an unfamiliar function
-can be approached: by inspecting its declarations and reasoning compositionally 
-about arguments, i.e. splices and parameters.
-Finally, livelits integrate cleanly with live evaluation of programs with holes \cite{HazelnutLive},
-whereas many projectional editors do not support live evaluation or do 
-not integrate liveness into projections.
+However, these systems generate entire editors, whereas livelit definitions 
+are defined in libraries. 
+Furthermore, livelit definitions are governed by a type and binding discipline.
+% In many ways, an unfamiliar livelit can be approached much like an unfamiliar function
+% can be approached: by inspecting its declarations and reasoning compositionally 
+% about arguments, i.e. splices and parameters.
+Finally, livelits are uniquely live, building on Hazelnut Live \cite{HazelnutLive}.
 
-A number of notebook systems, including Mathematica, Jupyter \cite{Guo13}, and others, 
+A number of notebook systems, including Mathematica, Jupyter \cite{Guo13}, and others, do 
 support the insertion of simple widgets like sliders that respond live to changes.
+These systems inspired our approach but they are not compositional: only constant values can be constructed, as with Graphite.
 
-Related to projectional editors are ``syntax-recognizing editors'' \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84} and 
+Related to projectional editors are 
 a variety of systems that generate visualizations from code \cite{DBLP:conf/chi/Lerner20,koschke2003software,urquiza2004survey}. Livelits differ 
 from these systems in directionality: visualization systems are given code and generate
-a visualization, whereas livelits are asked to create code where there would otherwise
+a visualization, whereas livelits are generate code where there would otherwise
 be a hole.
@@
 Conal Elliott's work on tangible functional programming \cite{DBLP:conf/icfp/Elliott07} similarly explored a
-system that provides visual feedback about the result of a computation, 
-but the computation itself could be constructed using only a fixed structural syntax
-manifest as a series of connected windows.
+system allows editing a graphical representation of code in compositional ways, 
+but the editing representation itself is fixed as a series of connected windows.
+Only the visual representation is customizable.
 
@@
 live feedback. Moreover, the system does not enforce any hygiene or typing principles:
-the user is entirely responsible for syntactic and semantic correctness, and the 
-underlying semantics have not been formally defined.
+the user is entirely responsible for syntactic and semantic correctness.
 

### 2021-04-06 pldi21/abstract.tex
revisions up through sec 3

@@
 \begin{abstract}
-    Although text editing is powerful, 
+    Text editing is powerful, but 
     some types of expressions are more naturally represented and manipulated
     graphically.
-    % Examples include expressions that compute 
-    % colors,
-    % animation parameters,
-    % musical sequences,
-    % tabular data, 
-    % plot parameters,
-    % GUI widgets, 
-    % and various other domain-specific data structures.
+    Examples include expressions that compute 
+    colors,
+    music,
+    animations,
+    tabular data, 
+    plots,
+    GUI widgets, 
+    mathematical diagrams, 
+    and other domain-specific data structures.
     This paper introduces \emph{live literals}, or \emph{livelits}, 
     which allow
-    clients to fill holes of these types  
+    clients to fill holes of types like these   
     by directly manipulating a     
-    provider-defined GUI embedded persistently into otherwise 
-    symbolic code. 
-    Uniquely, livelits are compositional: the GUI can itself contain
-    spliced expressions, which have full editor support and can in turn contain other livelits. 
-    Splices are typed and the system 
-    ensures that livelits treat splices hygienically. 
-    Livelits are also uniquely live: 
-    they can offer {immediate feedback} about the run-time implications 
+    user-defined GUI embedded persistently into code. 
+    Uniquely, livelits are \emph{compositional}: a livelit GUI can itself embed 
+    spliced expressions, which are typed, lexically scoped, enjoy full editor support and can in turn embed 
+    other livelits. 
+    Livelits are also uniquely \emph{live}: 
+    a livelit can provide {continuous feedback} about the run-time implications 
     of the client's choices 
@@
     hole that the livelit is tasked with filling.
-    This paper introduces livelits with case studies that exercise these novel capabilities. 
-    We implement livelits in Hazel, 
+    We integrate livelits into Hazel, 
     a live programming environment able to 
-    typecheck and run programs with holes. 
-    We then define a typed lambda calculus that captures the essence of livelits as 
+    typecheck and run programs with holes, 
+    and describe several case studies that exercise these novel capabilities. 
+    We then define a minimal typed livelit calculus, which precisely 
+    specifies how livelits operate as 
     live graphical macros.     
-    The macro expansion process has been mechanized in Agda.
+    The metatheory of macro expansion has been mechanized in Agda.
 \end{abstract}

### 2021-04-06 pldi21/case-studies.tex
revisions up through sec 3

@@
 These case studies have been implemented
-in Hazel, a browser-based live programming environment for a dialect of Elm.
+in Hazel, a browser-based live programming environment for a 
+variant of Elm \cite{Elm}.
 Elm is an industrial pure typed functional language in the
@@
 The most common end-user application for this task is the spreadsheet, because
-it allows the instructor to record grades using a natural tabular interface,
-visualize this data in one of a finite number of plot styles,
-and perform basic computations,
+it allows an instructor to (1) record grades using a natural tabular interface,
+(2) visualize this data in one of a finite number of plot styles,
+and (3) perform basic computations,
 with results updated live.
 However, these affordances are limited.
-It is difficult to package up common operations
+It is difficult to package common operations
 into reusable libraries, interact with the data using domain-specific visualizations,
 and perform complex, unanticipated operations
-(e.g. preparing the data in an idiosyncratic format demanded by the university's grading system).
+(e.g. preparing the data in an idiosyncratic format demanded by the university registrar).
 
 General-purpose programming languages
-can handle these scenarios, but users
-lose the ability to receive live feedback and
-directly manipulate data and visualizations in the editor.
+\emph{can} handle these scenarios, but users
+lose the ability to 
+directly manipulate data and visualizations
+and receive live (in-editor) feedback.
 
@@
 that implements a tabular user interface. The formula bar
-allows the selected cell to be filled with an arbitrary Hazel expression.
-For the sake of demonstration, we show a cell that has been filled using another livelit,
-\li{\$slider}, in combination with symbolic manipulation.
-The table itself displays not the expression itself but rather its value, just as in a spreadsheet.
+allows the selected cell to be filled with an arbitrary Hazel expression,
+here a floating point expression that adds together individual problem scores,
+one of which is expressed using another livelit,
+\li{\$slider}.
+The table itself displays not the expression itself but rather its value, \li{93.}, just as in a spreadsheet.
 
 Next, the instructor computes \li{averages}
-for each student by applying \li{compute_averages}, a helper function
-defined in a library  (not shown) shared between multiple courses.
+for each student by applying \li{compute_weighted_averages}, a helper function
+defined in a library  (not shown) shared between courses.
 
-Next, the instructor wants to ``eyeball'' reasonable \li{cutoffs} between letter grades
+Next, the instructor wants to ``eyeball'' reasonable letter grade \li{cutoffs} 
 by directly manipulating a domain-specific livelit, \li{\$grade_cutoffs}, that provides draggable ``paddles''
-superimposed on a live visualization of the distribution of \li{averages}, which is provided as a livelit parameter.
-% The value of \li{cutoffs} is a labeled 4-tuple containing each cut-off.
+superimposed on a live visualization of the distribution of \li{averages}, 
+which is provided as a livelit parameter.
+The value of \li{cutoffs} is a labeled 4-tuple containing each cut-off.
 
@@
 \subsection{Livelit Expansion}\label{sec:livelit-expansion}
-Livelit invocations
-expand to expressions.
+Livelit invocations are expressions 
+that are given meaning by expansion.
 For example, the expansion of Fig.~\ref{fig:grading} is:
@@
    ("Bob", [61., 64., 98., 70., 85.]),
-   ("Ciri", [75., 81., 73., 82., 79.]),
+   ("Carol", [75., 81., 73., 82., 79.]),
    (* ... *) ]) in
@@
 \subsection{Compositionality}\label{sec:splicing-and-parameterization}
-Livelits are compositional: they can work with sub-expressions
+Livelit invocations are compositional: they can embed sub-expressions
 in the form of parameters and splices.
@@
 so writing \li{\$uslider} in expression position will display as a ``missing livelit parameter'' error.
-(In Hazel, erroneous expressions
+In Hazel, erroneous expressions
 are automatically placed inside holes and do not prevent other parts of the program from evaluating
-\cite{HazelnutLive}.)
+\cite{HazelnutLive}.
+A livelit invocation can also indicate that no 
+expansion is available with a custom error message, e.g. due to non-sensical bounds,
+which is also displayed to the user (not shown).
 % \footnote{\label{footnote:typing}}
@@
 \subsubsection{Splices}\label{sec:splices}
-Spliced expressions, or \emph{splices}, appear directly inside the livelit GUI.
+Spliced sub-expressions, or \emph{splices}, appear directly inside the livelit GUI.
 Splices can be filled with Hazel expressions of any form, including other livelit invocations.
@@
 Unlike parameters, the number of splices can change
-as the user interacts with the livelit, e.g. when changing the number of rows or columns in a \li{\$dataframe}.
+as the user interacts with the livelit, e.g. when adding or removing rows or columns in a \li{\$dataframe}.
 % The cell itself displays the live value of the spliced expression---
@@
 and the remaining cells have expected type \li{Float}.
-Hazel surfaces and uses the expected type when the cursor is in the splice \cite{hazeltutor}.
+Hazel displays and uses the expected type when the cursor is on the splice \cite{hazeltutor}.
 % If an expression of invalid type is entered, it will display in an error hole as usual,
@@
 
-\subsubsection{Hygienic Composition}\label{sec:hygiene}
+\subsubsection{Binding Discipline}\label{sec:hygiene}
 
 Ensuring that clients can reason about binding while leaving expansions
- invisible
- requires a hygiene discipline that enforces \emph{capture avoidance}
-and \emph{context independence} \cite{TLMs,adamsHygiene}.
+ abstract
+ requires enforcing two key properties: \emph{capture avoidance}
+and \emph{context independence} \cite{TLMs,adamsHygiene,DBLP:conf/popl/ClingerR91}.
 
 \textbf{Capture Avoidance.}
-Splices and parameters can appear anywhere in the expansion.
-This becomes problematic when
-they appear under a binder, e.g. in the body of a function or \li{let} binding.
-Na\"ively, this could cause inadvertent capture of the bound variable by a free variable
+Splices and parameters are passed into the expansion as function arguments, 
+so they cannot capture any bindings internal to the expansion. All bindings 
+are lexically scoped to the livelit invocation site.
+
+Consider an alternative where 
+splices could directly appear anywhere within the expansion, 
+e.g. in the body of a generated function or \li{let} binding.
+Na\"ively, this could cause inadvertent capture of the bound variables in the expansion by a free variable
 in the parameter or splice. For example, consider a livelit that generates an expansion
@@
   let len = strlen <splice1> in
-Some (<splice2> + len)
+  if len > 0 then Some (<splice2> + len) else None 
 \end{lstlisting}
@@
 
-To avoid this situation, parameters and splices are placed in the expansion
+To avoid this situation, parameters and splices must be placed in the expansion
 in a capture-avoiding manner: variables in splices
@@
 rather than bindings that are hidden inside the expansion.
-We discuss how this is implemented in Sec.~\ref{sec:expansion}.
+In other macro systems, e.g. that of \citet{TLMs}, 
+capture avoiding substitution is used. 
+% This allows splices to appear 
+% multiple times and also appear in branches that may or may not evaluate. 
+To simplify reasoning about live evaluation, discussed below, we take 
+an even more restrictive approach by passing splices in as function 
+arguments. In an eagerly evaluated language, this ensures that 
+splice evaluation is not conditional on computations invisible to the 
+client (such as the length check above), 
+and it also ensures that evaluation will not inadvertently 
+occur multiple times or not occur at all. 
+% We discuss how this is implemented in Sec.~\ref{sec:expansion}.
+
+The trade-off is that this prevents the expression of new 
+binding constructs and even control flow constructs as livelits.
+We take this as a strength, in that it simplifies 
+reasoning for client programmers. However, this is an independent 
+design decision: 
+more permissive designs that nevertheless 
+maintain the critical capture avoidance property are possible.
+
 % This is implemented by alpha-renaming bindings internal to the expansion as necessary.
@@
 These values can be used to evaluate
-splices that use the corresponding variables, such as the cell selected in Fig.~\ref{fig:grading}.
+splices that use these variables, such as the cell selected in Fig.~\ref{fig:grading}.
 

### 2021-04-06 pldi21/intro.tex
authors, as submitted
revisions up through sec 3

@@
 \section{Introduction}\label{sec:intro}
-Text-based program editors are flexible and expressive user interfaces
+Text-based program editors are flexible and powerful user interfaces, 
 so it is little wonder that they remain dominant decades after the teletype.
@@
 
-As a simple example, consider a record type
+Consider, as a simple example, a record type
 classifying RGBA-encoded colors. 
 It is possible to select a particular color by entering
-an expression of this type in a text editor, e.g. \li{\{ r: 255, g: 178, b: 45, a: 100 \}}. 
-The problem with this textual user interface for color selection is that 
-it offers no live feedback about which color has been selected 
-and limited editing affordances for tweaking the selected color.
+an expression of this type in a text editor, e.g. \li{\{ r: 57, g: 107, b: 57, a: 92 \}}. 
+The problem with this \emph{de facto} textual user interface for color selection is that 
+it offers no live feedback about which color has been expressed 
+and limited editing affordances for choosing a different color.
 Analagous critiques apply to strictly textual user interfaces for 
 countless other data structures,
-such as vector graphics,
+e.g. vector graphics,
 animation parameters,
@@
 GUI widgets and layouts,
-tabular data, 
+tabular numeric data, 
 plots,
@@
 biological neuron models, 
-mathematical diagrams, 
-and so on.
-
-% It is difficult for the programmer, or anyone subsequently reading or modifying the code, to know which color is represented
-% and to interactively tweak that color.
+and mathematical diagrams.
 
@@
 have largely eschewed general-purpose programming environments 
-in favor of more specialized graphical end-user applications, like %
+in favor of specialized graphical end-user applications, e.g. %
 image and video editors, music composition software, level design tools, 
-and bespoke GUIs written by students or lab technicians, 
-in large part because these applications 
-take seriously the need for domain-specific forms of live feedback, 
-graphical data representations, and 
+and bespoke GUIs written by students or lab technicians.
+This is in large part because these applications 
+take seriously the need for live feedback, domain-specific 
+non-textual data representations, and 
 direct manipulation affordances, 
-e.g. color palettes, visual timelines, interactive plots, and maps.
+e.g. color palettes, visual timelines, plots, and maps.
 
@@
 otherwise directly constructed game map, 
-or to define a function that computes portions of an 
-otherwise directly constructed vector graphic,
-or to transform a directly constructed musical sequence 
-by passing it through a series of symbolically defined functions.
-Moreover, it is difficult to add new affordances or to compose
+or to define functional combinators to compute portions of an 
+otherwise directly constructed musical composition. 
+Some applications include \emph{ad hoc} abstraction mechanisms for 
+the most common such use cases, e.g. named color swatches, 
+but users cannot themselves define new affordances, either programmatic or graphical, 
+nor compose
 affordances in ways that the application developer did not anticipate.
-Users cannot easily make even simple changes like replacing a numeric text box in a dialog with a slider,
-much less more ambitious changes like installing an alternative visual interface for expressing geospatial data queries 
-into a database frontend.
+For example, 
+users cannot make even simple changes like replacing a 
+numeric text box with a slider,
+much less more ambitious changes like importing an alternative 
+visual interface for expressing geospatial data queries 
+into a civic database front-end. 
+% In short, end users and programmers speak entirely different languages, 
+% leaving end users to rely helplessly on programmers, 
+% and programmers unable to bring their skills to bear 
+% when using end-user applications.
 
@@
 
-This paper aims to resolve this tension between
+This paper aims to bridge the gap between
 programmatic and direct manipulation user interfaces by designing 
 a programming environment that
-is able to surface GUIs when working with types for which
-they are useful, while retaining full support for symbolic program manipulation
+is able to surface a GUI when entering an expression of a type for which
+it is useful, while retaining full support for symbolic program manipulation
 and the abstraction and composition mechanisms
-available in modern general-purpose programming languages.
+available in modern general-purpose languages, both   
+in the spaces between these GUIs and compositionally within these GUIs.
 
@@
 
-Of course, we are not the first to integrate direct manipulation interfaces
-into symbolic programming environments.
+We are not the first to integrate GUIs 
+into programming.
 % Prior work on projectional editing
@@
 Graphite allows a library provider to associate a GUI, called a \emph{palette}, with a type 
-(via a Java class annotation).
+(via a Java class annotation), here \li{Color}.
 Wherever an expression of this type is needed,
-i.e. wherever there is a \emph{hole} of that type in the program
+i.e. wherever there is a \emph{hole} of this type 
 (as determined by Eclipse's online parser and typechecker),
 the environment offers the client the option, via the code completion menu,
-to interact with the palette.
-Once the interaction is finished, the palette generates a
-Java expression to fill the hole.
-Figure~\ref{fig:graphite}, adapted from this prior work, demonstrates a color palette invoked using Graphite.
-When the user presses the \li{Enter} key, the Java expression \li{new Color(173, 173, 173, 85)} is inserted at the cursor and the palette disappears.
-Several related systems, such as the  
-\textbf{mage} system for the Jupyter notebook environment \cite{DBLP:conf/uist/KeryRHMWP20}
-behave fundamentally similarly.
+to generate it using the palette.
+Once the user presses \li{Enter} to indicate that the 
+interaction is finished, the palette generates a
+Java expression to fill the hole, here \li{new Color(57, 107, 57, 235)}.
+Several other systems, such as the  
+\textbf{mage} system for Jupyter notebooks \cite{DBLP:conf/uist/KeryRHMWP20}
+behave analagously.
 Projectional editing, the visual macro system in Racket \cite{interactive-visual-syntax}, and a number of other designs
-also confront this general problem of integrating GUIs and code.
+also confront this general problem of integrating GUIs with symbolic code
+in broadly similar ways.
 
-\newpage
 \citet{Graphite} evaluated Graphite by surveying 473 developers 
@@
 as evidence for, and a showcase of, 
-the value of this class of mechanisms for integrating GUIs into code.
+the value of this broad class of mechanisms for integrating GUIs with code.
 \vspace{-1mm}
@@
 \label{sec:contributions}
-We turn our attention in this paper to several fundamental technical
-deficiencies that limit GUI providers and clients using these prior systems.
+We turn our attention in this paper to a number of fundamental technical
+deficiencies that limit both GUI providers and clients using these prior mechanisms.
 To address these, 
@@
 Livelits are unique in achieving all of the following properties.
-(We describe which subset of these properties 
+(Sec.~\ref{sec:related-work} describes which subset of these  
 are achieved by prior systems,
-including those just mentioned, 
-in Sec.~\ref{sec:related-work}.)
+including those just mentioned.)
 
@@
 \llproperty{Decentralized Extensibility}
-    Providers define livelits in libraries, and 
-    clients invoke livelits by name. Livelit names, e.g. \li{\$color},  are prefixed by \li{\$},    
+    Providers define livelits in libraries. 
+    Clients invoke livelits by name. Livelit names, e.g. \li{\$color}, 
+    are prefixed by \li{\$} (pronounced ``lit''),    
     to distinguish them from variables.
-    % We call this \emph{decentralized extensibility}.
+    We call this \emph{decentralized extensibility}
+    to distinguish it from systems that are not extensible or 
+    that can only be extended via editor extensions or editor generation.
     % Both mage and Racket's visual syntax system are similarly extensible, 
@@
     % so there is a tighter association between 
-    % Other systems, discussed in Sec.~\ref{sec:related-work}, 
-    % are either non-extensible, or cannot be extended by library providers 
-    % (only by, for example, separately installed editor extensions, or by 
-    % regenerating the editor using a language workbench.)
+    % as distinct from systems that 
+    % are either non-extensible or cannot be extended by library providers.
 
@@
   % to persist GUI state that is not included in the generated code.}
-  Livelits are persistent elements of the syntax tree. They operate as  
+  Livelit invocations are expressions, i.e. they  are persistent elements of the syntax tree. They operate as  
   graphical literals, rather than as the ephemeral code generation GUIs of Graphite and \textbf{mage}. 
@@
   where only the model needs to be persisted.
-  The dynamic meaning of a livelit is determined by macro expansion.
+  The dynamic meaning of a livelit is determined by a macro expansion step.
   % We chose the word ``literal'' rather than ``palette'' because,  persistence, livelits
@@
 
-\llproperty{Hygienic Composition}
+\llproperty{Compositionality}
 % Prior systems have limited or no support for {entering sub-expressions within the GUI}, 
@@
 %
-  Livelits support sub-expressions directly in the GUI, which we call \emph{splices} (after
+   Livelit GUIs can embed sub-expressions, which we call \emph{splices} (after
    \citet{TLMs}).
   Fig.~\ref{fig:color} demonstrates splicing: the RGBA components  
-  are splice editors, so the client can define a variable, \li{baseline},
-  to relate the color components 
+  are each splice editors, so the client can define a variable, \li{baseline},
+  to relate the color components (here, to explore greens by offsetting the green component past the baseline)
   and use a slider livelit inline to specify the alpha component.
@@
 
-  Crucially, composition is strictly 
-  governed by a hygiene discipline that ensures
-  (1) \textbf{capture avoidance}, i.e. that variables that the client uses in splices 
-  will not capture expansion-internal bindings; and 
+  Crucially, composition is  
+  governed by a binding discipline that ensures 
+  (1) \textbf{capture avoidance}, i.e. that variables that the client uses in splices, like \li{baseline},  
+  are lexically scoped to the livelit invocation site (so they cannot inadvertently capture expansion-internal bindings, which can therefore be left abstract); and 
   (2) \textbf{context independence}, i.e. that the livelit 
-  can be invoked in any program context.%without pre-condition or conflict.
+  can be invoked and operate in any lexical context (so the client need not worry about naming conflicts or manage hidden library dependencies).
+  This binding discipline is a form of ``macro hygiene'' \cite{TLMs, adamsHygiene, DBLP:conf/popl/ClingerR91}, 
+  though note that we take a particularly restrictive approach to hygiene compared to, for example, the Racket macro system \cite{DBLP:conf/popl/Flatt16}: 
+  new binding and control flow constructs intentionally cannot be expressed to allow clients to understand splices as   
+  function parameters (and indeed this is how they are internally handled).
 
-\llproperty{Parameterization} Livelits can form parameterized families.
+ %without pre-condition or conflict.
+
+\llproperty{Parameterization} Livelits can also take parameters directly, forming parameterized families.
   For example, \li{\$slider} in Fig.~\ref{fig:color} is parameterized by the slider's bounds.
@@
 
-\llproperty{Typing} Each livelit specifies the type of expansions it generates, 
-and parameters and splices also specify types, 
-so livelits are compatible with type-driven methodologies and tools.
-Together with the hygiene discipline, this allow clients to reason abstractly about expansions, i.e.  
-without inspecting the expansions or livelit implementations directly.
+\llproperty{Typing} Each livelit must specify an expansion type, 
+and parameters and splices must also specify types, 
+so livelits are compatible with type-driven development.
+Together with the binding discipline, 
+this allow clients to reason abstractly, i.e.  
+without inspecting the expansion or livelit implementation.
+% in the manner of a function application.
 
@@
   using a two-phased variant of the semantics for 
-  live programming with typed holes developed by \citet{HazelnutLive}.
-  We support live evaluation even for livelits 
-  that appear inside a function. Multiple function calls lead 
+  Hazelnut Live \cite{HazelnutLive}.
+  Live evaluation is supported even for livelits 
+  that appear inside a function: multiple function calls 
+  can lead 
   to multiple closures that the client 
-   can select between.
+   selects between.
 
-\paragraph{Outline.} We begin in Sec.~\ref{sec:case-studies} by introducing
+\paragraph{Outline.} Sec.~\ref{sec:case-studies} introduces
 livelits from the perspective of client programmers. 
-Our examples are organized into case studies 
-and are chosen to demonstrate the novel contributions of this paper.
-In Sec.~\ref{sec:livelit-definitions}, 
-we consider the livelit provider's perspective by introducing livelit
+% Our examples are organized into case studies 
+% that demonstrate the novel contributions of this paper.
+Sec.~\ref{sec:livelit-definitions} then 
+considers the livelit provider's perspective by introducing livelit
 definitions with a detailed example.
-In Sec.~\ref{sec:livelit-calculus}, we define the \emph{typed livelit calculus}. 
-We have mechanically specified the central mechanism, livelit expansion, and proven the associated metatheorems in Agda.
+Sec.~\ref{sec:livelit-calculus} defines the \emph{typed livelit calculus}. 
+We have mechanically specified the central mechanism, livelit expansion, 
+and proven the associated metatheorems in Agda.
 This calculus 
@@
 because we believe livelits can be integrated into a wide variety of programming systems. 
-In Sec.~\ref{sec:implementation}, we provide a more detailed account of our two implementations of livelits.
+Sec.~\ref{sec:implementation} provides a more detailed account of our implementation of livelits.
 Our primary implementation, used in the screenshots in the paper,
-is integrated into Hazel, a live programming environment designed 
+is integrated into Hazel, a live functional programming environment designed 
 around hole-driven development. 
-We have also prototyped livelits within a standard text editor. 
+We have also prototyped livelit editing within a standard text editor. 
 Additionally, we discuss factors that must be considered when integrating livelits into languages with side effects.
-In Sec.~\ref{sec:related-work}, we compare livelits to related work using the design properties outlined above 
+Sec.~\ref{sec:related-work} compares livelits to related work using the design properties outlined above 
 as a rubric.
-Finally, we conclude in Sec.~\ref{sec:discussion} after a discussion of present limitations and future work.
+We conclude in Sec.~\ref{sec:discussion} after a discussion of present limitations and several directions for future work.

### 2021-04-06 pldi21/livelit-definitions.tex
revisions up through sec 3

@@
 
-  captures { }
+  context { }
 
-  let init : UpdateMonad(Model) = do 
+  let init : UpdateCmd(Model) = do 
     r <- new_splice(`Int`, Some(`0`))
@@
 
-  let view : Model -> ViewMonad(Html(Action)) = 
+  let view : Model -> ViewCmd(Html(Action)) = 
     fun model -> do 
@@
         | _ -> 
-          (* indetermine color shown as X *)
+          (* indeterminate color shown as X *)
           None 
@@
   let update : 
-      Model -> Action -> UpdateMonad(Model) =
+      Model -> Action -> UpdateCmd(Model) =
     fun model (ClickOn c) -> do 
@@
 Line 2 of Fig.~\ref{fig:color-impl} is \li{\$color}'s declaration,
-which defines its name and its {expansion type}. 
-Livelit parameters would also appear here. 
+which defines its name and its {expansion type}, \li{Color}, defined on Line 1. 
+Livelit parameters would also appear here as shown in the declarations in Sec.~\ref{sec:expansion-typing}. 
 The declaration is part of the client interface, 
@@
 \subsection{Livelit Implementations}
-The curly braces delimit the livelit's implementation. 
+The curly braces delimit the livelit's implementation, which is not intended to be seen 
+by clients. 
+Every livelit implementation must include all of the definitions shown in Fig.~\ref{fig:color-impl}, 
+which we will discuss in turn below. 
 Livelits are implemented using a variation on the functional model-view-update
-architecture popularized by the Elm language, upon which Hazel is based \cite{ElmArchitecture}. We add a fourth component,
-expansion generation. In addition, we add a simple monadic framework (\emph{a la} Haskell \cite{marlow2010haskell}) to provide the necessary  
-interface between the livelit and the programming environment, while retaining
-a pure functional programming model. (We discuss imperative languages in Sec.~\ref{sec:imperative-langs}.)
-Livelit implementations can themselves invoke other livelits (see Sec.~\ref{sec:livelit-contexts}).
+architecture popularized by Elm \cite{ElmArchitecture}. We add a fourth component,
+expansion generation. In addition, we use a simple monadic framework (\emph{a la} Haskell \cite{marlow2010haskell}) to provide the necessary  
+interface between the livelit and the editor, while retaining
+a pure functional programming model (monadic commands, like \li{new_splice} discussed below, are executed by the editor).
 
 \subsubsection{Model}\label{sec:model}
-The state of a livelit invocation is encoded in its model value. 
-Line 3 of Fig.~\ref{fig:color-impl} specifies the livelit's \emph{model type},
+The state of a livelit's GUI is determined by its model value. 
+Line 3 of Fig.~\ref{fig:color-impl} specifies the corresponding \emph{model type},
 here a labeled 4-tuple of \emph{splice references}, one for each of the four splices
 that appear in the GUI in Fig.~\ref{fig:color}.
-The model is how the livelit state is persisted in the syntax tree, so 
+The model is how the GUI state is persisted in the syntax tree, so 
  the system requires that the model type supports automatic serialization
- (so, no functions in models).
+ (so no functions can appear in models).
 
@@
 when the livelit is first invoked in the editor.
-It is a command in the \li{UpdateMonad}, discussed below, that returns the 
+It is a command in the \li{UpdateCmd} monad, further discussed in Sec.~\ref{sec:def-update}, that returns the 
 initial model value after generating four new splices 
@@
   new_splice : (Typ, Maybe(Exp)) 
-             -> UpdateMonad(SpliceRef)
+               -> UpdateCmd(SpliceRef)
 \end{lstlisting}
 This command creates a splice
-of the given type and, optionally, specifies its initial contents. 
+of the given type and, optionally, its initial contents. 
 It returns a splice reference, which uniquely identifies that splice.
-Bolded types  
+In this section, bolded types  
 are defined in the standard library. The \li{Typ} and \li{Exp} types 
-encode the syntax of Hazel's types and expressions.
-These two arguments are expressed here using quasiquotes, e.g. \li{`0`} \cite{bawden1999quasiquotation}. 
+encode the syntax of Hazel's types and expressions 
+and we use quasiquotation, e.g. \li{`0`}, as the introduction forms \cite{bawden1999quasiquotation}. 
 % If no initial expression is provided, the splice starts as  
@@
 
-The system checks that the splice type and initial content are valid 
-assuming only the parameters and explicitly specified set of captured bindings on Line 6.
-Here, the capture set is empty because \li{Int} is a built in type
-and the initial content is closed. 
-This serves to ensure \emph{context independence}: the livelit needs to make no 
-assumptions about the typing context at invocation sites.
-We use an explicit capture set, rather than implicitly capturing all bindings 
+To ensure \emph{context independence}, 
+the system checks that the splice type and initial content are valid 
+assuming only the parameters and explicitly specified \li{context} on Line 6.
+Here, the context is empty because \li{Int} is a built in type
+and the initial expressions are integer literals. 
+We use an explicit context, rather than implicitly capturing all bindings 
 at the definition site, 
 to ensure that private bindings are not unintentionally  
-leaked to clients as detailed by \citet{TLMs}.
+leaked to clients \cite{TLMs}.
 % Formally, captured variables can be understood as parameters that have been 
@@
 specifies a single user-initiated action: clicking on a color using
-the right side of Fig.~\ref{fig:color}. Actions are emitted
-from event handlers (e.g. click handlers) defined in the computed \li{view}, 
-and actions are consumed by the \li{update} function, causing a change to the model. 
-Let us discuss each of these functions in turn in Sec.~\ref{sec:def-view}-\ref{sec:def-update}.
+the right half of Fig.~\ref{fig:color}. Actions are emitted
+from event handlers (e.g. click handlers) defined in the computed \li{view}, Sec.~\ref{sec:def-view}, 
+and actions are consumed by the \li{update} function, Sec. \ref{sec:def-update}, causing a change to the model.
 
 \subsubsection{View}\label{sec:def-view}
-The \li{view} function computes the view given a model and access to the commands in 
-the \li{ViewMonad}. The computed view is a value of type \li{Html(Action)}.  
-The type family \li{Html(a)} provides a simple immutable
-encoding of an HTML tree, where the type parameter \li{a} is the type of actions that 
-are emitted by event handlers that can be attached to elements of the tree, e.g.
+The \li{view} function computes the view given the model and access to the commands in 
+the \li{ViewCmd} monad (which is distinct from the \li{UpdateCmd} monad). The computed view is a value of type \li{Html(Action)}.  
+This type  provides a simple immutable
+encoding of an HTML element, where the type parameter is the type of actions that 
+are emitted by event handlers that can be attached to elements, e.g.
 \li{on_click} and so on. 
-We elide the details of the particular user interface in Fig.~\ref{fig:color}
-to focus on three key editor mechanisms exposed by \li{ViewMonad}: live evaluation,
+We elide the details of the particular user interface in Fig.~\ref{fig:color}, 
+but note that livelit implementations can themselves invoke other  
+livelits, e.g. the \li{view} function could use livelits for 
+expressing user interface widgets and layouts (not shown). 
+Instead, we focus on three mechanisms exposed by \li{ViewCmd}: live evaluation,
 splice editors, and result rendering.
@@
 As discussed in Sec.~\ref{sec:live-evaluation}, 
-the view can depend on the result of evaluating a splice 
+the view can depend on the result of evaluating a splice or a parameter
 under the closure the client has selected.
+(Parameters, not shown in this example but discussed in Sec.~\ref{sec:parameterization}, operate like splices and have type \li{SpliceRef}
+within the livelit definition.)
 The interface between the view and the live evaluator 
@@
 \begin{lstlisting}[numbers=none,xleftmargin=0cm]
-eval_splice : SpliceRef -> ViewMonad(Maybe(Result))
+eval_splice : SpliceRef -> ViewCmd(Maybe(Result))
 \end{lstlisting}
 
-%eval_fun : (Exp, List(SpliceRef)) -> ViewMonad(Maybe(Result))
+%eval_fun : (Exp, List(SpliceRef)) -> ViewCmd(Maybe(Result))
 
 The \li{None} case 
-arises when evaluation is not possible, e.g. because no closures are collected.
+arises when evaluation is not possible, e.g. because no closures are collected 
+or because no value has been collected for a variable used in the splice. 
+In practice, there is an implicit hole at the end of each cell in Hazel, so 
+livelits will have at least one closure containing results for at least 
+the top-level bindings. Function argument values may not be available 
+if that function has not been applied.
 % because the free variables in the selected closure overlap with the free 
@@
 % We discuss this further in Sec.~\ref{sec:calculus-closure-collection}.
-If a result is available, the \li{Result} type distinguishes two possibilities:
+
+If available, \li{Result} distinguishes two possibilities:
 \begin{lstlisting}[numbers=none]
@@
 enough information to determine a color. The livelit 
-indicates this indeterminacy with an ``X'' over the preview.
+indicates this situation by disabling the color preview.
 
@@
 to handle a list of notes where certain notes are missing, i.e. holes, by 
-simply playing silence or some default sound when it encounters them.
+playing silence.
 This behavior is highly domain-specific, so each livelit provider must decide 
@@
 \begin{lstlisting}[numbers=none]
-editor : (SpliceRef, Dim) -> ViewMonad(Html(a))
+editor : (SpliceRef, Dim) -> ViewCmd(Html(a))
 \end{lstlisting}
@@
 a rendered evaluation result. For example, each of the cells in the \li{\$dataframe}
-livelit in Fig.~\ref{fig:grading} shows the evaluation result for the corresponding 
+livelit in Fig.~\ref{fig:grading} show the evaluation result for the corresponding 
 cell. Only the formula bar at the top is an editor. To support this, the \li{view}
-function can use the \li{result_view} command, which mirrors the \li{eval} command:
+function can use the \li{result_view} command, which mirrors the \li{editor} command:
 \begin{lstlisting}[numbers=none]
-result_view : 
-  (SpliceRef, Dim) -> ViewMonad(Maybe(Html(a)))
+result_view : (SpliceRef, Dim) 
+              -> ViewCmd(Maybe(Html(a)))
 \end{lstlisting}
-
 \subsubsection{Update}\label{sec:def-update}
@@
 
-% We discussed in Sec.~\ref{sec:model} that the \li{UpdateMonad}
+% We discussed in Sec.~\ref{sec:model} that the \li{UpdateCmd}
 % allows new splices to be generated in response to user actions. 
@@
 \begin{lstlisting}[numbers=none]
-  set_splice : (SpliceRef, Exp) -> UpdateMonad ()
+  set_splice : (SpliceRef, Exp) -> UpdateCmd ()
 \end{lstlisting}
-The system checks that the expression must satisfy the splice type 
-and onlys make use of the capture set, as in Sec.~\ref{sec:model}.
+As with \li{new_splice} described in Sec.~\ref{sec:model},
+the system maintains context independence by checking the expression against the splice type 
+and only allows use of the specified \li{context}.
 
@@
 Changes to splices can also cause the view to be recomputed, because the view might 
-evaluate the splices. The \li{update} function does not itself 
-have the ability to request evaluation, because the model should not depend directly  
+evaluate the splices. The \li{UpdateCmd} monad does not itself 
+have the ability to request evaluation (\li{eval_splice}), because the model should not depend directly  
 on which closure the user has selected. 
-%Of course, the \li{view} might emit 
-%result-dependent actions when appropriate.
+Of course, the \li{view} might emit 
+result-dependent actions when appropriate.
 
@@
 i.e. an expression of the expansion type, here \li{Color}.
-The \li{expand} function determines this expansion based on the model.
+The \li{expand} function determines this expansion based on the model 
+(which encodes the GUI state, and so usually contains information, like splice references
+and widget states, 
+that should not persist into run-time as part of the expansion).
 
-The expansion can include splices, but the system does not make the contents of a splice 
-available directly as a value of type \li{Exp}. Instead, \li{expand} returns a \emph{parameterized expansion}
-paired with a list of \li{SpliceRef}s. 
-The {parameterized expansion}
- is an encoding (of type \li{Exp}) of a function 
- that takes an argument for each listed \li{SpliceRef}s. 
+The expansion can include the spliced expressions, 
+but the system does not make the expression  
+available directly as a value of type \li{Exp}. 
+Instead, \li{expand} returns a \emph{proto-expansion}, of type \li{Exp}, 
+paired with a list of \li{SpliceRef}s (which can come from 
+both splices and parameters as discussed above).
+The {proto-expansion}
+ is an encoding (i.e. a value of type \li{Exp}) of a function 
+ (here curried)
+ that takes an argument for each listed \li{SpliceRef}. 
  That argument is of the corresponding splice 
 type, which was provided when the splice was initialized. 
- The return type of the parameterized expansion is the expansion type.
- Here, the parameterized expansion for \li{\$color} encodes a function of type 
+ The return type of the proto-expansion is the expansion type.
+ So here, the proto-expansion for \li{\$color} must be a function of type 
  \li{Int -> Int -> Int -> Int -> Color}.
 
-This parameterization strategy makes it simple to enforce hygiene: the parameterized expansion 
-can depend only on the livelit parameters and capture set, whereas the splices are entered by the client and so they can 
+Note that Hazel does \emph{not} statically check the definition of \li{expand}
+to ensure that the encoded proto-expansion has this type. Instead, the proto-expansion 
+is \emph{validated} at each livelit invocation site, with errors reported to the client. 
+A \emph{typed quotation} system as in, e.g., MetaOCaml \cite{DBLP:conf/flops/Kiselyov14}, could be adapted to allow for definition-site verification 
+in lieu of client-site validation, as discussed in \cite{TLMs}, but we leave the integration 
+of such a system into Hazel as future work. This would be non-trivial:  
+the type of the quotation depends on the type of each splice in 
+the splice list, which may be of model-dependent length (e.g. in the case of \li{\$dataframe}).
+
+This parameterization strategy makes it simple to enforce the binding discipline described in Sec.~\ref{sec:hygiene}. 
+The proto-expansion 
+can depend only on the \li{context}, whereas the splices are entered by the client and so they can 
 depend only on the client site typing context. The use of function application ensures
-that splices are capture avoiding. We consider this more formally in the next section.
+that splices are capture avoiding as discussed in Sec.~\ref{sec:hygiene}. 
+We consider this more formally next.

### 2021-04-06 pldi21/related-work.tex
revisions up through sec 3

@@
 and the reasoning principles are closely related. However, the  
-approach to hygiene we take is cleaner by its use of function application 
+approach to capture avoidance we take is both more restrictive and cleaner by its use of function application 
 rather than direct insertion of splices (and could perhaps be ported to those formalisms).
@@
 that does not support full compositionality as described here, where livelits 
-can appear within other livelits. Splicing is not hygienic, nor is there any 
+can appear within other livelits. Splicing is not strictly hygienic, though 
+Racket's scope management machine can be used by macros to voluntarily 
+maintain a binding discipline \cite{DBLP:conf/popl/Flatt16}. There is not any 
 consideration of typing. Parameterization and partial application is also impossible, because invocation 
 is via an editor command rather than a syntactic mechanism. Finally,
-there is no support for liveness.
+there is no support for liveness: macro evaluation occurs in the editor 
+environment, not the run-time environment of the program being written, 
+so splice evaluation would not work correctly (except in trivial cases
+where the splice was, e.g., a closed expression, or used only standard 
+library constructs).
 

### 2021-04-07 pldi21/case-studies.tex
made code in text consistent with new fig1

@@
 \li{\$slider}.
-The table itself displays not the expression itself but rather its value, \li{93.}, just as in a spreadsheet.
+The table itself displays not the expression itself but rather its value, \li{80.}, just as in a spreadsheet.
 
@@
 \begin{lstlisting}[xleftmargin=0.2cm]
+let q1_max = 36. in 
 let grades = Dataframe (
   ["A1", "A2", "A3", "Midterm", "Final"],
-  [("Alice", [24. +. 36. +. 33.,
+  [("Andrew", [q1_max +. 24. +. 20.,
                92., 83.5, 95., 88.]),
-   ("Bob", [61., 64., 98., 70., 85.]),
-   ("Carol", [75., 81., 73., 82., 79.]),
+   ("Cyrus", [61., 64., 98., 70., 85.]),
+   ("David", [75., 81., 73., 82., 79.]),
    (* ... *) ]) in
 let averages = compute_averages grades weights in
-let cutoffs = (.A 90., .B 80., .C 70., .D 60.) in
+let cutoffs = (.A 86., .B 76., .C 67., .D 48.) in
 format_for_university

### 2021-04-07 pldi21/implementation.tex
second round

@@
 although the developers are aiming to maintain a text-like experience (this effort is orthogonal to our own). 
+To maintain this guarantee, Hazel inserts empty holes where there are missing terms, 
+and non-empty holes as markers of errors. To maintain this invariant in the presence of livelits, 
+new non-empty holes are needed to mark each of the failure modes suggested by the premises of the \rulename{ELivelit} rule in Fig.~\ref{fig:expansion}. 
+In particular, non-empty holes mark (1) the invocation of an unbound livelit; (2) the invocation of a livelit 
+with a model value of the wrong type; 
+(3) a run-time error during evaluation of \li{init}, \li{update}, or \li{expand} functions (which manifests as a run-time hole); (4) 
+expansion validation failure (i.e. that the parameterized expansion is of the wrong type given the splice list).  
+Errors in view generation are not considered semantic errors (they display as error messages where the livelit GUI would have appeared.)
 

### 2021-04-07 pldi21/livelit-calculus.tex
revisions up through sec 3
second round

@@
 
+In order to communicate the semantics of livelits
+independently
+of the specifics of the Hazel environment and the web platform, 
+we now specify a \emph{simply typed livelit calculus}.
+
+Fig.~\ref{fig:syntax} specifies the syntax of the livelit calculus.
+Programs are written as \emph{unexpanded expressions}, $\hat e$, which are \emph{expanded} to
+\emph{external }(or \emph{expanded}) \emph{expressions}, $e$, before being \emph{elaborated}
+to \emph{internal expressions}, $d$, for evaluation. All three sorts are classified
+by the same types, $\tau$. We include partial functions, products, sums, and recursive
+types, all in their standard form \cite{pfpl}, but this specific type structure is not critical.
+Any language expressive
+enough to encode its own abstract syntax 
+would be a suitable basis for a livelit calculus.
+
+We begin in Sec.~\ref{sec:external-and-internal-lang} with a terse
+overview of the external and internal languages,
+which are adapted from Hazelnut Live \cite{HazelnutLive}.
+We then detail livelit expansion (Sec.~\ref{sec:calculus-expansion}) and 
+liveness via closure collection (Sec.~\ref{sec:calculus-closure-collection}). 
+% and parameterization (Sec.~\ref{sec:calculus-parameterization}).
+
+\subsection{Background: External and Internal Language}\label{sec:external-and-internal-lang}
+The external and internal languages are straightforward adaptations of the
+external and internal languages of \emph{Hazelnut Live},
+a typed lambda calculus that assigns static and dynamic meaning to programs with holes,
+notated $\hehole{u}$ where $u$ is a \emph{hole name} \cite{HazelnutLive}.
+We omit non-empty holes (which internalize type inconsistencies \cite{Hazelnut}) and type holes
+(which operate like the unknown type from gradual type theory \cite{Siek06a,Hazelnut}).
+These mechanisms are orthogonal to livelits and are included in our implementation.
+
+External expressions, e, are governed by a typing judgement of the form, $\hasType{\Gamma}{e}{\tau}$,
+where the typing context, $\Gamma$,
+is a finite set of typing assumptions of the form $x : \tau$ \cite{pfpl}.
 \begin{figure}
+    \vspace{-3px}
     \[
@@
         \mathsf{Splice} & \psi & ::= & \hsplice{\hat e}{\tau}
-    \end{array}
-    \]
-    \caption{Syntax of types, $\tau$, unexpanded expressions, $\hat{e}$, expanded expressions, $e$, and internal expressions, $d$.
+    \end{array}\]
+    \caption{Syntax of types, $\tau$, unexpanded expressions, $\hat{e}$, external expressions, $e$, and internal expressions, $d$.
     Here, $x$ ranges over variables, $u$ over hole names, and $\livelitname{a}$ over livelit names.
     We write $\splat{\psi_i}$ for a finite sequence of $n \geq 0$ splices,
-    and $\sigma$ for finite substitutions of $n \geq 0$ internal expressions for variables, $[d_1/x_1, \cdots, d_n/x_n]$.
+    and $\sigma$ for finite substitutions of $n \geq 0$ internal expressions for variables, $[d_i/x_i]_{i < n}$.
     We elide standard forms
@@
 
-In order to communicate the semantics of livelits
-independently
-of the specifics of the Hazel environment and the web platform, 
-we now specify a simply typed \emph{livelit calculus}.
-
-Fig.~\ref{fig:syntax} specifies the syntax of the livelit calculus.
-Programs are written as \emph{unexpanded expressions}, $\hat e$, which are \emph{expanded} to
-\emph{external expressions}, $e$, before being \emph{elaborated}
-to \emph{internal expressions}, $d$, for evaluation. All three sorts are classified
-by the same types, $\tau$. We include partial functions, products, sums, and recursive
-types, all in their standard form \cite{pfpl}, but this specific type structure is not critical.
-Any language expressive
-enough to encode its own abstract syntax 
-would be a suitable basis for a livelit calculus.
-
-We begin in Sec.~\ref{sec:external-and-internal-lang} with a terse
-overview of the external and internal languages,
-which are adapted from Hazelnut Live \cite{HazelnutLive}.
-We then detail livelit expansion (Sec.~\ref{sec:calculus-expansion}) and 
-liveness via closure collection (Sec.~\ref{sec:calculus-closure-collection}). 
-% and parameterization (Sec.~\ref{sec:calculus-parameterization}).
-
-\subsection{Background: External and Internal Language}\label{sec:external-and-internal-lang}
-The external and internal languages are straightforward adaptations of the
-external and internal languages of \emph{Hazelnut Live},
-a typed lambda calculus that assigns static and dynamic meaning to programs with holes,
-notated $\hehole{u}$ where $u$ is a \emph{hole name} \cite{HazelnutLive}.
-We omit non-empty holes (which internalize type inconsistencies \cite{Hazelnut}) and type holes
-(which operate like the unknown type from gradual type theory \cite{Siek06a,Hazelnut}).
-These mechanisms are orthogonal to livelits and are included in our implementation.
-
-External expressions, e, are governed by a typing judgement of the form, $\hasType{\Gamma}{e}{\tau}$,
-where the typing context, $\Gamma$,
-is a finite set of typing assumptions of the form $x : \tau$ \cite{pfpl}.
-
 The internal language is a contextual type theory \cite{Nanevski2008}, i.e. the typing judgement is
-of the form $\hasTypeD{\Delta}{\Gamma}{d}{\tau}$ where $\Delta$ is a finite set of hole typing
-assumptions of the form $u :: \tau[\Gamma]$.
-We need this hole typing context only for the internal language because, although hole names are assumed
+of the form $\hasTypeD{\Delta}{\Gamma}{d}{\tau}$ where the hole context $\Delta$ is a finite set of hole typing
+assumptions of the form $u :: \tau[\Gamma]$ which mean that hole $u$ must be filled with an expression of type $\tau$ under $\Gamma$.
+We need a hole typing context only for the internal language because, although hole names are assumed
 unique in the external language, they can be duplicated during evaluation of internal expressions.
-The hole context ensure that each closure associated with hole $u$, written $\dehole{u}{\sigma}$,
-shares a type and can be filled
-with expressions valid under a shared context.
+% The hole context ensures that each closure associated with hole $u$, written $\dehole{u}{\sigma}$,
+% shares a type and can be filled
+% with expressions valid under a shared context.
 
@@
 
-External expressions are given dynamic meaning by typed elaboration to internal expressions, $d$,
-according to the judgement $\elabs{\Gamma}{e}{d}{\tau}{\Delta}$.
-The main purpose of elaboration is to initialize the substitution $\sigma$ on each hole closure,
-which operates to capture
-the substitutions that have occurred around that hole during evaluation. The key rule is:
+External expressions elaborate to internal expressions, $d$,
+by the judgement $\elabs{\Gamma}{e}{d}{\tau}{\Delta}$.
+The main purpose of elaboration is to initialize the substitution $\sigma$ on each hole closure, $\dehole{u}{\sigma}$, 
+which takes the form $[d_i/x_i]_{i<n}$ and serves to capture
+the substitutions that occur around the hole during evaluation. The key rule is:
 \begin{mathpar}
@@
 \end{mathpar}
+In this declarative formulation, the type $\tau$ is arbitrary (in Hazelnut Live, elaboration is specified algorithmically \cite{HazelnutLive}).
 The substitution is initially the identity substitution, $\idof{\Gamma}$, i.e. the
@@
 
-All of these judgements are adapted directly from Hazelnut Live,
-differing only in that we use a declarative rather than an algorithmic (bidirectional)
-formulation for simplicity. Rather than restating the rules, we simply state the key governing
-metatheorems and refer the reader to the prior work and our Agda mechanization (Sec.~\ref{sec:agda}) for the full details \cite{HazelnutLive}.
+Rather than restating the remaining rules, we simply state the key governing
+metatheorems and defer to the prior work and our Agda mechanization (Sec.~\ref{sec:agda}) for the details \cite{HazelnutLive}.
 
-First, we have that elaboration preserves typing.
+Elaboration preserves typing.
 \begin{theorem}[Typed Elaboration]
@@
 
-Next, we have that evaluation of a closed expression with holes produces a final (i.e. irreducible) preserves typing.
+Evaluation of a closed well-typed expression with holes results in a final (i.e. irreducible) expression of the same type.
 \begin{theorem}[Preservation]
@@
 The rules for the mirrored 
-forms like variables and functions, both shown in Fig.~\ref{fig:expansion},
-are simple.
+forms like \rulename{EVar} and \rulename{EFun}, both shown in Fig.~\ref{fig:expansion},
+are straightforward.
 
@@
 Livelit definitions are collected in the livelit context, $\Phi$, which
-maps livelit names to livelit definitions of the following form:
+maps livelit names $\livelitname{a}$ to livelit definitions of the following form:
 \[ \livelitCtxEntry{a}{\taut{expand}}{\taut{model}}{\dtxt{expand}} \]
@@
 
-\subsubsection{Hygienic Livelit Expansion}
+\subsubsection{Livelit Expansion}
 \begin{figure}
@@
         }
-
+        \hspace{15px} 
         \inferrule[ELam]{
@@
             \expands{\Phi}{\Gamma}{\hlam{x}{\hat e}}{\hlam{x}{e}}{\tarr{\taut{in}}{\taut{out}}}
-        }~~~~~~~~\cdots
+        }\hspace{15px}\cdots
 
@@
         % }
-        \inferrule[EApLivelit]{
+        \inferrule[ELivelit]{
             \livelitCtxEntry{a}{\taut{expand}}{\taut{model}}{\dtxt{expand}} \in \Phi\\\\
@@
     \end{figure}
-Unexpanded expressions are unique in that they include livelit invocations:
+Unexpanded expressions include livelit invocations:
  \[\haplivelit{u}{a}{\dtxt{model}}{\splices}\]
 Here, $\livelitname{a}$ names the livelit
- being applied. Livelits can be understood as filling holes, so $u$ identifies the hole
+ being invoked. Livelits can be understood as filling holes, so $u$ identifies the hole
  that is, conceptually, being filled.
@@
 itself (unexpanded, so it may contain other livelits) and $\tau_i$ is the type of that splice,
-as determined when the livelit definition first requested the splice (discussed in Sec.~\ref{sec:model}).
+as determined when the livelit definition first requested the splice (as discussed in Sec.~\ref{sec:model}).  
+Note that we do not formally model the edit actions that change the model or splice list here;
+we focus on a single snapshot of the editor state.
 
-Rule \rulename{EApLivelit} performs livelit expansion. Its premises, in order, operate as follows:
+Rule \rulename{ELivelit} performs livelit expansion. Its premises, in order, operate as follows:
 \begin{enumerate}[itemsep=3px,leftmargin=*]
-    \item \textbf{Lookup.} The first premise looks up the livelit name in the livelit context.
-    \item \textbf{Model Validation.} The second premise serves to ensure that the model value, $\dtxt{model}$, is of the
+    \item \textbf{Lookup.} The first premise looks up the livelit definition in the livelit context.
+    \item \textbf{Model Validation.} The second premise checks that the model value, $\dtxt{model}$, is of the
     specified model type, $\taut{model}$.
     \item \textbf{Expansion.} The third premise applies the expansion function, $\dtxt{expand}$, to the model value, $\dtxt{model}$,
-    producing the encoded parameterized expansion, $\dtxt{encoded}$, which, by the definitions given above, is of type $\mathsf{Exp}$.
+    producing the \emph{encoded parameterized expansion}, $\dtxt{encoded}$, which, by the definitions and theorems given above, is of type $\mathsf{Exp}$.
     \item \textbf{Decoding.} The fourth premise decodes $\dtxt{encoded}$, producing the \emph{parameterized expansion}, $\etxt{pexpansion}$.
+The isomorphism between encodings and external expressions ensures that decoding cannot fail.
     \item \textbf{Expansion Validation.} The fifth premise checks that {parameterized expansion} is a function that returns a value of the expansion type, $\taut{expand}$, when applied (in curried fashion, though this is not critical)
-    to the splices, whose types, $\splat{\tau_i}$, are explicitly recorded in the livelit.
-    We must take care to ensure that the parameterized expansion is \emph{context independent},
-    i.e. that it cannot depend on the particular bindings available in the call site typing context, $\Gamma$.
+    to the splices, whose types, $\splat{\tau_i}$, are given in the splice list.
+    Validation can fail, in which case expansion fails. (In Hazel, validation failure is instead marked by a non-empty hole.)
+
+    We ensure that the parameterized expansion is \emph{context independent},
+    i.e. that it cannot depend on the particular bindings available in the call site typing context, $\Gamma$, 
     % Context independence is one facet of what is colloquially known as \emph{hygiene} \cite{TLMs}.
-    In this simple formulation of the system, we maintain context independence by
+     by
     requiring that the parameterized expansion be
-    closed. Consequently, any necessary helper functions used in the expansion must be provided explicitly by the client
+    closed. Consequently, any necessary helper functions used in the expansion must be provided by the client
     via a splice.
-    In Sec.~\ref{sec:livelit-definitions}, we discussed how the use of explicit capture sets eliminates this client burden.
-    For simplicity, we omit capture sets from the calculus, but see \cite{TLMs} for a formal treatment.
+
+    In Sec.~\ref{sec:livelit-definitions}, we discussed how the use of an explicit definition site \li{context} eliminates this client burden.
+    The explicit context is formally just a value (tupled, typically) that can be passed as an additional argument to the parameterized expansion alongside the splices, 
+    so we omit it from the calculus, but see \cite{TLMs} for a full treatment.
+
     \item \textbf{Splice Expansion.} The sixth premise inductively expands each of the spliced expressions in the same context as the livelit
@@
 The typed expansion process is governed by the following metatheorem, which establishes that the expansion
-is indeed an external expression of the indicated type.
-
+is indeed an external expression of the indicated type (i.e. the expansion type, in the case of livelit invocations).
 \begin{theorem}[Typed Expansion]
@@
 %     The cases involving the standard forms follow by straightforward induction.
-%     The only interesting case is the \rulename{EApLivelit} case.
+%     The only interesting case is the \rulename{ELivelit} case.
 %     In this case, we have that $e = \hap{\etxt{pexpansion}}{\splat{e_i}}$.
@@
 
-When composed with the Typed Elaboration theorem and the Type Safety properties established for the internal
+When composed with the Typed Elaboration theorem and the type safety of the internal
 language, we achieve end-to-end type safety: every well-typed unexpanded
@@
 \label{sec:agda}
-We have mechanized the external and internal language and 
-typed expansion and proven the aforementioned theorems
+The supplement mechanically proves these theorems 
 using the Agda proof assistant, building on the Agda mechanization 
-of Hazelnut Live \cite{HazelnutLive}. This mechanization is available in the supplement.
+of Hazelnut Live \cite{HazelnutLive}.
 
@@
 where each livelit invocation expands to an empty hole applied to its splices. In other words,
-the hole is in place of the parameterized expansion, which we instead record in a \emph{cc-context}, $\Omega$, on the side.
-The key rule for the cc-expansion judgement, $\ccexpands{\Phi}{\Gamma}{\hat e}{e}{\tau}{\Omega}$, is:
+a hole appears in place of the parameterized expansion (but not splices).
+On the side, we generate a \emph{cc-context}, $\Omega$,  that maps each livelit hole to the \emph{elaboration} of its parameterized expansion, $\Omegaitem{u}{\dtxt{pexpansion}}$.
+The key rule for cc-expansion, $\ccexpands{\Phi}{\Gamma}{\hat e}{e}{\tau}{\Omega}$, is:
 \begin{mathpar}
-\inferrule[CCApLivelit]{
+\inferrule[CCLivelit]{
     \{ \ccexpands{\Phi}{\Gamma}{\hat e_i}{e_i}{\tau_i}{\Omega_i} \}_{i < n}\\
     \expands{\Phi}{\Gamma}{\haplivelit{u}{a}{\dtxt{model}}{\splat{\hsplice{\hat e_i}{\tau_i}}}}{
-        \hap{\etxt{pexpansion}}{\splat{e_i}}
+        \hap{\etxt{pexpansion}}{\splat{e'_i}}
     }{\taut{expand}}\\
-    \elabs{ }{\etxt{pexpansion}}{\dtxt{pexpansion}}{\tarr{\splat{\tau_i}}{\taut{expand}}}{\Delta}\\
+    \elabs{ }{\etxt{pexpansion}}{\dtxt{pexpansion}}{\tarr{\splat{\tau_i}}{\taut{expand}}}{ }\\
     \Omega = \cup_{i<n} \Omega_i \cup \{\Omegaitem{u}{\dtxt{pexpansion}}\}
@@
 In this case, however, the parameterized expansion is necessarily closed due to
-the context independence discipline we maintain in Rule \rulename{EApLivelit},
+the context independence discipline we maintain in Rule \rulename{ELivelit},
 so hole filling amounts to syntactic replacement.
@@
 
-In a language with side effects, one could evaluate 
-cc-expansions in an alternative mode where the full expansion of each livelit invocation is evaluated 
-in the order that corresponding livelit hole, and any expressions blocked on it are resumed immediately. 
-The results would need to be recorded in memory for use in the filling phase. This would ensure that
-side effects happen in the same order and only once. %A formal specification is left as future work.
+In an imperative language with non-commutative side effects, resumption is not sound. We conjecture that one can specify an 
+alternative evaluation mode where the full expansion of each livelit invocation is evaluated 
+in the order that corresponding livelit hole is encountered during evaluation, with the closure recorded in a memory. 
+This would ensure that
+side effects happen in the same order and only once. 
+% It would also allow substitutions that only occur conditionally on another livelit invocation's value to be recorded (as specified, these dependencies). 
+%A formal specification is left as future work.
 Alternatively, an imperative language might use a different mechanism to make
-environments available to livelits, e.g. by environment logging \emph{a la} Lamdu \cite{lamdu}.
-This is more limited in situations where a livelit is never evaluated, as sometimes 
-occurs during development of a livelit in a function which has yet to be applied, 
-but does not require cc-expansion. We leave further exploration of this design space 
+environments available to livelits, e.g. by environment snapshotting \emph{a la} Lamdu \cite{lamdu}.
+This is more limited in situations where a livelit invocation appears in a function which has yet to be applied
+ (where collected livelit closures essentially 
+ capture the function's closure).
+
+Livelits invocations in branches that are not taken or that appear under unused variables do not have closures collected for them with our approach
+(though in practice there is an implicit ``trailing hole'' after each cell in Hazel which means the latter situation is infrequent). 
+Livelit invocations that determine which of several branches are taken prevent substitutions conditional on those branches from being recorded in 
+downstream closures because resumption operates only within the proto-closures. 
+This design gives clients the ability to control closure collection in branches that are not of interest (and therefore control the cost of closure collection), 
+but may sometimes cause unexpected behavior. 
+We leave further exploration of this design space 
 to future work.

### 2021-04-07 pldi21/livelit-definitions.tex
second round

@@
 available directly as a value of type \li{Exp}. 
-Instead, \li{expand} returns a \emph{proto-expansion}, of type \li{Exp}, 
+Instead, \li{expand} returns a \emph{parameterized expansion}, of type \li{Exp}, 
 paired with a list of \li{SpliceRef}s (which can come from 
 both splices and parameters as discussed above).
-The {proto-expansion}
+The {parameterized expansion}
  is an encoding (i.e. a value of type \li{Exp}) of a function 
@@
 type, which was provided when the splice was initialized. 
- The return type of the proto-expansion is the expansion type.
- So here, the proto-expansion for \li{\$color} must be a function of type 
+ The return type of the parameterized expansion is the expansion type.
+ So here, the parameterized expansion for \li{\$color} must be a function of type 
  \li{Int -> Int -> Int -> Int -> Color}.
@@
 Note that Hazel does \emph{not} statically check the definition of \li{expand}
-to ensure that the encoded proto-expansion has this type. Instead, the proto-expansion 
+to ensure that the encoded parameterized expansion has this type. Instead, the parameterized expansion 
 is \emph{validated} at each livelit invocation site, with errors reported to the client. 
-A \emph{typed quotation} system as in, e.g., MetaOCaml \cite{DBLP:conf/flops/Kiselyov14}, could be adapted to allow for definition-site verification 
-in lieu of client-site validation, as discussed in \cite{TLMs}, but we leave the integration 
-of such a system into Hazel as future work. This would be non-trivial:  
+A \emph{typed quotation} system as in, e.g., MetaOCaml \cite{DBLP:conf/flops/Kiselyov14}, could be adapted to allow for definition-site verification, as discussed in \cite{TLMs}, but 
+note that   
 the type of the quotation depends on the type of each splice in 
-the splice list, which may be of model-dependent length (e.g. in the case of \li{\$dataframe}).
+the splice list.
 
 This parameterization strategy makes it simple to enforce the binding discipline described in Sec.~\ref{sec:hygiene}. 
-The proto-expansion 
+The parameterized expansion 
 can depend only on the \li{context}, whereas the splices are entered by the client and so they can 

### 2021-04-07 pldi21/related-work.tex
second round

@@
 
-Recent work on a interactive visual macro system in Racket 
+Recent work on a interactive visual macro system in Dr. Racket 
 is quite similar in spirit to our work \cite{interactive-visual-syntax}.
@@
 can appear within other livelits. Splicing is not strictly hygienic, though 
-Racket's scope management machine can be used by macros to voluntarily 
+Racket's scope management machinery can be used by macros to voluntarily 
 maintain a binding discipline \cite{DBLP:conf/popl/Flatt16}. There is not any 
@@
 where the splice was, e.g., a closed expression, or used only standard 
-library constructs).
+library constructs). That said, we conjecture that it is
+likely possible to implement many of the mechanisms we describe in this paper 
+perhaps in some form as a layer atop existing mechanisms in Dr. Racket.
 
@@
 However, these systems generate entire editors, whereas livelit definitions 
-are defined in libraries. 
+are decentralized in libraries. 
 Furthermore, livelit definitions are governed by a type and binding discipline.
@@
 from these systems in directionality: visualization systems are given code and generate
-a visualization, whereas livelits are generate code where there would otherwise
+a visualization, whereas livelits generate code where there would otherwise
 be a hole.

### 2021-04-07 pldi21/implementation.tex
tweaks about text editor integration

@@
 Livelit interaction causes the serialized model in the text buffer to be changed, which updates the view.
-This proof-of-concept is not at feature parity with 
+This proof-of-concept is focused on this interaction so it is not at feature parity with the 
+mechanism as described here or 
 our main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84,interactive-visual-syntax} 
@@
 there are syntax errors.
+There are other systems that integrate visual syntax into an otherwise textual editor as well, notably Dr.
+Racket's recently introduced visual macro system \cite{interactive-visual-syntax}, discussed in the next section.
 

### 2021-04-13 pldi21/discussion.tex
second round
updates in respond to shepherd

@@
   patterns, and entire modules. Prior work on literal macros has explored this \cite{TLMs}.
+  (Note that Racket's visual macro system generates
+  arbitrary syntax, so it can already be used in 
+  this manner, albeit with no sort-specific semantic guarantees.)
   % Support for type and module splices, too, would likely be useful.
@@
 
-   The strict hygiene discipline has, we believe, substantial 
+   The strict binding discipline has, we believe, substantial 
    benefits---programmers will inevitably encounter unfamiliar livelits, and 
@@
 
-\newpage
 % \begin{itemize}
@@
 % \end{itemize}
+
+\section*{Acknowledgements}
+
+(Acknowledgements are being prepared and will be included in the version sent to the publisher.)

### 2021-04-13 pldi21/implementation.tex
updates in respond to shepherd

@@
 We have also developed a 
-proof-of-concept implementation of livelits in a textual program editor, Sketch-n-Sketch \cite{sns-pldi,sns-uist},
-which recently added support for the necessary mechanisms \cite{DBLP:journals/pacmpl/LubinCOC20}.
+proof-of-concept implementation of livelit interaction in a textual program editor, Sketch-n-Sketch \cite{sns-pldi,sns-uist}.
 Livelit interaction causes the serialized model in the text buffer to be changed, which updates the view.
-This proof-of-concept is focused on this interaction so it is not at feature parity with the 
-mechanism as described here or 
-our main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84,interactive-visual-syntax} 
+This proof-of-concept is focused on this interaction loop so it is not at feature parity with the 
+ main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84,interactive-visual-syntax} 
 is sufficient to support livelits, albeit with some gaps in availability when 

### 2021-04-13 pldi21/livelit-calculus.tex
updates in respond to shepherd

@@
 
-In order to communicate the semantics of livelits
+In order to specify the semantics of livelits
 independently
@@
 to \emph{internal expressions}, $d$, for evaluation. All three sorts are classified
-by the same types, $\tau$. We include partial functions, products, sums, and recursive
+by the same types, $\tau$.
+
+We include partial functions, products, sums, and recursive
 types, all in their standard form \cite{pfpl}, but this specific type structure is not critical.
@@
 enough to encode its own abstract syntax 
-would be a suitable basis for a livelit calculus.
+would be a suitable basis.
 
-We begin in Sec.~\ref{sec:external-and-internal-lang} with a terse
-overview of the external and internal languages,
-which are adapted from Hazelnut Live \cite{HazelnutLive}.
-We then detail livelit expansion (Sec.~\ref{sec:calculus-expansion}) and 
-liveness via closure collection (Sec.~\ref{sec:calculus-closure-collection}). 
+% We begin in Sec.~\ref{sec:external-and-internal-lang} with a terse
+% overview of the external and internal languages,
+% which are adapted from Hazelnut Live \cite{HazelnutLive}.
+% We then detail livelit expansion (Sec.~\ref{sec:calculus-expansion}) and 
+% liveness via closure collection (Sec.~\ref{sec:calculus-closure-collection}). 
 % and parameterization (Sec.~\ref{sec:calculus-parameterization}).
@@
 External expressions elaborate to internal expressions, $d$,
-by the judgement $\elabs{\Gamma}{e}{d}{\tau}{\Delta}$.
+according to the declarative elaboration judgement $\elabs{\Gamma}{e}{d}{\tau}{\Delta}$.
+(Hazelnut Live specifies elaboration algorithmically \cite{HazelnutLive}.)
 The main purpose of elaboration is to initialize the substitution $\sigma$ on each hole closure, $\dehole{u}{\sigma}$, 
 which takes the form $[d_i/x_i]_{i<n}$ and serves to capture
-the substitutions that occur around the hole during evaluation. The key rule is:
+the substitutions that occur around the hole during evaluation. The key rule for elaboration at type $\tau$ is:
 \begin{mathpar}
@@
 \end{mathpar}
-In this declarative formulation, the type $\tau$ is arbitrary (in Hazelnut Live, elaboration is specified algorithmically \cite{HazelnutLive}).
+% In this declarative formulation, the type $\tau$ is arbitrary .
 The substitution is initially the identity substitution, $\idof{\Gamma}$, i.e. the

### 2021-04-13 pldi21/livelit-definitions.tex
updates in respond to shepherd

@@
 \noindent
-We now take the perspective of a livelit provider.
+We will now take the perspective of a livelit provider.
 Fig.~\ref{fig:color-impl}
-defines \li{\$color} from Fig.~\ref{fig:color}. We omit certain 
+defines \li{\$color} from Fig.~\ref{fig:color}, 
+which is our prototypic example of a livelit 
+definition. 
+We omit certain  
 incidental details and use unimplemented syntactic sugar, including
-Haskell-style \li{do} notation and quasiquotation, for presentation.
+Haskell-style \li{do} notation \cite{marlow2010haskell} and quasiquotation \cite{bawden1999quasiquotation}, for presentation.
 
 Livelit definitions are scoped and packaged like 
-any other definition. They consist of a declaration and an implementation. 
+any other definition. Each \emph{definition} consists of a \emph{declaration} and an \emph{implementation}. 
+In Hazel, a ``template'' declaration and implementation is generated 
+as soon as the user types ``\li{livelit\ }''.
+
+\subsection{Livelit Declarations} 
 Line 2 of Fig.~\ref{fig:color-impl} is \li{\$color}'s declaration,
 which defines its name and its {expansion type}, \li{Color}, defined on Line 1. 
-Livelit parameters would also appear here as shown in the declarations in Sec.~\ref{sec:expansion-typing}. 
+Both are required. 
+Livelit parameters can also optionally appear here as shown in the declarations in Sec.~\ref{sec:expansion-typing}. 
 The declaration is part of the client interface, 
@@
 The curly braces delimit the livelit's implementation, which is not intended to be seen 
-by clients. 
-Every livelit implementation must include all of the definitions shown in Fig.~\ref{fig:color-impl}, 
-which we will discuss in turn below. 
-Livelits are implemented using a variation on the functional model-view-update
+by clients. Each livelit implementation must define types \li{Model} and \li{Action}, 
+values \li{init}, \li{view}, \li{update}, and \li{expand} with the signatures specified in Fig.~\ref{fig:color}, 
+and a \li{context}, which is a listing of definition-site bindings that the livelit can make use of as described below. All of these are included in the generated template.
+ These constitute a variation on the pure functional model-view-update
 architecture popularized by Elm \cite{ElmArchitecture}. We add a fourth component,
-expansion generation. In addition, we use a simple monadic framework (\emph{a la} Haskell \cite{marlow2010haskell}) to provide the necessary  
-interface between the livelit and the editor, while retaining
-a pure functional programming model (monadic commands, like \li{new_splice} discussed below, are executed by the editor).
+expansion generation. In addition, we use a monadic framework (\emph{a la} Haskell \cite{marlow2010haskell}) to provide a pure   
+ interface between the livelit and the editor: monadic commands, like \li{new_splice} discussed below, are executed by the editor,
+and \li{do} notation provides syntactic sugar for monadic bind.
 
@@
  the system requires that the model type supports automatic serialization
- (so no functions can appear in models).
+ (so functions cannot appear in models).
 
@@
 i.e. an expression of the expansion type, here \li{Color}.
-The \li{expand} function determines this expansion based on the model 
-(which encodes the GUI state, and so usually contains information, like splice references
-and widget states, 
-that should not persist into run-time as part of the expansion).
+The \li{expand} function determines the expansion given the model. 
+The two are necessarily distinct: the model encodes the GUI state, 
+and so usually tracks ephemeral widget state, like which tab is active in a tabbed GUI, 
+as well as splice references, which refer only indirectly to the spliced expressions themselves, discussed below. 
+This information is not needed at run-time as part of the expansion.
+Only the model is persisted when a Hazel program is saved, because the expansion 
+can be regenerated by calling \li{expand} whenever needed.
 
-The expansion can include the spliced expressions, 
-but the system does not make the expression  
-available directly as a value of type \li{Exp}. 
-Instead, \li{expand} returns a \emph{parameterized expansion}, of type \li{Exp}, 
+The expansion can include spliced expressions, 
+but the system does not make these expressions  
+available directly as values of type \li{Exp}. 
+Instead, the expansion must treat splices parametrically. 
+In particular, \li{expand} returns an encoded \emph{parameterized expansion}, of type \li{Exp}, 
 paired with a list of \li{SpliceRef}s (which can come from 
@@
 The {parameterized expansion}
- is an encoding (i.e. a value of type \li{Exp}) of a function 
+  is a function 
  (here curried)
@@
  So here, the parameterized expansion for \li{\$color} must be a function of type 
- \li{Int -> Int -> Int -> Int -> Color}.
+ \li{Int -> Int -> Int -> Int -> Color} because the splices 
+ in the splice list were each of type \li{Int}.
+
+This parameterization strategy makes enforcing the binding discipline described in Sec.~\ref{sec:hygiene}
+straightforward. 
+Context independence is maintained by allowing the parameterized expansion 
+to depend only on bindings given in the explicit \li{context}, which 
+the system maintains bound relative to the definition site. 
+The splices can 
+depend only on the client site typing context. The use of function application ensures
+that splices are capture avoiding. 
+% We consider this more formally next.
 
 Note that Hazel does \emph{not} statically check the definition of \li{expand}
-to ensure that the encoded parameterized expansion has this type. Instead, the parameterized expansion 
-is \emph{validated} at each livelit invocation site, with errors reported to the client. 
+to ensure that the encoded parameterized expansion has the necessary type in the specified \li{context}. Instead, the parameterized expansion 
+is only \emph{validated} at each livelit invocation site, with errors reported to the client. 
 A \emph{typed quotation} system as in, e.g., MetaOCaml \cite{DBLP:conf/flops/Kiselyov14}, could be adapted to allow for definition-site verification, as discussed in \cite{TLMs}, but 
@@
 the type of the quotation depends on the type of each splice in 
-the splice list.
-
-This parameterization strategy makes it simple to enforce the binding discipline described in Sec.~\ref{sec:hygiene}. 
-The parameterized expansion 
-can depend only on the \li{context}, whereas the splices are entered by the client and so they can 
-depend only on the client site typing context. The use of function application ensures
-that splices are capture avoiding as discussed in Sec.~\ref{sec:hygiene}. 
-We consider this more formally next.
+the splice list, which has arbitrary length in general (e.g. the 
+number of splices in the \li{\$dataframe} livelit 
+depends on how many rows and columns the user has added).

### 2021-04-13 pldi21/discussion.tex
acks

@@
 
-(Acknowledgements are being prepared and will be included in the version sent to the publisher.)
+The authors would like to thank
+Pavel Panchekha and 
+the anonymous POPL 2021 and PLDI 2021 referees 
+for thoughtful feedback that has significantly improved this paper. 
+This material is based upon work supported by the National Science Foundation 
+under Grant No. 1814900 and 1817145.

### 2021-04-13 pldi21/abstract.tex
updated to latest style files, added required location data

@@
     plots,
-    GUI widgets, 
-    mathematical diagrams, 
+    diagrams, 
     and other domain-specific data structures.
@@
     Uniquely, livelits are \emph{compositional}: a livelit GUI can itself embed 
-    spliced expressions, which are typed, lexically scoped, enjoy full editor support and can in turn embed 
+    spliced expressions, which are typed, lexically scoped, and can in turn embed 
     other livelits. 
@@
     because the system continuously gathers closures associated with the 
-    hole that the livelit is tasked with filling.
+    hole that the livelit is filling.
     We integrate livelits into Hazel, 
-    a live programming environment able to 
-    typecheck and run programs with holes, 
-    and describe several case studies that exercise these novel capabilities. 
-    We then define a minimal typed livelit calculus, which precisely 
+    a live hole-driven programming environment, 
+    and describe case studies that exercise these novel capabilities. 
+    We then define a simply typed livelit calculus, which  
     specifies how livelits operate as 

### 2021-04-13 pldi21/discussion.tex
updated to latest style files, added required location data

@@
 
-The authors would like to thank
+The authors thank
 Pavel Panchekha and 
-the anonymous POPL 2021 and PLDI 2021 referees 
+ anonymous referees at TyDe 2019, POPL 2021 and PLDI 2021  
 for thoughtful feedback that has significantly improved this paper. 

### 2021-04-13 pldi21/intro.tex
update grade-cutoffs figure + tweak subfigure sizes to make same height
fixed flow around fig 1
updated to latest style files, added required location data

@@
 countless other data structures,
-e.g. vector graphics,
-animation parameters,
-musical sequences,
-audio filters,
+e.g. audio filters, 
+vector graphics,
+% animation parameters,
+% musical sequences,
 board game states,
-GUI widgets and layouts,
+% GUI widgets and layouts,
 tabular numeric data,
-plots,
+% plots,
 geospatial data,
-neural network diagrams, 
-biological neuron models, 
+% neural network diagrams,
+neurobiological circuits,
 and mathematical diagrams.
@@
 image and video editors, music composition software, level design tools,
-and bespoke GUIs written by students or lab technicians.
+and bespoke GUIs written by lab technicians.
 This is in large part because these applications
@@
 direct manipulation affordances,
-e.g. color palettes, visual timelines, plots, and maps.
+e.g. checkboxes, sliders, color palettes, visual timelines, interactive plots, and maps.
 
@@
 \begin{figure*}
-  \begin{minipage}[t]{0.37\textwidth}
+  \begin{minipage}[t]{0.38\textwidth}
     \begin{subfigure}[t]{\linewidth}
@@
   \hspace{10.5mm}
-  \begin{subfigure}[t]{0.50\textwidth}
+  \begin{subfigure}[t]{0.55\textwidth}
   \begin{snugshade}
@@
   \end{subfigure}
-  \vspace{-4mm}
+  \vspace{-5mm}
    \caption{Introductory Examples}

### 2021-04-13 pldi21/discussion.tex
acks and ccs xml

@@
 
-The authors thank
+We thank
 Pavel Panchekha and  
  anonymous referees at TyDe 2019, POPL 2021 and PLDI 2021  
-for thoughtful feedback that has significantly improved this paper. 
+for thoughtful feedback that has improved this paper. 
 This material is based upon work supported by the National Science Foundation 

### 2021-04-14 pldi21/intro.tex
cleanup spacing issues, add alt descriptions, misc cleanup

@@
       \includegraphics[width=\linewidth]{graphite-color-palette-green.png}
+      \Description{A reproduction of a Graphite color palette appearing as a pop above Java code where the cursor is in a hole of type \li{Color}.}
       \vspace*{-5mm}
@@
       \includegraphics[width=\linewidth]{slider-color-livelits.png}
+      \Description{An application of the \li{\$color} livelit to compute a shade of green as described in the text.}
     \end{subfigure}
@@
     \includegraphics[width=\linewidth]{grade-cutoff-livelit.png}
+    \Description{A case study involving several livelits which are described in the text.}
   \end{subfigure}
@@
 the value of this broad class of mechanisms for integrating GUIs with code.
-\vspace{-1mm}
 \subsection{Contributions}
@@
 
-\newcommand{\llproperty}[1]{\vspace{5px}\noindent\textbf{#1}.}
+\newcommand{\llproperty}[1]{\vspace{5px plus 5px minus 3px}\noindent\textbf{#1}.}
 

### 2021-04-14 pldi21/livelit-calculus.tex
cleanup spacing issues, add alt descriptions, misc cleanup
added link to Agda mechanization github, clarifications

@@
     }
+    \Description{Syntax}
     \label{fig:syntax}
@@
     \caption{Expansion}
+    \Description{Expansion}
     \label{fig:expansion}
@@
 Note that we do not formally model the edit actions that change the model or splice list here;
-we focus on a single snapshot of the editor state.
+we focus on a single snapshot of the editor state. We leave an action semantics for livelits, 
+following the Hazelnut action semantics \cite{Hazelnut}, as future work.
 
@@
     to the splices, whose types, $\splat{\tau_i}$, are given in the splice list.
-    Validation can fail, in which case expansion fails. (In Hazel, validation failure is instead marked by a non-empty hole.)
+    Validation can fail, in which case expansion fails. (In Hazel, validation failure is instead marked by a non-empty hole and an appropriate error message is shown.)
 
@@
     In Sec.~\ref{sec:livelit-definitions}, we discussed how the use of an explicit definition site \li{context} eliminates this client burden.
-    The explicit context is formally just a value (tupled, typically) that can be passed as an additional argument to the parameterized expansion alongside the splices, 
+    The explicit context can formally be modeled as just a value (tupled, typically) that can be passed as an additional argument to the parameterized expansion alongside the splices, 
     so we omit it from the calculus, but see \cite{TLMs} for a full treatment.
@@
 \label{sec:agda}
-The supplement mechanically proves these theorems 
+We have mechanically proven these theorems 
 using the Agda proof assistant, building on the Agda mechanization 
 of Hazelnut Live \cite{HazelnutLive}. 
+This is the first mechanization of the literal macro expansion 
+process 
+(previous work on textual literal macros had only paper proofs \cite{TLMs}). 
+The mechanization is available
+in the artifact and online at the following URL:
+
+\begin{center}
+\url{https://github.com/hazelgrove/hazelnut-livelits-agda/}
+\end{center}
 

### 2021-04-14 pldi21/discussion.tex
cleanup spacing issues, add alt descriptions, misc cleanup
final content tweaks

@@
   %
-  The arithmetical symbols are written diagrams and the geometrical figures are graphic formulas.
+  The arithmetical symbols are written diagrams and the geometrical figures are graphic formulas.\vspace{3px}
   %
   }
-
-  \vspace{3pt}
 
@@
   patterns, and entire modules. Prior work on literal macros has explored this \cite{TLMs}.
-  (Note that Racket's visual macro system generates
+  Note that Racket's visual macro system generates
   arbitrary syntax, so it can already be used in 
-  this manner, albeit with no sort-specific semantic guarantees.)
+  this manner, albeit with no sort-specific semantic guarantees.
   % Support for type and module splices, too, would likely be useful.

### 2021-04-14 pldi21/implementation.tex
final content tweaks

@@
 with a model value of the wrong type; 
-(3) a run-time error during evaluation of \li{init}, \li{update}, or \li{expand} functions (which manifests as a run-time hole); (4) 
-expansion validation failure (i.e. that the parameterized expansion is of the wrong type given the splice list).  
+(3) a run-time error during evaluation of \li{init}, \li{update}, or \li{expand} functions (which would manifest as run-time holes); (4) 
+expansion validation failure (i.e. that the parameterized expansion is of the wrong type given the splice list and expansion type).  
 Errors in view generation are not considered semantic errors (they display as error messages where the livelit GUI would have appeared.)
@@
 We have also developed a 
-proof-of-concept implementation of livelit interaction in a textual program editor, Sketch-n-Sketch \cite{sns-pldi,sns-uist}.
-Livelit interaction causes the serialized model in the text buffer to be changed, which updates the view.
-This proof-of-concept is focused on this interaction loop so it is not at feature parity with the 
- main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84,interactive-visual-syntax} 
-is sufficient to support livelits, albeit with some gaps in availability when 
+proof-of-concept implementation of livelit interaction in a textual program editor, Sketch-n-Sketch \cite{sns-deuce}.
+The livelit GUI appears in a pop up window when requested by the user. Interactions with this GUI cause the serialized model in the text buffer to be changed, which updates the view.
+This proof-of-concept is not at feature parity with the Hazel  
+ implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84,interactive-visual-syntax} 
+is sufficient to support livelits, albeit with gaps in availability when 
 there are syntax errors.
 There are other systems that integrate visual syntax into an otherwise textual editor as well, notably Dr.
-Racket's recently introduced visual macro system \cite{interactive-visual-syntax}, discussed in the next section.
+Racket's recent visual macro system \cite{interactive-visual-syntax}, discussed in the next section.
 
@@
 Side effects pose a challenge on two fronts.
+
 If they occur in a livelit implementation,
+as is possible for livelits implemented using Hazel's support for Javascript-based 
+livelits, 
 then the state exists at edit-time. 
-The most viable approach would be to retain the model-view-update
-architecture and specify that any transient mutable state will be reset between updates.
+To ensure that the GUI's state persists and remains valid between reloads, 
+it may then be necessary to persist 
+the relevant portions of the edit-time state 
+alongside the program, as in Smalltalk systems \cite{Goldberg84smalltalk-80-}, or else provide for a state re-initialization protocol
+that is called when a program is re-loaded from persistent memory.
+It may also be necessary to more carefully control the access that a livelit 
+has to the editor itself: many participants in the Graphite survey indicated 
+that they did not want palettes, which were able to make network requests at edit-time, 
+to be able to access 
+proprietary or sensitive source code \cite{Graphite}. 
+In any case, the expansion should ideally be a pure function of the model even in an imperative language, 
+or run as an isolated process each time, 
+to ensure that program 
+behavior does not depend on expansion-time editor state.
+
 If side effects occur in livelit expansions, 
@@
 care to avoid unsoundness, as described in Sec.~\ref{sec:calculus-closure-collection}.
-% However, these problems are surmountable, and we look forward to 
-% efforts to integrate livelits into imperative languages.
+In addition, continuous live evaluation as in Hazel is intrinsically problematic 
+in an imperative setting because running code on each keystroke can cause arbitrary side
+effects, e.g. network requests or I/O.
+
+These problems are surmountable with suitable compromises, e.g. limiting continuous 
+evaluation for code with external side effects, and we look forward to 
+efforts to integrate livelits into imperative languages.
 

### 2021-04-14 pldi21/related-work.tex
final content tweaks

@@
 
+This structural delimitation of splices is reminiscent of work on 
+\emph{language boxes} \cite{DBLP:conf/sle/RenggliDN09}, which focused on 
+combining different notations, primarily textual, using structural delimiters inserted 
+explicitly using a special-purpose editor.
+
 Recent work on a interactive visual macro system in Dr. Racket 
@@
 library constructs). That said, we conjecture that it is
-likely possible to implement many of the mechanisms we describe in this paper 
-perhaps in some form as a layer atop existing mechanisms in Dr. Racket.
-
-Our structural delimitation of splices is reminiscent of work on 
-\emph{language boxes} \cite{DBLP:conf/sle/RenggliDN09}, which focused on 
-combining different notations, primarily textual, using structural delimiters inserted 
-explicitly using a special-purpose editor.
+possible to implement many of the mechanisms we describe in this paper 
+in some form as a layer atop these existing mechanisms in Dr. Racket.
 
@@
 However, these systems generate entire editors, whereas livelit definitions 
-are decentralized in libraries. 
+are decentralized in libraries and lexically scoped. 
 Furthermore, livelit definitions are governed by a type and binding discipline.
@@
 % about arguments, i.e. splices and parameters.
-Finally, livelits are uniquely live, building on Hazelnut Live \cite{HazelnutLive}.
+Finally, livelits are uniquely live, building on hole closure tracking 
+from Hazelnut Live \cite{HazelnutLive}.
 
@@
 Related to projectional editors are 
-a variety of systems that generate visualizations from code \cite{DBLP:conf/chi/Lerner20,koschke2003software,urquiza2004survey}. Livelits differ 
+a variety of systems that generate visualizations from code \cite{DBLP:conf/chi/Lerner20,koschke2003software,urquiza2004survey,Goldberg84smalltalk-80-}. Livelits differ 
 from these systems in directionality: visualization systems are given code and generate
@@
 but the editing representation itself is fixed as a series of connected windows.
-Only the visual representation is customizable.
+Only the visualizations are customizable.
 
